### 2019-06-25 Final_doctoral_lan.tex
all files 6/25/2019

\documentclass[14pt]{article}
%	options include 12pt or 11pt or 10pt
%	classes include article, report, book, letter, thesis
\usepackage{setspace}
\usepackage{listings}
\usepackage[a4paper, total={6in, 8in}]{geometry}
\usepackage{pgfgantt}
\usepackage{afterpage,lscape}
\usepackage{bbm}
\usepackage{amsmath, amsthm, amscd, amsfonts}
\usepackage{graphicx}
\usepackage{pstricks}
\usepackage{natbib}
\setcitestyle{authoryear,open={(},close={)}}
\usepackage{multirow}
\usepackage{tocbibind}
\usepackage{cleveref}
\usepackage{textcomp}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{property}{Property}
\newtheorem{definition}{Definition}
\newtheorem{remark}{Remark}
\usepackage{setspace}
\usepackage[titletoc,title]{appendix}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{epstopdf}
\usepackage{listings}
\usepackage{longtable}
\renewcommand{\baselinestretch}{1.5} 
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}

\usepackage{geometry} % to change margins
\usepackage{pdflscape} % provides the landscape environment
\usepackage{ragged2e} % provides \RaggedLeft

\usepackage{fancyhdr}

\pagestyle{fancy}
\fancyhf{}
\rhead{{\textsf{Vahid Nassiri}}}
\lhead{{\textsf{Final Doctoral Plan: Data Splitting and its Applications}}}
\rfoot{Page \thepage}

\renewcommand{\headrulewidth}{1pt}

\begin{document}
\begin{titlepage}

\begin{center}
\includegraphics[scale=1]{KUL.eps}\\
\vspace*{.3cm}
{\sc{\LARGE{Final Doctoral Plan}}}\\
\vspace*{2cm}
{\Huge{\textbf{Data Splitting and its Applications}}}

\vspace*{4cm}
{\LARGE{\textbf{Vahid Nassiri}}}\\
\vspace*{1cm}
{\Large{I-BioStat, KU Leuven}}\\
\vspace*{2cm}
{\Large{Supervisor:}}\\
\vspace*{0.2cm}
{\LARGE{\textbf{Prof. Dr. Geert Molenberghs}}}\\
\vspace*{0.5cm}
{\Large{Co-supervisor:}}\\
\vspace*{0.2cm}
{\LARGE{\textbf{Prof. Dr. Geert Verbeke}}}
\vspace*{2.3cm}

{\Large{April 2017}}

\end{center}

\end{titlepage}
\newpage
{\large{

\textbf{Title}: {\Large{Data Splitting and its Applications}}
\vspace*{0.5cm}

\textbf{Name of applicant}: {\Large{Vahid Nassiri}} (r0292876)
\vspace*{0.5cm}

\textbf{Supervisor}:

{\Large{Prof. Dr. Geert Molenberghs,}}

I-BioStat, Universiteit Hasselt, Diepenbeek, Belgium.

I-BioStat, KU Leuven, Leuven, Belgium

\vspace*{0.5cm}

\textbf{Co-supervisor:}

{\Large{Prof. Dr. Geert Verbeke,}}

I-BioStat, KU Leuven, Leuven, Belgium,

I-BioStat, Universiteit Hasselt, Diepenbeek, Belgium.

\vspace*{0.5cm}
\textbf{Address}:

I-BioStat, KU Leuven
Kapucijnenvoer 35 3000 Leuven, Belgium.

Tel: +32 (0) 16 37 46 96,

Email: vahid.nassiri@kuleuven.be

\vspace*{0.5cm}
\textbf{Available Grant}: IWT- SBO/ Exascience/ IAP

\vspace*{0.5cm}
\textbf{Doctoral School Programme}: Patient Related and Public Health Research
\vspace*{0.5cm}

\textbf{Full-time or part-time PhD project}: it is a full-time PhD Project

\vspace*{0.5cm}
\textbf{Starting date}: 1 May 2014

\vspace*{0.5cm}
\textbf{Expected end date}: 30 April 2018}}

\newpage

\section{Introduction and problem statement}
\label{sec_intro}
Sample size has always been an issue in statistical analysis. While for many years small sample sizes were a central issue, in recent years, large sample sizes too might confront the statistical analyst with serious problems. When large to huge sample sizes concur with complex models, prohibitive computation times, convergence issues, and the mere impossibility to analyze the data with the preferred inferential technique (e.g., full maximum likelihood) can ensue.

Furthermore, while in the simplest designs all measurements are independent, many designs lead to hierarchical (a.k.a., clustered, correlated, dependent) data \citep{laird1982,zeger1986,verbeke2009, molenberghs2005}. By cluster here we mean a set of measurements repeatedly collected for one single unit (e.g., subject, household, trial, etc.). Therefore, the number of repetitions refers to number of measurements available per cluster. Examples include: patients measured repeatedly over time, several patients attending the same general practitioner or clinic, several links clicked by a specific user, genetic data available for a person, etc. Some important examples of clustered data are repeated measurements, longitudinal data, multi-center studies, and meta-analysis. Accounting for such correlations remains a challenge today, when the total number of clusters is large and/or there are a large number of repetitions per cluster. This would sometimes result in prohibitive computations.

Consider a study which measures (at most) $m$ outcomes for $N$ clusters, each $n_i$ times. Let $y_{rij}$ be the $j$th measurement taken on the $i$th cluster for the $m$th outcome, $i=1,\ldots,N$, $r=1,\ldots,m$ and $j=1,\ldots,n_{ri}$. A general scheme of such data is presented in Figure~\ref{fig_scheme}. As it was mentioned, fitting a model to such data could become challenging when the size of dataset make it eligible for the title \emph{big data}. A dataset could be entitled as \emph{big data} in different situations. In case of the clustered data of interest in this PhD research, big data considered to be happened in at least one of these situations,
\begin{itemize}
\item When the sample size, $N$, becomes very large,
\item When the cluster sizes become very large:
\begin{itemize}
\item When the number of measurements per outcome for some clusters, $n_{ri}$'s, becomes very large,
\item When the number of outcomes, $m$, becomes very large.
\end{itemize}
\end{itemize}
\begin{figure}
\centering
\includegraphics[width=\textwidth]{scheme_new.eps}
\caption{General scheme of the data we consider through this PhD research.} 
\label{fig_scheme}
\end{figure} 
A large sample size could increase the computation time dramatically. On the other hand, even with small sample sizes, having large clusters may increase the computation time. Doing the computations and achieving convergence would become more challenging when the clusters are highly unbalanced (a wide range of cluster sizes). Also, having too many outcomes would increase the number of parameters in the model, to the point which it could not be feasible any more.

In this PhD research, such situations are considered and we are trying to propose methodologies to deal with such problems using the so-called data splitting. In a nutshell, data splitting means to split the difficult (sometimes infeasible) job of modelling the big data by splitting it into smaller chunks of manageable sub-data, analyze each of them separately, and then combine the results of these analyses using appropriate combinations rules. This would break a difficult task into several embarrassingly parallel easier tasks.

The rest of this report is structured as follows. In Section~\ref{sec_objective} the main objectives of this PhD research will be briefly reviewed. The datasets motivating our research will be presented in Section~\ref{sec_data_sets}. Section~\ref{sec_idea} describes the general idea behind the methodologies we are developing. The results obtained so far will be presented in Section\ref{sec_results_so_far}. The work in progress as well as a timeline for finishing them will be given in Section~\ref{sec_progress}. Finally, a provisional structure for the upcoming PhD thesis is proposed in Section~\ref{sec_structure}.

\section{Main objectives}
\label{sec_objective}
Considering the problem stated in Section~\ref{sec_intro}, our main objective is to develop a unified framework which helps researchers to deal with big data in the situations we have described. We expect this framework to have the following properties.

\begin{itemize}
\item The developed framework should be as independent as possible from the modelling approach. In other words, it should be compatible with various modelling approaches, e.g.,  linear, generalized linear, non-linear, etc.
\item As the end-user of this product will be researchers of different fields with different tastes of data analysis systems who are dealing with big data, the developed methodologies should preferably be easy to understand and implement in different platforms, e.g., \textsf{R}, \textsf{SAS}, \textsf{Matlab}, \textsf{Python}, \textsf{Julia}, etc.
\item It goes without saying that a combination of the situations we have described in Section~\ref{sec_intro} could happen for a single set of data. For example, a dataset would consist of a large number of clusters (large $N$), and each of these clusters could be large as well (large $n_{ri}$'s), see Amazon book ratings in Section~\ref{sec_data_sets}. Or large number of outcomes of interest can happen with large number of clusters, see divorce in Flanders dataset in Section~\ref{sec_data_sets}. Therefore, our proposed methodologies for each situation should have the possibility of being combined to deal with a combination of the mentioned situations.
\item This unified approach is going to be implemented in freely available statistical package \textsf{R}. So every interested person can use it without the trouble of implementation. This implementation should be done as efficient as possible.
\item Finally, we should be able to apply our proposed methodology to solve the problems involving our motivating datasets. Considering the diversity of them, that would prove the usefulness of our approach to deal with 21st century real world big data problems.
\end{itemize}

\section{Motivating datasets}
\label{sec_data_sets}
A brief summary of the datasets motivating our research is presented in this section.
\paragraph*{A developmental toxicity study:} This study investigates the dose-response
relationship in mice of the potentially hazardous chemical compound di(2-ethylhexyl)phthalate
(DEHP), used in vacuum pumps and as plasticizers for numerous plastic devices made of polyvinyl chloride. The developmental toxicity study, conducted in timed-pregnant
mice during the period of major organogenesis has attracted
much interest in the toxicity of DEHP. The doses selected for the study were 0, 0.025, 0.05, 0.1,
and 0.15\%, corresponding to a DEHP consumption of 0, 44, 91, 191, and 292 mg/kg/day, respectively.
The dams were sacrificed, slightly prior to normal delivery, and the status of uterine implantation sites recorded. A total of 1082 live fetuses were dissected from the uterus, anesthetized, and examined for external, visceral, and skeletal malformations, as well as for body
weight. Our focus will be on the continuous weight outcome. Evidently, fetuses are clustered within mothers; hence the implied association needs to be accommodated in the analysis. As each mother has different number of fetuses, we have clusters of different sizes.

\paragraph*{Divorce in Flanders:} This dataset is collected in Centre for Sociological Research, KU Leuven. The data concerned on studying and comparing the personality among
three generations in about 4000 Flemish families. Each family would at most consist of mother, father, step-mother, step-father, one child, grand mother, and grand-father. The personality is measured using the so-called
Big Five Inventory (BFI) score. These scores are built upon 5 factors which in total are measured through 44 items. The usual factor analysis would lose most of the information in the data. Therefore, an idea is to model different combinations of the 5 aspects of personality for each family role using a mixed model. This means, for each family we will have up to 35 responses. The large size of families would add to the difficulty of the problem.

\paragraph*{Hearing data:} To evaluate the hearing performance of a subject, hearing threshold sound pressure levels (measured in dB) are considered. By definition, a hearing threshold is the lowest signal intensity possible to perceive by a subject at a specific frequency. In a study considered in \cite{verbeke2009} and \cite{Verbeke2006}, hearing threshold measured in 11 different frequencies between 125 Hz and 8000 Hz for left and right ears obtained on 603 male participants from the Baltimore Longitudinal Study of Ageing, BLSA. The number of visits for each subject which varies between 1 to 15 are unequally spaced. Modelling these 22 frequencies together is a challenge for the methodologies we developed in this PhD thesis.

\paragraph{Leuven diabetes project:} In order to study how well the diabetes is controlled, three outcomes were measured for each patient at baseline ($T_0$) and one year later ($T_1$). These three outcomes of interest were LDL-cholesterol (low-density lipoprotein cholesterol, mg/dl), HbA1c (glycosylated hemoglobin, \%), and SBP (systolic blood pressure, mmHg). Although these three variables are continuous, the interest was in expert-specified cut-off values of these outcomes. The cut-off values are defined as $\{<100, [100,115), [115, 130),\geq 130 \}$ (mg/dl) for LDL-cholesterol, $\{<7, [7, 8), \geq 8\}\%$ for HbA1c, and $\{ \leq 130, (30,140],(140,160],>160\}$mmHg for SBP. Modelling these 3 ordinal outcomes together provides a suitable platform to study the performance of our proposed methodologies dealing with non-Gaussian data.

\paragraph*{Amazon book ratings:} This data consist of the ratings of the books on the Amazon website which are recently made publicly available. The dataset contains $2,330,066$ books, which are rated at least $1$ and at most $21,398$ times. This is an example of clustered data with large highly unbalanced clusters and a large sample size.

\paragraph*{Clinical Trials in Schizophrenia:} These data were collected from five double-blind randomized clinical trials to compare the effects of two treatments for chronic schizophrenia: risperidone and conventional antipsychotic agents. Subjects who received doses of risperidone (4–6 mg/day) or an active control (haloperidol, perphenazine, zuclopenthixol) have been included in the analysis. Patients were clustered within country, and longitudinal measurements were made on each subject over time. The number of patients ranges from 9 to 128 per country with a total of 2039. The
positive and negative syndrome scale (PANSS) was used to asses the global condition of a patient. This scale is constructed from 30 items, each taking values between 1 and 7, giving an overall range of 30 to 210. PANSS provides an operationalized, drug-sensitive instrument, which is useful for both typological and dimensional assessment of schizophrenia. Depending on the trial, treatment was administered for a duration of 48 weeks with at most 12 monthly measurements. Because not all subjects received treatment at the same time points and, not the same amount, the final dataset is unbalanced. Our analysis shows AR(1) structure could be a good candidate for modelling the covariance matrix of the random effects, which is useful to illustrate some of the methodologies we have developed.

\paragraph*{Age-related macular degeneration (ARMD):} The data come from a clinical trial in age-related macular degeneration (ARMD), which is a disease of the retina that causes severe central vision loss. The data consist of a multi-center study (17 centers) to evaluate the effect of an experimental treatment based on interferon-$\alpha$ in a group of patients who were randomly allocated to either placebo or interferon-$\alpha$. The outcome of interest was the change in visual acuity over time. Using standard vision charts, the visual acuity was measured by the number of letters that were correctly read by the patient. The data is used in the context of surrogate endpoint evaluation. The goal of this analyses is to examine if  visual acuity at week 24 could be an appropriate surrogate for the one at week 52, the so-called true endpoint. To this end, the surrogate and the true endpoint have to be modelled jointly. Because of the bivariate nature of the model, coupled with the multi-centre nature of the study, a mixed model is used, by which surrogate evaluation measures are estimated. The data however is highly unbalanced and only has 17 clusters. This makes the iterative algorithms to face with difficulty in convergence. We try to use the methodologies developed in this PhD research to deal with this problem.

\paragraph*{Leuven eyes study:} This is an extensive observational study of glaucoma performed at the ophthalmology department of KU Leuven. The dataset so far consists of 141 variables measured in 614 subjects. As measuring all of these variables was not possible for all of the subjects, this dataset exhibits considerable missingness. Considering the size of dataset and amount of missing values, applying multiple imputations and determining the number of imputations is challenging in this change.

\section{Data splitting: the general idea}
\label{sec_idea}
As it was briefly mentioned in Section~\ref{sec_intro}, the data splitting approach consists of 3 steps.
\begin{enumerate}
\item Splitting the data.
\item Performing the analysis on each sub-data.
\item Efficiently combine the results from each sub-data to form an overall estimate with its precision.
\end{enumerate}
Let us briefly go over our approaches to deal with each of these steps.

The first step of this procedure is determined by type of the big data we are dealing with. We divide the data splitting methods based on two main characteristics: 1- independent or dependent sub-samples, 2- random or structured sub-sampling.

Data splitting can be done horizontally or vertically. If the splitting is done at the subject level (each subject will be fully presented in sub-samples), it is called horizontal splitting. The reason of using such an appellation can be observed in Figure~\ref{fig_scheme} (left). Splitting at the subject level means splitting the full data into smaller pieces using horizontal slices. Considering the assumption of having independent clusters, this will result in independent sub-samples. Unless, the sub-samples are overlapped, i.e., one subject appears in more than one sub-sample\footnote{Note that, while perfectly possible, the overlapped horizontal splitted sub-samples are not considered in this thesis. One can look in vast literature of bootstrap methods for such an approach.}.

On the other hand, when splitting is done within each subject (part of the measurements for each subject are presented in each sub-sample), it is called vertical splitting. The right-side of Figure~\ref{fig_scheme} shows the reason behind this title. Splitting the measurements within the subject means to take the sub-samples vertically. Such sub-samples will be dependent, since they are measured for a common subject.

Each of horizontal or vertical splitting can be done at random or based on a pre-specified structure.  Here we briefly describe the four different combinations of vertical/horizontal and random/structured data splitting.

\begin{itemize}
\item \textbf{Random horizontal data splitting.} This approach is useful when the number of clusters is large, and preferably the clusters are roughly of equal size, i.e., (almost-) balanced design. This will produce several sub-data with smaller sample sizes which are easier to handle.
\item \textbf{Structured horizontal data splitting.} This approach is useful when some specific clusters are more convenient to deal with when they are together. Therefore, instead of taking random sub-samples, the sub-sampling can be done based on this pre-specified structure. An important example of this approach which is considered in this PhD research is splitting based on the cluster sizes, i.e., forming sub-samples consist of clusters of equal size. This would lead to balanced sub-samples. We have shown existing of closed-form solutions, as well as, faster convergence can be achieved using this approach.
\item \textbf{Random vertical data splitting.} This approach is useful when we are dealing with data with large number of measurements for each cluster per outcome (large $n_{ri}$'s). In this situation one can take sub-samples of each cluster with a manageable size $n_{f(ri)} < n_{ri}$. Obviously, this can be done by taking overlapped (sub-sampling with replacement) or non-overlapped (sub-sampling without replacement) sub-samples.
\item \textbf{Structured vertical data splitting.} This approach is useful when the number of outcomes of interest, $m$, is large. While fitting a model to $m$ outcomes could become heavy or infeasible, fitting models to all possible pairs (triples, etc.) of these outcomes is feasible. That means each sub-sample should consist of all of the measurements for each subject regarding the pair of interest. Therefore, the sub-samples are taken using a pre-specified structure.
\end{itemize}
Obviously, according to type of big data one is dealing with, all of the the combinations of random/structured and horizontal/vertical data splitting can be freely combined. The combination rules then should be combined accordingly. This enables the proposed framework to deal with a combination of the big data situations described in Section~\ref{sec_intro}.

After splitting the data into appropriate sub-samples, the next step is to perform the necessary analysis on these sub-samples. The analysis performed on each sub-sample does not need to be similar to one another. For example, consider 3 outcomes of interest of types continuous, binary, and ordinal. If sub-samples are taken using a structured vertical splitting approach, and each sub-sample consists of pairs of outcomes, the model fitted to each sub-sample will depend on the type of the responses presenting there. Furthermore, the explanatory variables of the models within each sub-sample does not need to be identical across sub-samples. Therefore, such a framework is compatible with different modelling approaches.

The last step of this procedure is combining the results obtained from various sub-samples. As the models fitted within each sub-sample have some common parameters, each parameter of the initial model could possibly be estimated more than once after second step. We propose to use a weighted average of them to obtain a unique estimation for each parameter. The critical part is to define such appropriate weights. They have to be computed proportional to the number of clusters within each sub-sample. Also, if needed, size of these clusters should be involved. We have also computed optimal combination weights in order to minimize the final combined variance.

Computing the precision of such an estimator is another important aspect. Computing this precision depends on whether the sub-samples are dependent or independent. For independent sub-samples, one can simply use an average of the obtained variances from each sub-sample. For dependent sub-samples, this averaged variance should be corrected for the information the overlapped sub-samples share. This correction is done using a sandwich estimator \citep{Verbeke2006, Verbeke2007} and multiple ouputation \citep{hoffman2001, follmann2003}.

\section{Results obtained so far}
\label{sec_results_so_far}
We try to present the results obtained so far through the published, accepted, and submitted papers. The working papers and projects will be presented in Section~\ref{sec_progress}.

\begin{itemize}
\item[--]{\textsf{Lisa Hermans, Vahid Nassiri, Geert Molenberghs, Michael G. Kenward, Wim Van der Elst, Marc Aerts, and Geert Verbeke}} , {\it Clusters with random size: maximum likelihood versus weighted estimation}, Submitted to Statistica Sinica.
\end{itemize}
This paper considers the case of structured horizontal splitting for a dataset consists of vectors of random variables with multivariate Gaussian distribution with compound-symmetry (CS) covariance structure of different sizes. The CS covariance structure considers a common variance for all of the components in the random vector. Furthermore, it assumes all pairs of variables are equally correlated (a common correlation coefficient). We have proposed to split the large sample into sub-samples formed by all clusters of equal sizes. Closed-form solutions are obtained to find the parameter estimates within each balanced sub-sample. The appropriate weights are obtained to efficiently combine estimates from different sub-samples. The closed-form estimates and efficient combination weights are also implemented in \textsf{R}. Simulation results show the derived closed-forms reduce the computation time up to 30,000 times comparing with iterative procedures in \textsf{Proc MIXED} in \textsf{SAS}.

\begin{itemize}
\item[--]{\textsf{Lisa Hermans, Vahid Nassiri, Geert Molenberghs, Michael G. Kenward, Wim Van der Elst, Marc Aerts, and Geert Verbeke}} , {\it Fast, Closed-form, and Efficient Estimators for Hierarchical Models with AR(1) Covariance and Unequal Cluster Sizes}, Communications in Statistics - Simulation and Computation, 2017, DOI: 10.1080/03610918.2017.1316395.
\end{itemize}
This paper considers the same problem as in the previous one, except here the covariance structure is assumed to be autoregressive of order 1, AR(1). In this structure, unlike the CS, the correlation between pairs of variables depends on their distance, i.e., the correlation between $i$th and $j$th components is defined as $\sigma^2 \rho^{|i-j|}$ ($\sigma^2>0$, $-1\leq\rho\leq +1$). The closed-form, fast and efficient estimates are obtained for a balanced data following Gaussian distribution with AR(1) structured covariance matrix. The appropriate weights are computed to efficiently combine these estimates and their variances. These closed-form estimates and efficient combination weights are also implemented in \textsf{R}. Simulation results show using the derived closed-forms, the computation time reduced up to 700 times comparing with iterative procedures in \textsf{Proc MIXED} in \textsf{SAS}.

\begin{itemize}
\item[--]{\textsf{Vahid Nassiri, Anna Ivanova, Geert Molenberghs, and Geert Verbeke}} , {\it Fast precision estimation in high-dimensional multivariate joint models}, To appear in Biometrical Journal.
\end{itemize}
In this paper the case of structured vertical splitting is considered when the number of outcomes of interest becomes large. We have proposed to take sub-samples formed by pairs (if needed triples, etc.) of the outcomes. Efficient combination rules were obtained and evaluated using simulations for linear and generalized linear models. The proposed methodology was also implemented in \textsf{R}. Simulation results shows up to 2500 times faster computations comparing with existing procedures.

\begin{itemize}
\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}} , {\it Finite information limit variance-covariance structures: Is the entire dataset needed for analysis?}, In High Performance Computing \& Simulation (HPCS), 2016 International Conference on (pp. 736-742). IEEE.
\end{itemize}
We have introduced the concept of finite information limit (FIL) estimator in this paper. This concerns the situation that amount of information within each cluster is finite. Therefore, after some point adding new observations would not substantially improve efficiency. This means, in case of FIL estimators, one only needs to analyze a small fraction of data to obtain results as efficient as analyzing the full dataset. This concept is introduced in this paper, and some procedures are proposed to detect this property as well as determining the size of this fraction. Comparing with fitting the model to the full data, simulation results show about 1500 times faster computations using this idea.

\begin{itemize}
\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}} , {\it A simple, principled rule to determine the number of multiple imputations}, Submitted to The American Statistician.
\end{itemize}
We have discussed the need to take several sub-samples of the original data to deal with big data. In case that these sub-samples are taken with replacement, infinite number of sub-samples can be taken. Furthermore, for without replacement sub-samples it is very well possible that taking all possible sub-samples needs enormous amount of time. Thus, the need to use only a part of these sub-samples would emerge. This paper considers the problem of determining the number of sub-samples. Of course, as this problem is similar to the problem of determining the number of imputations when dealing with missing data using multiple imputations, the paper was written with that topic as a target. But the idea is general and can be applied in many situations. It has been used in the FIL paper, and will be used in the IMO paper (see Section~\ref{sec_progress}).

\begin{itemize}
\item[--]{\textsf{Vahid Nassiri, Anik\'{o} Lovik, Geert Molenberghs, and Geert Verbeke}} , {\it On using multiple imputation for factor analysis of incomplete data}, Submitted to Behavior Research Methods.
\end{itemize}
When it comes to factor analysis, principal component analysis, or other methodlogies dealing with  eigendecomposition of the covariance matrix, analyzing several datasets in place of the original one would cause some problems at the combination step (Step 3 in our proposed procedure). The reason lies in the fact that not all of the obtained factor loadings would follow the same order. In other words, there is no guarantee that the eigenvector corresponding to the largest eigenvalue of one of these sets of data is comparable with another. This could happen when missing data are multiply imputed, when sub-samples of clusters are taken to overcome computation issues with larger clusters, or when a multilevel factor analysis is of interest, but only at within cluster level. This paper considers this problem and proposed to first estimate the covariance matrix and then perform the factor analysis on the single estimated covariance matrix. The consequences of such an approach are discussed and some criteria are proposed to decide whether this is an appropriate solution or not.

\section{Work in progress}
\label{sec_progress}

In this section, we try to explain the work in progress through briefly describing working papers and projects.

\begin{itemize}
\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}} , {\it Iterative multiple outputation: a simple procedure to deal with large clustered data}, In progress.
\end{itemize}
Inspired by finite information limit estimators, we are trying to extend this idea to cases where this property is not hold, or the size of the one sub-sample needed to achieve it is large comparing with available computation power. In this situation, we would suggest to take several smaller random sub-samples. That would involve appropriate rules to determine the number of sub-samples, as well as, combining the results after fitting the models to such sub-samples. Some simulations are already performed, also \textsf{R} functions are written to efficiently take the sub-samples and combine the results of analyses performed on them. Works are on going to apply this methodology on some of our motivating data. The paper would hopefully be submitted before beginning of next academic year.

\begin{itemize}
\item[--]{\textsf{Vahid Nassiri, Anik\'{o} Lovik, Geert Molenberghs, and Geert Verbeke}} , {\it A simple multilevel factor analysis}, In progress.
\end{itemize}
The use of the sub-sampling from different clusters is not restricted to big data. This can also be useful in situations where the intra-cluster correlation is nuisance. Consider the divorce in Flanders data in Section~\ref{sec_data_sets}. There, the main interest, at validation stage, is to verify the BFI structure among the 44 items of the questionnaire removing the correlation imposed by the fact that some of the participants are coming from the same family. This paper considers this problem and try to propose a solutions based on random vertical splitting. The manuscript is in its final levels and we expect it to be submitted within this academic year.

\begin{itemize}
\item[--]{\textsf{Vahid Nassiri, Alvaro Florez Poveda, Luis Adrian Quintero Sarmiento, Wim Van der Elst, Geert Molenberghs, and Geert Verbeke}} , {\it Comparing different approaches for mixed-effects modelling of unbalanced clustered data}, In progress.
\end{itemize}
Although the methodologies we proposed and developed in this thesis are targeting big data, sometimes very small data and very large data suffer from the same issues. One common problem of large and small data is the convergence issues. In this article we try to study the performance of our proposed method to deal with small highly unbalanced data and compare it with other solutions developed in I-BioStat.

\begin{itemize}
\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}} , {\it Data splitting: an \textsf{R} package to analyze big clustered data.}, In progress.
\end{itemize}
As it was mentioned a couple of times, for many of the proposed and developed methods \textsf{R} functions are prepared to make the implementation easier for the end-user. We are planning to make generic functions out of them and collect them in the form an \textsf{R} package which will be publicly available. This will need more time and effort, but we hope it can be accomplished by the end of the PhD.

The last final step in the last year of this PhD research will be preparing the final manuscript of the PhD thesis. Figure~\ref{fig_gant} shows a Gantt chart of the fourth year PhD in order to complete the mentioned tasks.

\begin{figure}
\centering
\begin{ganttchart}[vgrid={draw=none,draw=none},%
             %today=15,%
             %today offset=.5,%
             %today label=Heute,%
             %progress=today,%
             x unit=0.7cm,
             y unit title=1cm,
             y unit chart=0.5cm,
             bar incomplete/.append style={fill=red},%
             progress label text=  {\quad\pgfmathprintnumber[precision=0,verbatim]{#1}\%},
             milestone label font=\tiny,
             group label font=\tiny,
             title label font=\tiny,
             bar label node/.style={text width=5cm,align=right,font=\scriptsize\RaggedLeft,anchor=east},
             milestone label node/.style={text width=3cm,align=right,font=\scriptsize\RaggedLeft,anchor=east},
             group label node/.style={text width=4cm,align=right,font=\scriptsize\RaggedLeft,anchor=east}
             ]{1}{12}
\gantttitlecalendar*[compress calendar,time slot format=isodate]{2017-05-1}{2018-4-30}{year, month} \\
\gantttitlelist{1,...,12}{1}\\
\ganttgroup{Total Duration}{1}{12} \\
  %%%%%%%%%%%%%%%%%Phase-1
\ganttgroup{Finishing research papers}{1}{6} \\
\ganttbar{Paper 1}{1}{1} \\
\ganttlinkedbar{Paper 2}{2}{3} \ganttnewline
\ganttlinkedbar{Paper 3}{4}{6} \ganttnewline

%%%%%%%%%%%%%%%%%Phase-2
\ganttgroup{Preparing the \textsf{R} package}{4}{9} \\
  %%%%%%%%%%%%%%%%%Phase-3
\ganttgroup{Writing the thesis manuscript}{9}{12} \\
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{ganttchart}
\label{fig_gant}
\caption{Gantt chart of the provisional fourth year PhD plan}
\end{figure}

\section{Provisional structure of the thesis}
\label{sec_structure}

\begin{itemize}
\item Chapter 1: Introduction and the state of art
\item Chapter 2: Motivating datasets
\item Chapter 3: Problem statement and preliminaries
\item Chapter 4: Horizontal splitting
\item Chapter 5: Vertical splitting
\item Chapter 6: Related topics
\item Chapter 7: Conclusions
\end{itemize}

\bibliography{bib_final_plan}{}
\bibliographystyle{abbrvnat}

\end{document}

### 2019-06-25 Final_doctoral_plan.tex
all files 6/25/2019

--- 2019-06-25 Final_doctoral_lan.tex
@@
 By cluster here we mean a set of measurements repeatedly collected for one single unit (e.g., subject, household, trial, etc.).
-Therefore, the number of repetitions refers to number of measurements available per cluster.
+Therefore, the number of repetitions refers to the number of measurements available per cluster.
 Examples include: patients measured repeatedly over time, several patients attending the same general practitioner or clinic, several links clicked by a specific user, genetic data available for a person, etc.
@@
 A dataset could be entitled as \emph{big data} in different situations.
-In case of the clustered data of interest in this PhD research, big data considered to be happened in at least one of these situations,
+In case of the clustered data of interest in this PhD research, data are considered to be big if at least one of the following applies:
 \begin{itemize}
@@
 Doing the computations and achieving convergence would become more challenging when the clusters are highly unbalanced (a wide range of cluster sizes).
-Also, having too many outcomes would increase the number of parameters in the model, to the point which it could not be feasible any more.
+Also, having too many outcomes would increase the number of parameters in the model, to the point at which it could not be feasible any more.
 
-In this PhD research, such situations are considered and we are trying to propose methodologies to deal with such problems using the so-called data splitting.
+In this PhD research, such situations are considered and we aim to propose methodologies to deal with such problems using the so-called data splitting.
 In a nutshell, data splitting means to split the difficult (sometimes infeasible) job of modelling the big data by splitting it into smaller chunks of manageable sub-data, analyze each of them separately, and then combine the results of these analyses using appropriate combinations rules.

### 2019-06-25 Final_doctoral_plan02.tex
all files 6/25/2019

--- 2019-06-25 Final_doctoral_plan.tex
@@
 \usepackage{listings}
-\usepackage[a4paper, total={6in, 8in}]{geometry}
+\usepackage[a4paper, total={6.5in, 9in}]{geometry}
 \usepackage{pgfgantt}
@@
 \usepackage{longtable}
-\renewcommand{\baselinestretch}{1.5} 
+\renewcommand{\baselinestretch}{1.4} 
 \usepackage[english]{babel}
@@
 In a nutshell, data splitting means to split the difficult (sometimes infeasible) job of modelling the big data by splitting it into smaller chunks of manageable sub-data, analyze each of them separately, and then combine the results of these analyses using appropriate combinations rules.
-This would break a difficult task into several embarrassingly parallel easier tasks.
+This would break a difficult task into several, often so-called embarrassingly parallel, simpler tasks.
 
@@
 \label{sec_objective}
-Considering the problem stated in Section~\ref{sec_intro}, our main objective is to develop a unified framework which helps researchers to deal with big data in the situations we have described.
+Considering the problem stated in Section~\ref{sec_intro}, our main objective is to develop a unified framework that helps researchers to deal with big data in the situations we have described.
 We expect this framework to have the following properties.
@@
 
-\paragraph*{Divorce in Flanders:} This dataset is collected in Centre for Sociological Research, KU Leuven. The data concerned on studying and comparing the personality among
+\paragraph*{Divorce in Flanders:} This dataset is collected at Centre for Sociological Research, KU Leuven. The data concerned on studying and comparing the personality among
 three generations in about 4000 Flemish families.
@@
 
-\paragraph{Leuven diabetes project:} In order to study how well the diabetes is controlled, three outcomes were measured for each patient at baseline ($T_0$) and one year later ($T_1$).
+\paragraph{Leuven diabetes project:} In order to study how well diabetes is controlled, three outcomes were measured for each patient at baseline ($T_0$) and one year later ($T_1$).
 These three outcomes of interest were LDL-cholesterol (low-density lipoprotein cholesterol, mg/dl), HbA1c (glycosylated hemoglobin, \%), and SBP (systolic blood pressure, mmHg).
@@
 
-\paragraph*{Amazon book ratings:} This data consist of the ratings of the books on the Amazon website which are recently made publicly available.
+\paragraph*{Amazon book ratings:} These data consist of the ratings of the books on the Amazon website which are recently made publicly available.
 The dataset contains $2,330,066$ books, which are rated at least $1$ and at most $21,398$ times.
@@
 \label{sec_idea}
-As it was briefly mentioned in Section~\ref{sec_intro}, the data splitting approach consists of 3 steps.
+As it was briefly mentioned in Section~\ref{sec_intro}, the data splitting approach consists of 3 steps:
 \begin{enumerate}
@@
 If the splitting is done at the subject level (each subject will be fully presented in sub-samples), it is called horizontal splitting.
-The reason of using such an appellation can be observed in Figure~\ref{fig_scheme} (left).
+The reason of using such a terminology can be observed in Figure~\ref{fig_scheme} (left).
 Splitting at the subject level means splitting the full data into smaller pieces using horizontal slices.
 Considering the assumption of having independent clusters, this will result in independent sub-samples.
-Unless, the sub-samples are overlapped, i.e., one subject appears in more than one sub-sample\footnote{Note that, while perfectly possible, the overlapped horizontal splitted sub-samples are not considered in this thesis.
+Unless the sub-samples are overlapped, i.e., one subject appears in more than one sub-sample\footnote{Note that, while perfectly possible, the overlapped horizontal splitted sub-samples are not considered in this thesis.
 One can look in vast literature of bootstrap methods for such an approach.}.
@@
 \begin{itemize}
-\item \textbf{Random horizontal data splitting.} This approach is useful when the number of clusters is large, and preferably the clusters are roughly of equal size, i.e., (almost-) balanced design.
+\item \textbf{Random horizontal data splitting.} This approach is useful when the number of clusters is large, and preferably the clusters are roughly of equal size, i.e., an (almost) balanced design.
 This will produce several sub-data with smaller sample sizes which are easier to handle.
@@
 In this situation one can take sub-samples of each cluster with a manageable size $n_{f(ri)} < n_{ri}$.
-Obviously, this can be done by taking overlapped (sub-sampling with replacement) or non-overlapped (sub-sampling without replacement) sub-samples.
+Obviously, this can be done by taking overlapping (sub-sampling with replacement) or non-overlapping (sub-sampling without replacement) sub-samples.
 \item \textbf{Structured vertical data splitting.} This approach is useful when the number of outcomes of interest, $m$, is large.
@@
 The last step of this procedure is combining the results obtained from various sub-samples.
-As the models fitted within each sub-sample have some common parameters, each parameter of the initial model could possibly be estimated more than once after second step.
+As the models fitted within each sub-sample have some common parameters, each parameter of the initial model could possibly be estimated more than once after the second step.
 We propose to use a weighted average of them to obtain a unique estimation for each parameter.
 The critical part is to define such appropriate weights.
-They have to be computed proportional to the number of clusters within each sub-sample.
+They have to be computed proportionally to the number of clusters within each sub-sample.
 Also, if needed, size of these clusters should be involved.
@@
 \label{sec_results_so_far}
-We try to present the results obtained so far through the published, accepted, and submitted papers. The working papers and projects will be presented in Section~\ref{sec_progress}.
+We present the results obtained so far through the published, accepted, and submitted papers. The working papers and projects will be presented in Section~\ref{sec_progress}.
 
@@
 \item[--]{\textsf{Lisa Hermans, Vahid Nassiri, Geert Molenberghs, Michael G.
-Kenward, Wim Van der Elst, Marc Aerts, and Geert Verbeke}} , {\it Clusters with random size: maximum likelihood versus weighted estimation}, Submitted to Statistica Sinica.
+Kenward, Wim Van der Elst, Marc Aerts, and Geert Verbeke}}, {\it Clusters with random size: maximum likelihood versus weighted estimation}, Submitted to Statistica Sinica.
 \end{itemize}
@@
 \item[--]{\textsf{Lisa Hermans, Vahid Nassiri, Geert Molenberghs, Michael G.
-Kenward, Wim Van der Elst, Marc Aerts, and Geert Verbeke}} , {\it Fast, Closed-form, and Efficient Estimators for Hierarchical Models with AR(1) Covariance and Unequal Cluster Sizes}, Communications in Statistics - Simulation and Computation, 2017, DOI: 10.1080/03610918.2017.1316395.
+Kenward, Wim Van der Elst, Marc Aerts, and Geert Verbeke}}, {\it Fast, Closed-form, and Efficient Estimators for Hierarchical Models with AR(1) Covariance and Unequal Cluster Sizes}, Communications in Statistics - Simulation and Computation, 2017, DOI: 10.1080/03610918.2017.1316395.
 \end{itemize}
@@
 In this structure, unlike the CS, the correlation between pairs of variables depends on their distance, i.e., the correlation between $i$th and $j$th components is defined as $\sigma^2 \rho^{|i-j|}$ ($\sigma^2>0$, $-1\leq\rho\leq +1$).
-The closed-form, fast and efficient estimates are obtained for a balanced data following Gaussian distribution with AR(1) structured covariance matrix.
+The closed-form, fast and efficient estimates are obtained for balanced data following Gaussian distribution with AR(1) structured covariance matrix.
 The appropriate weights are computed to efficiently combine these estimates and their variances.
@@
 \begin{itemize}
-\item[--]{\textsf{Vahid Nassiri, Anna Ivanova, Geert Molenberghs, and Geert Verbeke}} , {\it Fast precision estimation in high-dimensional multivariate joint models}, To appear in Biometrical Journal.
+\item[--]{\textsf{Vahid Nassiri, Anna Ivanova, Geert Molenberghs, and Geert Verbeke}}, {\it Fast precision estimation in high-dimensional multivariate joint models}, To appear in Biometrical Journal.
 \end{itemize}
@@
 \begin{itemize}
-\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}} , {\it Finite information limit variance-covariance structures: Is the entire dataset needed for analysis?}, In High Performance Computing \& Simulation (HPCS), 2016 International Conference on (pp.
+\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}}, {\it Finite information limit variance-covariance structures: Is the entire dataset needed for analysis?}, In High Performance Computing \& Simulation (HPCS), 2016 International Conference on (pp.
 736-742).
@@
 \end{itemize}
-We have introduced the concept of finite information limit (FIL) estimator in this paper.
+We have introduced the concept of a finite information limit (FIL) estimator in this paper.
 This concerns the situation that amount of information within each cluster is finite.
@@
 \begin{itemize}
-\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}} , {\it A simple, principled rule to determine the number of multiple imputations}, Submitted to The American Statistician.
+\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}}, {\it A simple, principled rule to determine the number of multiple imputations}, Submitted to The American Statistician.
 \end{itemize}
@@
 In case that these sub-samples are taken with replacement, infinite number of sub-samples can be taken.
-Furthermore, for without replacement sub-samples it is very well possible that taking all possible sub-samples needs enormous amount of time.
+Furthermore, for sub-sampling without replacement, it is very well possible that taking all possible sub-samples needs enormous amount of time.
 Thus, the need to use only a part of these sub-samples would emerge.
@@
 \begin{itemize}
-\item[--]{\textsf{Vahid Nassiri, Anik\'{o} Lovik, Geert Molenberghs, and Geert Verbeke}} , {\it On using multiple imputation for factor analysis of incomplete data}, Submitted to Behavior Research Methods.
+\item[--]{\textsf{Vahid Nassiri, Anik\'{o} Lovik, Geert Molenberghs, and Geert Verbeke}}, {\it On using multiple imputation for factor analysis of incomplete data}, Submitted to Behavior Research Methods.
 \end{itemize}
-When it comes to factor analysis, principal component analysis, or other methodlogies dealing with  eigendecomposition of the covariance matrix, analyzing several datasets in place of the original one would cause some problems at the combination step (Step 3 in our proposed procedure).
+When it comes to factor analysis, principal component analysis, or other methodologies dealing with  eigendecomposition of the covariance matrix, analyzing several datasets in place of the original one would cause some problems at the combination step (Step 3 in our proposed procedure).
 The reason lies in the fact that not all of the obtained factor loadings would follow the same order.
@@
 
-In this section, we try to explain the work in progress through briefly describing working papers and projects.
+We now give an overview of work in progress.
 
 \begin{itemize}
-\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}} , {\it Iterative multiple outputation: a simple procedure to deal with large clustered data}, In progress.
+\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}}, {\it Iterative multiple outputation: a simple procedure to deal with large clustered data}, In progress.
 \end{itemize}
-Inspired by finite information limit estimators, we are trying to extend this idea to cases where this property is not hold, or the size of the one sub-sample needed to achieve it is large comparing with available computation power.
+Inspired by finite information limit estimators, we are trying to extend this idea to cases where this property does not hold, or the size of the one sub-sample needed to achieve it is large comparing with available computation power.
 In this situation, we would suggest to take several smaller random sub-samples.
@@
 \begin{itemize}
-\item[--]{\textsf{Vahid Nassiri, Anik\'{o} Lovik, Geert Molenberghs, and Geert Verbeke}} , {\it A simple multilevel factor analysis}, In progress.
+\item[--]{\textsf{Vahid Nassiri, Anik\'{o} Lovik, Geert Molenberghs, and Geert Verbeke}}, {\it A simple multilevel factor analysis}, In progress.
 \end{itemize}
@@
 \begin{itemize}
-\item[--]{\textsf{Vahid Nassiri, Alvaro Florez Poveda, Luis Adrian Quintero Sarmiento, Wim Van der Elst, Geert Molenberghs, and Geert Verbeke}} , {\it Comparing different approaches for mixed-effects modelling of unbalanced clustered data}, In progress.
+\item[--]{\textsf{Vahid Nassiri, Alvaro Florez Poveda, Luis Adrian Quintero Sarmiento, Wim Van der Elst, Geert Molenberghs, and Geert Verbeke}}, {\it Comparing different approaches for mixed-effects modelling of unbalanced clustered data}, In progress.
 \end{itemize}
@@
 \begin{itemize}
-\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}} , {\it Data splitting: an \textsf{R} package to analyze big clustered data.}, In progress.
+\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}}, {\it Data splitting: an \textsf{R} package to analyze big clustered data.}, In progress.
 \end{itemize}
-As it was mentioned a couple of times, for many of the proposed and developed methods \textsf{R} functions are prepared to make the implementation easier for the end-user.
+As mentioned before, for many of the proposed and developed methods \textsf{R} functions have been prepared to make the implementation easier for the end-user.
 We are planning to make generic functions out of them and collect them in the form an \textsf{R} package which will be publicly available.

### 2019-06-25 Final_doctoral_plan03.tex
all files 6/25/2019

--- 2019-06-25 Final_doctoral_plan02.tex
@@
 \label{sec_intro}
-Sample size has always been an issue in statistical analysis.
+Sample size has always been an issue in statistical analyses.
 While for many years small sample sizes were a central issue, in recent years, large sample sizes too might confront the statistical analyst with serious problems.
@@
 Furthermore, while in the simplest designs all measurements are independent, many designs lead to hierarchical (a.k.a., clustered, correlated, dependent) data \citep{laird1982,zeger1986,verbeke2009, molenberghs2005}.
-By cluster here we mean a set of measurements repeatedly collected for one single unit (e.g., subject, household, trial, etc.).
+By `cluster' we mean a set of measurements repeatedly collected for one single unit (e.g., subject, household, trial, etc.).
 Therefore, the number of repetitions refers to the number of measurements available per cluster.
@@
 Some important examples of clustered data are repeated measurements, longitudinal data, multi-center studies, and meta-analysis.
-Accounting for such correlations remains a challenge today, when the total number of clusters is large and/or there are a large number of repetitions per cluster.
+Accounting for correlations remains a challenge today, when the total number of clusters is large and/or there are a large number of repetitions per cluster.
 This would sometimes result in prohibitive computations.
@@
 Consider a study which measures (at most) $m$ outcomes for $N$ clusters, each $n_i$ times.
-Let $y_{rij}$ be the $j$th measurement taken on the $i$th cluster for the $m$th outcome, $i=1,\ldots,N$, $r=1,\ldots,m$ and $j=1,\ldots,n_{ri}$.
+Let $y_{rij}$ be the $j$th measurement taken on the $i$th cluster for the $r$th outcome, $i=1,\ldots,N$, $r=1,\ldots,m$ and $j=1,\ldots,n_{ri}$.
 A general scheme of such data is presented in Figure~\ref{fig_scheme}.
-As it was mentioned, fitting a model to such data could become challenging when the size of dataset make it eligible for the title \emph{big data}.
+As it was mentioned, fitting a model to such data could become challenging when the size of dataset makes it eligible for the title \emph{big data}.
 A dataset could be entitled as \emph{big data} in different situations.
@@
 
-In this PhD research, such situations are considered and we aim to propose methodologies to deal with such problems using the so-called data splitting.
-In a nutshell, data splitting means to split the difficult (sometimes infeasible) job of modelling the big data by splitting it into smaller chunks of manageable sub-data, analyze each of them separately, and then combine the results of these analyses using appropriate combinations rules.
-This would break a difficult task into several, often so-called embarrassingly parallel, simpler tasks.
+In this PhD research, we aim to propose methodologies to deal with such problems using the so-called data splitting.
+In a nutshell, data splitting means to split the difficult (sometimes infeasible) job of modelling the big dataset by splitting it into smaller chunks of manageable sub-data, analyze each of them separately, and then combine the results of these analyses using appropriate combinations rules.
+This would break a difficult task into several, often so-called embarrassingly parallel by computer scientists, simpler tasks.
 
@@
 \item The developed framework should be as independent as possible from the modelling approach.
-In other words, it should be compatible with various modelling approaches, e.g.,  linear, generalized linear, non-linear, etc.
+In other words, it should be compatible with various modelling approaches, e.g.,  linear, generalized linear, non-linear models, etc.
 \item As the end-user of this product will be researchers of different fields with different tastes of data analysis systems who are dealing with big data, the developed methodologies should preferably be easy to understand and implement in different platforms, e.g., \textsf{R}, \textsf{SAS}, \textsf{Matlab}, \textsf{Python}, \textsf{Julia}, etc.
@@
 For example, a dataset would consist of a large number of clusters (large $N$), and each of these clusters could be large as well (large $n_{ri}$'s), see Amazon book ratings in Section~\ref{sec_data_sets}.
-Or large number of outcomes of interest can happen with large number of clusters, see divorce in Flanders dataset in Section~\ref{sec_data_sets}.
+Also, large number of outcomes of interest can happen with large number of clusters, see divorce in Flanders dataset in Section~\ref{sec_data_sets}.
 Therefore, our proposed methodologies for each situation should have the possibility of being combined to deal with a combination of the mentioned situations.
@@
 Evidently, fetuses are clustered within mothers; hence the implied association needs to be accommodated in the analysis.
-As each mother has different number of fetuses, we have clusters of different sizes.
+As each mother has a different number of fetuses, we have clusters of different sizes.
 
-\paragraph*{Divorce in Flanders:} This dataset is collected at Centre for Sociological Research, KU Leuven. The data concerned on studying and comparing the personality among
+\paragraph*{Divorce in Flanders:} This dataset is collected at the Centre for Sociological Research, KU Leuven. The data concerned on studying and comparing the personality among
 three generations in about 4000 Flemish families.
@@
 These scores are built upon 5 factors which in total are measured through 44 items.
-The usual factor analysis would lose most of the information in the data.
+The usual factor analysis would lose most of the hierarchical information in the data.
 Therefore, an idea is to model different combinations of the 5 aspects of personality for each family role using a mixed model.
@@
 By definition, a hearing threshold is the lowest signal intensity possible to perceive by a subject at a specific frequency.
-In a study considered in \cite{verbeke2009} and \cite{Verbeke2006}, hearing threshold measured in 11 different frequencies between 125 Hz and 8000 Hz for left and right ears obtained on 603 male participants from the Baltimore Longitudinal Study of Ageing, BLSA.
+In a study considered in \cite{verbeke2009} and \cite{Verbeke2006}, hearing thresholds measured in 11 different frequencies between 125 Hz and 8000 Hz for left and right ears obtained on 603 male participants from the Baltimore Longitudinal Study of Ageing, BLSA.
 The number of visits for each subject which varies between 1 to 15 are unequally spaced.
@@
 Because not all subjects received treatment at the same time points and, not the same amount, the final dataset is unbalanced.
-Our analysis shows AR(1) structure could be a good candidate for modelling the covariance matrix of the random effects, which is useful to illustrate some of the methodologies we have developed.
+Our analysis shows the AR(1) structure could be a good candidate for modelling the covariance matrix of the random effects, which is useful to illustrate some of the methodologies we have developed.
 
@@
 \item Splitting the data.
-\item Performing the analysis on each sub-data.
-\item Efficiently combine the results from each sub-data to form an overall estimate with its precision.
+\item Performing the analysis on each sub-dataset.
+\item Efficiently combine the results from each sub-dataset to form an overall estimate with its precision.
 \end{enumerate}
@@
 
-The first step of this procedure is determined by type of the big data we are dealing with.
+The first step of this procedure is determined by the type of big data we are dealing with.
 We divide the data splitting methods based on two main characteristics: 1- independent or dependent sub-samples, 2- random or structured sub-sampling.
@@
 Considering the assumption of having independent clusters, this will result in independent sub-samples.
-Unless the sub-samples are overlapped, i.e., one subject appears in more than one sub-sample\footnote{Note that, while perfectly possible, the overlapped horizontal splitted sub-samples are not considered in this thesis.
+Unless the sub-samples are overlapping, i.e., one subject appears in more than one sub-sample\footnote{Note that, while perfectly possible, the overlapping horizontally splitted sub-samples are not considered in this thesis.
 One can look in vast literature of bootstrap methods for such an approach.}.
@@
 \item \textbf{Random horizontal data splitting.} This approach is useful when the number of clusters is large, and preferably the clusters are roughly of equal size, i.e., an (almost) balanced design.
-This will produce several sub-data with smaller sample sizes which are easier to handle.
+This will produce several sub-datasets with smaller sample sizes which are easier to handle.
 \item \textbf{Structured horizontal data splitting.} This approach is useful when some specific clusters are more convenient to deal with when they are together.
 Therefore, instead of taking random sub-samples, the sub-sampling can be done based on this pre-specified structure.
-An important example of this approach which is considered in this PhD research is splitting based on the cluster sizes, i.e., forming sub-samples consist of clusters of equal size.
+An important example of this approach which is considered in this PhD research is splitting based on the cluster sizes, i.e., forming sub-samples consisting of clusters of equal size.
 This would lead to balanced sub-samples.
-We have shown existing of closed-form solutions, as well as, faster convergence can be achieved using this approach.
+We have shown the existence of closed-form solutions, leading to faster convergence of model fitting procedures.
 \item \textbf{Random vertical data splitting.} This approach is useful when we are dealing with data with large number of measurements for each cluster per outcome (large $n_{ri}$'s).
@@
 \end{itemize}
-Obviously, according to type of big data one is dealing with, all of the the combinations of random/structured and horizontal/vertical data splitting can be freely combined.
+Obviously, according to the type of big data one is dealing with, all of the the combinations of random/structured and horizontal/vertical data splitting can be freely combined.
 The combination rules then should be combined accordingly.
@@
 For example, consider 3 outcomes of interest of types continuous, binary, and ordinal.
-If sub-samples are taken using a structured vertical splitting approach, and each sub-sample consists of pairs of outcomes, the model fitted to each sub-sample will depend on the type of the responses presenting there.
+If sub-samples are taken using a structured vertical splitting approach, and each sub-sample consists of pairs of outcomes, the model fitted to each sub-sample will depend on the type of the responses present in the pair at hand.
 Furthermore, the explanatory variables of the models within each sub-sample does not need to be identical across sub-samples.
@@
 The closed-form estimates and efficient combination weights are also implemented in \textsf{R}.
-Simulation results show the derived closed-forms reduce the computation time up to 30,000 times comparing with iterative procedures in \textsf{Proc MIXED} in \textsf{SAS}.
+Simulation results show the derived closed-forms reduce the computation time up to 30,000 times compared to iterative procedures in \textsf{Proc MIXED} in \textsf{SAS}.
 
@@
 This paper considers the same problem as in the previous one, except here the covariance structure is assumed to be autoregressive of order 1, AR(1).
-In this structure, unlike the CS, the correlation between pairs of variables depends on their distance, i.e., the correlation between $i$th and $j$th components is defined as $\sigma^2 \rho^{|i-j|}$ ($\sigma^2>0$, $-1\leq\rho\leq +1$).
+In this structure, unlike the CS, the correlation between pairs of observations depends on their distance, i.e., the correlation between $i$th and $j$th components is defined as $\sigma^2 \rho^{|i-j|}$ ($\sigma^2>0$, $-1\leq\rho\leq +1$).
 The closed-form, fast and efficient estimates are obtained for balanced data following Gaussian distribution with AR(1) structured covariance matrix.
@@
 These closed-form estimates and efficient combination weights are also implemented in \textsf{R}.
-Simulation results show using the derived closed-forms, the computation time reduced up to 700 times comparing with iterative procedures in \textsf{Proc MIXED} in \textsf{SAS}.
+Simulation results show using the derived closed-forms, the computation time reduced up to 700 times compared to iterative procedures in \textsf{Proc MIXED} in \textsf{SAS}.
 
@@
 The proposed methodology was also implemented in \textsf{R}.
-Simulation results shows up to 2500 times faster computations comparing with existing procedures.
+Simulation results shows up to 2500 times faster computations compared to existing procedures.
 
 \begin{itemize}
-\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}}, {\it Finite information limit variance-covariance structures: Is the entire dataset needed for analysis?}, In High Performance Computing \& Simulation (HPCS), 2016 International Conference on (pp.
+\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}}, {\it Finite information limit variance-covariance structures: Is the entire dataset needed for analysis?}, In High Performance Computing \& Simulation (HPCS) 2016 International Conference (pp.
 736-742).
@@
 We have introduced the concept of a finite information limit (FIL) estimator in this paper.
-This concerns the situation that amount of information within each cluster is finite.
+This concerns the situation that the amount of information within each cluster is finite.
 Therefore, after some point adding new observations would not substantially improve efficiency.
-This means, in case of FIL estimators, one only needs to analyze a small fraction of data to obtain results as efficient as analyzing the full dataset.
+This means, in case of FIL estimators, one only needs to analyze a small fraction of the data to obtain results as efficient as analyzing the full dataset.
 This concept is introduced in this paper, and some procedures are proposed to detect this property as well as determining the size of this fraction.
-Comparing with fitting the model to the full data, simulation results show about 1500 times faster computations using this idea.
+Compared to fitting the model to the full data, simulation results show about 1500 times faster computations using this idea.
 
@@
 Of course, as this problem is similar to the problem of determining the number of imputations when dealing with missing data using multiple imputations, the paper was written with that topic as a target.
-But the idea is general and can be applied in many situations.
+The idea is general and can be applied in many situations.
 It has been used in the FIL paper, and will be used in the IMO paper (see Section~\ref{sec_progress}).
@@
 
-We now give an overview of work in progress.
+We now give an overview of work in progress with a provisional timeline, see Figure~\ref{fig_gant}.
 
@@
 \end{itemize}
-Inspired by finite information limit estimators, we are trying to extend this idea to cases where this property does not hold, or the size of the one sub-sample needed to achieve it is large comparing with available computation power.
+Inspired by finite information limit estimators, we are trying to extend this idea to cases where this property does not hold, or the size of the one sub-sample needed to achieve it is large compared to available computation power.
 In this situation, we would suggest to take several smaller random sub-samples.
@@
 Some simulations are already performed, also \textsf{R} functions are written to efficiently take the sub-samples and combine the results of analyses performed on them.
-Works are on going to apply this methodology on some of our motivating data.
-The paper would hopefully be submitted before beginning of next academic year.
+Work is ongoing to apply this methodology on some of our motivating data.
+We expect to submit the paper before beginning of next academic year.
 
@@
 This paper considers this problem and try to propose a solutions based on random vertical splitting.
-The manuscript is in its final levels and we expect it to be submitted within this academic year.
+The manuscript is almost finalized and we expect it to be submitted within this academic year.
 

### 2019-06-25 Final_doctoral_plan04.tex
all files 6/25/2019

--- 2019-06-25 Final_doctoral_plan03.tex
@@
 \end{itemize}
-When it comes to factor analysis, principal component analysis, or other methodologies dealing with  eigendecomposition of the covariance matrix, analyzing several datasets in place of the original one would cause some problems at the combination step (Step 3 in our proposed procedure).
+When it comes to factor analysis, principal component analysis, or other methodologies dealing with eigenvalue-eigenvector decomposition of the covariance matrix, analyzing several datasets in place of the original one would cause some problems at the combination step (Step 3 in our proposed procedure).
 The reason lies in the fact that not all of the obtained factor loadings would follow the same order.

### 2019-06-25 appendix.tex
all files 6/25/2019

\section{Fieller's method and Delta method}
\label{app1}

Here, the Fieller's method to construct a confidence interval for (\ref{nf_var_min_cs}) as well as the Delta method  are discussed. The ratio which we require a confidence interval for is:
$$n_f=\frac{\hat{\sigma^2}}{\widehat{\tau} \varepsilon}.$$
Suppose $n_1$ is a small-feasible sub-sampling size chosen to estimate the parameters in the model (e.g., $n_1=5$). From \cite{Iddi2011}, one finds:

\begin{multline}
%\label{cov_sigma2_tau}
\mathrm{Var}  \left(
\begin{array}{c}
\widehat{\sigma^2}\\
\widehat{\tau\varepsilon}
\end{array} \right) =  \\ \frac{2 \sigma^4}{Nn_1(n_1-1)}  \left(
\begin{array}{cc}
n_1 & -\varepsilon \\
-\varepsilon & \varepsilon^2\frac{\sigma^4 + 2(n_1-1) \tau \sigma^2 + n_1 (n_1-1) \tau^2}{\sigma^4}
\end{array} \right) \\ = \left(
\begin{array}{cc}
s_{11} &  s_{12} \\ s_{12} & s_{22}
\end{array} \right).
\end{multline}

The Fieller's confidence interval for (\ref{nf_var_min_cs}) can be calculated following these steps.
\begin{equation}
\begin{aligned}
C_1^2 =&\frac{s_{11}}{\widehat{\sigma^2}^2}, \\
C_2^2=& \frac{s_{22}}{\left(\varepsilon\widehat{\tau}\right)^2},\\
r= & \frac{s_{12}}{\sqrt{s_{11} s_{22}}},\\
A= &C_1^2 + C_2^2 - 2r C_1C_2,\\
B=& t^2 C_1^2 C_2^2 (1-r^2),\\
L=& \frac{\widehat{\sigma^2}}{\varepsilon\widehat{\tau}} \frac{1-z_{\alpha/2}^2 r C_1C_2 - z_{\alpha/2}\sqrt{A-B}}{1-z_{\alpha/2}^2C_2^2},\\
U=& \frac{\widehat{\sigma^2}}{\varepsilon\widehat{\tau}} \frac{1-z_{\alpha/2}^2 r C_1C_2 + z_{\alpha/2}\sqrt{A-B}}{1-z_{\alpha/2}^2C_2^2}.
\end{aligned}
\end{equation}
One may take $n_f=\max\{L,U\}$. Of course, in case were $z_{\alpha/2}^2 C_2^2 <1$, there would be no finite interval. As it was mentioned, one may also use the Delta method as follows:

\begin{equation}
\label{var_delta_method}
\mathrm{Var}(\frac{\widehat{\sigma^2}}{\varepsilon \widehat{\tau}}) \approx \frac{1}{\varepsilon^2 \widehat{\tau}^2} \left( s_{11} - 2 \frac{\sigma^2}{\tau} s_{12} + \frac{\sigma^4}{\tau^2} s_{22} \right).
\end{equation}

\section{Random vertical data splitting for compound-symmetry structure}
\label{app2}
Assume a single cluster of size $N$ with multivariate normal distribution and compound-symmetry structure for its covariance matrix, see Section~\ref{sec_CS_effect}. Suppose we take $M_0$ sub-samples of size $n_0$ from this cluster. Following \cite{hoffman2001}, the parameter $\mu$ should be estimated within each sub-sample, $\widehat{\mu}^{(m)}$ ($m=1,\ldots,M_0$) and then the overall estimate can be obtained by averaging these estimates. Furthermore, one needs the within and between sub-samples variabilities to compute the overall variability, see (\ref{mo_cov}).

Within each sub-sample the variance of $\widehat{\mu}^{(m)}$ can be computed using the variance formula for a mean estimator in CS-multivariate normal \cite{Iddi2011} :

\begin{equation}
\label{within}
S_W= \frac{\sigma^2}{n_0} \left[ 1+ (n_0-1) \rho \right ].
\end{equation}
Estimating the between variability requires computing $\mathrm{E}\left\{ (\widehat{\mu}^{(m)} - \bar{\mu})^2 \right\}$, where $\bar{\mu}$ is the average of $\widehat{\mu}^{(m)}$'s ($m=1,\ldots,M_0$). Set $S_m$ as the index of members of sub-sample $m$ and $S_{m'}$ the rest of the sample, then,
\begin{equation*}
\widehat{\mu}^{(m)} - \bar{\mu} = \frac{1}{n_0}\left\{(1-\frac{1}{M_0})\sum_{i\in S_m} y_i - \frac{1}{M_0} \sum_{m=1}^ {M_0} \sum_{i\in S_{m'}} y_i \right\}.
\end{equation*}
Therefore,
\begin{equation}
\label{T1_T4}
\mathrm{E}\left\{(\widehat{\mu}^{(m)} - \bar{\mu})^2 \right\} = \frac{1}{n_0} \left( T_1 + T_2 + T_3+ T_4\right),
\end{equation}
where,
\begin{equation}
\label{eq_T1_T4}
\begin{cases}
T_1= \left( 1- \frac{1}{M_0} \right) \mathrm{E} \left [ \sum_{i\in S_m} y_i \right]^2\\
T_2= \frac{1}{M_0^2} (M_0-1) \mathrm{E} \left[ \sum_{i \in S_m} y_i \right]^2\\
T_3=  -2 \left(1-\frac{1}{M_0} \right) \frac{1}{M_0} (M_0-1) \mathrm{E} \left\{ \left(\sum_{i\in S_{m'}} y_i \right) \left( \sum_{i'\in S_{m'}} y_{i'} \right) \right\} \\
T_4= 2 \frac{1}{M_0} \frac{(M_0-1) (M_0-2)}{2} \mathrm{E} \left\{ \left(\sum_{i \in S_m} y_i \right) \left(\sum_{i'\in S_{m'}} y_{i'} \right) \right\}.
\end{cases}
\end{equation}
From (\ref{eq_T1_T4}) one may find,
\begin{equation}
\label{sum_T}
\begin{cases}
T_1+ T_2 = \frac{M_0-1}{M_0} n_0 \sigma^2 \left [1+ (n_0-1) \rho \right]\\
T_3 + T_4= - \frac{M_0-1}{M_0} \frac{\sigma^2 n_0^2}{N} \left[ 1+ (N_1) \rho \right].
\end{cases}
\end{equation}
Therefore, the between variability can be computed as follows,
\begin{equation}
\label{between}
S_B=\mathrm{E} \left\{ (\widehat{\mu}^{(m)} - \bar{\mu} )^2 \right\} = \frac{M_0-1}{M_0} \left[\frac{1+(n_0-1)\rho}{n_0} - \frac{1+(N_1)\rho}{N} \right].
\end{equation}
Using (\ref{within}) and (\ref{between}), the total variability can be computed as follows,
\begin{equation}
\label{total_variability}
S_T= \frac{1}{M_0} \sigma^2 \frac{1+ (n_0-1) \rho }{n_0} + \frac{M_0-1}{M_0} \sigma^2 \frac{1+ (N-1) \rho}{N}
\end{equation}
Comparing the variance in (\ref{total_variability}) with the variance when estimating $\mu$ using the full sample would give,
\begin{equation}
\label{ARE}
\mathrm{ARE} = \frac{M_0-1}{M_0} + \frac{1}{M_0} \frac{N}{n_0} \frac{1+(n_0-1) \rho}{1+ (N-1) \rho}
\end{equation}
For example, if sub-sampling is done only once ($M_0=1$), then,
\begin{equation}
\label{ARE_M1}
\mathrm{ARE} = \frac{N}{n_0} \frac{1+(n_0-1) \rho}{1+(N-1) \rho},
\end{equation}
The same calculation is possible for $M_0=2$,
\begin{equation}
\label{are_M02}
\mathrm{ARE}= \frac{1}{2} + \frac{1}{2} \frac{N}{n_0} \frac{1+(n_0-1) \rho}{1+ (N-1) \rho}.
\end{equation}
By considering the desired ARE as $1+\epsilon$, one may find $n_0$.

\section{Combination rule for $p$-values in LES dataset analysis (IMI)}

The Kruskal-Wallis test statistic asymptotically follows a $\chi^2$ distribution. Therefore, in order to find the combined $p$-value, we may follow the procedure proposed by \cite{li1991}. More details can be found in \cite{rubin2004} and \cite{enders2010}. The combination is done using {\tt{micombine.chisquare}} in package {\tt{miceadds}} in \textsf{R}, \cite{miceadds}.

\begin{itemize}
	\item \textbf{Averaging.} Compute the test statistic for each imputed data and take their average: $\overline{\chi^2} = \frac{1}{M} \sum_{m=1}^M \chi^2_m$.
	\item \textbf{Relative variance increase.} $r=\left(1+\frac{1}{M} \right) \frac{\sum_{m=1}^M (\sqrt{\chi^2_m} - \sqrt{\overline{\chi^2}})^2}{M-1}$
	\item \textbf{Test statistic.} $D=\frac{\frac{\overline{\chi^2}}{\kappa} - \frac{M+1}{M-1}r}{1+r}$, where $\kappa$ is the degrees-of-freedom of the $\chi^2_m$. For a $\chi^2$-test of independence, it is $(k_1-1)(k_2-1)$ where $k_1$ and $k_2$ are the number of levels of two categorical variables. Take five diagnosis groups and variable \textsf{Eye} with two levels, then $\kappa=(5-1)(2-1)=4$. For the Kruskal-Wallis test, $\kappa$ is the number of groups minus 1. So, when comparing values of a continuous variable among three types of diagnosis, $\kappa=2$. When this comparison is done for all five types, $\kappa=4$. 
	\item \textbf{Computing $p$-value.} $p$-value$=P(F_{\kappa,\nu} > D)$, where $F$ is the $F$-distribution and $\nu=\kappa^{-3/M} (M-1) \left(1+\frac{1}{r}\right)^2$
\end{itemize}

\section{The surrogate model}
The model takes the form:
\begin{equation}
\label{model}
\left\{
\begin{array}{l}
S_{ij}=\mu_S + m_{S_i} + (\alpha + a_i) Z_{ij} + \varepsilon_{S_{ij}},\\
T_{ij}= \mu_T + m_{T_i} + (\beta + b_i) Z_{ij} + \varepsilon_{T_{ij}},
\end{array} \right.
\end{equation}
where $S_{ij}$ and $T_{ij}$ are the surrogate and true endpoints, and $Z_{ij}$ is the treatment indicator (1 for active; 0 for placebo), respectively. Index $i$ refers to the trial and $j$ to the subject. Further, $\mu_{S}$ and $\mu_{T}$ are fixed intercepts and $m_{S_i}$ and $m_{T_i}$ are  random intercepts, for surrogate and true endpoint, respectively. The fixed treatment effects are $\alpha$ (for $S$) and $\beta$ (for $T$), and the centre-specific treatment effects are $a_i$ (for $S$) and $b_i$ (for $T$). 
Random intercepts and slopes are assumed to follow
$(m_{S_i}, m_{T_i}, a_i, b_i)'\sim N(0,D)$,
where 
\begin{equation}
\label{D}
D = \left(
\begin{array}{cccc}
d_{SS} & & &\\
d_{ST} & d_{TT} & &\\
d_{Sa} & d_{Ta} & d_{aa} &\\
d_{Sb} & d_{Tb} & d_{ab} & d_{bb}
\end{array} \right).
\end{equation}
Furthermore, for the error terms,
\begin{equation}
\label{Sigma}
\left(\begin{array}{c}
\varepsilon_{S_{ij}}\\
\varepsilon_{T_{ij}}
\end{array} \right)
\sim N(0,\Sigma),\; \Sigma = \left(
\begin{array}{cc}
\sigma_{SS} & \\
\sigma_{ST} & \sigma_{TT}
\end{array} \right).
\end{equation}
The parameters of interest for the surrogate evaluation are trial- and individual-level coefficients of determination, $R^2_{\mbox{\tiny trial}}$ and $R^2_{\mbox{\tiny indiv}}$, respectively:
\begin{equation}
\label{R2}
R^2_{\mbox{\tiny trial}} = \frac{1}{d_{ab}}
\left(
\begin{array}{c}
d_{Sb}  \\
d_{ab}
\end{array} \right)^{\prime} 
\left(
\begin{array}{cc}
d_{SS} & d_{Sa}\\
d_{Sa} & d_{aa}
\end{array} \right)^{-1}
\left(
\begin{array}{c}
d_{Sb}  \\
d_{ab}
\end{array} \right),\quad R^2_{\mbox{\tiny indiv}}=\frac{\sigma_{ST}^2}{\sigma_{SS} \sigma_{TT}}.
\end{equation}

### 2019-06-25 imi.tex
all files 6/25/2019

\subsection{Introduction}

Since Rubin's seminal work on multiple imputation (MI) \citep{rubin1978,rubin1979,rubin1987},
the method has been broadly applied, methodologically extended, and expanded towards ever more areas of application  \citep{carpenter2008,buuren12,carpenter2012,wim2015}.  {\color{black}{Multiple imputation is a commonly used approach to analyze incomplete data. A growing literature and increasing number of software implementations have contributed to the spread of the method. One of the attractions of MI is its very good to excellent performance even with a relatively small number of imputed datasets. This was important when Rubin created the method, about forty years ago, in view of, among others, the US Census. It still is today because of ever increasing data streams.}}

Broadly, MI replaces a missing value with several plausible values, sampled from an appropriate predictive distribution for the missing values, given observed information. The method produces several completed datasets to replace the initial partially observed dataset.

An evident practical question is how many {\color{black}{imputed datasets}}, $M$ say, are sufficient for reliable results, knowing that full efficiency would be reached for $M=+\infty$. Precision should be balanced against computational expense. It has been stated repeatedly, and practically confirmed, that a small number of imputations oftentimes gives very acceptable results. Sources like \cite{rubin1987} and the classic \cite{littlerubin2002} quote values as low as 2--5. While attractive, especially when faced with large and complex databases, such low numbers may not always apply. Features that would require larger numbers of imputation include: increasing amounts of missing information, and the need for hypothesis testing rather than merely parameter and precision estimation. It is reassuring that \cite{schafer1997} indicated that, even under undesirable circumstances, it is rare for $M$ needing to be in excess of twenty. This observation was based on extensive simulation, rather than theory.

Some of the approaches for dealing with selecting the number of imputations are reviewed in Section~\ref{sec_review}. Our proposal to handle the choice is presented in 
Section~\ref{sec_proposal}. Section~\ref{sec_simulations} reports some simulations, while real-life applications are offered in Section~\ref{sec_case_studies}.

\subsection{Number of imputed datasets: A review}
\label{sec_review}
Upon producing multiply imputed datasets, a standard analysis is routinely applied to each of the completed datasets. Let $\bftheta$ be the parameter vector of interest and  $\widehat{\bftheta}_m$  its estimate from the $m$th imputed dataset. If $M$ is the number of {\color{black}{imputed datasets}}, then $\tilde{\bftheta}$, the multiple imputation estimator is defined as:
\begin{equation}
\label{est}
\widetilde{\bftheta}=\frac{1}{M} \sum_{m=1}^M \widehat{\bftheta}_m.
\end{equation}
If $\widehat{\Sigma}_m$ is the estimated variance-covariance of $\widehat{\bftheta}_m$, then the within-imputation variability is:
\begin{equation}
\label{M}
\widehat{W}=\frac{1}{M}\sum_{m=1}^M \widehat{\Sigma}_m,
\end{equation}
and the between-imputation variability is: 
\begin{equation}
\label{B}
\widehat{B} = \frac{1}{M-1}\sum_{m=1}^M \left (\widehat{\bftheta}_m - \widetilde{\bftheta} \right)\left (\widehat{\bftheta}_m - \widetilde{\bftheta} \right)'.
\end{equation}
Then, asymptotically, for the sample size as well as $M$ going to infinity, we have 
\begin{equation}
\label{large_M}
\widetilde{\bftheta}\sim N(\bftheta,\widehat{B}+\widehat{W}).
\end{equation} 
For finite $M$, the variance of the normal distribution in (\ref{large_M}) becomes \citep{littlerubin2002}:
\begin{equation}
\label{finite_M}
\widehat{V}=(1+M^{-1}) \widehat{B} + \widehat{W}.
\end{equation}
The term $\widehat{B}/M$ takes into account the increased variability stemming from finite $M$. Obviously, for $M\rightarrow \infty$, this extra term vanishes. Consider
{\color{black}{$r=\frac{1}{q}\left(\mathrm{tr}(\widehat{B}\widehat{W}^{-1})\right)(1+M^{-1})$ and $\nu = (M-1)(1+r^{-1})^2$ with $\mathrm{tr}(A)$ indicating the trace of the matrix $A$, and $q$ the length of parameter vector $\bftheta$}}. For the $k$th element of $\bftheta$ we have:
\begin{equation}
\label{theta_approx}
\widetilde{\theta_k} \approx \theta_k+ t_{\nu} \sqrt{\widehat{V_{kk}}},
\end{equation}
where $t_{\nu}$ is Student's $t$ distribution with $\nu$ degrees-of-freedom and $\widehat{V}$ can be computed using (\ref{finite_M}). These expressions are derived and discussed in detail by  \cite{rubin1987} or \cite{carpenter2012}. The degrees-of-freedom $\nu$ are computed by \cite{rubin1986_df}. Small sample degrees-of-freedom are reviewed by \cite{wagstaff2011}. Note that (\ref{theta_approx}) applies to the scalar case.

Now, if $I_M$ is the information based on $M$ imputations and $I_{\infty}$ is the information for $M\rightarrow+\infty$, then,
\begin{equation}
\label{relative_information}
\frac{I_M}{I_{\infty}} = \left( 1+ \frac{\gamma}{M} \right) ^{-1},\; \gamma=\frac{r+2/(\nu+3)}{r+1},
\end{equation}
where $\gamma$ in (\ref{relative_information}) can be regarded as the fraction of missing information. \cite{rubin1987} suggested that for many applications with a moderate amount of missingness, 3--5 imputations might well be sufficient. For example, using these expressions, with $10\%$ missingness, $M=3$ would provide about $97\%$ efficiency when compared with an infinite number of {\color{black}{imputed datasets}}. It is not surprising that this rule of thumb is relatively broadly accepted in practice. For example, according to SAS/STAT(R) 9.4 User's Guide, the default number of imputations in the MI procedure is set equal to 5.

 {\color{blue}{However, in the latest version of SAS (SAS/STAT(R) 14.1)  the default number of imputed datasets ({\tt{NIMPUTE}}) is increased to 25. Furthermore, following \cite{white2011}, an option is provided in this version of SAS' {\tt{PROC MI}} to set the number of imputed datasets equal to the fraction of incomplete cases ({\tt{NIMPUTE=PCTMISSING}}). This indicates a level of acceptance among commercial software developers for the need for larger $M$'s, also alternative procedures to determine it.}}

A practical difficulty with this heuristic is the need for $\gamma$. 
The quantity should be estimated to begin with, but obviously, the quality of this estimate is hugely affected by the number of {\color{black}{imputed datasets}} itself. Furthermore, as \cite{bodner2008} also suggests, the rule for the number of {\color{black}{imputed datasets}} would take into account three parameters:  $\widetilde{\bftheta}$, its variance, and the degrees-of-freedom of the Student's $t$ distribution. When one is interested in controlling the $p$-values when conducting hypothesis tests, or other quantities, then perhaps more imputations are needed \citep{harel2003,carpenter2008}.

\cite{carpenter2012} distinguished between two cases. Their proposal is to use a small number of {\color{black}{imputed datasets}} when the inference is clear-cut, but if it is less clear-cut and an accurate estimate of the $p$-value or $\gamma$ is needed, they proposed to set $M=100$.

Given that for small $M$, the approximation in (\ref{theta_approx}) may not be accurate enough, \cite{royston2004} proposed an iterative procedure to select $M$ such that the confidence level remains at a selected level. He proposed to select $M$ such that the coefficient of variation of $t_{\nu}\sqrt{\widehat{V_{kk}}}$ for the worst-case parameter becomes less than the type I error rate $\alpha$; this author used the conventional $\alpha=0.05$.

%\cite{hershberger2003} have derived a rather surprising result, motivated by structural equations problems. They treated the number of imputations as following a standard uniform distribution.  They brought into the equation a measure of required precision, much like in survey sampling theory: they placed a cap on the difference between the ``true" and predicted number of imputations. In their logic, with a confidence level of 95\% and an error between true and predicted number of imputations at most 10\%, they found that 510 imputations are needed. We consider their method less practically relevant because important aspects, such as the amount of missing information, target of inference, etc., are not taken into account. \cite{von2005} has refuted the proposal of \cite{hershberger2003} as well, but we mention it here as an example of different ways of looking at the problem of determining the number of imputations.

\cite{graham2007} have performed a simulation study to investigate the effect of $M$ on various characteristics, such as power, mean squared error, fraction of missingness, etc. \cite{bodner2008} explored various factors that are affected by increasing $M$. His findings were used in \cite{lu2014} who considered the case of longitudinal data and then focused on determining $M$ to stabilize MI-based inferences. Stability in these articles is defined in terms of four conditions regarding the conditional standard errors of the MI estimator, the test statistics, the missingness fraction, as well as the coefficient of variation of the half confidence interval. {\color{blue}{Under some circumstances, these authors proposed to use as many as $200$ imputed datasets.}}

{\color{black}{\cite{royston2009} proposed a jackknife procedure \citep{efron1981} to estimate $\sqrt{B/M}$, the so-called Monte Carlo error. These authors then propose to select the sufficient number of imputed datasets based on achieving a pre-defined level of precision. Their approach is implemented in the command {\ttfamily{mim}} in \textsc{Stata}.}}

\subsection{Number of {\color{black}{imputed datasets}}: an alternative proposal}
\label{sec_proposal}
As we have seen, in most of the proposed rules for choosing $M$, the comparison is always between characteristics under a given, finite $M$ on the one hand, and $M\rightarrow+\infty$ on the other. In contrast, our proposal is to compare the quantity of interest with its successor (i.e., under $M$ versus $M+1$). As the estimated parameter and its variance using MI will converge to some asymptotic values as $M\rightarrow\infty$, monitoring this convergence is insightful. This implies an iterative perspective, which is convenient for practice. {\color{black}{We call this procedure the iterative multiple imputation (imi).}} The {\color{black}{imi}} procedure is formulated as follows:
\begin{enumerate}
{\color{black}{\item \textbf{Start.} Select an initial number of imputed datasets, $M_0$, {\color{blue}{$\widetilde{\bftheta}_{M_0} = \sum_{i=1}^{M_0} \widehat{\bftheta}_i/ M0.$}}
\item \textbf{Update.} For $m>M_0$,}} \begin{equation}
\label{update}
\widetilde{{\bftheta}}_{m+1}=\frac{m\widetilde{\bftheta}_m + \widehat{\bftheta}_{m+1}}{m+1}.
\end{equation}
\item \textbf{Distance.} Compute: $d_{m+1}=d(\widetilde{\bftheta}_{m+1},\widetilde{\bftheta}_{m})$ using an appropriate distance.
\item \textbf{Stopping rule.} {\color{black}{$d_{j} < \varepsilon$ for $j=m+1,\ldots,m+k_0$.}}
\end{enumerate}
{\color{black}{Here, $M_0$ is an integer indicating the initial number of imputations. For $M_0=2$, the stopping rule will be examined from the beginning, but in situations where the user knows a minimum number of imputations are needed (based on proportion of missing data, etc.) a larger $M_0$ can be used.}}  $\widehat{\bftheta}_m$ is the estimated parameter in the $m$-th iteration and $\widetilde{\bftheta}_m=\sum_{i=1}^ M \widehat{\bftheta}_m / M$. Note that $\widetilde{\bftheta}$ can be replaced with other quantities of interest, e.g., $p$-values. Clearly, for various quantities of interest, the corresponding combination rule should be applied in (\ref{update}). {\color{black}{Since the convergence of this procedure is not monotone, one needs to determine an integer $k_0$ as the number of successive steps that the stopping rule should be validated. This would prevent an early declaration of convergence. Again, $k_0=1$ would stop the first time this criterion is met, but a larger $k_0$ is recommended. Our proposal is to use $k_0=3$ (see Sections \ref{sec_simulations} and \ref{sec_conclusions}).

As can be seen, the number of imputed datasets in the proposed iterative procedure has two components: a deterministic part: $M_0+k_0$ and a stochastic part which is determined by the iterations. Our implementation of this procedure in \textsc{R} package {\ttfamily{imi}} makes it possible to determine these two parameters interactively after observing the computation time of imputing the dataset and fit the model for $M0=2$ times.}}

{\color{black}{According to the convergence rate of the multivariate central limit theorem \citep{gotze1991,berckmoes2016}, roughly speaking, the convergence of $\widetilde{\bftheta}$ to $\theta$ as $M\rightarrow\infty$ in (\ref{large_M}) depends on 3 different aspects: sample size, dimension of the random vector (parameter space), and the third moment of the components of the random vector. In our iterative procedure, the sample size consists of two aspects: the number of imputed datasets, $M$, also the proportion of missing data. A larger $M$ makes the convergence faster, while a larger proportion of missing data makes it slower. On the other hand, another important issue is the dimension of the parameter space, the more parameters to estimate the more slowly the convergence will be achieved. The third moment highlights the role of the model, for example, the convergence for a linear model would be different from a logistic regression (see Section~\ref{sec_simulations}). Therefore, an appropriate method for determining $M$ should consider all these aspects. In other words, it should be sensitive to changes in any of these factors. As our proposed method uses the goal of the analyses (parameter estimate, hypothesis testing, etc.) in the process of determining $M$, it includes such aspects. Simulation results in Section~\ref{sec_simulations} support this claim as well.}}

One important aspect of the proposed iterative procedure is the choice of an appropriate distance in Step 3. One immediate choice is to use an Euclidean distance, as follows:
\begin{equation}
\label{relative_euc}
d_{m+1}^{\mathrm{Euc}}=\sqrt{\left(\widetilde{\bftheta}_{m+1}- \widetilde{\bftheta}_{m}\right)^T\left(\widetilde{\bftheta}_{m+1}- \widetilde{\bftheta}_{m}\right)}.
\end{equation}
%$d_{m+1}^{\mathrm{Euc}}$ depends on the length of the vector of interest. Therefore, one may use $d_{m+1}^{\mathrm{Euc}}/k$ instead, where $k$ is the length of vector $\bftheta$. 
Alternatively, one may try to make the largest difference smaller, then an $\ell_{\infty}$-norm would be the appropriate distance. 
%While $d_{m+1}^{\mathrm{Euc}}/k$ is a compromise distance measure, $\ell_{\infty}$-norm is more stringent. 
Obviously, for {\color{black}{$q=1$}} these two are equivalent:
\begin{equation}
\label{infinity_norm}
d_{m+1}^{\ell_{\infty}}= \max\left\{ |\widetilde{\bftheta}_{m+1}- \widetilde{\bftheta}_{m}|\right\}.
\end{equation}
The common problem with $\ell_p$-norm type distance measures in (\ref{relative_euc})--(\ref{infinity_norm}) is the fact that they would emphasize the elements in the parameter vector which have larger magnitude. Also, they are not robust against changes in units. {\color{black}{This can also be seen from the general definition of the $\ell_p$-norm of a vector $\mathbf{x} = (x_1,\ldots,x_k)$, $\|\mathbf{x}\|_p$:
\begin{equation*}
\|\mathbf{x}\|_p = \left(|x_1|^p + \ldots+ |x_k|^p\right) ^{1/p}, p\leq 1.
\end{equation*}
When $p\rightarrow\infty$, one can show:
\begin{equation*}
\|\mathbf{x}\|_{\infty}=\max\{|x_1|,\ldots,|x_k| \}.
\end{equation*}}}
As an alternative, we propose using the Mahalanobis distance \citep{mahalanobis1936}: 
\begin{equation}
\label{mal_dis}
d_{m+1}^{\mathrm{Mah}} = \sqrt{\left(\widetilde{\bftheta}_{m+1}- \widetilde{\bftheta}_{m}\right)^T S^{-1} \left(\widetilde{\bftheta}_{m+1}- \widetilde{\bftheta}_{m}\right)}.
\end{equation}
{\color{black}{For using (\ref{mal_dis}), one needs to define an appropriate $S$. One can show:
\begin{equation}
\label{diff_cov}
\widetilde{{\bftheta}}_{m+1}-\widetilde{{\bftheta}}_m = \frac{\widehat{\bftheta}_{m+1}- \widetilde{{\bftheta}}_m}{m+1}.
\end{equation}
Considering the fact that $\widehat{\bftheta}_{m+1}$ and $\widetilde{{\bftheta}}_m$ are independent given the observed part of the sample, one may use:
\begin{equation}
\label{diff_cov}
\mathrm{Cov}\left(\frac{\widehat{\bftheta}_{m+1}- \widetilde{\bftheta}_m}{m+1} \right)=\frac{1}{(m+1)^2}\left[\mathrm{Var}(\widehat{\bftheta}_{m+1})+[\mathrm{Var}(\widetilde{\bftheta}_{m})\right] = \frac{1}{(m+1)^2}[\mathrm{Var}(\widehat{\theta}_{m+1})+\widehat{W}_m].
\end{equation}
The division by $m+1$ in (\ref{diff_cov}) makes this an inappropriate candidate for $S$, as it changes every time, while we need to monitor convergence of the procedure, relative to a sufficiently stable yardstick.

Other choices for $S$ include the within, between, or combined covariance matrix of the parameters at step $m+1$, i.e., $\widehat{W}_{m+1}$, $\widehat{B}_{m+1}$, or $\widehat{V}_{m+1}$, respectively; here $\widehat{V}$ is computed as in \ref{finite_M}.}}
%Considering the fact that $\widetilde{\bftheta}_{m+1}$ and $\widetilde{\bftheta}_{m}$ are independent given the observed part of the sample, one may use,
%\begin{equation}
%\label{mal_all}
%S=\widehat{V}_{m+1}+\widehat{V}_{m},
%\end{equation}
%where $\widehat{V}$ is computed as in \ref{finite_M}. Another choice for $S$ would only use the between variability of $\widehat{\theta}_m$'s $(m=1,\ldots,M)$,
%\begin{equation}
%\label{mal_between}
%S=\widehat{B}_{m+1} + \widehat{B}_{m}.
%\end{equation}
{\color{black}{Our simulation results show that $\widehat{B}_{m+1}$ cannot be an appropriate choice since being normalized by the variability of the differences, the resulting Mahalanobis distance is not sensitive to the fraction of missing data. This will be discussed in more detail in Section \ref{sec_simulations}.}}

{\color{black}{An appropriate distance should be sensitive to the fraction of missing data, but robust against rescaling model parameters. Both $\widehat{W}$ and $\widehat{V}$ serve these purposes, but our simulation results show that (see Section~\ref{sec_simulations}) for some large values of variance components, $\widehat{W}$ would show chaotic behavior from one iteration to another in some cases, while $\widehat{V}$ balances between and within variability, hence behave more stably. Therefore, our proposal for an appropriate distance in Step 3 is the Mahalanobis distance with $S=\widehat{V}_{m+1}$.}} In cases where the quantity of interest is unitless, e.g., a correlation coefficient, a coefficient of variation, a coefficient of determination, a $p$-value, etc., using $\ell_p$-norm type distances such as in (\ref{relative_euc})--(\ref{infinity_norm}) is recommended, especially when obtaining the variance of the quantity of interest is not straightforward. Of course, one may use any tailored distance according to the problem at hand, {\color{black}{this}} would not change the proposed iterative procedure.

{\color{black}{As mentioned earlier, the convergence speed of the imi procedure depends on various factors such as the model, the parameter space, and the fraction of missing data. Also, the choice of the distance in Step 3 of the procedure plays a determining role when it comes to deciding when to stop. Therefore, formulating a fixed universal threshold for $\epsilon$ in Step 3 of the proposed iterative procedure seems not to be possible. However, based on extensive simulation results, when using a Mahalanobis-types distance as in (\ref{mal_dis}) with $S=\widehat{V}_{m+1}$, considering $\epsilon=0.05$ leads to a liberal choice of $M$. If one wants to select a conservative $M$, we propose to set $\epsilon=0.01$. In the same setting, such choices of $\epsilon$ would select $M$'s that are approximately in agreement with Rubin's classic suggestion, e.g., to use $M=5$ for $10\%$ missing data (See Table~\ref{tab_mvn}).}}

{\color{black}{In general, for selecting $\epsilon$}}, one needs to consider the nature of the quantity of interest when deciding on $\epsilon$. For example, when dealing with $p$-values, one needs to ensure that increasing the {\color{black}{number of imputed datasets}} would not change the result of the test, so the distance between two $p$-values from imputed datasets number $m$ and $m+1$, near or smaller than the significance, $\alpha$, should be smaller than, for example, $\alpha/10$; see Section~\ref{sec_les}.     {\color{black}{Therefore, in case of $p$-values, we propose to use a Euclidean distance with $\epsilon=\alpha/10$. Note that, given the null hypothesis, the $p$-value is uniformly distributed. Therefore, for an appropriate $\epsilon$ and up to a constant coefficient, using a Mahalanobis distance is equivalent to the Euclidean distance.}}

 {\color{blue}{As \cite{wald1939} pointed out, the two central procedures in theory of statistics are parameter estimation and hypothesis testing. To summarize our discussion above, we proposed to use a different $\epsilon$ for each of these procedures. In case of parameter estimation one may use $0.05$ or $0.01$, and in case of hypothesis testing (using $p$-values) one may use $\alpha/10$.}}

%When the fraction of missingness is large, it is possible that the convergence plot becomes too noisy. In such cases one may use an appropriately smoothed curve to decide on the number of imputations. To avoid the possible dangers of overcorrection, one may consider different smoothing methods and compare their results as a sensitivity analysis. In case different smoothing methods lead to almost the same conclusion, a final decision on $M$=1 can be made; see Section~\ref{sec_armd}.

\subsection{Simulation study}
\label{sec_simulations}

{\color{black}{In order to study our proposed method, two simulation plans are considered. One with a multivariate normal vector, and the other with a logistic regression model. Using these two simulation settings, the performance of our proposed procedure for both parameter estimation and hypothesis testing will be studied. The results will be displayed, compared, and discussed, after presenting the plans.}}

{\color{black}{
\subsubsection{Simulation plan}

A simulation study is undertaken for a multivariate normal vector as well as data generated from a logistic regression model. Two types of covariance structures are considered for the multivariate normal model: compound-symmetry (CS) and first--order autoregressive (AR(1)). Consider $\BY_i$ a vector of length $k$, then $\BY_i\sim N(\mu \mathbf{1}_{k}, \Sigma)$. Under CS, $\Sigma=\sigma^2I_{k}+\tau J_{k}$, for $\sigma^2,\tau >0$ and $I_{k}$ and $J_{k}$ the identity and all-ones matrices, respectively. For AR(1), we have $\Sigma=\sigma^2 C$ where the $(i,j)$ element of $C$ is defined as $\rho^{|i-j|}$, for $\sigma^2>0,-1\leq \rho \leq +1$.

All data are generated for $\mu=0$. The length of each random vector is set to 5, and the sample size is considered to be 100. For each covariance structure in the multivariate normal case 18 different scenarios are considered as combinations of the following settings:

\begin{itemize}
	\item Two proportions of missing data are used: $10\%$ and $70\%$;
	\item Three $\sigma^2$ values are used: $1,16,64$;
	\item Three $\rho$ values are used: $0.2, 0.5, 0.8$. In case of CS, the corresponding $\tau$ is computed given the specified $\rho$ using $\tau=\rho\sigma^2/(1-\rho)$.
\end{itemize}
Logistic regression data are generated with parameters $(\beta_0,\beta_1,\beta_2)=(0.2,-2,0.5)$, where the design matrix is generated using each of the 9 considered CS settings. The parameters of the CS and AR(1) models are estimated using the results in \cite{hermans2017_cs} and \cite{hermans2017_ar1}, respectively. The logistic regression is fitted using \textsc{R} function {\tt{glm}}. The proportion of missing values are the same as in the multivariate normal scenarios ($10\%$ and $70\%$).

The missing data for all of these scenarios are generated under a missing at random (MAR) mechanism using the function {\ttfamily{ampute}} in the \textsc{R} package {\ttfamily{mice}}{\color{blue}; see {\cite{ampute} and \cite{schouten2018}. The function {\tt{ampute}} implements a multivariate approach to generate missing data by assigning a weight to each observation. In case of a missing at random (MAR) mechanism, these weights only depend on the observed part of the sample. Based on the assigned weights, a logistic distribution will be used to compute the probability of missingness. An observation with a larger weight has more chance to be missing.}}

Each incomplete dataset is imputed $500$ times using a multivariate normal predictive model in the \textsc{R} package {\tt{Amelia2}}, and each scenario is repeated 100 times as well.

%Let us briefly describe how the function {\ttfamily{ampute}} works. It starts by considering different schemes of missingness. A scheme of missingness is a vector of $1$'s and $0$'s. The length of this vector is the same as the number of variables in the dataset. For a specific subject, $1$ means that the corresponding variable is recorded, and $0$ means it is possibly missing. Corresponding to each scheme of missingness, a set of weights will be considered that specifies the importance of each variable in determining whether a subject should contain missing values or not. Under an MAR mechanism, the variable with a missing value ($0$ in the scheme matrix) should have weight $0$, since it should not play a role in determining whether a missing should happen or not, this will be determined by observed variables. For each subject, based on the observed values and allocated weights, a weighted sum of scores is computed. These scores are used in a logit function to determine the probability of being missing or not. Obviously, this would lead to occurrence of various missing data patterns. More details can be found in \cite{ampute}.
%

To compute the distance between two steps in our iterative procedure, we use five different measures: Euclidean distance (\ref{relative_euc}), $\ell_{\infty}$-norm (\ref{infinity_norm}), Mahalanobis distance (\ref{mal_dis}) with $S$ selected as within ($\widehat{W}$), between ($\widehat{B}$), and combined covariance matrix ($\widehat{V}$) of the estimated parameter vector.

There are two main outcomes of interest from different considered scenarios. First, the convergence plot for different distances. Second, the $M$ for which the iterative procedure terminates. Such $M$ is computed using seven values $\epsilon= 0.005, 0.01, 0.02, 0.03, 0.04$ and $0.05$, and for each $\epsilon$, three successive validation steps are considered: $k_0=1,3$ and $5$.

In order to evaluate our proposed method in case of $p$-values, one-sample $t$-tests are applied with $H_0:\mu=d_0$ on the first column of the data generated from each setting of CS and AR(1) covariance structure. In addition, paired $t$-tests are also performed for testing $H_0:\mu_1-\mu_2=d_0$ for the first and the second columns of data generated using each setting of CS and AR(1) covariance matrices. To see the effect of the test value on the convergence, we have considered $d_0=0$ and $d_0=10$. As the $p$-value is a scalar, we use Euclidean distance (\ref{relative_euc}) to measure distance of two successive steps. In case of $p$-values, again the convergence plots are provided; also, the sufficient number of imputed datasets is computed for two values of $\epsilon=0.05/10$ and $0.05/100$. For each $\epsilon$, three successive validation steps are considered: $k_0=1,3$ and $5$.

\subsubsection{Simulation results}
As displaying the results of 18 different scenarios takes a lot of space, here we discuss all of the outcomes but only display the results for one of them. The rest of the results will be available in the \textsc{R} package {\tt{imi}} via 9 lists of datasets, the function {\tt{imi.make.plots}} can be used to produce similar plots as in Figure~\ref{fig_mvn} and Figure~\ref{fig_ttest} for other scenarios. Further information can be found in the documentation of the package.

Figure~\ref{fig_mvn} shows the convergence plots for the scenario with $\sigma^2=16$ and $\rho=0.5$. As one may see, while using all of the distance functions a decreasing trend can be observed, but as expected Euclidean and $\ell_{\infty}$ norms, as well as Mahalanobis distance with $S=\widehat{B}$ are not performing well. On the other hand, the latter with both $S=\widehat{W}$ and $S=\widehat{V}$ performs fine. {\color{blue}{Note that for $70\%$ missing data for some of the imputed datasets, quasi-complete separation occurs when fitting logistic regression. In such cases, a new imputed dataset is generated.}}

An interesting observation from Figure~\ref{fig_mvn} is the different convergence rate for logistic regression compared with the multivariate normal case. As one may see, for $10\%$ missing data, the logistic regression converges almost with the same speed as $70\%$ missing data in case of a multivariate normal vector parameter estimates. That highlights the effect of the model (thus not only the proportion of missing data) when determining a sufficient number of imputed datasets.

\begin{figure*}
	\centering
	\includegraphics[width=\textwidth]{figMMest.eps}
	\caption{Convergence plots for multivariate normal random vectors parameter estimates with CS and AR(1) covariance matrix structures (with $\sigma^2=16$ and $\rho=0.5$), as well as a logistic regression model, using five different distance functions: Euclidean norm, $\ell_{\infty}$-norm, and Mahalanobis distance with within, between, and combined covariance matrices over imputed sets of data for two proportion of missing data: $10\%$ and $70\%$. Each point in the plot is the average over 100 replications. {\color{blue}{The three horizontal dashed lines are $\epsilon = 0.05, 0.01, 0.001$, respectively.}}}
	\label{fig_mvn}
\end{figure*}

Figure~\ref{fig_ttest} shows the convergence plots using $M=500$ for the $p$-values of one sample and paired $t$-tests for two different test values (0, 10). The distance between two $p$-values is computed using the Euclidean distance. As one may see, one important factor that should be taken into account when determining $M$ is the test value. When it is far from the actual parameter convergence is achieved for a much smaller $M$.

\begin{figure*}
	\centering
	\includegraphics[width=\textwidth]{figttestMM.eps}
	\caption{Convergence plots for the $p$-values of one sample (first row) and paired (second row) $t$-tests for different values of test statistics. The data are generated from a (multivariate) normal distribution with mean $0$ and CS and AR(1) covariance matrices (with $\sigma^2=16$ and $\rho=0.5$). The missing values are generated with proportions $10\%$ and $70\%$. {\color{blue}{The two horizontal dashed lines are $\epsilon = 0.005, 0.0005$, respectively.}}}
	\label{fig_ttest}
\end{figure*}

To explore the selected $M$ using various $\epsilon$'s and $k_0$'s to evaluate our proposals ($\epsilon=0.01$ or $0.05$, and $k_0=3$), Tables~\ref{tab_mvn} and \ref{tab_ttest} show the selected $M$ for parameter estimation in multivariate normal and logistic regression (Table~\ref{tab_mvn}) and $p$-values of one sample and paired $t$-tests (Table~\ref{tab_ttest}). As one may see, again the effect of proportion of missing data and the model is verified on the selected sufficient $M$. {\color{blue}{While a dataset with a larger proportion of missing data needs a larger $M$, the number of sufficient imputed datasets in case of logistic regression is different from multivariate normal, even for a smaller proportion of missing data. This would highlight the role of a procedure that accounts for all of these aspects.}}

Also, it seems that our proposals for $\epsilon$ and $k_0$ are sensible in terms of comparing with Rubin's classical rule suggesting 3--5 imputed datasets for $10\%$ missing data. In case of $p$-values it seems $\epsilon=0.005$ together with $k_0=3$ works acceptably fine.

{\color{blue}{An interesting observation when comparing the selected $M$ for different values of $\rho$ is the effect of \emph{missing information}. For the same fraction of missing data, when the correlation is larger, the selected $M$ is smaller. That would suggest that what really matters is not only the fraction of missing data, but the fraction of missing information.  As an extreme case, consider two variables $X$ and $Y$ in a given dataset, suppose $X$ is complete and $Y$ is partially missing. Assume further that $Y_i = X_i$ ($\rho = 1$), for every subject. In this case one imputation would be sufficient, no matter the fraction of missing items. Figure~\ref{fig_missinfo} shows the selected $M$ for different scenarios with $\sigma^2 = 16$ and $\rho = 0.5$ and $0.8$, and $k_0 = 3$
. As one may see, in most of the cases, for the same proportion of missing data and model, the selected $M$ is smaller when $\rho$ is larger.}}

\begin{table}[ht]
	\caption{Mean, standard deviation (STD) for selected $M$ given different $\epsilon$ and $k_0$ values, and number of times $M>500$ (out of 100 replications) for different models and different $\epsilon$'s using the Mahalanobis-type distance with $S=\widehat{V}$. The results are presented for 3 different successive validation steps $k_0=1, 3, 5$.}
	\label{tab_mvn}
	\centering
\resizebox{\textwidth}{!}{	\begin{tabular}{lcccccccccc}
		\hline
\multirow{2}{*}{Model}	&	\multirow{2}{*}{$\epsilon$} & \multicolumn{3}{c}{$k_0=1$} &  \multicolumn{3}{c}{$k_0=3$} &  \multicolumn{3}{c}{$k_0=5$} \\ 
	&	& Mean & STD & $M>500$ & Mean & STD & $M>500$ & Mean & STD & $M>500$\\

  \hline
\multirow{6}{*}{CS-$10\%$}  &0.005 & 13.82 & 5.02 & 0.00 & 29.16 & 8.54 & 0.00 & 37.63 & 9.24 & 0.00 \\ 
  &0.01 & 7.15 & 3.08 & 0.00 & 15.38 & 4.88 & 0.00 & 19.31 & 6.01 & 0.00 \\ 
  &0.02 & 3.88 & 1.96 & 0.00 & 6.98 & 3.05 & 0.00 & 8.62 & 3.64 & 0.00 \\ 
  &0.03 & 2.67 & 1.44 & 0.00 & 4.36 & 2.22 & 0.00 & 5.31 & 2.51 & 0.00 \\ 
  &0.04 & 2.09 & 1.06 & 0.00 & 3.27 & 1.71 & 0.00 & 3.63 & 1.91 & 0.00 \\ 
  &0.05 & 1.68 & 0.84 & 0.00 & 2.42 & 1.51 & 0.00 & 2.78 & 1.80 & 0.00 \\ \hline
\multirow{6}{*}{CS-$70\%$}   &0.005 & 30.71 & 10.28 & 0.00 & 71.12 & 13.25 & 0.00 & 94.98 & 14.39 & 0.00 \\ 
  &0.01 & 18.12 & 5.24 & 0.00 & 37.93 & 8.01 & 0.00 & 50.62 & 9.72 & 0.00 \\ 
  &0.02 & 10.06 & 3.74 & 0.00 & 20.73 & 5.33 & 0.00 & 25.04 & 5.41 & 0.00 \\ 
  &0.03 & 6.45 & 2.59 & 0.00 & 13.65 & 3.95 & 0.00 & 16.85 & 4.39 & 0.00 \\ 
  &0.04 & 4.81 & 2.10 & 0.00 & 9.61 & 3.03 & 0.00 & 12.24 & 3.78 & 0.00 \\ 
  &0.05 & 3.94 & 1.94 & 0.00 & 7.67 & 2.70 & 0.00 & 9.40 & 3.18 & 0.00 \\ \hline
\multirow{6}{*}{AR(1)-$10\%$}  &0.005 & 18.86 & 7.24 & 0.00 & 39.05 & 11.13 & 0.00 & 46.60 & 12.93 & 0.00 \\ 
  &0.01 & 10.37 & 4.28 & 0.00 & 18.72 & 6.58 & 0.00 & 23.81 & 8.44 & 0.00 \\ 
  &0.02 & 4.99 & 2.33 & 0.00 & 9.85 & 3.57 & 0.00 & 11.74 & 4.29 & 0.00 \\ 
  &0.03 & 3.45 & 1.59 & 0.00 & 5.96 & 2.77 & 0.00 & 7.44 & 3.33 & 0.00 \\ 
  &0.04 & 2.40 & 1.22 & 0.00 & 4.15 & 2.29 & 0.00 & 4.67 & 2.54 & 0.00 \\ 
  &0.05 & 2.03 & 1.07 & 0.00 & 3.02 & 1.65 & 0.00 & 3.64 & 2.15 & 0.00 \\ \hline
\multirow{6}{*}{AR(1)-$70\%$}  &0.005 & 44.30 & 13.06 & 0.00 & 97.12 & 17.89 & 0.00 & 123.94 & 20.28 & 0.00 \\ 
  &0.01 & 23.71 & 8.08 & 0.00 & 51.85 & 9.49 & 0.00 & 64.14 & 11.04 & 0.00 \\ 
  &0.02 & 13.55 & 4.93 & 0.00 & 26.79 & 6.21 & 0.00 & 33.91 & 6.82 & 0.00 \\ 
  &0.03 & 9.28 & 3.36 & 0.00 & 17.81 & 3.97 & 0.00 & 21.94 & 5.11 & 0.00 \\ 
  &0.04 & 6.76 & 2.65 & 0.00 & 13.59 & 3.34 & 0.00 & 16.67 & 3.93 & 0.00 \\ 
  &0.05 & 5.44 & 2.26 & 0.00 & 10.84 & 2.98 & 0.00 & 13.12 & 3.17 & 0.00 \\ \hline
\multirow{6}{*}{Logreg-$10\%$}  &0.005 & 39.06 & 16.12 & 0.00 & 71.73 & 24.19 & 0.00 & 90.08 & 29.32 & 0.00 \\ 
  &0.01 & 19.95 & 8.71 & 0.00 & 38.31 & 14.02 & 0.00 & 46.67 & 15.70 & 0.00 \\ 
  &0.02 & 11.14 & 5.26 & 0.00 & 20.38 & 7.42 & 0.00 & 25.47 & 9.62 & 0.00 \\ 
  &0.03 & 7.31 & 3.60 & 0.00 & 13.06 & 5.50 & 0.00 & 16.28 & 6.07 & 0.00 \\ 
  &0.04 & 5.36 & 2.82 & 0.00 & 9.98 & 4.10 & 0.00 & 11.72 & 4.71 & 0.00 \\ 
  &0.05 & 4.60 & 2.56 & 0.00 & 7.85 & 3.38 & 0.00 & 9.53 & 3.78 & 0.00 \\ \hline
\multirow{6}{*}{Logreg-$70\%$}  &0.005 & 61.33 & 26.34 & 0.00 & 140.31 & 53.29 & 0.00 & 174.82 & 67.77 & 0.00 \\ 
  &0.01 & 36.68 & 15.25 & 0.00 & 76.05 & 25.85 & 0.00 & 94.65 & 32.48 & 0.00 \\ 
  &0.02 & 21.35 & 7.88 & 0.00 & 40.75 & 13.77 & 0.00 & 51.16 & 16.31 & 0.00 \\ 
  &0.03 & 14.68 & 6.37 & 0.00 & 28.97 & 8.96 & 0.00 & 34.92 & 10.65 & 0.00 \\ 
  &0.04 & 11.76 & 4.29 & 0.00 & 22.35 & 6.44 & 0.00 & 26.23 & 7.47 & 0.00 \\ 
  &0.05 & 10.01 & 4.00 & 0.00 & 18.99 & 5.44 & 0.00 & 21.77 & 6.26 & 0.00 \\ 
   \hline
\end{tabular}}
\end{table}

\begin{table}[ht]
	\caption{Mean, standard deviation (STD) for selected $M$ given different $\epsilon$ and $k_0$ values, and number of times $M>500$  for different models and different $\epsilon$'s for $t$-test and paired $t$-test with test value $\mu=0$. The results are presented for 3 different successive validation steps $k_0=1, 3, 5$.}
	\label{tab_ttest}
	\centering
\resizebox{\textwidth}{!}{	\begin{tabular}{clcccccccccc}
		\hline
\multirow{2}{*}{Test}&	\multirow{2}{*}{Model}&\multirow{2}{*}{$\epsilon$} & \multicolumn{3}{c}{$k_0=1$} &  \multicolumn{3}{c}{$k_0=3$} &  \multicolumn{3}{c}{$k_0=5$} \\ 
&&& Mean & STD & $M>500$ & Mean & STD & $M>500$ & Mean & STD & $M>500$ \\ 
		\hline
\multirow{8}{*}{$\mu1=0$} &\multirow{2}{*}{CS-$10\%$}  &0.005 & 2.90 & 2.55 & 0.00 & 8.17 & 6.74 & 0.00 & 11.32 & 8.93 & 0.00 \\ 
 & &5e-04 & 12.78 & 9.70 & 0.00 & 50.41 & 32.28 & 0.00 & 83.97 & 53.85 & 0.00 \\ 
&\multirow{2}{*}{CS-$70\%$}  &0.005 & 5.41 & 3.85 & 0.00 & 20.22 & 10.08 & 0.00 & 29.20 & 14.79 & 0.00 \\ 
 & &5e-04 & 21.75 & 12.38 & 0.00 & 119.70 & 57.55 & 0.00 & 211.70 & 93.22 & 0.00 \\ 
 &\multirow{2}{*}{AR(1)-$10\%$}  &0.005 & 2.76 & 2.03 & 0.00 & 7.55 & 6.39 & 0.00 & 10.55 & 9.08 & 0.00 \\ 
 & &5e-04 & 12.73 & 10.51 & 0.00 & 53.61 & 34.37 & 0.00 & 88.68 & 63.04 & 0.00 \\ 
  &\multirow{2}{*}{AR(1)-$70\%$} &0.005 & 5.91 & 4.34 & 0.00 & 20.98 & 10.62 & 0.00 & 30.73 & 15.85 & 0.00 \\ 
  &&5e-04 & 22.88 & 15.28 & 0.00 & 120.90 & 57.80 & 1.00 & 214.83 & 97.30 & 1.00 \\ \hline

\multirow{8}{*}{$\mu1-\mu_2=0$} &\multirow{2}{*}{CS-$10\%$}    &0.005 & 3.32 & 2.24 & 0.00 & 10.34 & 7.32 & 0.00 & 15.16 & 11.30 & 0.00 \\ 
  &&5e-04 & 15.75 & 11.31 & 0.00 & 64.97 & 37.27 & 0.00 & 110.12 & 58.83 & 0.00 \\ 
 &\multirow{2}{*}{CS-$70\%$}  &0.005 & 6.62 & 4.14 & 0.00 & 25.41 & 13.59 & 0.00 & 36.49 & 19.45 & 0.00 \\ 
  &&5e-04 & 25.94 & 17.69 & 0.00 & 157.88 & 74.58 & 6.00 & 243.37 & 124.17 & 6.00 \\ 
  &\multirow{2}{*}{AR(1)-$10\%$}  &0.005 & 4.02 & 2.78 & 0.00 & 11.46 & 7.58 & 0.00 & 16.18 & 10.01 & 0.00 \\ 
 & &5e-04 & 14.37 & 10.65 & 0.00 & 74.38 & 40.09 & 0.00 & 134.99 & 70.34 & 0.00 \\ 
   &\multirow{2}{*}{AR(1)-$70\%$}&0.005 & 8.31 & 5.11 & 0.00 & 29.14 & 14.83 & 0.00 & 43.30 & 19.48 & 0.00 \\ 
 & &5e-04 & 31.39 & 17.23 & 0.00 & 160.09 & 72.00 & 2.00 & 288.81 & 112.58 & 2.00 \\ 
	\end{tabular}}
\end{table}

}}

\begin{figure*}
	\centering
	\includegraphics[width=\textwidth]{figMisisngInformation.eps}
	\caption{Averaged selected $M$ with $k_0 = 3$ for data generated from a multivariate normal with mean 0 and covariance matrix with CS and AR(1) structures with  $\sigma^2 =16$ and $\rho = 0.5, 0.8$, as well as a logistic regression model. The proportions of missing data are $10\%$ and $70\%$. Each point in the plot is the average over 100 replications.}
	\label{fig_missinfo}
\end{figure*}

{\color{black}{

\subsection{Applications}
\label{sec_case_studies}
In this section, two applications of multiple imputations are considered: fitting a logistic regression to incomplete data, as well as combining $p$-values from a one-sample $t$-test for a set of incomplete data. The number of imputed datasets will be determined using the proposed iterative procedure via the \textsc{R} package {\ttfamily{imi}}.

\subsubsection{Leuven Eye Study. Logistic regression}
\label{sec_les}
The Leuven Eye Study \citep[LES;][]{abegao2015}, is an extensive observational study of glaucoma performed at the ophthalmology department of UZ Leuven. The dataset so far consists of 141 variables measured for 585 subjects. As this study was performed in a clinical setting, it was not feasible to measure all these variables for all of the subjects. Because of that, missingness is a considerable problem. Among these 141 variables, 130 of them are selected to be used in multiple imputation. An analysis was performed to study which risk factors are relevant for the binary outcome defined as normal vs. glaucoma. The risk factors selected for the model include both structural and functional measurements of the eye (cup to disc ratio, corneal thickness, visual acuity), previous medical history (gender, sleep apnea, rhythm disorder, having had cataract surgery, being medicated with statins or calcium channel blockers) and more variable biometric measurements, like intra-ocular pressure and diastolic blood pressure.

Figure~\ref{fig_conv_real} (bottom-left) shows a histogram of the number of missing values (out of 585 patients) in the LES dataset. As one may see, the number of missing values is diverse. Also the patterns of missing values could be different from variable to variable. Therefore, determining $M$ using classical or simulation-based approaches could be a challenge here. Using our proposal, it could be more convenient to determine the sufficient number of imputed datasets.

Considering the size of the dataset as well as various variable types, in order to impute the missing values the fully conditional specification (FCS) approach \citep[and \citealp{van2006}]{van2007,buuren12} is employed. For continuous variables, predictive mean matching, \cite{little1988}, is used for generating the imputed datasets. Also, a binary logistic regression predictive model is used to impute binary variables. In the specific case of one variable with three levels, a polytomous logistic regression predictive model is used. 
In order to create a sufficient number of imputed datasets we use the function {\ttfamily{imi.glm}} in the \textsc{R} package {\ttfamily{imi}}.

%\noindent{\ttfamily{regressors=c("Sleepapnea", "C\_vs\_D","RithmDisor","DBP","VA\_log\_mar","IOP\_DCT","Phaco",
%		"Statinany",\\"CCBany","CCT","SEX") \\
%		resp='Diagnosis5.bin'\\
%		les.glm=imi.glm (les.new.mi2,family=binomial(),M0='manual',max.M=500,epsilon=0.01,method='auto',resp,\\regressors,
%		conv.plot=TRUE,
%		dis.method='mahalanobis',mah.scale='within',successive.valid='manual',\\max.iter.glm=1000)}}
%
%As the parameters {\ttfamily{M0}} and {\ttfamily{successive.valid}} are both set as {\ttfamily{manual}} the software acts interactively and first imputes the data twice and report them time and ask to determine these two parameters afterwards:
%
%\noindent{\ttfamily{"The time it takes (in seconds) to impute the data two times and fit the model to them is: 181.24"\\
%		What is your choice of initial number of imputations?2\\
%		What is your choice for successive steps validation?3}}
%	
%As one may see, 
We have selected 2 for the initial number of imputations ($M_0$). Also, the convergence criterion should be successively validated 3 times to terminate the procedure ($k_0=3$). Figure~\ref{fig_conv_real} (bottom-right) shows the convergence plot for this model on the LES data. As one may see, with $\epsilon=0.05$ (the liberal choice) and $k_0=3$ we need $M=58$ imputed datasets, while with the conservative choice, $\epsilon=0.01$, one may need to generate $M=250$ imputed datasets.

\begin{figure*}
	\centering
	\includegraphics[width=\textwidth]{fig_real_conv_gray.eps}
	\caption{Convergence plots for $p$-values of $t$-test on cholesterol data (top) with $\mu_0=200$ (left) and $\mu_0=220$ (right) test values, and LES logistic regression (bottom-right), with $\epsilon=0.05, 0.01$, $M_0=2$, and $k_0=3$. The dashed vertical line shows the selected $M$ for different values of $\epsilon$. The histogram for the number of missing values in LES dataset (bottom-left) is also presented.}
	\label{fig_conv_real}
\end{figure*}

\subsubsection{Cholesterol data. One sample $t$-test $p$-values}

The cholesterol dataset includes cholesterol levels for 28 patients treated at a Pennsylvania medical center. The cholesterol levels have been recorded on day 2, day 4 and day 14 after an attack for each patient. However, there are 9 missing values for day 14. This dataset was analyzed by \citet[Chapter 5]{schafer1997} and is publicly available in \textsc{R} package {\ttfamily{norm2}}.

Here we use our \textsc{R} implemented function {\ttfamily{imi.t.test}} to sufficiently impute this dataset and perform the $t$-test on it for two different test-values ($\mu_0=200, 220$).

%This can be done using the command {\ttfamily{imi.t.test (cholesterol,M0='manual',max.M=500,epsilon=0.05/100,method='mvn',
%		x=names(cholesterol)[3], y = NULL, alternative='two.sided',mu=200,
%		paired = FALSE, var.equal = FALSE, 
%		conf.level = 0.95,conv.plot=TRUE,successive.valid='manual')}}. 
%
%As both of {\tt{M0}} and {\ttfamily{succeessive.valid}} are set as {\ttfamily{'manual'}} the software will impute the data and perform the test and the following message will be shown:
%
%\noindent{\tt{"The time it takes (in seconds) to imput the data once and perform the test is less than 0.001 seconds"\\
%		What is your choice of initial number of imputations?2\\
%		What is your choice for successive steps validation?3}}

Again, we ask for two initial imputations and then 3 successive validation steps ($M_0=2, k_0=3$). Figure~\ref{fig_conv_real} (top) shows the convergence plot for $\mu_0=200$ (left) and $\mu_0=220$ (right). In each case the vertical dashed line shows the selected $M$ for different values of $\epsilon$. As one may see, when testing $H_0: \mu=200$, for $\epsilon=0.05/10$ and $k_0=3$, only $M=9$ imputed datasets are sufficient, changing $\epsilon$ to $0.05/100$ will increase $M$ to 28. When testing $H_0: \mu=220$, however, more imputations are needed. For $\epsilon=0.05/10$ we get $M=37$ and for $\epsilon=0.05/100$ we obtain $M=239$.

%First of all, to provide a descriptive analysis of the data, we need to find the mean of each of these 104 variables. Figure \ref{fig_les} (top-right) shows the convergence plot for the Mahalanobis distance of the estimated mean vector. As one may see, using $M=50$, the distance between two successive $M$'s becomes as small of 0.01, which is acceptable.
%
%The LES data divides the subjects into 5 diagnosis groups: primary open-angle glaucoma (POAG), normal--tension glaucoma (NTG), ocular hypertension (OHT), glaucoma suspects, and healthy. Another aspect of this study is to test whether there is a significant difference among the five diagnostic types for each of the variables in the study. For the 104 continuous variables, we use a Kruskal-Wallis test \citep{kruskal1952}. The 50 $p$-values are combined using the combination rule proposed in \cite{li1991}. 
%
%As $p$-values are unitless, $\ell_p$-norm distances are appropriate. Two types of distances are used: the Euclidean distance divided by the number of variables (104), which is called mean Euclidean distance here, and the $\ell_{\infty}$-norm. As mentioned in Section~\ref{sec_proposal}, the mean Euclidean distance is a compromise distance measure, while the $\ell_{\infty}$-norm, which looks at the maximum absolute difference, is stringent.
%
%Figure~\ref{fig_les} (bottom--left and --middle) shows the convergence plot using these two distances. As one may see, using mean Euclidean distance (bottom-left), $M=50$ imputations would make the distance between $p$-values of two successive $M$'s as small as 0.0004, which is acceptable. But for the $\ell_{\infty}$-norm distance (bottom-middle) this would not become smaller than $0.02$. For small $p$-values (around and smaller than $\alpha=0.05$), this might lead to unwanted changes in decision. 
%
%Fortunately, looking at the variables of which the $\ell_{\infty}$-norm distance between $p$-values obtained using $M=49$ and $M=50$ is larger than 0.005 ($\alpha/10$, for $\alpha=0.05$) shows that they are at least as large as $0.109$. Thus, a maximum difference of $0.02$ would not change the decision. Therefore, we may conclude that $M=50$ is also enough for combining $p$-values in this case. Figure~\ref{fig_les} (bottom-right) shows the histogram of variables with $\ell_{\infty}$-norm distance larger than 0.005 for their $p$-values between $M=49$ and $M=50$ imputations.
%
%\begin{figure}[t]
%\centering
%\includegraphics[trim=0 40 0 10,width=\textwidth]{fig_les_conv2.eps}
%\caption{Leuven eye study. The histogram of the number of missing observations in 104 continuous variables (top-left), convergence plot of the Mahalanobis distance between two estimated means for continuous variables (top-right), the mean Euclidean (bottom-left) and $\ell_{\infty}$-norm (bottom-middle) distances between $p$-values of Kruskal-Wallis tests, and the $p$-values using $M=50$ imputations for variables with their $\ell_{\infty}$-norm distance between $p$-values using $M=49$ and $M=50$ was larger than 0.005. Note that all the y-axes are on the log-scale.} 
%\label{fig_les}
%\end{figure}

%\subsection{ARMD trial. A surrogate endpoint evaluation model}
%\label{sec_armd}
%The data come from a clinical trial in age-related macular degeneration (ARMD), which is a disease of the retina that causes severe central vision loss. \cite{pharmacological1997} conducted a multi-center study (17 centers) to evaluate the effect of an experimental treatment based on interferon-$\alpha$ in a group of patients who were randomly allocated to either placebo or interferon-$\alpha$. The outcome of interest was the change in visual acuity over time. Using standard vision charts, the visual acuity was measured by the number of letters that were correctly read by the patient. For our purpose, we use a subset of data such that each centre has at least 5 patients, resulting in a total of 119 patients. Table~\ref{tab_freq} shows the observed centre-size frequencies. The data can be obtained from the \emph{Surrogate package} in \textsf{R} \citep{surrogate2016}.
%\begin{table}[t]
%\centering
%\caption{Frequency of centres based on the number of patients.}
%\label{tab_freq}
%
%\vspace*{2mm}
%
%\begin{tabular}{lcccccc}
%  \hline\hline
%Number of patients & 5 & 6 & 7 & 8 & 9 & 18 \\ 
%Number of centres &   7 &   2 &   4 &   1 &   2 &   1 \\ 
%   \hline\hline
%\end{tabular}
%\end{table}
%
%\cite{burzykowski2006} and \cite{alonso2015} analyzed this dataset in the context of surrogate endpoint evaluation. The goal of their analyses was to examine if  visual acuity at week 24 could be an appropriate surrogate for the one at week 52, the so-called true endpoint. To this end, the surrogate and the true endpoint have to be modelled jointly.
%Because of the bivariate nature of the endpoint, coupled with the multi-centre nature of the study, a mixed model is used, by which surrogate evaluation measures are estimated; see also \cite{buyse2000}. 
%
%As is clear from Table~\ref{tab_freq}, the data are highly unbalanced, presenting the observed-likelihood model fitting procedure with serious convergence issues. \cite{wim2015} proposed to treat these unbalanced data as an incomplete set, where the data in each centre are considered relative to the maximally observed centre size. MI can then be used to balance the data, i.e., to make all centres of size 18 in this case. The results of \cite{wim2015} showed substantial improvement in convergence. However, determining the number of imputations remains an issue. Here, we use the proposed iterative MI method to study the effect of number of imputations on the final estimate. The parameters of interest are the trial and individual coefficients of determination, $R^2_{\mbox{\tiny trial}}$ and $R^2_{\mbox{\tiny indiv}}$. For more details on the model and these parameters; see Appendix II.
%
%Figure~\ref{fig_conv_armd} shows the convergence results for estimating these quantities using the Euclidean distance with $M=1000$ imputations, generating using PROC MI in SAS. Again, as the coefficient of determination is unitless, this distance is appropriate. As Figure~\ref{fig_conv_armd} shows, the convergence plot is very noisy for both parameters. This could be expected, because the centre-size distribution shows that there is a large fraction of missing data, when the problem is cast as one of missing-data. To obtain a smoother convergence plot, a local polynomial regression (loess) curve \citep{cleveland1991} is fitted. To avoid the danger of overcorrection, this curve is fitted using three different smoothing parameter $(\lambda=0.01, 0.1, 0.5)$. As one may see in Figure~\ref{fig_conv_armd}, the three curves are leading to almost the same conclusion about $M$.
%
%As one may see, for stabilizing the convergence plot, some hundreds of imputations are needed, but using 100 imputations, the difference between two iterations is acceptably small. This number is much larger than usually recommended, even though not as large as what is recommended by \cite{hershberger2003}.
%
%\begin{figure}[t]
%\centering
%\includegraphics[trim=0 100 0 120,width=\textwidth]{fig_armd.eps}
%\caption{ARMD data. Euclidean distance (log-scale) for estimated $R^2_{\mbox{\tiny trial}}$ and $R^2_{\mbox{\tiny indiv}}$ for ARMD data using 1000 imputations. The dashed, dotted and dotted-dashed lines are the fitted smooth loess curve for smoothing parameter $\lambda=0.5, 0.1$, and $0.05$, respectively.} 
%\label{fig_conv_armd}
%\end{figure}

}}

\subsection{Discussion and Concluding Remarks}
\label{sec_conclusions}
%Multiple imputation is an appealing and extensively used method to analyze incomplete data, or, even more broadly, data that can be cast in a missing-data framework \citep{wim2015}. There is widely held and largely justified wisdom that a small number of imputation suffices for many practical purposes, even though \cite{schafer1997} states that the number depends on the amount of missing information, the data type under study, and the inferential purpose. We have presented a very simple {\color{blue}{procedure}}, that is based on conventional large-sample convergence results, by merely using the fact that multiple imputation is in itself a sampling mechanism. It is based on comparing the estimates under $M$ and $M+1$ imputations, but obvious extensions towards using three or more subsequent estimates can be used as well. This would prevent early stopping when two consecutive quantities of interest are unusually close. Thus, when convergence is suggested, to play safe, one could still go on for a while, and wait until convergence has been confirmed. Stopping too early would lead to less precise estimates, or unwanted decision making, while the main problem with too high an $M$ is the computational cost.

Multiple imputation is an appealing and extensively used method to analyze incomplete data, or, even more broadly, data that can be cast in a missing-data framework \citep{wim2015}. There is widely held and largely justified wisdom that a small number of imputation suffices for many practical purposes, even though \cite{schafer1997} states that the number depends on the amount of missing information, the data type under study, and the inferential purpose. We have presented a very simple {\color{black}{procedure}}, that is based on conventional large-sample convergence results, by merely using the fact that multiple imputation is in itself a sampling mechanism. It is based on comparing the estimates under $M$ and $M+1$ imputations, {\color{black}{the procedure stops when the distance between two steps become smaller than a pre-defined $\epsilon$. One needs to select an initial number of imputations as well as number of steps the stopping rule should be validated.}} This would prevent early stopping when two consecutive quantities of interest are unusually close. Thus, when convergence is suggested, to play safe, one could still go on for a while, and wait until convergence has been confirmed. Stopping too early would lead to less precise estimates, or unwanted decision making, while the main problem with too high an $M$ is the computational cost, {\color{black}{also when the distance between two steps becomes very small, due to a small $\epsilon$ (or equivalently selecting a large $M$), some numerical underflow issues could occur.}}

{\color{black}{There are two main parameters that are needed to be specified before applying our procedure on an incomplete dataset. The stopping rule for the distance between two steps, $\epsilon$, and the number of steps this criterion should be successively validated, $k_0$. Based on our numerical experiences, we suggest $k_0=3$ validation steps is enough in most of the applications. For $\epsilon$, the choice depends on the chosen distance which itself depends on the purpose of the analysis: estimation or hypothesis testing. For estimation purposes, i.e., estimating the parameters and their precision, we suggest to use a Mahalanobis distance with $S=\widehat{V}_{M+1}$, then $\epsilon=0.05$ as a liberal choice and $\epsilon=0.01$ as a conservative choice. For $p$-values at $100(1-\alpha)\%$ level of confidence, we suggest to use the Euclidean distance with $\alpha/10$ as a liberal choice, and as a conservative choice our proposal is to use $\alpha/100$. Of course, as discussed in Section~\ref{sec_proposal}, the convergence rate depends on various aspects (proportion of missing data, the model, dimension of the parameter space). Therefore, it is very well possible that for less conventional applications one needs different tailor-made values for $\epsilon$ and $k_0$.}}

%For example, hypothesis testing for binary data with a large amount of incompleteness, will require more imputations. A dissenting voice is  \cite{hershberger2003}, who derived a survey-sampling-based rule that prescribes hundreds of imputations. A criticism one may have towards the former work is that the guidelines remain qualitative and are difficult to use by the less experienced user.  \cite{hershberger2003} does not take into account obviously important factors such as the amount of missing information.

Simulations and case studies, the latter with admittedly a large fraction of missingness, indicate that it might be needed, in many practical settings, to generate a number of imputations well above what common wisdom prescribes.

Our method has several advantages. First, it is applicable in any context where multiple imputation is useful, irrespective of the amount of missing information, the model used, or the target of inference. Second, imputed data sets can be generated one by one, and each time one can easily decide about the need to continue adding new imputed data sets, depending on the research question(s) to be answered. Third, the procedure can easily be automated since `expert judgement' is not needed. Finally, no post-hoc sensitivity assessment to the number of imputations is required, allowing the stopping criterion to be specified prior to the data analysis or even prior to the data collection.

{\color{black}{Furthermore, our implementation of this procedure in \textsc{R} package {\ttfamily{imi}} could make it more available for different applications. The current version of this package includes the $t$-test (one sample, two samples and paired), as well as linear and generalized linear regression models. For each application one has the possibility to generate a sufficient number of imputed datasets. In case of existence of some already generated imputed sets of data, one can examine their sufficiency for the desired analyses. When they are not sufficient it is also possible to generate new imputed datasets till $M$ becomes sufficiently large. The considered interactive options allow the user to select $M_0$ and/or $k_0$ based on the time it takes to generate $2$ imputed datasets and perform the analyses on them.}}

%\section*{Acknowledgement}

### 2019-06-25 intro.tex
all files 6/25/2019

\chapter{Introduction and problem statement}
\label{sec_intro1}

Small sample sizes would make wrong or statistically inaccurate conclusions. Therefore, in designing any statistical study, an important step is to estimate the sample size that would enable the researcher to draw valid conclusions. While a small sample size prevents reliable or meaningful statistical conclusions, large sample sizes has rarely been an issue, in fact a large sample size was a desirable aspect of a study. However, in 21st century, having large sample sizes also, becomes a trouble-maker. By emerging technologies that make the task of data collection easy and fast, the size of samples in many studies becomes larger and larger, and this came to the point that, the power of a typical computer would fail to analyze such huge datasets.

There are several issues related to a large datasets, including: how to store them, how to load them, and how to analyze them. Our interest in this thesis is in the last one. Although the last issue, the same as the two others could be solved using ideas coming from computer science, or simply, increasing the computation power of the hardware we use could solve the issue, we believe statistical methods and ideas can provide solutions that are easy to understand, capable of being implemented in different situations, and also accurate. This becomes more important when we realize increasing the computation power and size of available datasets are highly and positively correlated.

In Table \ref{tab_comptime1} we compare the computation when the sample sizes $n=10, 10^2, 10^3, 10^4, 10^5, 10^6, 10^7$, for some simple computational tasks including:

\begin{itemize}
\item Generating a random sample from standard normal of size $n$.
\item Finding the mean of the generated sample.
\item Finding the variance of the generated sample.
\item Fitting a linear model to the generated sample as the predictor and $y = 1 + ex + \epsilon$, where $\epsilon\sim N(0, 0.01)$.
\item Fitting a MAD (mean absolute deviation) regression to the same data.
\end{itemize}

\begin{table}[ht]
\centering
\begin{tabular}{lccccc}
  \hline
Sample size & Generating a normal sample & Mean & Variance & Linear model & MAD regression \\ 
  \hline
 10.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
 100.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
 1000.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.00 \\ 
 10000.00 & 0.00 & 0.00 & 0.00 & 0.00 & 0.06 \\ 
 100000.00 & 0.00 & 0.00 & 0.00 & 0.01 & 3.75 \\ 
 1000000.00 & 0.08 & 0.01 & 0.00 & 0.32 & 764.37 \\ 
 10000000.00 & 1.05 & 0.01 & 0.06 & 3.46 & 60558.83 \\ 
   \hline
\end{tabular}
\caption{Computation time (in seconds) of different methods for different sample sizes.} 
\label{tab_comptime1}
\end{table}

As we may see, for some analyses, the computation time increases linearly as the sample size increases, but for others, it would even speeds down with a sub-linear rate. Once the cause of expensive computation time is the sample size, an effective solution could be to split the very large sample into smaller sub-samples and then analyze them separately. As each sub-sample has a smaller size, and analyzing these different sub-samples are independent of each other (can be run in parallel), the overall computation time can be decreased.

Splitting a sample into smaller chunks, then analyzing each part separately, and finally combine the results of these analyses is a known methodology to deal with large samples. Some authors have called this \emph{Software Alchemy}, \cite{matloff2014}. This idea is also at the heart of MadReduce methodology \citep{dean2008}. The split-apply-combine strategy in the R packages {\tt{plyr}} and {\tt{dplyr}} also uses this idea \citep{wickham2011}. All these would show the importance of this approach. Furthermore, Figure \ref{fig_bigdata} shows a plot of number of results found using Google scholar when searching the exact phrase "big data" upto different years. We obviously can see the increasing interest in the topic in the recent years.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{bigdata.eps}
\caption{Number of items found by Google scholar when searching the exact phrase "big data" up to different years.} 
\label{fig_bigdata}
\end{figure}

However, most of these implementations would consider independent sub-samples, or non-clustered data. Our goal in this thesis to extend the existing methodology to the cases where the subjects would form clusters of correlated data. This type of data is very common in medical studies in particular, but they also can be observed in any longitudinal or multilevel study. Some examples of such data are measurement on offsprings of the same animal. Data collected from family members. Data collected from different countries in a meta-analytic setting. Measuring body temperature of the same person several times, ratings of the same book by different people, measuring hearing ability of elderly people for several frequencies, measuring primary vital signs (body temperature, heart rate, respiratory rate, and blood pressure) for each person, etc.

As we have seen in Table~\ref{tab_comptime1}, even for non-clustered data and simple models, increasing the sample size could cause very expensive computation times. The time needed to fit a MAD regression with one predictor to a sample of size 10 million is around 18 hours. Now, if we have clustered data that needs more complicated models, things would get even more difficult. In case of clustered data, the sample size can be seen from different angles.

\section{Large clustered data}

Consider a study which measures (at most) $m$ outcomes for $N$ clusters, each $n_i$ times. Let $y_{rij}$ be the $j$th measurement taken on the $i$th cluster for the $r$th outcome, $i=1,\ldots,N$, $r=1,\ldots,m$ and $j=1,\ldots,n_{ri}$. As a simple example, take $m=4$ responses as the vital signs: body temperature, heart rate, respiratory rate, and blood pressure, measured for $n_1 = n_2 = n_3 = n_4 = 100$ hospitalized patients. And assume these measurements are done every $4$ hours. So, $y_{111}$ is the body temperature of the first patient measured for the first time).

A general scheme of such clustered data is presented in Figure~\ref{fig_scheme}. A suitable model for such data should appropriately respect the correlation structures imposed by the fact that several measurements are done on the same subject. While even for moderate sample sizes, fitting such models could already become challenging, we may face a greater challenge when the size of dataset makes it eligible for the title \emph{big data}.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{scheme_new.eps}
\caption{General scheme of the clustered data we consider through this thesis.} 
\label{fig_scheme}
\end{figure}

As it was mentioned, for non-clustered data big would mean a large number of subjects. However, in our case, a dataset could be entitled as \emph{big data} in different situations. In case of the clustered data of interest in this thesis, data are considered to be big if at least one of the following applies:
\begin{itemize}
\item When the sample size, $N$, becomes very large,
\item When the cluster sizes become very large:
\begin{itemize}
\item When the number of measurements per outcome for some clusters, $n_{ri}$'s, becomes very large,
\item When the number of outcomes, $m$, becomes very large.
\end{itemize}
\end{itemize}

Note that, \emph{big} would be different for $N$, $n_{ri}$, and $m$. For example, in case of a MAD regression, we have seen a dataset of size $N=100,000$ could be analyzed in $0.001$h, increasing the sample size to $N = 1000,000$ would increase the computation time to $0.2$h, which is not fast, but still feasible, but $N=10,000,000$ would take around $17$ hours which could be too much. Therefore, in the case of a simple MAD regression, we may consider $N>10,000,000$ as \emph{big}.

When it comes to clustered data, however, there is a different story. Some 100's could already be big in case of cluster size, $n_{ri}$ (see Amazon book ratings data ({\bf{give reference to the section}})), and even some 10's is \emph{big} when we consider the number of outcomes $m$ (see Hearing data)({\bf{give reference to the section}}).

It goes without saying that, in reality, any combination of the above situations could happen. Our aim in this thesis to provide appropriate methodology to: 1- split the data into smaller chunks, 2- analyze each split, 3- combine these analyses in an efficient way. Our methodology should consider the nature of the data. While the mere task of splitting would already breaks down the difficult (sometimes infeasible) problem into smaller easier and feasible to analyze pieces, sometimes the structure of the data would allow for some \emph{smart} other than random splitting which would then provide more gain. Our aim in this thesis is to cover all of the situations which a dataset of clustered subjects could be called \emph{big}. We will go over our proposal for splitting/analyzing/combining steps for each of these cases.

In developing these methodologies we will make sure to keep two promises:

\begin{itemize}
\item While each methodology can deal with one situation, it should be easily possible to combine methodologies to deal with a combined situation.

\item The splitting and combining steps of our methodology should be as independent as possible from the analyses step, so one can easily (or with a minor modification) use it for its desired analysis.

\end{itemize}

\section{Data splitting: a unified approach}

Data splitting is a general idea to deal with big, massive, and large data. Large here could be called to any of the cases we have mentioned above, or a combination of them. Our general approach to deal with this situation consists of three main steps which we would call them together Data splitting.

\begin{enumerate}
\item \textbf{Splitting}: in this step, the data should be splitted into smaller chunks in a way that analyzing each of them becomes more convenient that the dataset as a whole. This convenience would at least come from having sub-samples of smaller sizes, but it could be possible to perform the splitting in a way that some other nice features (balancedness, for example) that the original data does not own, can be created in all or some of these sub-samples.

2- \textbf{Analyzing}: in this step, each split of data will be analyzed using the conventional methods. Therefore, using \emph{data splitting} approach, one can use its own favorite methodologies which are only suitable for small to moderate sample sizes. Typically, with \emph{data splitting} approach, the only information one needs to extract and keep from analyzing each split of data are: the estimated parameter vector, and their covariance matrix (if there is a need to provide standard errors of such estimates).

3- \textbf{Combining}: within this unified framework, each parameter in the model could be estimated several times. As our interest is to have one single set of parameter estimates, the last step in this paradigm is to combine all of analysis results from the second step into one single set. This should be done via appropriate combination rules. The combination rule is simply a weighted average of the results from different splits. Therefore, there is the need for proposing appropriate weights. Also, when it comes to combine covariance matrices, the fact that these results are coming from different (possible dependent) sub-samples should be well respected.

\end{enumerate}

In the following sections we will review different splitting approaches, also their appropriate combination rules. WE will also give examples of our motivating datasets where each of these approaches could be beneficial.

\section{Different types of splitting}

We have already discussed the MapReduce, or split-apply-combine approaches. The main splitting technique that is implemented using such approaches is to split the data: 1- randomly, 2- at subject level. Throughout this these we would call the splitting at the subject level \emph{horizontal splitting}. The reason for that can be seen in the Figure~\ref{fig_scheme}. If we illustrate the splits by separating lines, such lines will be horizontal. On the other hand, if the splitting is done for the data within a subject, then the lines should be drawn vertically. This case is specific to the clustered data where we have more than one observation for each subject.

While random splitting (assigning members of each split at random) is an effective way in many situations, there are cases where assigning members of each split based on a predefined structure has an added value. We call such splitting approach structural splitting as opposed to random splitting. In the following we will give an overview of the four types of splitting, and present examples where each of them could be useful. We will also briefly go over analyzing and combining steps also.

\subsection{Random horizontal splitting}

Random horizontal splitting, as its name suggested tries to assign different subjects to different splits at random. There are several ways to do this task, the way we have used in our implementations is as follows, for a sample of size $N$,

\begin{enumerate}
\item To split the dataset into $M$ sub-samples, take $m = [{N/M}]$ as the size of each split to make sub-samples of roughly the same sizes.
\item Generate a random sample of size $N$ from standard normal.
\item Order the generated sample.
\item Divide the ordered indexes into (M-1) parts of size $m$, and the $m$th chunk will include all the remaining indexes. These will be the index of subjects that should be assigned to each split.
\end{enumerate}

Note that, $M$ should be carefully selected. Two main points to consider when selecting $M$ are first to make sure the resulting sub-samples are small enough to keep the computation time at an acceptable level, but at the same time one needs to be careful that the sub-samples stay large enough, so the methodologies can still be applied on them in a valid way.

Assume $\widehat{\theta}_1, \ldots, \widehat{\theta}_M$ are the estimated parameters from $M$ sub-samples, the data splitting estimate of this parameter can be computed as follows:

\begin{equation}
\label{eq_estimate}
\widetilde{\theta} = \sum_{m=1}^M \widehat{\theta_m},
\end{equation}
where $\sum_{m=1}^M w_m = 1$. An appropriate weights could be assigned proportional to the size of each split. We would call such weights proportional weights in this thesis:

\begin{equation}
\label{eq_prop_w} 
w_{prop, i} = \frac{m_i}{N}.
\end{equation}

Of course, in case of random splitting, as the split sizes are roughly the same, one can simple assign equal weights to different sub-samples. Such weights are called equal weights in this thesis:

\begin{equation}
\label{eq_equal_w}
w_{equal,i} = \frac{1}{M}.
\end{equation}

The combination rule in case of horizontal splitting is simpler than vertical splitting. This comes from the fact that the sub-samples are independent in this case, therefore, one does not need to be concerned about the correlations between the estimates coming from different sub-samples. For a general weight $w_i$, consider $\mathrm{Var}(\widehat{\theta}_i) - \sigma^2_i$, then the variance of data splitting estimator can be computed as follows:

\begin{equation}
\label{eq_variance1}
\mathrm{Var}_{horizontal}(\widetilde{\theta}) = \sum_{m=1}^M w_m^2 \sigma_m^2.
\end{equation}

As having a smaller variance is a desired property for an estimator, Equation (\ref{eq_variance1}) would give an idea about an alternative way of assigning weights to our estimates: the weights should be computed in a way that the variance in (\ref{eq_variance1}) becomes minimized. In this thesis, such weights are called optimal weights. We define the optimal weights as the minimizer of the following objective function:

\begin{equation}
\label{eq_objective_opt_w}
Q = \sum_{m=1}^M w_m^2 \sigma_m^2 - \lambda \left(\sum_{m=1}^M w_m -1 \right),
\end{equation}

where $\lambda$ is a Largange multiplier to make sure that the sum of estimated optimal weights is still 1. Solving the optimization problem in (\ref{eq_objective_opt_w}) would lead to the following optimal weights,

\begin{equation}
\label{eq_opt_w}
w_{opt, i} = \frac{1/\sigma^2_i}{\sum_{m=1}^M 1/\sigma^2_m},
\end{equation}

As one may see, we may give more weights to a sub-sample with smaller variance.

In using optimal weights, one needs to be concerned about the fact that, optimal weights are including parameters from model, which then needs to be replaced by an estimate (we will discuss different ways of doing it in the coming chapters), and these estimates will come with their own variability. So, if the variability these estimates will bring together with them becomes larger than the variability that the optimal weights could decrease, then it is better to use the simpler parameter-free weights.

An example in our motivation datasets that random horizontal splitting can be beneficial is the Divorce in Flanders dataset, where clusters are formed by family members. While each family is at most of size 7, the number of total available families is large enough to make conventional software to fail. By splitting this large sample randomly and at the family level (horizontally), we could be able to fit the models and answer the research questions.

\subsection{Structured horizontal splitting}

As we have discussed, random horizontal splitting has been used by many researchers in different contexts to deal with big data issue. However, a randomly horizontal split would only decrease the sample size. So all of our gain is due to having smaller sample sizes. We have observed that they are cases where splitting using a pre-defined structure would provide some added benefits on top of just having smaller sub-samples.

It is a well-known fact that a balanced study deign works better than an unbalanced one. By balanced we mean, the number of available data for each cluster is the same. In fact, most of the studies are planned to be balanced, but in practice they would rarely end up like that. One can see several reasons for that, for example, the required materials are not enough, some patients die, some animals fail to follow the instructions, etc. Therefore, while having a balanced set of data is desired and even planned, we will usually have an unbalanced data to analyze.

Although most of the data in practice are unbalanced, one can usually find balanced sub-sets of those data. Take our example on a developmental toxicity study. There, the clusters are made of fetuses within litters. Obviously, different mothers would have different number of babies: an inevitably happened unbalanced data. However, it is not like that the number of fetuses within litter from every single mother is different from the others. Table \ref{tab_litter_size} shows unique number of fetuses in different litters with their frequencies. As one may see, we have a limited number of cluster sizes. Therefore, although the complete data is unbalanced, it is formed by balanced sub-sets. An structured horizontal splitting would assign all clusters of the same size to one sub-sample.

\begin{table}[ht]
\centering
\label{tab_litter_size}
\caption{Unique cluster sizes and their frequency in a developmental toxicity case study for the response DEHP.}
\begin{tabular}{lcccccccccccc}

  \hline
Cluster size & 9 & 10 & 15 & 8 & 12 & 13 & 14 & 6 & 7 & 3 & 11 & 16 \\ 
  Frequency & 10 & 9 & 3 & 12 & 20 & 18 & 6 & 4 & 5 & 4 & 9 & 4 \\ 
   \hline
\end{tabular}
\end{table}

As it was mentioned, there are several benefits for a balanced dataset, and in the special case of clustered data, we have shown that a closed-form solution exists when the cluster sizes are balanced, whereas for unbalanced data one needs to fit the models using iterative procedures. For very large datasets, the iterative procedures could be prohibitive and slow. While, a closed-form solution can be computed much faster. We have observed tens of thousands time faster computation using structured horizontal splitting based on the cluster sizes. Two cases are studied in details: clustered data with a compound-symmetry (CS) covariance structure, and clustered data with an autoregressive of order 1 (AR1) covariance structure. It worth to note that, even if the closed-form does not exist, with a balanced dataset one would have a faster and more stable convergence compared to unbalanced data.

The two considered covariance structures would cover a wide range of applications: where the correlation between to  members of a cluster is always the same (compound-symmetry), and when it decreases when the distance between the two observations becomes larger (AR1). For these two cases, the existance of the closed-form solution are shown using sufficient complete statistics. Then these closed-form solutions are obtained, also various kinds of weights, including optimal weights are computed for each case. On top of the already discussed weights, as we have clusters of all different sizes in each sub-sample here, we have proposed another type of weights, as follows,

\begin{equation}
\label{eq_size_prop_w}
w_{size-prop, i} = \frac{m_i n_i}{\sum_{k=1}^ M m_k n_k},
\end{equation}

In some cases that the cluster size is also important (other than the number of subjects in each sub-sample), such weights would perform better. We have shown that AR(1) is one of these cases, while for the CS, proportional weights would work fine.

Note that, the structured horizontal splitting idea is very general and all the introduced methodologies regarding different weights, combination rules, etc., can be used for any suitable structure of splitting (as long as they are horizontal), but in this thesis we only consider the cases where the structure is defined based on size of the clusters.

\subsection{Random vertical splitting}

The horizontal splitting techniques that we have introduced and discussed so far would be beneficial when the number of clusters ($N$) is large, but the cluster size is still small or moderate, so by breaking down the large $N$ into $M$ smaller sub-samples, we can manage to solve our problem. But there are examples where the size of some or all of the clusters is also very large. And as we have mentioned, large for a cluster size has a much smaller magnitude compared with large for the number of clusters. In such cases, horizontal splitting would not help much, because each cluster in our sub-samples is still large, so by horizontal splitting we just breakdown one problem into $M$ problems.

As n example, take the Amazon book ratings. There, we have the ratings for many books given by different buyers. Therefore, each book could possibly be rated by thousands of people. This will for very large clusters where even the simple models would fail to converge. In such a case, horizontal splitting cannot solve our problem. To solve the issue here, we have to perform the splitting within the subject. For example, if there are 5000 people rating a book, in order to make smaller clusters in our sub-samples, we have to split these 5000 people, e.g., into 100 sub-samples, each of size 50.

As long as the interest is in the parameter estimate, the combination rules in (\ref{eq_estimate}) is still valid. But the difficulty will appear in deriving the combination rule for the variances. As the data from different clusters are shared among different sub-samples, in other words, one cluster is present in more than one sub-sample, the estimated parameters from different sub-samples are correlated, and obviously, when combining the variances, these correlations should be taken into account.

One way to deal with the combination rule of variance in case of vertical splitting is to look at our sample splitting based model fitting in the pseudo-likelihood methodology, rather than the likelihood. In fact, once the dataset is analyzed in whole we use the conventional likelihood theory by finding the estimate of our parameters by maximizing the likelihood function. But when the data is splitted into some sub-samples, each of these sub-samples are analyzed using a likelihood theory, but their combination is not the original likelihood function anymore. It is called pseudo-likelihood. Roughly speaking, pseudo-likelihood theory tries to replace the complex or difficult likelihood function with something simpler or easier. But maximizing both likelihood or pseudo-likelihood functions would follow the same goal, i.e., estimate the same set of parameters, but with different precision.

From Cram\'{e}r-Rao's lower bound we know that the lower bound for variance can be achieved by a maximum likelihood estimator. In other words, any other estimator (including maximum pseudo-likelihood) would come with a variance larger than the one from a maximum likelihood. Therefore, having the easier problem to solve comes at the price of losing a bit of accuracy.

Let us define the pseudo-likelihood in plain English. Assume we have a sample $\mathbf{y}$ and using that we want to estimate the parameters vector $\theta$ via the log-likelihood function $\ell(\theta; \mathbf{y}$. Via sample splitting, we split the sample $\mathbf{y}$ into $M$ sub-samples $\mathbf{y_1}, \ldots, \mathbf{y_M}$, then feed them to the same likelihood function and estimate the parameter vector $\theta$ $M$ times. Under pseudo-likelihood paradigm, one can look at this as estimating a parameter vector $\Theta = (\theta_1, \theta_2, \ldots_3, \theta_M)$ using the sample $\mathbf{Y} = (\mathbf{y_1}, \ldots, \mathbf{y_M})$ via maximizing the pseudo-likelihood function $p\ell(\Theta, \mathbf{Y})$. Note that, $\theta_1 = \theta_2 = \theta_3 = \ldots = \theta_M$.

Now that the problem is defined under a pseudo-likelihood theory, all the results obtained there can be carried over to the data splitting as well. If we consider matrix $A$ that combines the estimates, i.e., $\widetilde{\theta} = A\widehat{\Theta}$, then the most important result we can present is:

\begin{equation}
\label{eq_clt_pseudolikelihood}
\sqrt{N}(\widehat{\theta} - \theta) = \sqrt{N}(A\widehat{\Theta} - A\Theta) \stackrel{d}{\sim} N\left(\mathbf{0}, A I_0^{-1} I_1 I_0^{-1} A^{T}\right),
\end{equation}

where $T$ means the transpose, and $I_0$ and $I_1$ are defined as follows,
\begin{equation}
\label{eq_hessian_gradient_pl}
I_0(\Theta) = \mathrm{E} \left(\frac{\partial^2 p\ell(\Theta; \mathbf{Y})}{\partial \Theta^T \partial \Theta} \right),\; I_1(\Theta) = \left[\left(\frac{\partial p\ell(\Theta)}{\partial \theta} \right)^T \frac{\partial p\ell(\Theta)}{\partial \theta}  \right].
\end{equation}

More details on pseudo-likelihood interpretation of the data splitting can be found in \cite{Iddi2011}. However, a main difficulty to find the variance in (\ref{eq_clt_pseudolikelihood}) if the need to cluster-wise hessian and gradient of the pseudo-likelihood function. This could easily become expensive. Another difficulty with this approach comes from the fact that, depending on the derivative and hessian means this variance estimator needs to computed case by case. For example, the results obtained for a logistic regression cannot used for another model without the necessary modification. To overcome these difficulties, we have considered an alternative solution proposed by \cite{hoffman2001} in different context. Their approach is called within cluster re-sampling, which was later on extended and renamed to \emph{multiple outputation} by \cite{follmann2003}.

In order to explain the motivation behind the multiple outputation technique, consider clustered data where the cluster consists of blood pressure of a patient visiting a hospital several times. The objective of the study is the severity of a disease which is highly correlated with blood pressure (more severe the disease, the larger the blood pressure). As the blood pressure is measured every time the patient is visiting the hospital, a larger cluster size, means the patient has visited the hospital more, which itself means the patient had more issues with the blood pressure, so a more severe state of the disease. In other words, in such situations the size of the cluster is also correlated with the response and objective of the study. Such situations are also called \emph{informative cluster size} in the literature. The correlation of cluster size and the response should be respected in the model, otherwise it could cause issues. Multiple outputation has been proposed to deal with this problem.

Within cluster resampling proposed by \cite{hoffman2001} to deal with the so-called informative cluster size, when it is a nuisance parameters, i.e., we are not interested to study the correlation of the response and cluster size, we just want estimates of other parameters which are not affected by this correlation. \cite{hoffman2001} proposed to take sub-samples of size 1 from each cluster repeatedly to form a sample on non-clustered data every time. Then the conventional methods can be used for each of these non-clustered data. The results then can be combined via combination rules which are similar to (\ref{eq_estimate}) for the estimates. But for the variance \cite{hoffman2001} proposed the following combination rule:

\begin{equation}
\label{eq_combination_variance}
\mathrm{Var}(\widetilde{\theta}) = W - \left(1 + \frac{1}{M}\right) B,
\end{equation}
where $W$ and $B$ are within and between sub-samples variances respectively. These quantities are defined as follows,
\begin{equation}
\label{eq_W_B_Var}
W = \frac{1}{M} \sum_{i=1}^M \sigma^2_m,\; B = \frac{1}{M-1} \sum_{i=1}^M \left(\widehat{\theta}_i - \widetilde{\theta} \right)^2.
\end{equation}

As one may see, the combination rule for the variance in (\ref{eq_combination_variance}) is also intuitively interesting also. It states, the variance of the data splitting estimator is an average of the variance in different sub-samples, but when we subtract it from the sample variance of estimated parameter from different sub-samples. Looking at this combination rule, one might also understand the reason behind naming this method multiple outputation by \cite{follmann2003}. That comes with various similarities of this technique with multiple imputation. Although, the similarities are technically rather than literally.

First let us say a word on multiple imputation itself. Multiple imputation (MI) is an effective way to deal with missing data issue. The traditional missingness remedies would try to either ignore that missing part of the sample, or replace it with a single plausible value. This way, one would treat a missing before and now imputed value the same as a value which is originally observed. In order to respect the fact that imputed values are actually imputed and as well could be any other value, \cite{rubin1978} proposed to replace each missing value with several values and not just one single value.

Now we may look at similarities of MI and MO. As it was mentioned, multiple imputation creates artificial data and augment it to the incomplete sample to make it complete, on the contrary, but in a similar way, multiple outputation would every time put some part of the same out of the analysis. Therefore, multiple \textbf{im}putation increases the sample size by including new artificial data, while multiple \textbf{out}putation decreases the sample size by selecting only a subset of them. But in both of the methods, we will have several datasets to analyze rather than the one single original dataset. So, in both methods we estimate each parameter several times. This is also reflected in the variance estimator. Let us remind the combination rule for variance from \cite{rubin1987},

\begin{equation}
\label{eq_comb_var_MI}
\mathrm{Var}(\widehat{\theta}_{MI}) = W + (1 + \frac{1}{M}) B.
\end{equation}

As one may see, the only difference of (\ref{eq_comb_var_MI}) and (\ref{eq_combination_variance}) is in the fact that MI would add the between datasets variability of the estimated parameter to the averaged variance, while MO would subtract this. And intuitively, that would also make sense, in case of MI we artificially increase the sample size, so we would get a smaller variance which needs a correction (to become larger, as it should be), but in case of MO we decrease the sample size, so a larger variance would be obtained, which is the also corrected in the same was as MI just with a different sign.

Because of the discussed similarity and for the vast use of multiple imputation in practice, we have developed part of contributions with the target to be used in data splitting for the case of multiple imputation. As for many cases, the conclusions and results from MI are also valid for MO, and vice versa.

We have extended the idea in within cluster re-sampling and multiple outputation to it them suitable for our purpose: analyzing clustered data with big clusters. The so-called iterative multiple outputation (IMO) procedure is proposed to fulfil this role. IMO is proposed as follows,

\begin{enumerate}
\item \textbf{Start.} Select an initial number of sub-samples, $M_0$, and sub-sampling size $m$. Take $M_0$ sub-samples of size $m$, fit the model to each and obtain $\widehat{\theta}_i$ and its variance $\Sigma_{\widehat{\theta}_i}$ ($i=1,\ldots,M_0$). Then compute 
\begin{equation}
\label{comb_rules}
\widetilde{\bftheta}_{M_0} = \sum_{i=1}^{M_0} \widehat{\bftheta}_i,\;\; \Sigma_{\widetilde{\theta_{M_0}}}=\widehat{W}_{M_0} - \left( \frac{M_0+1}{M_0} \right) \widehat{B}_{M_0}.
\end{equation}

			\item \textbf{Update.} For $m>M_0$, \begin{equation}
	\label{update}
	\widetilde{{\bftheta}}_{m+1}=\frac{m\widetilde{\bftheta}_m + \widehat{\bftheta}_{m+1}}{m+1},\; \Sigma_{\widetilde{\theta}_{m+1}}=\widehat{W}_{m+1} - \left( \frac{m+1}{m} \right) \widehat{B}_{m+1}.
	\end{equation}
	\item \textbf{Distance.} Compute: $d_{m+1}=d(\widetilde{\bftheta}_{m+1},\widetilde{\bftheta}_{m})$ using an appropriate distance.
	\item \textbf{Stopping rule.} $d_{j} < \varepsilon$ for $j=m+1,\ldots,m+k_0$.
\end{enumerate}
Where for $m=r$ we have,
\begin{equation}
\widehat{W}_r= \frac{\sum_{i=1}^r \sigma_{\widehat{\theta}_i}}{r},\; \widehat{B}_r= \frac{\sum_{i=1}^r (\widehat{\theta}_i - \widetilde{\theta}_r)(\widehat{\theta}_i - \widetilde{\theta}_r)'}{r-1}.
\end{equation}

We may remark that, there are cases only $M=1$ sub-sample would be sufficient to obtain results almost as efficient as analyzing the full dataset. We have studies this matter in detail and called such estimators \textit{finite information limit estimators}. We have also proposed procedures to detect this property.

A very useful dataset to illustrate the usefulness of random vertical splitting is the Amazon's book ratings dataset. This set of data consists of ratings of several books (from 1 to 5) on Amazon website. Each book is rated by different number of people, which could be as small as 1 and as large as 20,000. So we will deal with clustered data of very variable and large sizes.

\subsection{Structured vertical splitting}

So far we have discussed random horizontal and random vertical splitting, as well as structured horizontal splitting. In case of the vertical splitting also, pre-defining some structures to take the sub-samples other thank taking them just at random could be beneficial. We have already seen one application of structured splitting in case of horizontal splitting, which enabled us to form balanced sub-samples. Defining an structure is not always for computation efficiency. Sometimes, without such structure, estimating the model parameters is not possible.

We have considered an important case of jointly modeling several outcomes via random effects as a useful example for structured vertical splitting. Such a problem can easily become infeasible when the number of outcomes becomes large (even tens of them). \cite{fieuws2006, fieuws2007} proposed to analyze all possible pairs of all outcomes instead of analyzing them simultaneously. Then use appropriate combination rules to find the parameters of the model of interest. To illustrate the problem, also to show how structural vertical splitting is similar to the approach of \cite{fieuws2006, fieuws2007}. Also, to see how it is different from random vertical splitting, we present a simple example.

Imagine we have 3 outcomes to be jointly modeled. A random effects approach would fit a separate model to each outcome, then let them to be correlated by defining a joint distribution for their random effects. This joint distribution is usually a normal distribution where its covariance matrix would capture the dependencies between different outcomes. Let us elaborate more here, if the three separate models (all linear) for the three outcomes are defined using the following general form:

\begin{equation}
\label{eq_SV_models}
\begin{cases}
y_{1ij} = \beta_1 + b_{1i} + \epsilon_{1ij} \\
y_{2ij} = \beta_2 + b_{2i} + \epsilon_{2ij} \\
y_{3ij} = \beta_3 + b_{3i} + \epsilon_{3ij},
\end{cases}
\end{equation}
we let the three random intercepts to be normally distributed as follows,
\begin{equation}
\label{eq_SV_D}
\begin{bmatrix}
b_{1i} \\
b_{2i}\\
b_{3i}\\
\end{bmatrix} \sim N \left ( 
\begin{bmatrix}
0 \\
0\\
0\\
\end{bmatrix}, 
D=\begin{bmatrix}
D_{11} & D_{12} & D_{13}\\
 & D_{22} & D_{23}\\
&& D_{33}\\
\end{bmatrix}
\right).
\end{equation}
Therefore, the parameter of interest, $\bftheta$, is as follows:
\begin{equation}
\label{eq_SV_theta}
\bftheta= (\beta_1,\beta_2,\beta_3, D_{11},D_{22},D_{33},D_{12}, D_{13}, D_{23}).
\end{equation}
As we have $3$ outcomes, the number of pairs will be $M=3\times (3-1) /2 =3$. However, unlike random vertical splitting where usually all parameters of interest are estimated in each sub-sample, here, because of the special structure we have imposed on the sub-samples, each of them would estimate part of parameter vector. The parameters estimated in each pair ($\bftheta^{(s)}$) are as follows:
\begin{equation}
\label{eq_SV_pairs}
\begin{cases}
\bftheta^{(1)}=\theta_{(y_{1ij},y_{2ij})} = (\beta_{1_1},\beta_{2_1},D_{11_1},
,D_{22_1},D_{12_1}) \\
\bftheta^{(2)}=\theta_{(y_{1ij},y_{3ij})} = (\beta_{1_2},\beta_{3_2},D_{11_2},
,D_{33_2},D_{13_2}) \\
\bftheta^{(3)}=\theta_{(y_{2ij},y_{3ij})} = (\beta_{2_3},\beta_{3_3},D_{22_3},
,D_{33_3},D_{23_3}),
\end{cases}
\end{equation}
as one may see in (\ref{eq_SV_pairs}), each $\bftheta^{(s)}$ ($s=1,2,3)$ is a subset of $\bftheta$ in (\ref{eq_SV_theta}. Now, the stacked vector combining all parameters estimated in different sub-samples $\boldsymbol{\Theta}'=(\bftheta^{(1)},\bftheta^{(2)},\bftheta^{(3)})$ can be constructed as follows:
\begin{equation}
\label{eq_SV_Theta}
\boldsymbol{\Theta}'=(\beta_{1_1},\beta_{2_1},D_{11_1},
,D_{22_1},D_{12_1}, \beta_{1_2},\beta_{3_2},D_{11_2},
,D_{33_2},D_{13_2},\beta_{2_3},\beta_{3_3},D_{22_3},
,D_{33_3},D_{23_3}).
\end{equation}
As one may see, the covariance parameters ($D_{12},D_{13}, D_{23}$) are only estimated once using their corresponding pair, but the rest of parameters are appearing in two out of three pairs. Therefore, to combine these estimates we cannot just average all estimates as we did in (\ref{eq_estimate}). The weight matrix to perform the appropriate combination can be defined as follows,
\vspace*{2mm}
\renewcommand{\kbldelim}{(}% Left delimiter
\renewcommand{\kbrdelim}{)}% Right delimiter
% \resizebox{\textwidth}{!}{  
\[
  A = \kbordermatrix{
    & \beta_{1_1}&\beta_{2_1}&D_{11_1}
	&D_{22_1}&D_{12_1}& \beta_{1_2}&\beta_{3_2}&D_{11_2}
	&D_{33_2}&D_{13_2}&\beta_{2_3}&\beta_{3_3}&D_{22_3}
	&D_{33_3}&D_{23_3} \\
\beta_1 & 1 & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\ 
  \beta_2 & 0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\ 
  \beta_3 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 \\ 
  D_{11} & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\ 
  D_{22} & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\ 
  D_{33} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 1 & 0 \\ 
  D_{12} & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\ 
  D_{13} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\ 
  D_{23} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 
  }
\] 
%}

\vspace*{2mm}
\noindent Obviously, $\sum_i A_i$ (with $A_i$ the $i$th row of $A$) shows the number of times each parameter is estimated. Therefore, by defining $W$ with each of its rows computed as $A_i / \sum_i A_i$, and pre-multiplying it by $\widehat{\boldsymbol{\Theta}}$ would provide the appropriate averaging. The same goes for the variance combination rule. \cite{fieuws2006, fieuws2007} used a pseuo-likelihood approach to find the combined covariance matrix (the same as in \ref{eq_clt_pseudolikelihood}), we could show using \ref{eq_combination_variance} with appropriate weights works the same, but much faster.

As one may see, only the covariance matrix of the random effects have $6$ parameters to be estimated, if one decides to add random slopes also, then the number of parameters will increase to $(6\times7)/2 = 21$. Once the number of responses becomes slightly large, the number of parameters will dramatically increase. This would face all conventional methods with serious problems, and the problem can be unfeasible. Figure \ref{fig_strucVert} shows how the number of parameters only in the covariance matrix of random effects would increase when the number of responses increases. As we may see in this figure, even for less than 100 responses, the number of parameters becomes extremely large.

\begin{figure}
\centering
\includegraphics[width=\textwidth]{strucVert.eps}
\caption{Number of parameters in the covariance matrix of jointly normally distributed random effects for different number of responses using only random intercept or random intercept and slope models.} 
\label{fig_strucVert}
\end{figure}

Therefore, there is a main difference between the applications we have proposed for random vertical splitting, and structured vertical splitting. While both of them are proposed to deal with large clusters, the issue with random vertical splitting is a large sample size, while structured vertical splitting deals with a high-dimensional parameter space. In other words, in case of structured random splitting we are dealing with the so-called curse of dimensionality \citep{donoho2000}. In fact, exactly for this \emph{big} parameter space, we had to impose a structure on it, so in every sub-sample a sub-set of this parameter vector could be estimated. The smart idea of modeling several outcomes by all their sub-sets of two would break down the parameter space by breaking down the sample itself. That is why, in case of random splitting, every sub-sample usually estimates all parameters, while this is not often the case for structured vertical splitting.

We have two examples in our motivational datasets which are used to illustrate the applications of structured vertical splitting. We have used \emph{Leuven diabetes project} data as a basis to perform simulation studies to evaluate our proposed way of combining variances with what pseudo-likelihood theory would give. Furthermore, the \emph{Hearing data} is used as an example of high-dimensional parameter space where the conventional methods would fail.

\section{Supplementary topics}

As it was mentioned, due to several technical similarities between random vertical data splitting (iterative multiple outputation) and multiple imputation, and due to wide acceptance of MI in different research fields, we have developed part of our required methodology in MI language rather than language of IMO. However, the conclusions are valid in both cases. In this section we briefly go over these contributions which will be presented in Chapter \textbf{NUMBER OF CHAPTERS}.

\subsection{How many sub-samples?}

The iterative multiple outputation (IMO) procedure that we have proposed for random vertical splitting displays a stopping rule at step 4. This is used to determine the number of sub-samples. Let us briefly explain the idea behind this stopping rule in this section. Obviously, more sub-samples are better in case of obtaining a more precise estimate. But on the other hand, as the sub-samples are with replacement, on can take an infinite number of sub-samples, so there is a need for a rule to stop the IMO procedure.

As we have seen in (\ref{eq_estimate}), the combination rule of data splitting for parameter estimate is averaging. From laws of large numbers we know that under general regularity conditions, as the sample size increases, the average converges to a finite quantity. We have seen that also via central limit theorem in (\ref{eq_clt_pseudolikelihood}). As we average over sub-samples, sample size here means the number of sub-samples. Therefore, we would expect, after some point, increasing the number of sub-samples would not impressively change the final results. But it is important to come up with appropriate criteria and stopping rule to stop the sub-sampling at the right moment. It is important for two main reasons: 1- if we stop early, we would fail to provide precise and correct estimates, 2- if we stop too late our initial intention for saving computation time could be compromised.

We have already explained the connection of multiple imputation and IMO, also how they would suffer from the same issues. In multiple imputation also, a main issue is to determine the number of imputed datasets. Therefore, we have studied this question in detail via simulation studies and theoretical discussion. Various distance function (Mahalanobis, Euclidean, etc.) were examined for different situations and as a result we proposed the iterative multiple imputation procedure. The obtained results are all valid in case of IMO, therefore, we have used the same stopping rule in IMO and in IMI.

In order to illustrate IMI, we have used two datasets, the \emph{Leuven Eye Study} (LES) and the \emph{Age-related macular degeneration} (ARMD). LES is collected via one of the largest observational studies for glaucoma. However, the dataset includes a lot of missing data. Practically, no fully observed line exists in this dataset, so without MI the analysis was not possible.

The second dataset, which is also ophthalmology related, come from a clinical trial that tries to compare two methods to diagnose the age related mascular degenaration which causes central vision loss. The final dataset is highly unbalanced, and only includes 17 clusters. \cite{van2016} proposed to make the clusters balanced by imputation them using MI.

For both of these datasets, determining a correct and sufficient number of imputed datasets is very important, as due to large amount of missing values, as well as small sample size in ARMD, the traditional 5 imputed datasets recipe would not be sufficient.

\subsection{Special case: where combination rule fails}

A fundamental rule which should be followed when using the combination rule in (\ref{eq_estimate}) is to average parameters which are counterparts of the same parameter in the model for the full sample. In some applications, finding such counter parts is not a straightforward task. An example of such situations is principal component analysis (PCA). Methods like PCA are working based on eigenvalue decomposition of the covariance (or correlation) matrix. Eigenvalues of a matrix $\Sigma$ can be ontained by solving the following equation,

$$|\Sigma-\lambda I|,$$

where $|.|$ for a matrix denotes the determinant of it and $I$ is the identity matrix of the same order as $\Sigma$. The roots of this equation are the eigenvalues. Obviously, there is no natural ordering among roots of a polynomial but as in PCA one can show that \citep{mardiamultivariate} these eigenvalues are actually variances of their corresponding principal components, therefore, a PC with a larger eigenvalue explains a larger propportion of variance.

The issue arises when we use data splitting or multiple imputation where the eigenvalue decomposition should be done for several sub-samples or several imputed datasets. There is no guarantee that the largest eigenvalue from first datasets corresponds to the largest eigenvalue from the second dataset. The problem of combining results of PCA or exploratory factor analysis at the factor (or principal components) level has been studies in \cite{lovik2017combining}.

Our alternative solution here to first estimate the covariance matrix as out parameter of interest, and then perform the eigenvalue decomposition on this single estimated covariance matrix. We have studied this idea in detail and investigated different aspects of it, also compared it with other alternative solutions. Again, the problem is considered in the setting of multiple imputation, but all the findings are also valid for data splitting. This is because, the issue in this case is that the dataset under study is replaced with several datasets. Now these several datasets could come from data splitting or multiple imputation, that would not change the problem, hence, the proposed solutions.

\section{Software implementation}

Our main concern is R, but for comparison reasons, we have also used SAS.

### 2019-06-25 thesisMain.tex
all files 6/25/2019

\documentclass[oneside]{scrbook}

\title{A Sample Thesis} 
\author{A.N. Other} 
\date{July 2013} 
\titlehead{A Thesis submitted for the degree of Doctor of Philosophy} 
\publishers{School of Something\\University of Somewhere}

\begin{document} 
\maketitle

\frontmatter 
\tableofcontents 
\listoffigures 
\listoftables

\chapter{Acknowledgements}

I would like to thank my supervisor, Professor Someone. This 
research was funded by the Imaginary Research Council.

\chapter{Abstract}

A brief summary of the project goes here.

\mainmatter

\include{intro}

\backmatter

\end{document}

### 2019-06-25 verticalStructured.tex
all files 6/25/2019

\section{Fast precision estimation in high‐dimensional multivariate joint models}

\subsection{Introduction}
\label{sec_intro}

There are different approaches to deal with modelling multivariate clustered data. Authors like \cite{pawitan1993} proposed to analyze each response conditioned on all the others. As an alternative, one may specify a marginal model for each response separately and then join them using an appropriate way. Among such ways are using a latent variable \citep{catalano1992}, copulas \citep{nelsen2007}, or random effects \citep{laird1982}. The latter is of interest in this short note.

%In the joint modelling approach of multivariate repeated measurements data using random effects, a mixed model is considered for each response separately. Then, the random effects of these models are considered to follow a multivariate distribution which would reflect the association between different responses.

Let $y_{rij}$ be the $j$th measurement taken on the $i$th cluster for the $r$th outcome, $i=1,\ldots,N$, $r=1,\ldots,m$ and $j=1,\ldots,n_{ri}$. Obviously, the number of measurements available for subject $i$ is $\sum_{r=1}^m n_{ri}$. Note that the various outcomes do not need to be of the same kind (continuous, binary, count data, etc.). Also there is no need to have sequences of the same length for each cluster or each outcome. Let $\mathbf{Y_{ri}}$ be the $n_{ri}$ measurements taken on cluster $i$ for outcome $r$. The model assumes that each $\mathbf{Y_{ij}}$ satisfies a mixed model:
\begin{equation}
\label{mixed_model}
\mathbf{Y_{ri}} = \boldsymbol{\mu_{ri}} + \epsilon_{ri},
\end{equation}
where
\begin{equation}
\label{mu_function}
\mathbf{\mu_{ri}} = \mathbf{h}(X_{ri} \boldsymbol{\beta_r} + Z_{ri} \mathbf{b_{ri}}).
\end{equation}
As one may see in (\ref{mu_function}), each outcome can have its own set of fixed effects. In order to capture the associations between different responses, the mixed model approach for joint modelling of the $m$ outcomes allows for correlation between the random effects of each outcome. Let $\mathbf{b_i} = (\mathbf{b_{1i}},\ldots,\mathbf{b_{mi}})'$; the joint model assumes:
\begin{equation}
\label{random_effect_dist}
\mathbf{b_i}=
\begin{bmatrix}
b_{1i} \\
b_{2i}\\
\vdots\\
b_{mi}\\
\end{bmatrix} \sim N \left ( 
\begin{bmatrix}
0 \\
0\\
\vdots\\
0\\
\end{bmatrix}, 
D=\begin{bmatrix}
D_{11} & D_{12} & \ldots & D_{1m}\\
D_{21} & D_{22} & \ldots & D_{2r}\\
\vdots & \vdots & \ddots & \vdots\\
D_{m1} & D_{r2} & \ldots & D_{mm}\\
\end{bmatrix}
\right).
\end{equation}
The matrix $D_{rs}$ ($r,s=1,\ldots,m$) represents the covariance between $\mathbf{b_{ri}}$ and $\mathbf{b_{si}}$.

While using the joint modelling approach for the multivariate clustered data is a flexible methodology, it can easily become intractable when the number of outcomes becomes large. As the matrix D contains the association between all outcome sequences, its dimension can quickly grow large. Therefore, it is easily possible that such a flexible joint model cannot be used in practice.

In order to overcome this drawback, \cite{Verbeke2006} and \cite{Verbeke2007} proposed a pseudo-likelihood based approach which considers all possible pairs (or higher order subsets) of the $r$ outcomes and fit them separately. Obviously, using this methodology some parameters are estimated more than once. Averaging multiply estimated parameters gives a final estimate for each parameter. To find a correct estimate of the variance, \cite{Verbeke2006} and \cite{Verbeke2007} proposed a sandwich-type correction of the form $J^{-1}KJ^{-1}$, where $J$ and $K$ are formed based on cluster-wise Hessians and gradients of the log-pseudo-likelihood function, respectively.

The need for cluster-wise derivatives of the first and second order makes computing this sandwich estimator difficult and possibly expensive. \cite{pair_lin} and \cite{kundu} provided {\tt{SAS}} implementations for the case of linear mixed models. Also, \cite{pair_gen} considered the case of generalized linear models using {\tt{Proc NLMIXED}} in {\tt{SAS}}. However, the cluster-wise calculation can become very expensive for large samples. Also, the focus of these implementations is on the fixed effects only. Implementing the variance estimation for, e.g., matrix $D$ (\ref{random_effect_dist}) is far more complicated. Also, obtaining the right precision quantities in commonly available software like {\tt{R}} demands new implementations. Furthermore, a different model needs its own implementation which includes evaluating gradients and Hessians of the log-(pseudo-)likelihood function.

\cite{hoffman2001} proposed within-cluster re-sampling to analyze clustered data when the intra-class correlation is considered a nuisance. Their idea was explored in \cite{follmann2003} who gave it the name multiple outputation (MO). MO consists of repeatedly taking sub-samples of size $1$ from each cluster to form a sub-sample of independent observations, to then apply the standard methodologies for such data. The estimated parameters  and their precisions from several sub-samples will be combined using appropriate rules. \cite{hoffman2001} have shown that their variance estimator is consistent.

{\color{black}{Note that using pairs of variables instead of analyzing them jointly is a trick which is not restricted to likelihood-based methods. It has been used in different contexts. For example, \cite{aas2009} have used pair-copulas to model complex multivariate dependencies. Especially with Gaussian outcomes, pairwise approaches (e.g., vine copulas) allow retrieval of the full multivariate association structure, without the need for full joint modelling.}}

In this short note we extend the idea of MO to compute the variance of a pairwise-fitted multivariate joint model as a faster and easier way. The MO-based correction for the variance will be presented in Section~\ref{sec_method}. A simulation study will compare its performance with a sandwich correction in Section~\ref{sec_sim}. The two corrections are compared in analyzing a real dataset in Section~\ref{sec_application}, and finally the paper is concluded in Section~\ref{sec_conclusions}.

\subsection{A fast variance estimator based on multiple outputation}
\label{sec_method}
Consider $y^{(s)}_{rij}$ as the $j$th measurement on $i$th cluster for $r$th outcome in the $s$th sub-sample, where $s=1,\ldots ,M$, $j=1,\ldots,n_{ri}$, $i\in \{1,\ldots,N\}$, and $r\in \{1,\ldots,m\}$. Also consider $\boldsymbol{\theta}^{(s)}$ a vector of size $p_s$ ($s=1,\ldots,M$), containing parameters estimated using the variables involved in the $s$th sub-sample, obviously, $\boldsymbol{\theta}^{(s)} \subseteq \boldsymbol{\theta}$. Let $\boldsymbol{\Theta}' = (\boldsymbol{\theta}^{(1)},\ldots,\boldsymbol{\theta}^{(M)})$ the stacked vector combining all parameters estimated in different sub-samples. $\widehat{\boldsymbol{\Theta}}$ replaces each $\boldsymbol{\theta}^{(s)}$ by its estimate. Also, $\boldsymbol{\Sigma}_{\boldsymbol{\Theta}}$ is a vector of length $\sum_{s=1}^M\times p_s$, with its elements the variance of each element of $\widehat{\boldsymbol{\Theta}}$.

The MO in the current paper extends the one in \cite{hoffman2001} and \cite{follmann2003} in two main directions. First, the sub-samples in our approach are of size larger than one. The second aspect of our proposed extension of MO concerns the sub-sampling scheme. The MO in the sense of \cite{hoffman2001} and \cite{follmann2003} takes random sub-samples of size $1$ from each cluster many times to form sub-samples of independent observations. In our approach, the sub-samples are not taken at random but based on a pre-defined structure which implies a fixed number of needed sub-samples.  {\color{black}{{One may note that the sub-sampling in our proposal is performed at the response level. Therefore, as it is not necessary to measure every response for all subjects, each subject will be presented in the sub-samples which the responses presented there are measured for it.}}}

For example, using a pairwise approach, each sub-sample consists of all the observations available for each subject corresponding to the two outcomes in the pair under analysis. Therefore, having $m$ outcomes, we need to take $M=m(m-1)/2$ sub-samples, each with a size larger than one. Note that, in a pairwise approach, $M=m(m-1)/2$ is the number of all possible sub-samples. Taking less sub-samples would fail to estimate some parameters. On other hand, taking more sub-samples is not possible without repeating some already taken sub-samples.  {\color{black}{{As it was remarked, it is not necessary for every subject to be present in every sub-sample. In a pairwise approach, suppose we only have three responses $y_1,y_2,y_3$ and for a particular subject only $y_1$ is measured. Then this subject will only be present in pairs $(y_1,y_2)$ and $(y_1,y_3)$.}}}

%Note that in a pairwise modelling approach, each sub-sample consists of the measurements corresponding to one pair of outcomes. Therefore, having $m$ outcomes, we have $M=m(m-1)/2$ sub-samples. For example, the sub-sample corresponding to the first pair (first and second responses) would consists of $\sum_{j=1}^K (n_{1j}+n_{2j})$ observations. This can be considered an extension of MO in the sense of \cite{hoffman2001} and \cite{follmann2003} that takes sub-samples of size one from each cluster.

As one may have already noticed, some of the parameters in $\boldsymbol{\theta}$ would have more than one counterpart in $\boldsymbol{\Theta}$. In order to bring the $\boldsymbol{\Theta}$ back to the desired parameter space, one may take the average of possible several counterparts of each element of $\boldsymbol{\theta}$ in $\boldsymbol{\Theta}$. This can be done using an appropriate weight matrix pre-multiplied by $\boldsymbol{\Theta}$. Consider $A$, a $(p\times \sum_{s=1}^M p_s)$ matrix where its $(\ell,k)$th element, $\ell=1,\ldots,p$, $k=1,\ldots, (p\times \sum_{s=1}^M p_s)$, is defined as follows:

\begin{equation}
\label{matrix_A}
A_{\ell,k} = \left\{
\begin{array}{rl}
1 & \text{if } \boldsymbol{\Theta}_{k} \text{ is a {\color{black}{component}} of } \boldsymbol{\theta}_{\ell},\\
0 & \text{otherwise},
\end{array} \right.
\end{equation}

where $\boldsymbol{\Theta}_{k}$ and $\boldsymbol{\theta}_{\ell}$ are the $k$th and $\ell$th elements in $\boldsymbol{\Theta}$ and $\boldsymbol{\theta}$, respectively. If $W$ is defined as $A$ but with the rows divided by their sum then the combination rules for combining the estimates from $M$ sub-samples are as follows,
%\begin{equation}
%\label{comb_rules}
%\begin{cases}
%\widetilde{\boldsymbol{\theta}} = W \widehat{\boldsymbol{\Theta}},\\
%\mathrm{diag}\left\{\mathrm{Var}(\widetilde{\boldsymbol{\theta}} )\right\}=  W \boldsymbol{\Sigma}_{\boldsymbol{\Theta}} - \mathrm{diag}\left\{W \left(\widehat{\boldsymbol{\Theta}}-\bar{\boldsymbol{\Theta}} \right) \left(\widehat{\boldsymbol{\Theta}}-\bar{\boldsymbol{\Theta}} \right)' W'\right\},
%\end{cases}
%\end{equation}

\begin{equation}
\label{comb_rules}
\begin{cases}
\widetilde{\boldsymbol{\theta}} = W \widehat{\boldsymbol{\Theta}},\\
{\color{black}{\mathrm{diag}\left\{\mathrm{Var}(\widetilde{\boldsymbol{\theta}} )\right\}=  W \boldsymbol{\Sigma}_{\boldsymbol{\Theta}} - W \mathrm{diag}\left\{\left(\widehat{\boldsymbol{\Theta}}-\bar{\boldsymbol{\Theta}} \right) \left(\widehat{\boldsymbol{\Theta}}-\bar{\boldsymbol{\Theta}} \right)' \right\}}},
\end{cases}
\end{equation}

where $A'$ is the transpose of $A$, and $\bar{\boldsymbol{\Theta}} = \widehat{\boldsymbol{\Theta}} - A' \tilde{\boldsymbol{\theta}}$. For the proof, see \citet[][Appendix 2]{hoffman2001}. These authors have shown that $I^{1/2} ( \widetilde{\boldsymbol{\theta}} - \boldsymbol{\theta}) \xrightarrow{d} N(0, \Sigma)$, where $\xrightarrow{d}$ means convergence in distribution and $\mathrm{Var}(\widetilde{\boldsymbol{\theta}} )$ is a consistent estimator of $\Sigma$.

We may note that the combination rules in (\ref{comb_rules}) are only estimating the diagonal elements of the covariance matrix of $\widetilde{\boldsymbol{\theta}}$.  If one is interested in estimating the off-diagonal elements as well, the combination rules are the same. One would only need to construct $\boldsymbol{\Sigma}_{\boldsymbol{\Theta}}$, which takes into account the off-diagonal elements also, then an appropriate $W$ can be used to combine the estimated variances and covariances obtained from each sub-sample. The corresponding vector of unique parameters in this covariance matrix would be $p(p+1)/2$, where each sub-sample would estimate part of them.

{\color{black}{As one may see in (\ref{comb_rules}), the combination rule for the variance has a subtraction, so as pointed out by both \cite{hoffman2001} and \cite{follmann2003}, theoretically, it is possible to estimate a non-positive definite covariance matrix. Although the chance for such an occurrence is small, and we have not observed it even once in our simulations, it is still necessary to discuss this matter. We can consider three situations in which such a phenomenon would occur: (1) the sample size, $N$, is not sufficiently large, (2) the number of sub-samples, $M$, is not sufficiently large, and (3) the size of each sub-sample is not sufficiently large.

\cite{hoffman2001} proposed to increase the number of sub-samples, $M$, to solve this problem. As it was pointed out, unlike the usual MO, in our approach the number of sub-samples is fixed. Therefore, in case of a non-positive definite covariance estimate, we suggest to take larger sub-samples each time. Note that taking such a strategy is not possible in the usual MO, since it takes one observation per cluster, so the sub-sampling size is fixed. Taking larger sub-samples in our case means to consider triples, quadruples, etc., instead of pairs. We expect the problem of non-positive definite covariance estimate to vanish when taking larger sub-samples.

When maximum likelihood estimator (MLE) is used, \cite{follmann2003} proposed a variance estimator based on MLE approximation, using first order Taylor expansion of the score function. This approach requires the evaluation of the subject-wise gradient and Hessian of the log-likelihood function, which is similar to the requirements of the sandwich estimator. Therefore, in case where even with larger sub-samples a non-positive definite covariance estimate persists, we suggest to use the sandwich estimator instead of MO.}}

\subsection{Simulations}
\label{sec_sim}
In this section the proposed idea will be explored and examined via two simulations studies. Sub-section~\ref{subsec_lin} considers joint modelling of linear mixed models, and Sub-section~\ref{subsec_genlin} takes a joint model of ordinal data in a generalized linear mixed models context.

\subsubsection{Gaussian responses}
\label{subsec_lin}
Consider the linear case with identity link in (\ref{mu_function}), as follows:
\begin{equation}
\label{lin_mixed}
\BY_i= X_i \beta + Z_i \Bb_i + \epsilon_i,
\end{equation}
where $X_i$ is the design matrix regarding fixed effects, $\beta$ represents the fixed effects, $Z_i$ is the design matrix related to the random effects, and $\Bb\sim N(0,D)$ shows the random effects. The errors are represented by $\epsilon_i\sim N(0,R)$. In order to evaluate the proposed method in a simulation study, consider the model in (\ref{lin_mixed}) with parameters:

\begin{equation}
\label{sim_params}
\beta=(5,5,5,5), D= \left(
\begin{array}{cccc}
60 & 40 & 25 & 40 \\
& 50 & 35& 30 \\
& & 45 & 40 \\
&&&55
\end{array} \right), R= 196 I_{n_i},
\end{equation}
where $I_{n_i}$ is the identity matrix of order $n_i$, with $n_i$ the length of the $i$-th cluster. {\color{black}{This means that both of the $X_i$ and $Z_i$ are formed by columns of $1$'s.}} The data are generated from such model 100 times, each consists of 100 clusters of size $n_i=5$. The mixed model is fitted to the full sample as well as each of the 6 pairs using Proc MIXED in {\tt{SAS}}. Then, the estimates and variances are combined using (\ref{comb_rules}). Also, the sandwich estimator is calculated using {\tt{\%jointpair}} macro in {\tt{SAS}} in \cite{pair_lin}.

Figure \ref{fig_sim} shows the simulation results. As one may see, the relative efficiency of estimating fixed effects using sandwich and MO corrections are very similar. Furthermore, the relative efficiency for estimating parameters in the $D$ matrix are also presented. These parameters are only estimated using the MO correction, as the implementation was not available for sandwich-correction. The interesting result here is the computation time. As one may see, using MO-correction the computation can be done more than 2500 times faster. On the other hand, in case that computing the full likelihood model is possible (as it is the case in our simulation study), MO-correction can also possibly be computed faster than the full-likelihood. Therefore, it is not only an alternative for infeasible likelihoods, but also, it can be used to accelerate computations that are strictly speaking doable.

%\begin{figure}
%\centering
%\includegraphics[width=\textwidth,trim=0 110 0 120]{fig_sim.eps}
%\caption{Histogram of computation time using IMO and sandwich corrections.} 
%\label{fig_scatter2}
%\end{figure}

\begin{figure}[htb]
\begin{center}
\includegraphics[width=\textwidth]{fig_sim_eq_new.eps}
\caption{Simulations {\color{black}{(Gaussian outcomes)}}. Top: Relative efficiency of estimated fixed effects using MO-correction (left) and sandwich-correction (middle), estimated $D$-matrix using MO-correction (right) compared to the full model. {\color{black}{The relative efficiency is defined as the ratio of the variance of the alternative method and the full likelihood, e.g., for MO $\mathrm{Var}(\widetilde{\theta}_{\mathrm{MO}})/\mathrm{Var}(\widehat{\theta}_{\mathrm{full}})$ is computed}}. Bottom: Histograms of computation time (in seconds) using sandwich and MO corrections (first and second). Histograms of computation time (in seconds) using sandwich and MO corrections (third and fourth) relative to the full model computation time.}
\label{fig_sim}
\end{center}
\end{figure}

{\color{black}{
In order to see how our proposal performs comparing with sandwich estimator when the correlation between the random effects is small, the simulation above is repeated, but this time the off-diagonal elements of $D$ matrix in (\ref{sim_params}) are divided by $10$. Figure~\ref{fig_sim_small} shows the relative efficiencies of the variances obtained using MO and sandwich corrections. As one may see, these relative efficiencies are again very close to $1$.

\begin{figure}[htb]
\begin{center}
\includegraphics[width=\textwidth]{fig_sim_small.eps}
\caption{{{Simulations (Gaussian outcomes). Top: Relative efficiency of estimated fixed effects using MO-correction (left) and sandwich-correction (right). Bottom: estimated $D$-matrix using MO-correction compared to the full model. The relative efficiency is defined as the ratio of the variance of the alternative method and the full likelihood, e.g., for MO $\mathrm{Var}(\widetilde{\theta}_{\mathrm{MO}})/\mathrm{Var}(\widehat{\theta}_{\mathrm{full}})$ is computed.}}}
\label{fig_sim_small}
\end{center}
\end{figure}
}}

{\color{black}{

\subsubsection{Non-Gaussian responses}
\label{subsec_genlin}

In order to investigate the performance of our proposal for non-Gaussian responses and more complex models a simulation is done based on the Leuven diabetes project. This dataset is analyzed in \cite{ivanova2015}, details about this study can be found there. Here we only review parts necessary for our simulations. In order to study how well the diabetes is controlled, three outcomes were measured for each patient at baseline ($T_0$) and one year later ($T_1$). These three outcomes of interest were LDL-cholesterol (low-density lipoprotein cholesterol, mg/dl), HbA1c (glycosylated hemoglobin, \%), and SBP (systolic blood pressure, mmHg). Although these three variables are continuous, the interest was in expert-specified cut-off values of these outcomes. The cut-off values are defined as $\{<100, [100,115), [115, 130),\geq 130 \}$ (mg/dl) for LDL-cholesterol, $\{<7, [7, 8), \geq 8\}\%$ for HbA1c, and $\{ \leq 130, (30,140],(140,160],>160\}$mmHg for SBP.

In order to analyze these 3 ordinal outcomes together, a joint modelling approach with a proportional odds mixed model (POMM) for each response was considered in \cite{ivanova2015}. The POMM for each response is as follows:
\begin{equation}
\label{sim2_model}
\mathrm{logit}[P(Y_{rij}<k)] = \xi_{r0k} + \xi_{r1} t_{ij} + \xi_{r2} X_{i} + b_{ri} + \epsilon_{rij},
\end{equation}
where $r=1,2,3$ indicates the outcome of interest, and $k=1,\ldots,K-1$ with $K$ as the number of levels of the corresponding outcome of interest. Furthermore, $X_i$ shows the gender of the subject, and $b_{ri}$'s are the random intercepts corresponding to each outcome. Following the mixed model approach for joint modelling of multiple outcomes, we assume the three random intercepts to be normally distributed with the following covariance matrix.

\begin{equation}
\label{sim2_random_effect_dist}
\begin{bmatrix}
b_{1i} \\
b_{2i}\\
b_{3i}\\
\end{bmatrix} \sim N \left ( 
\begin{bmatrix}
0 \\
0\\
0\\
\end{bmatrix}, 
D=\begin{bmatrix}
D_{11} & D_{12} & D_{13}\\
 & D_{22} & D_{23}\\
&& D_{33}\\
\end{bmatrix}
\right).
\end{equation}
The parameter values for generating the data were obtained by fitting the joint model to the Leuven diabetes dataset. The data were generated $200$ times using {\tt{SAS Proc IML}}. Each time the parameters in the $3$ pairs were estimated using adaptive Gaussian quadrature \citep{pinheiro1995} with $3$ quadrature points in {\tt{SAS Proc NLMIXED}} and the combined variance was estimated using sandwich and MO corrections. Considering the complexity of the model, in order to do the simulation quicker, the codes were run on VSC cluster in Leuven, Belgium.

The results of the simulations are summarized in Table~\ref{tab_sim2}. Column 3 of Table~\ref{tab_sim2} shows the parameter values which were used to generate the data. The last three columns show the means and standard deviations over 200 replications for estimated variance using MO, sandwich, and their relative efficiency, respectively. As one may see, for all of the variables the estimated variances are very close, therefore, one can conclude that our proposal performs well also in a complex model with non-Gaussian responses.

\begin{table}[ht]
\centering
{{
\caption{Simulations (non-Gaussian outcomes). The estimated model for three ordinal outcomes. The first columns indicates different outcomes, and the second columnn shows the estimated effect. The data obtained parameter-values which were used to generate the data are presented in column 3. Columns four and five show the average (standard deviation) of standard error using MO and sandwich, obtained over 200 replications, respectively. The last column shows the relative efficiency of MO compared to sandwich: $\mathrm{Var}(\widetilde{\theta}_{\mathrm{MO}})/\mathrm{Var}(\widetilde{\theta}_{\mathrm{Sandwich}})$.}
\label{tab_sim2}
\resizebox{\textwidth}{!}{\begin{tabular}{lllccc}
  \hline
Outcome&Effect&Parameter & MO & Sandiwch & ARE (MO/Sandwich)  \\ 
  \hline \vspace*{1mm} 
\multirow{4}{*}{LDL-chol.}&Int. 1&$\xi_{101}=-0.740$ & 0.039 (0.006) & 0.039 (0.006) & 0.993 (0.009)  \\ \vspace*{1mm} 
&Int. 2&  $\xi_{102}=0.480$ & 0.038 (0.006) & 0.038 (0.006) & 0.998 (0.009)  \\ \vspace*{1mm} 
&Int. 3 & $\xi_{103}=1.580$ & 0.042 (0.006) & 0.041 (0.007) & 1.012 (0.010)  \\ \vspace*{1mm} 
&Time  &$\xi_{11}=1.050$ & 0.032 (0.004) & 0.032 (0.004) & 1.024 (0.008)  \\ \vspace*{1mm} 
&Gender  &$\xi_{12}=0.450$ & 0.048 (0.008) & 0.048 (0.008) & 0.989 (0.001) \\ \vspace*{1mm} 
&RI sd  &$\sqrt{D_{11}}=1.880$ & 0.036 (0.005) & 0.034 (0.005) & 1.105 (0.010)  \\ \vspace*{1mm} 
\multirow{3}{*}{HbA1c}&Int. 1  &$\xi_{201}=270$ & 0.042 (0.007) & 0.042 (0.007) & 0.997 (0.008)  \\ \vspace*{1mm} 
&Int. 2  &$\xi_{202}=2.610$ & 0.052 (0.009) & 0.051 (0.009) & 1.043 (0.010)  \\ \vspace*{1mm} 
&Time  &$\xi_{21}=1.040$ & 0.036 (0.004) & 0.035 (0.005) & 1.034 (0.010)  \\ \vspace*{1mm} 
&Gender  &$\xi_{22}=-0.090$ & 0.054 (0.009) & 0.054 (0.009) & 0.979 (0.001)  \\ \vspace*{1mm} 
&RI sd  &$\sqrt{D_{22}}=2.150$ & 0.042 (0.007) & 0.040 (0.007) & 1.131 (0.012)  \\ \vspace*{1mm} 
\multirow{4}{*}{SBP}&Int. 1   &$\xi_{301}=-0.190$ & 0.035 (0.006) & 0.035 (0.006) & 0.995 (0.009)  \\ \vspace*{1mm} 
&Int. 2   &$\xi_{302}=1.380$ & 0.038 (0.007) & 0.038 (0.007) & 1.012 (0.009)  \\ \vspace*{1mm} 
&Int. 3   &$\xi_{303}=3.700$ & 0.056 (0.009) & 0.055 (0.010) & 1.028 (0.015)  \\ \vspace*{1mm} 
&Time   &$\xi_{31}=0.510$ & 0.030 (0.003) & 0.030 (0.003) & 1.013 (0.005)  \\ \vspace*{1mm} 
&Gender   &$\xi_{32}=0.190$ & 0.044 (0.008) & 0.044 (0.008) & 0.991 (0.001)  \\ \vspace*{1mm} 
&RI sd  &$\sqrt{D_{22}}=1.620$ & 0.034 (0.004) & 0.033 (0.005) & 1.073 (0.015)  \\ \vspace*{1mm} 
LDL-col. vs. HbA1c&Cov RI's &$D_{12}=0.808$ & 0.067 (0.013) & 0.068 (0.014) & 0.961 (0.014)  \\ \vspace*{1mm} 
LDL-col. vs. SPB&Cov RI's  &$D_{13}=1.218$ & 0.057 (0.012) & 0.057 (0.012) & 0.994 (0.015)  \\ \vspace*{1mm} 
HbA1c vs. SBP&Cov RI's   &$D_{23}=1.393$ & 0.065 (0.014) & 0.065 (0.014) & 0.993 (0.016)  \\ 
   \hline
\end{tabular}} }}
\end{table}

}}

\subsection{Application}
\label{sec_application}
To evaluate the hearing performance of a subject, hearing threshold sound pressure levels (measured in dB) are considered. By definition, a hearing threshold is the lowest signal intensity possible to perceive by a subject at a specific frequency. In a study considered in \cite{Molenberghs2000} and \cite{Verbeke2006}, hearing threshold measured in 11 different frequencies between 125 Hz and 8000 Hz for left and right ears obtained on 603 male participants from the Baltimore Longitudinal Study of Ageing, BLSA \citep{shock1984}. The number of visits for each subject which varies between 1 to 15 are unequally spaced. In this illustration all subjects with missing values are ignored to make a fair comparison between MO and sandwich corrections.

In order to model different frequencies simultaneously, \cite{Verbeke2006} used a joint model approach. Because of several outcomes of interest, they have proposed a pairwise approach to jointly fit the model to the data. In this section we use this dataset to illustrate our proposal and compare it with the sandwich correction proposed in \cite{Verbeke2006}. Consider $y_i(t)$ as the hearing threshold for subject $i$ at time $t$ for some frequency. The mixed model fitted to this single response is as follows:

\begin{equation}
\label{application_model}
y_i(t)=(\beta_1 +\beta_2 Age_i + \beta_3 Age_i^2 + a_i) + (\beta_4 + \beta_5 \mathrm{Age}_i + b_i)t + \beta_6 V(t) + \epsilon_i(t),
\end{equation}
where $t$ represents time and $V(t)$ is a dummy variable indicating whether the measurement is taken at the first time visit to account for systematically higher response values at the first visit. Also, $Age$ indicates the age of the participant at first visit. Furthermore, $a_i$ and $b_i$ are the random intercepts and slopes, respectively; and $\epsilon$ is the error term. Using a joint modelling approach to model the hearing threshold at several frequencies one needs to let the random intercepts and slopes from various models to be jointly distributed. Here a multivariate normal distribution is considered.

For the sake of our illustration, we consider two models: once three outcomes are modelled simultaneously (frequencies 125 Hz, 250Hz, and 500 Hz for the right ear) and once 5 outcomes are considered (frequencies 125 Hz, 250 Hz, 500 Hz, 750 Hz, and 1000 Hz for the right ear). A joint model of $m$ outcomes using (\ref{application_model}) for each response will have $6m$ fixed effects to estimate, as well as $2k (2k+1)/2$ parameters in $D$. For $k=3$ this is 39 parameters (on top of the parameters in the error matrix). Taking $k=5$ the number of parameters in fixed effects and $D$ matrix is $85$. A pairwise approach is used to estimate the parameters of such models. The precision of the parameter estimates are computed using MO and sandwich corrections. Whenever possible ($k=3$) the model is fitted directly using the full likelihood.

Figure~\ref{fig_app} shows the ARE for estimating fixed effects using MO and the sandwich corrections. The left panel of this figure shows the results when $3$ responses are modelled jointly. In this case, estimating the parameters using the full likelihood was also possible, so the ARE's are computed relative to the variance obtained from fitting the model using full likelihood. As one may see in Figure~\ref{fig_app} (left), the MO correction gives slightly better results compared to the sandwich. Fitting the full likelihood to the 5-response model was not feasible. Therefore, the precision was only estimated using MO and sandwich corrections. Figure~\ref{fig_app} (right) shows the ARE (MO divided by sandwich) for the $30$ fixed effects of a 5-response model. As one may see, the ARE's are around $1$; nevertheless, MO still gives slightly more precise estimates for most of the parameters.

\begin{figure}[htb]
\begin{center}
\includegraphics[trim=0 75 0 20,width=\textwidth]{fig_hear_new.eps}
\caption{Application. Left: comparing the ARE (relative to the variance using full likelihood) of estimated fixed effects of MO and sandwich corrections in a model with three responses. Right: the ARE=Var(MO)/Var(sandwich) of estimated fixed effects in a model with 5 responses. The x-axes show the indexes of the fixed effects (18 parameters for 3-response model and 30 parameters for 5-response model.)}
\label{fig_app}
\end{center}
\end{figure}

While both MO and sandwich corrections give similar precision estimates, their computation time differs dramatically. Fitting a 3-response model using the full likelihood takes only $38.12$ seconds. Fitting such a model  with a pairwise approach (it needs fitting 3 pairs) with sandwich correction takes $2.29$ hours ($216.66$ times more than the full likelihood) using \cite{pair_lin}. The computation time using the MO correction is only $25.34$ seconds ($0.66$ times the full likelihood computation time).

When considering 5 responses, fitting the full likelihood in {\tt{Proc MIXED}} in {\tt{SAS}} give {\tt{ERROR: The SAS System stopped processing this step because of insufficient memory.}} Using a pairwise approach (with 10 pairs) and sandwich correction the computation time is $7.98$ hours, while the MO correction takes only $1.45$ minutes which is $331.03$ faster comparing with sandwich correction implemented in \cite{pair_lin}.

We may note that the implementation in \cite{pair_lin} is not fully efficient. Also, as fitting different pairs is a so-called embarrassingly parallel task, using parallel computation is straightforward. So a more efficient implementation of sandwich correction will become faster \citep[as an example]{ivanova2015}. But the same discussion goes for the MO correction. Using parallel computation would also make this approach considerably faster. The interesting point is that on a single-core laptop scale, the MO correction is fast enough, also straightforward to implement, which would avoid the need of using cluster computing, difficult implementations, etc.

{\color{black}{Note that there could be other approaches to deal with this problem. In some cases, smoothing splines have been considered in the context of mixed models. For example, \cite{verbyla1999} and \cite{ruppert2003} consider random smoothing splines for the random-effects structure, and couple this with efficient computation. See also \citet[pp. 381-384]{molenberghs2005}. Evidently, smoothing splines can also be used to specify the mean structure. Other useful tools are fractional polynomials, which can be used in the mean function, the residual variance function, etc. Examples of both of these uses can be found in \citet[pp. 478-479 and pp. 137-139, respectively]{Molenberghs2000}. }}

\subsection{Conclusions}
\label{sec_conclusions}

In this short note the idea of multiple ouputation was extended to provide a fast precision estimation in high-dimensional multivariate joint models. Such precision estimation can be used in the pairwise approach of \cite{Verbeke2006, Verbeke2007} to replace the timely expensive and complicate to compute and implement sandwich estimator. While both MO and sandwich corrections give asymptotically the same preicisons, the main advantages of our MO-based proposal compared with sandwich correction are as follows:
\begin{enumerate}
\item MO correction is considerably faster, even on a laptop scale using a single core. In our simulations (Section~\ref{sec_sim}) the time gain was more than 2500 times and in real data analysis (Section~\ref{sec_application}) it was about 350 times. As our simulations and real data analysis results show, MO correction can possibly be computed even faster than the full likelihood (when feasible).
\item The Mo correction can be computed using the usual outcomes of any standard software. i.e., estimates and their precision. Unlike the sandwich correction which needs the first and second order derivatives. Therefore, MO correction is straightforward to implement with no need of complicated computations in any one of the usual software packages. This property is true for fixed effects, variance components, or other types of parameters of interest. Computing the derivatives of log-(pseudo-)likelihood function can possibly more difficult e.g. for variance components comparing with the fixed effects.
\item Using the MO correction for estimating the precision in a pairwise approach \citep{Verbeke2006, Verbeke2007} each pair can be treated completely independent from the others. Therefore, if one cluster contains no missing values for one pair it will be used to estimate the parameters of that specific pairs. This is not true for the sandwich correction. The reason is the need to compute cluster-wise gradients and Hessians. 
\end{enumerate}

Considering the above advantages of MO correction over sandwich correction, it can be considered as an effective and efficient alternative for precision estimation when dealing with high-dimensional multivariate joint models.

### 2019-08-06 Titlepage.tex
adjusting header, appendix numbering, title page

\begin{titlepage}
{\fontfamily{phv}\selectfont
\vspace*{-2.9cm}
{\setstretch{0.85}
\centerline{KU Leuven \hfill }
\centerline{Biomedical Sciences Group \hfill }
\centerline{Faculty of Medicine \hfill } 
\centerline{Department of Public Health and Primary Care \hfill } }
\vspace*{-2.4cm}
\hspace{2cm}
\centerline{\phantom{tram 9}\hfill \includegraphics{./logodoctschool.png} }

\begin{minipage}{1.2\textwidth}													%%%
\begin{flushleft} %\large															%%%
%\emph{Autores:}\\	
\vspace{2cm}
\hspace{1cm}\textbf{\Large{DATA SPLITTING AND ITS APPLICATIONS}} \\

\vspace{0.7cm}
\hspace{1cm}\large{Vahid NASSIRI}
\end{flushleft}																	
\end{minipage}

\vspace{2cm}
\begin{minipage}{0.4\textwidth}                                                    %%%
\begin{flushleft}   \small                                                        %%%
\hspace{0.3cm}\underline{Jury}:
\begin{tabular}{llp{0.5cm}l}
                 \vspace{0.5cm}  & \vspace{0.5cm}  &&\\
                 \hspace{0.1cm} Promoter: & Prof. Dr. G. Molenberghs&& Dissertation presented in  \\
                 \hspace{0.1cm} Co-promoter: & Prof. Dr. G. Verbeke&&  partial fulfilment of the \\
                 \hspace{0.1cm} Chair: & Prof. Dr. K. Dierickx&&  requirements for the \\
                 \hspace{0.1cm} Secretary: & Prof. Dr. E. Lesaffre&& degree of Doctor in \\
                 \hspace{0.1cm} Jury members: & Prof. Dr. H. Geys&&  Biomedical Sciences\\
                                                & Prof. Dr. SHelena Geys &&   \\
                                                & Prof. Dr. Dimitris Rizopoulos && \\
                                                &  Prof. Dr. Katrijn Van Deun  && \\
                                                &  Prof. Dr. Marlies Lacante  &&\\
                                                &  Prof. Dr. Karl Meerbergen  &&\\
                                                &  Prof. Dr. Philippe Lemey  &&
               \end{tabular}%
                                                                %%%
\end{flushleft}                                                                     %%%
\end{minipage}                                                   %%%

\vspace{1.2cm}
\begin{center}
\makebox{Date: 18 December 2019}
\end{center}

}

\end{titlepage}
\clearpage

### 2019-11-11 presentationJnJ.tex
revesion (comments of HG) + presentation JnJ

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Beamer Presentation
% LaTeX Template
% Version 1.0 (10/11/12)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND THEMES
%----------------------------------------------------------------------------------------

\documentclass{beamer}

\mode<presentation> {

% The Beamer class comes with a number of default slide themes
% which change the colors and layouts of slides. Below this is a list
% of all the themes, uncomment each in turn to see what they look like.

%\usetheme{default}
%\usetheme{AnnArbor}
%\usetheme{Antibes}
%\usetheme{Bergen}
%\usetheme{Berkeley}
%\usetheme{Berlin}
%\usetheme{Boadilla}
%\usetheme{CambridgeUS}
%\usetheme{Copenhagen}
%\usetheme{Darmstadt}
%\usetheme{Dresden}
%\usetheme{Frankfurt}
%\usetheme{Goettingen}
%\usetheme{Hannover}
%\usetheme{Ilmenau}
%\usetheme{JuanLesPins}
%\usetheme{Luebeck}
\usetheme{Madrid}
%\usetheme{Malmoe}
%\usetheme{Marburg}
%\usetheme{Montpellier}
%\usetheme{PaloAlto}
%\usetheme{Pittsburgh}
%\usetheme{Rochester}
%\usetheme{Singapore}
%\usetheme{Szeged}
%\usetheme{Warsaw}

% As well as themes, the Beamer class has a number of color themes
% for any slide theme. Uncomment each of these in turn to see how it
% changes the colors of your current slide theme.

%\usecolortheme{albatross}
%\usecolortheme{beaver}
%\usecolortheme{beetle}
%\usecolortheme{crane}
%\usecolortheme{dolphin}
%\usecolortheme{dove}
%\usecolortheme{fly}
%\usecolortheme{lily}
%\usecolortheme{orchid}
%\usecolortheme{rose}
%\usecolortheme{seagull}
%\usecolortheme{seahorse}
%\usecolortheme{whale}
%\usecolortheme{wolverine}

%\setbeamertemplate{footline} % To remove the footer line in all slides uncomment this line
\setbeamertemplate{footline}[page number] % To replace the footer line in all slides with a simple slide count uncomment this line

\setbeamertemplate{navigation symbols}{} % To remove the navigation symbols from the bottom of all slides uncomment this line
}
\newcommand{\bftheta}{\mbox{\boldmath $\theta$}}
\newcommand{\BY}{\mbox{\boldmath $Y$}}
\newcommand{\bn}{\mbox{\boldmath $n$}}
\newcommand{\bc}{\mbox{\boldmath $c$}}
  \newcommand{\BZ}{\mbox{\boldmath $Z$}}
\usepackage{graphicx} % Allows including images
\usepackage{booktabs} % Allows the use of \toprule, \midrule and \bottomrule in tables
%\usepackage {tikz}
\usepackage{tkz-graph}
\GraphInit[vstyle = Shade]
\tikzset{
  LabelStyle/.style = { rectangle, rounded corners, draw,
                        minimum width = 2em, fill = yellow!50,
                        text = red, font = \bfseries },
  VertexStyle/.append style = { inner sep=5pt,
                                font = \normalsize\bfseries},
  EdgeStyle/.append style = {->, bend left} }
\usetikzlibrary {positioning}
%\usepackage {xcolor}
\definecolor {processblue}{cmyk}{0.96,0,0,0}
%----------------------------------------------------------------------------------------
%	TITLE PAGE
%----------------------------------------------------------------------------------------

\title{Data Splitting And Its Applications} % The short title appears at the bottom of every slide, the full title is only on the title page

\author{Vahid Nassiri} % Your name
% Date, can be changed to a custom date

\begin{document}

\begin{frame}
\titlepage % Print the title page as the first slide
\end{frame}

\begin{frame}{It's all about the sample size!}
\begin{figure}
\centering
\includegraphics[width=7cm]{dog.eps}
\end{figure} 
source: Gunshow, by KC Green.
\end{frame}

\begin{frame}{Big-data era}
\begin{figure}
\centering
\includegraphics[width=\textwidth]{bigdata.eps}
\end{figure} 
\end{frame}

\begin{frame}{Data structure}
\begin{figure}
\centering
\includegraphics[width=\textwidth]{scheme_new.eps}
\end{figure}

\end{frame}

\begin{frame}{Clustered big-data}
\begin{itemize}
\item When the sample size, $N$, becomes very large,
\item When the cluster sizes become very large:
\begin{itemize}
\item When the number of measurements per outcome for some clusters, $n_{ri}$'s, becomes very large,
\item When the number of outcomes, $m$, becomes very large.
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{Data splitting: a unified approach}

\begin{enumerate}
\item \textbf{Splitting:} split data into smaller chunks in a way that analyzing each of them is easier than the complete data.

\item \textbf{Analyzing}: perform the desired analysis on each split, preferably noting more than the parameter estimates and their covariance matrix should not be required.

\item \textbf{Combining}: the results from several splits should be combined into a single set of results in an appropriate way.

\end{enumerate}

\end{frame}

\begin{frame}{Different types of splitting}

\begin{figure}
\centering
\includegraphics[width=8cm]{long.eps}
\end{figure} 
\end{frame}

\begin{frame}{Random horizontal splitting}

\begin{figure}
\centering
\includegraphics[width=8cm]{long.eps}
\end{figure} 
\end{frame}

\begin{frame}{Structured horizontal splitting}

\begin{figure}
\centering
\includegraphics[width=5.5cm]{plot2.eps}
\end{figure} 
\end{frame}

\begin{frame}{How to combine?}

Assume $\widehat{\theta}_1, \ldots, \widehat{\theta}_M$ are the estimated parameters from $M$ sub-samples, the data splitting estimate of this parameter can be computed as follows:
$$\widetilde{\theta} = \sum_{m=1}^M w_m\widehat{\theta_m}.$$
And for the variance:
$$\mathrm{Var}_{horizontal}(\widetilde{\theta}) = \sum_{m=1}^M w_m^2 \sigma_m^2.$$
\end{frame}

\begin{frame}{What about weights?}
\begin{itemize}
\item Equal weights: $w_{equal,i} = \frac{1}{M}$
\item Proportional weights: $w_{prop, i} = \frac{m_i}{N}.$
\item Size proportional weights: $w_{size-prop, i} = \frac{m_i n_i}{\sum_{k=1}^ M m_k n_k}.$
\item Optimal weights: $w_{opt, i} = \frac{1/\sigma^2_i}{\sum_{m=1}^M 1/\sigma^2_m},$
\begin{itemize}
\item as minimizer of $Q = \sum_{m=1}^M w_m^2 \sigma_m^2 - \lambda \left(\sum_{m=1}^M w_m -1 \right).$
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}{But large clusters...}
\begin{itemize}
\item We have a dataset of book ratings from Amazon.com,
\item each book is rated by different number of people: as small as 1 and
as large as 20,000. \pause
\item Splitting at book level would not help much!
\item But one can split the data within each book.
\end{itemize}

\end{frame}

\begin{frame}{Random vertical splitting}

\begin{itemize}
\item Splitting data within the cluster can be done by sub-sampling:
\begin{itemize}
\item \textbf{Splitting}: a random sub-sample of a reasonable size is taken from each cluster.
\item \textbf{Analysis}: the analysis of interest will be performed on this dataset.
\item \textit{Iterations}: As a new split, a new sub-sample is taken
\item \textbf{Combining}: the same combination rule can be applied on parameter estimates. 
\end{itemize}
\item But the variance is a different story.
\end{itemize}

\end{frame}

\begin{frame}{Variance combination rule for vertical splitting}
$$\mathrm{Var}(\widetilde{\theta}) = W - \left(1 + \frac{1}{M}\right) B,$$

where $W$ and $B$ are within and between sub-samples variances.

$$W = \frac{1}{M} \sum_{i=1}^M \sigma^2_m,\;$$ 
$$B = \frac{1}{M-1} \sum_{i=1}^M \left(\widehat{\theta}_i - \widetilde{\theta} \right)^2.$$

\end{frame}

\begin{frame}{But how many sub-samples?}
\begin{enumerate}
\item \textbf{Start.} Select an initial number of sub-samples, $M_0$, and sub-sampling size $m$. Take $M_0$ sub-samples of size $m$, fit the model to each and obtain $\widehat{\theta}_i$ and its variance $\Sigma_{\widehat{\theta}_i}$ ($i=1,\ldots,M_0$). Then compute 
$$\widetilde{\bftheta}_{M_0} = \sum_{i=1}^{M_0} \widehat{\bftheta}_i,\;\; \Sigma_{\widetilde{\theta_{M_0}}}=\widehat{W}_{M_0} - \left( \frac{M_0+1}{M_0} \right) \widehat{B}_{M_0}.$$
			\item \textbf{Update.} For sub-sampling size $m>M_0$, $m>M_0$, 
$$\widetilde{{\bftheta}}_{m+1}=\frac{m\widetilde{\bftheta}_m + \widehat{\bftheta}_{m+1}}{m+1},$$
$$ \Sigma_{\widetilde{\theta}_{m+1}}=\widehat{W}_{m+1} - \left( \frac{m+1}{m} \right) \widehat{B}_{m+1}.$$
	\item \textbf{Distance.} Compute: $d_{m+1}=d(\widetilde{\bftheta}_{m+1},\widetilde{\bftheta}_{m})$ using an appropriate distance.
	\item \textbf{Stopping rule.} $d_{j} < \varepsilon$ for $j=m+1,\ldots,m+k_0$.
\end{enumerate}
\end{frame}

\begin{frame}{Finite information limit estimators}
\begin{itemize}
\item The amount of information in some clusters is finite
\item In such cases a few sub-samples would be sufficient
\item We have shown in some cases even one or two sub-samples will do the job!
\end{itemize}
\end{frame}

\begin{frame}{Structured vertical splitting?}
\begin{itemize}
\item Vertical splitting based on a pre-defined structure could also be beneficial
\item An important example of this application is modeling several outcomes of interest using random effects.
\end{itemize}
\end{frame}

\begin{frame}{Using random effects to jointly modeling several responses}
\begin{equation*}
\begin{cases}
y_{1ij} = \beta_1 + b_{1i} + \epsilon_{1ij} \\
y_{2ij} = \beta_2 + b_{2i} + \epsilon_{2ij} \\
y_{3ij} = \beta_3 + b_{3i} + \epsilon_{3ij},
\end{cases}
\end{equation*}
we let the three random intercepts to be normally distributed as follows,
\begin{equation*}
\label{eq_SV_D}
\begin{bmatrix}
b_{1i} \\
b_{2i}\\
b_{3i}\\
\end{bmatrix} \sim N \left ( 
\begin{bmatrix}
0 \\
0\\
0\\
\end{bmatrix}, 
D=\begin{bmatrix}
D_{11} & D_{12} & D_{13}\\
 & D_{22} & D_{23}\\
&& D_{33}\\
\end{bmatrix}
\right).
\end{equation*}
Therefore, the parameter of interest, $\bftheta$, is:
\begin{equation*}
\bftheta= (\beta_1,\beta_2,\beta_3, D_{11},D_{22},D_{33},D_{12}, D_{13}, D_{23}).
\end{equation*}
\end{frame}

\begin{frame}{What about more responses?}
\begin{figure}
\centering
\includegraphics[width=\textwidth]{strucVert.eps}
\end{figure}

\end{frame}

\begin{frame}{Yes, it is vertical, but where is the structure?}
\begin{figure}
\centering
\includegraphics[width=8cm]{long.eps}
\end{figure} 
\end{frame}

\begin{frame}{Back to the example}
\begin{equation*}
\begin{cases}
\bftheta^{(1)}=\theta_{(y_{1ij},y_{2ij})} = (\beta_{1_1},\beta_{2_1},D_{11_1},
,D_{22_1},D_{12_1}) \\
\bftheta^{(2)}=\theta_{(y_{1ij},y_{3ij})} = (\beta_{1_2},\beta_{3_2},D_{11_2},
,D_{33_2},D_{13_2}) \\
\bftheta^{(3)}=\theta_{(y_{2ij},y_{3ij})} = (\beta_{2_3},\beta_{3_3},D_{22_3},
,D_{33_3},D_{23_3}),
\end{cases}
\end{equation*}

\end{frame}

\begin{frame}{Where combination rule fails}

\begin{itemize}
\item A weighted average is a reasonable combination rule both theoretically and intuitively,
\item But it should only be applied on parameter vectors which are each others counterparts.
\item A place where this condition fails is principal component analysis.
\end{itemize}

\end{frame}

\begin{frame}
\begin{itemize}
\item Methods like PCA work based on eigenvalue decomposition of the covariance (or correlation) matrix. 
\item Eigenvalues of a matrix $\Sigma$ can be obtained by solving the following equation,
		$$|\Sigma-\lambda I|,$$ 
\item The first PC is the eigenvector corresponds to the largest eigenvalue.\pause
\item There is no guarantee that these are the same for data from two sub-samples.
\item Think in case of factor analysis where latent underlying factors are there.
\end{itemize}
\end{frame}

\begin{frame}{How to solve the issue}
\begin{itemize}
\item There has been several solutions proposed for this problem:
\begin{itemize}
\item From simple odd methods like averaging the individuals data-points from several sub-samples,
\item to complicated methods like rotating PC's (factor loadings) in a way to make the ones from different sub-samples as similar as possible.
\end{itemize}
\item Our proposed solution was a king of take-the-pencil-instead solution!\pause
\item We proposed to first estimate the covariance matrix based on data-splitting, and then perform PCA on it.
\item We have explored theoretical and practical aspects of this proposal,
\item Also, developed appropriate confidence intervals for the proportion of explained variance.
\end{itemize}
\end{frame}

\begin{frame}{Conclusions}
\begin{itemize}
\item We have developed a unified approach to deal with clustered big data based on data splitting. 
\item Data splitting is similar to methods such as Google's MapReduce, or dplyr's split-apply-combine strategy.
\item The proposed framework makes sure that:
\begin{itemize}
\item While each methodology can deal with one situation, it is easily possible to combine methodologies to deal with a combined situation.

\item The splitting and combining steps of our methodology are as independent as possible from the analyses step, so one can easily (or with a minor modification) use it for its desired analysis.

\end{itemize}
\item R packages {\tt{fastCS}}, {\tt{fastAR1}}, {\tt{fimi}}, {\tt{mifa}}, and {\tt{miscVSS}} are prepared to implemented the methodologies we have developed. They are all publicly and freely available via {\tt{https://github.com/vahidnassiri}}.

\end{itemize}

\end{frame}

\begin{frame}{Immediate future plans}
\begin{itemize}
\item Using developed framework to deal with small data issue.
\item Implementing an R package to perform various currently-implemented-separately methods in a unified fashion. 
\end{itemize}
\end{frame}

\begin{frame}
\begin{figure}
\centering
\includegraphics[width=9cm]{thanks.eps}
\end{figure} 
\end{frame}

\end{document}

### 2019-11-12 rebuttal.tex
more comments applied  + summary

% LaTeX rebuttal letter example. 
% 
% Copyright 2019 Friedemann Zenke, fzenke.net
%
% Based on examples by Dirk Eddelbuettel, Fran and others from 
% https://tex.stackexchange.com/questions/2317/latex-style-or-macro-for-detailed-response-to-referee-report
% 
% Licensed under cc by-sa 3.0 with attribution required.
% See https://creativecommons.org/licenses/by-sa/3.0/
% and https://stackoverflow.blog/2009/06/25/attribution-required/

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{lipsum} % to generate some filler text
\usepackage{fullpage}

% import Eq and Section references from the main manuscript where needed
% \usepackage{xr}
% \externaldocument{manuscript}

% package needed for optional arguments
\usepackage{xifthen}
% define counters for reviewers and their points
\newcounter{reviewer}
\setcounter{reviewer}{0}
\newcounter{point}[reviewer]
\setcounter{point}{0}

% This refines the format of how the reviewer/point reference will appear.
\renewcommand{\thepoint}{P\,\thereviewer.\arabic{point}}

% command declarations for reviewer points and our responses
\newcommand{\reviewersection}{\stepcounter{reviewer} \bigskip \hrule
                  \section*{Reviewer \thereviewer}}

\newenvironment{point}
   {\refstepcounter{point} \bigskip \noindent {\textbf{Reviewer~Point~\thepoint} } ---\ }
   {\par }

\newcommand{\shortpoint}[1]{\refstepcounter{point}  \bigskip \noindent 
	{\textbf{Reviewer~Point~\thepoint} } ---~#1\par }

\newenvironment{reply}
   {\medskip \noindent \begin{sf}\textbf{Reply}:\  }
   {\medskip \end{sf}}

\newcommand{\shortreply}[2][]{\medskip \noindent \begin{sf}\textbf{Reply}:\  #2
	\ifthenelse{\equal{#1}{}}{}{ \hfill \footnotesize (#1)}%
	\medskip \end{sf}}

\begin{document}

\section*{Response to the reviewers}
% General intro text goes here
We thank the reviewers for their critical assessment of our work. 
In the following we address their concerns point by point.

% Let's start point-by-point with Reviewer 1
\reviewersection

% Point one description 
\begin{point}
	Page 56: $b_{ij} (i=1,..,15, j=1,…,m_j)$ -> Are those totals correct?
\end{point}

% Our reply
\begin{reply}
	We thank the reviewer to bring up this point. Indeed, the way we have displayed the totals could be misleading. In fact, $i$ goes from 1 to 15, as we have 5 personality factors, each for mother, father, and child. Then $j$ runs for various items in the 44-item BFI questionnaire correspond to each of these aspects. The number of items correspond to each personality aspect are not equal. Now, instead of $m_j$ we show these by $m_i$ as they corresponds to factor-role $i$. This is changed in the revised version as follows:
	\begin{quote}
	$b_{ij}$'s ($i=1,\ldots,15, j=1,\ldots,m_i$) are the random effects (latent variables) each representing one factor-role and $m_i$ is the number of items corresponds to factor-role $i$.
	\end{quote}
\end{reply}

\begin{point}
\begin{itemize}
\item Page 65, +2: main concern in
\item Page 80: eq. (ourgenpl)
\item Page 133: iterative multiple outputation
\item Page 186: preicisions
\item Page 186: item 2: The MO correction
\end{itemize}

\end{point}

\begin{reply}
These are fixed in the revised version of the manuscript.
\end{reply}

\begin{point}
Update. For $m>M_0$ (maybe a bit confusing to use the index $m$ again here?)
\end{point}

\begin{reply}
We agree with the reviewer that it could be confusing here, and the sentence is corrected as `For sub-sampling size $m>M_0$' in the revised manuscript. In fact, $m$ is is not just the index, it is still the sub-sampling size which is increased by one unit in every iteration.
\end{reply}

\begin{point}
Page 137 bottom: a 5x3 matrix is given and then columns are ordered and presented on top of page 138. However, the link with the indices of the ordered columns and the original matrix is not clear.
\end{point}

\begin{reply}
The matrix matrix on page 138 show the index corresponds to each element in the sorted matrix on page 137. We agree with the reviewer that currently this connection is not sufficiently clear, so we have changed it in the revised manuscript as follows:

\begin{quote}
Order the columns of the matrix in the previous step. For example, the following matrix shows a permutation which rearranges columns of the matrix in the previous step in an ascending order:
\end{quote}
\end{reply}

\begin{point}
Page 174, last sentence of first paragraph may need rephrasing?
\end{point}

\begin{reply}

We agree with the reviewer on this point. This is changed as `Therefore, each subject is only used in the sub-samples that the responses presented there are measured for it.' in the revised manuscript.

\end{reply}

\begin{point}
Page 177, sub-section 6.1.3 is mentioned twice but it is really 2 different paragraphs within that subsection?
\end{point}

\begin{reply}
We thank the reviewer for pointing this out. This is now corrected as follows:

\begin{quote}
In this section the proposed idea will be explored and examined via two simulations studies. Sub-section~6.1.3 considers 1- joint modelling of linear mixed models, and 2- a joint model of ordinal data in a generalized linear mixed models context.
\end{quote}
\end{reply}

\reviewersection

\begin{point}
\begin{itemize}
\item P3, first line: add ‘that’ -> when we realize THAT increasing….
 \item Top p4: replace “are” with ‘can be done’ in “and analyzing these different sub-samples are independent of each other”
 \item 2nd paragraph p4: add ‘the’ in front of ‘MapReduce Methodology’
 \item P6, add ‘the’ in ‘size of THE dataset makes it eligible …’
 \item Bottom p7: remove ‘s’ in ‘breakS down’
 \item P13 top: replace ‘its’ by ‘his/her own approach’
 \item P13 top: insert ‘the’ in ‘Typically, with THE data splitting …’
 \item P13, third bullet: remover ‘of’ in ‘all of analysis results’
 \item P13, bottom: add ‘and’ in ‘we will review different splitting approaches, AND also their appropriate combination rules’
 \item P18 replace ‘that’ by ‘where’ in ‘An example in our motivation datasets WHERE random horizontal splitting can be beneficial is the Divorce in Flanders’
 \item P18, insert ‘s’ in ‘deign’ -> design
 \item P20, top: insert ‘is’ in ‘it IS worth …’
 \item P26, top: insert ‘the’ in ‘with THE missing data issue’
 \item P43, replace ‘that’ by ‘for which’ in ‘the ones FOR WHICH using our proposed methodology’
 \item P45, replace ‘Big Five Inventory (BFI) is questionnaire’ by ‘THE Big Five Inventory (BFI) is A questionnaire’
 \item P55: replace personalty by personality
 \item P59, bottom: replace corresponds by corresponding
 \item P65, second line: replace is by iN in ‘is this thesis’
 \item P80, above Equation 4.12, please fix the reference to Eq. in ‘(see also Eq. (ourgenpl)).’
 \item P 243, first line: insert ‘the’ in ‘during the drug development process’
 \item P 246, replace ‘created’ by ‘create’ in ‘can be used to created’
\end{itemize}
\end{point}

\begin{reply}
These are fixed in the revised version of the manuscript.
\end{reply}

\begin{point}
P3: something seems to miss with respect to explaining the predictor in the bullet point “Fitting a linear model to the generated sample as the predictor and y = 1 + 3x + \_, where \_ \_ N(0, 0.01).”
\end{point}

\begin{reply}
We thank the reviewer to bring this up, indeed, the current sentence is vague. We meant the generated sample in the fist example as the predictor, but now it is changed as follows in the revised manuscript:

\begin{quote}
 Fitting a linear model to a random sample from standard normal as the predictor $x$, and $y = 1 + 3x + \epsilon$, where $\epsilon\sim N(0, 0.01)$.
\end{quote}
\end{reply}

\begin{point}
P4: please check the sentence containing “but for others, it would even speeds down with a sub-linear rate” -> the ‘s’ in speedS is grammatically not correct and you probably mean ‘slow down’
\end{point}
\begin{reply}
Indeed, that was a mistake which is now corrected as follows:

\begin{quote}
As we may see, for some analyses, the computation time increases linearly as the sample size increases, but for others, it would even slow down with a sub-linear rate. 
\end{quote}
\end{reply}

\begin{point}
Chapter 3, section 3.1., first paragraph: I miss a reference to the BFI
\end{point}

\begin{response}
We totally agree with the reviewer. The revised manuscript includes necessary references now:

\begin{quote}
Divorce in Flanders (Mortelmans et al., 2011) dataset presents data
from Big Five Inventory (BFI) questionnaire (John and Srivastava, 1999;
Denissen et al., 2008) answered by Flemish families.
\end{quote}

\reviewersection

\begin{point}
I did not find a definition of the symbol J. Therefore, I could not verify the result of Equation 4.1 on page 75.
\end{point}

\begin{response}
$J$ indicates a matrix of ones. The revised manuscript now includes this information: `where $I$ indicates an identity matrix and $J$ indicates a matrix of ones.'
\end{response}

\end{document}

### 2019-11-12 summaryThesis.tex
more comments applied  + summary

--- 2019-06-25 Final_doctoral_plan03.tex
@@
 \usepackage{listings}
-\usepackage[a4paper, total={6.5in, 9in}]{geometry}
+\usepackage[a4paper, total={6.5in, 9in},top=37mm]{geometry}
 \usepackage{pgfgantt}
@@
 
-\pagestyle{fancy}
-\fancyhf{}
-\rhead{{\textsf{Vahid Nassiri}}}
-\lhead{{\textsf{Final Doctoral Plan: Data Splitting and its Applications}}}
-\rfoot{Page \thepage}
-
 \renewcommand{\headrulewidth}{1pt}
@@
 \begin{document}
-\begin{titlepage}
-
 \begin{center}
-\includegraphics[scale=1]{KUL.eps}\\
-\vspace*{.3cm}
-{\sc{\LARGE{Final Doctoral Plan}}}\\
-\vspace*{2cm}
-{\Huge{\textbf{Data Splitting and its Applications}}}
-
-\vspace*{4cm}
-{\LARGE{\textbf{Vahid Nassiri}}}\\
-\vspace*{1cm}
-{\Large{I-BioStat, KU Leuven}}\\
-\vspace*{2cm}
-{\Large{Supervisor:}}\\
-\vspace*{0.2cm}
-{\LARGE{\textbf{Prof. Dr. Geert Molenberghs}}}\\
-\vspace*{0.5cm}
-{\Large{Co-supervisor:}}\\
-\vspace*{0.2cm}
-{\LARGE{\textbf{Prof. Dr. Geert Verbeke}}}
-\vspace*{2.3cm}
-
-{\Large{April 2017}}
-
+{\Huge{Summary}}
 \end{center}
-
-\end{titlepage}
-\newpage
-{\large{
-
-\textbf{Title}: {\Large{Data Splitting and its Applications}}
-\vspace*{0.5cm}
-
-\textbf{Name of applicant}: {\Large{Vahid Nassiri}} (r0292876)
-\vspace*{0.5cm}
-
-\textbf{Supervisor}:
-
-{\Large{Prof. Dr. Geert Molenberghs,}}
-
-I-BioStat, Universiteit Hasselt, Diepenbeek, Belgium.
-
-I-BioStat, KU Leuven, Leuven, Belgium
-
-\vspace*{0.5cm}
-
-\textbf{Co-supervisor:}
-
-{\Large{Prof. Dr. Geert Verbeke,}}
-
-I-BioStat, KU Leuven, Leuven, Belgium,
-
-I-BioStat, Universiteit Hasselt, Diepenbeek, Belgium.
-
-\vspace*{0.5cm}
-\textbf{Address}:
-
-I-BioStat, KU Leuven
-Kapucijnenvoer 35 3000 Leuven, Belgium.
-
-Tel: +32 (0) 16 37 46 96,
-
-Email: vahid.nassiri@kuleuven.be
-
-\vspace*{0.5cm}
-\textbf{Available Grant}: IWT- SBO/ Exascience/ IAP
-
-\vspace*{0.5cm}
-\textbf{Doctoral School Programme}: Patient Related and Public Health Research
-\vspace*{0.5cm}
-
-\textbf{Full-time or part-time PhD project}: it is a full-time PhD Project
-
-\vspace*{0.5cm}
-\textbf{Starting date}: 1 May 2014
-
-\vspace*{0.5cm}
-\textbf{Expected end date}: 30 April 2018}}
-
-\newpage
-
-\section{Introduction and problem statement}
-\label{sec_intro}
+\vspace{1cm}
 Sample size has always been an issue in statistical analyses.
@@
 
-Furthermore, while in the simplest designs all measurements are independent, many designs lead to hierarchical (a.k.a., clustered, correlated, dependent) data \citep{laird1982,zeger1986,verbeke2009, molenberghs2005}.
+Furthermore, while in the simplest designs all measurements are independent, many designs lead to hierarchical (a.k.a., clustered, correlated, dependent) data.
 By `cluster' we mean a set of measurements repeatedly collected for one single unit (e.g., subject, household, trial, etc.).
@@
 
-Consider a study which measures (at most) $m$ outcomes for $N$ clusters, each $n_i$ times.
-Let $y_{rij}$ be the $j$th measurement taken on the $i$th cluster for the $r$th outcome, $i=1,\ldots,N$, $r=1,\ldots,m$ and $j=1,\ldots,n_{ri}$.
-A general scheme of such data is presented in Figure~\ref{fig_scheme}.
-As it was mentioned, fitting a model to such data could become challenging when the size of dataset makes it eligible for the title \emph{big data}.
-A dataset could be entitled as \emph{big data} in different situations.
-In case of the clustered data of interest in this PhD research, data are considered to be big if at least one of the following applies:
-\begin{itemize}
-\item When the sample size, $N$, becomes very large,
-\item When the cluster sizes become very large:
-\begin{itemize}
-\item When the number of measurements per outcome for some clusters, $n_{ri}$'s, becomes very large,
-\item When the number of outcomes, $m$, becomes very large.
-\end{itemize}
-\end{itemize}
-\begin{figure}
-\centering
-\includegraphics[width=\textwidth]{scheme_new.eps}
-\caption{General scheme of the data we consider through this PhD research.} 
-\label{fig_scheme}
-\end{figure} 
-A large sample size could increase the computation time dramatically.
-On the other hand, even with small sample sizes, having large clusters may increase the computation time.
-Doing the computations and achieving convergence would become more challenging when the clusters are highly unbalanced (a wide range of cluster sizes).
-Also, having too many outcomes would increase the number of parameters in the model, to the point at which it could not be feasible any more.
-
-In this PhD research, we aim to propose methodologies to deal with such problems using the so-called data splitting.
+In this thesis, we propose methodologies to deal with such problems using the so-called data splitting.
 In a nutshell, data splitting means to split the difficult (sometimes infeasible) job of modelling the big dataset by splitting it into smaller chunks of manageable sub-data, analyze each of them separately, and then combine the results of these analyses using appropriate combinations rules.
 This would break a difficult task into several, often so-called embarrassingly parallel by computer scientists, simpler tasks.
-
-The rest of this report is structured as follows.
-In Section~\ref{sec_objective} the main objectives of this PhD research will be briefly reviewed.
-The datasets motivating our research will be presented in Section~\ref{sec_data_sets}.
-Section~\ref{sec_idea} describes the general idea behind the methodologies we are developing.
-The results obtained so far will be presented in Section\ref{sec_results_so_far}.
-The work in progress as well as a timeline for finishing them will be given in Section~\ref{sec_progress}.
-Finally, a provisional structure for the upcoming PhD thesis is proposed in Section~\ref{sec_structure}.
-
-\section{Main objectives}
-\label{sec_objective}
-Considering the problem stated in Section~\ref{sec_intro}, our main objective is to develop a unified framework that helps researchers to deal with big data in the situations we have described.
-We expect this framework to have the following properties.
-
-\begin{itemize}
-\item The developed framework should be as independent as possible from the modelling approach.
-In other words, it should be compatible with various modelling approaches, e.g.,  linear, generalized linear, non-linear models, etc.
-\item As the end-user of this product will be researchers of different fields with different tastes of data analysis systems who are dealing with big data, the developed methodologies should preferably be easy to understand and implement in different platforms, e.g., \textsf{R}, \textsf{SAS}, \textsf{Matlab}, \textsf{Python}, \textsf{Julia}, etc.
-\item It goes without saying that a combination of the situations we have described in Section~\ref{sec_intro} could happen for a single set of data.
-For example, a dataset would consist of a large number of clusters (large $N$), and each of these clusters could be large as well (large $n_{ri}$'s), see Amazon book ratings in Section~\ref{sec_data_sets}.
-Also, large number of outcomes of interest can happen with large number of clusters, see divorce in Flanders dataset in Section~\ref{sec_data_sets}.
-Therefore, our proposed methodologies for each situation should have the possibility of being combined to deal with a combination of the mentioned situations.
-\item This unified approach is going to be implemented in freely available statistical package \textsf{R}.
-So every interested person can use it without the trouble of implementation.
-This implementation should be done as efficient as possible.
-\item Finally, we should be able to apply our proposed methodology to solve the problems involving our motivating datasets.
-Considering the diversity of them, that would prove the usefulness of our approach to deal with 21st century real world big data problems.
-\end{itemize}
-
-\section{Motivating datasets}
-\label{sec_data_sets}
-A brief summary of the datasets motivating our research is presented in this section.
-\paragraph*{A developmental toxicity study:} This study investigates the dose-response
-relationship in mice of the potentially hazardous chemical compound di(2-ethylhexyl)phthalate
-(DEHP), used in vacuum pumps and as plasticizers for numerous plastic devices made of polyvinyl chloride. The developmental toxicity study, conducted in timed-pregnant
-mice during the period of major organogenesis has attracted
-much interest in the toxicity of DEHP. The doses selected for the study were 0, 0.025, 0.05, 0.1,
-and 0.15\%, corresponding to a DEHP consumption of 0, 44, 91, 191, and 292 mg/kg/day, respectively.
-The dams were sacrificed, slightly prior to normal delivery, and the status of uterine implantation sites recorded.
-A total of 1082 live fetuses were dissected from the uterus, anesthetized, and examined for external, visceral, and skeletal malformations, as well as for body
-weight.
-Our focus will be on the continuous weight outcome.
-Evidently, fetuses are clustered within mothers; hence the implied association needs to be accommodated in the analysis.
-As each mother has a different number of fetuses, we have clusters of different sizes.
-
-\paragraph*{Divorce in Flanders:} This dataset is collected at the Centre for Sociological Research, KU Leuven. The data concerned on studying and comparing the personality among
-three generations in about 4000 Flemish families.
-Each family would at most consist of mother, father, step-mother, step-father, one child, grand mother, and grand-father.
-The personality is measured using the so-called
-Big Five Inventory (BFI) score.
-These scores are built upon 5 factors which in total are measured through 44 items.
-The usual factor analysis would lose most of the hierarchical information in the data.
-Therefore, an idea is to model different combinations of the 5 aspects of personality for each family role using a mixed model.
-This means, for each family we will have up to 35 responses.
-The large size of families would add to the difficulty of the problem.
-
-\paragraph*{Hearing data:} To evaluate the hearing performance of a subject, hearing threshold sound pressure levels (measured in dB) are considered.
-By definition, a hearing threshold is the lowest signal intensity possible to perceive by a subject at a specific frequency.
-In a study considered in \cite{verbeke2009} and \cite{Verbeke2006}, hearing thresholds measured in 11 different frequencies between 125 Hz and 8000 Hz for left and right ears obtained on 603 male participants from the Baltimore Longitudinal Study of Ageing, BLSA.
-The number of visits for each subject which varies between 1 to 15 are unequally spaced.
-Modelling these 22 frequencies together is a challenge for the methodologies we developed in this PhD thesis.
-
-\paragraph{Leuven diabetes project:} In order to study how well diabetes is controlled, three outcomes were measured for each patient at baseline ($T_0$) and one year later ($T_1$).
-These three outcomes of interest were LDL-cholesterol (low-density lipoprotein cholesterol, mg/dl), HbA1c (glycosylated hemoglobin, \%), and SBP (systolic blood pressure, mmHg).
-Although these three variables are continuous, the interest was in expert-specified cut-off values of these outcomes.
-The cut-off values are defined as $\{<100, [100,115), [115, 130),\geq 130 \}$ (mg/dl) for LDL-cholesterol, $\{<7, [7, 8), \geq 8\}\%$ for HbA1c, and $\{ \leq 130, (30,140],(140,160],>160\}$mmHg for SBP.
-Modelling these 3 ordinal outcomes together provides a suitable platform to study the performance of our proposed methodologies dealing with non-Gaussian data.
-
-\paragraph*{Amazon book ratings:} These data consist of the ratings of the books on the Amazon website which are recently made publicly available.
-The dataset contains $2,330,066$ books, which are rated at least $1$ and at most $21,398$ times.
-This is an example of clustered data with large highly unbalanced clusters and a large sample size.
-
-\paragraph*{Clinical Trials in Schizophrenia:} These data were collected from five double-blind randomized clinical trials to compare the effects of two treatments for chronic schizophrenia: risperidone and conventional antipsychotic agents.
-Subjects who received doses of risperidone (4–6 mg/day) or an active control (haloperidol, perphenazine, zuclopenthixol) have been included in the analysis.
-Patients were clustered within country, and longitudinal measurements were made on each subject over time.
-The number of patients ranges from 9 to 128 per country with a total of 2039.
-The
-positive and negative syndrome scale (PANSS) was used to asses the global condition of a patient.
-This scale is constructed from 30 items, each taking values between 1 and 7, giving an overall range of 30 to 210.
-PANSS provides an operationalized, drug-sensitive instrument, which is useful for both typological and dimensional assessment of schizophrenia.
-Depending on the trial, treatment was administered for a duration of 48 weeks with at most 12 monthly measurements.
-Because not all subjects received treatment at the same time points and, not the same amount, the final dataset is unbalanced.
-Our analysis shows the AR(1) structure could be a good candidate for modelling the covariance matrix of the random effects, which is useful to illustrate some of the methodologies we have developed.
-
-\paragraph*{Age-related macular degeneration (ARMD):} The data come from a clinical trial in age-related macular degeneration (ARMD), which is a disease of the retina that causes severe central vision loss.
-The data consist of a multi-center study (17 centers) to evaluate the effect of an experimental treatment based on interferon-$\alpha$ in a group of patients who were randomly allocated to either placebo or interferon-$\alpha$.
-The outcome of interest was the change in visual acuity over time.
-Using standard vision charts, the visual acuity was measured by the number of letters that were correctly read by the patient.
-The data is used in the context of surrogate endpoint evaluation.
-The goal of this analyses is to examine if  visual acuity at week 24 could be an appropriate surrogate for the one at week 52, the so-called true endpoint.
-To this end, the surrogate and the true endpoint have to be modelled jointly.
-Because of the bivariate nature of the model, coupled with the multi-centre nature of the study, a mixed model is used, by which surrogate evaluation measures are estimated.
-The data however is highly unbalanced and only has 17 clusters.
-This makes the iterative algorithms to face with difficulty in convergence.
-We try to use the methodologies developed in this PhD research to deal with this problem.
-
-\paragraph*{Leuven eyes study:} This is an extensive observational study of glaucoma performed at the ophthalmology department of KU Leuven.
-The dataset so far consists of 141 variables measured in 614 subjects.
-As measuring all of these variables was not possible for all of the subjects, this dataset exhibits considerable missingness.
-Considering the size of dataset and amount of missing values, applying multiple imputations and determining the number of imputations is challenging in this change.
-
-\section{Data splitting: the general idea}
-\label{sec_idea}
-As it was briefly mentioned in Section~\ref{sec_intro}, the data splitting approach consists of 3 steps:
+The data splitting approach consists of 3 steps:
 \begin{enumerate}
@@
 \end{enumerate}
-Let us briefly go over our approaches to deal with each of these steps.
 
-The first step of this procedure is determined by the type of big data we are dealing with.
-We divide the data splitting methods based on two main characteristics: 1- independent or dependent sub-samples, 2- random or structured sub-sampling.
-
-Data splitting can be done horizontally or vertically.
-If the splitting is done at the subject level (each subject will be fully presented in sub-samples), it is called horizontal splitting.
-The reason of using such a terminology can be observed in Figure~\ref{fig_scheme} (left).
-Splitting at the subject level means splitting the full data into smaller pieces using horizontal slices.
-Considering the assumption of having independent clusters, this will result in independent sub-samples.
-Unless the sub-samples are overlapping, i.e., one subject appears in more than one sub-sample\footnote{Note that, while perfectly possible, the overlapping horizontally splitted sub-samples are not considered in this thesis.
-One can look in vast literature of bootstrap methods for such an approach.}.
-
-On the other hand, when splitting is done within each subject (part of the measurements for each subject are presented in each sub-sample), it is called vertical splitting.
-The right-side of Figure~\ref{fig_scheme} shows the reason behind this title.
-Splitting the measurements within the subject means to take the sub-samples vertically.
-Such sub-samples will be dependent, since they are measured for a common subject.
-
-Each of horizontal or vertical splitting can be done at random or based on a pre-specified structure.
-Here we briefly describe the four different combinations of vertical/horizontal and random/structured data splitting.
-
+We could summarize various splitting strategies into four main approaches:
 \begin{itemize}
@@
 Therefore, instead of taking random sub-samples, the sub-sampling can be done based on this pre-specified structure.
-An important example of this approach which is considered in this PhD research is splitting based on the cluster sizes, i.e., forming sub-samples consisting of clusters of equal size.
+An important example of this approach which is considered in this thesis is splitting based on the cluster sizes, i.e., forming sub-samples consisting of clusters of equal size.
 This would lead to balanced sub-samples.
 We have shown the existence of closed-form solutions, leading to faster convergence of model fitting procedures.
-\item \textbf{Random vertical data splitting.} This approach is useful when we are dealing with data with large number of measurements for each cluster per outcome (large $n_{ri}$'s).
-In this situation one can take sub-samples of each cluster with a manageable size $n_{f(ri)} < n_{ri}$.
-Obviously, this can be done by taking overlapping (sub-sampling with replacement) or non-overlapping (sub-sampling without replacement) sub-samples.
-\item \textbf{Structured vertical data splitting.} This approach is useful when the number of outcomes of interest, $m$, is large.
-While fitting a model to $m$ outcomes could become heavy or infeasible, fitting models to all possible pairs (triples, etc.) of these outcomes is feasible.
+\item \textbf{Random vertical data splitting.} This approach is useful when we are dealing with data with large number of measurements for each cluster per outcome.
+In this situation one can take sub-samples of each cluster with a manageable size.
+\item \textbf{Structured vertical data splitting.} This approach is useful when the number of outcomes of interest is large.
+While fitting a model to several outcomes could become heavy or infeasible, fitting models to all possible pairs (triples, etc.) of these outcomes is feasible.
 That means each sub-sample should consist of all of the measurements for each subject regarding the pair of interest.
@@
 \end{itemize}
-Obviously, according to the type of big data one is dealing with, all of the the combinations of random/structured and horizontal/vertical data splitting can be freely combined.
-The combination rules then should be combined accordingly.
-This enables the proposed framework to deal with a combination of the big data situations described in Section~\ref{sec_intro}.
 
-After splitting the data into appropriate sub-samples, the next step is to perform the necessary analysis on these sub-samples.
-The analysis performed on each sub-sample does not need to be similar to one another.
-For example, consider 3 outcomes of interest of types continuous, binary, and ordinal.
-If sub-samples are taken using a structured vertical splitting approach, and each sub-sample consists of pairs of outcomes, the model fitted to each sub-sample will depend on the type of the responses present in the pair at hand.
-Furthermore, the explanatory variables of the models within each sub-sample does not need to be identical across sub-samples.
-Therefore, such a framework is compatible with different modelling approaches.
-
-The last step of this procedure is combining the results obtained from various sub-samples.
-As the models fitted within each sub-sample have some common parameters, each parameter of the initial model could possibly be estimated more than once after the second step.
-We propose to use a weighted average of them to obtain a unique estimation for each parameter.
-The critical part is to define such appropriate weights.
-They have to be computed proportionally to the number of clusters within each sub-sample.
-Also, if needed, size of these clusters should be involved.
-We have also computed optimal combination weights in order to minimize the final combined variance.
-
-Computing the precision of such an estimator is another important aspect.
-Computing this precision depends on whether the sub-samples are dependent or independent.
-For independent sub-samples, one can simply use an average of the obtained variances from each sub-sample.
-For dependent sub-samples, this averaged variance should be corrected for the information the overlapped sub-samples share.
-This correction is done using a sandwich estimator \citep{Verbeke2006, Verbeke2007} and multiple ouputation \citep{hoffman2001, follmann2003}.
-
-\section{Results obtained so far}
-\label{sec_results_so_far}
-We present the results obtained so far through the published, accepted, and submitted papers. The working papers and projects will be presented in Section~\ref{sec_progress}.
-
-\begin{itemize}
-\item[--]{\textsf{Lisa Hermans, Vahid Nassiri, Geert Molenberghs, Michael G.
-Kenward, Wim Van der Elst, Marc Aerts, and Geert Verbeke}}, {\it Clusters with random size: maximum likelihood versus weighted estimation}, Submitted to Statistica Sinica.
-\end{itemize}
-This paper considers the case of structured horizontal splitting for a dataset consists of vectors of random variables with multivariate Gaussian distribution with compound-symmetry (CS) covariance structure of different sizes.
-The CS covariance structure considers a common variance for all of the components in the random vector.
-Furthermore, it assumes all pairs of variables are equally correlated (a common correlation coefficient).
-We have proposed to split the large sample into sub-samples formed by all clusters of equal sizes.
-Closed-form solutions are obtained to find the parameter estimates within each balanced sub-sample.
-The appropriate weights are obtained to efficiently combine estimates from different sub-samples.
-The closed-form estimates and efficient combination weights are also implemented in \textsf{R}.
-Simulation results show the derived closed-forms reduce the computation time up to 30,000 times compared to iterative procedures in \textsf{Proc MIXED} in \textsf{SAS}.
-
-\begin{itemize}
-\item[--]{\textsf{Lisa Hermans, Vahid Nassiri, Geert Molenberghs, Michael G.
-Kenward, Wim Van der Elst, Marc Aerts, and Geert Verbeke}}, {\it Fast, Closed-form, and Efficient Estimators for Hierarchical Models with AR(1) Covariance and Unequal Cluster Sizes}, Communications in Statistics - Simulation and Computation, 2017, DOI: 10.1080/03610918.2017.1316395.
-\end{itemize}
-This paper considers the same problem as in the previous one, except here the covariance structure is assumed to be autoregressive of order 1, AR(1).
-In this structure, unlike the CS, the correlation between pairs of observations depends on their distance, i.e., the correlation between $i$th and $j$th components is defined as $\sigma^2 \rho^{|i-j|}$ ($\sigma^2>0$, $-1\leq\rho\leq +1$).
-The closed-form, fast and efficient estimates are obtained for balanced data following Gaussian distribution with AR(1) structured covariance matrix.
-The appropriate weights are computed to efficiently combine these estimates and their variances.
-These closed-form estimates and efficient combination weights are also implemented in \textsf{R}.
-Simulation results show using the derived closed-forms, the computation time reduced up to 700 times compared to iterative procedures in \textsf{Proc MIXED} in \textsf{SAS}.
-
-\begin{itemize}
-\item[--]{\textsf{Vahid Nassiri, Anna Ivanova, Geert Molenberghs, and Geert Verbeke}}, {\it Fast precision estimation in high-dimensional multivariate joint models}, To appear in Biometrical Journal.
-\end{itemize}
-In this paper the case of structured vertical splitting is considered when the number of outcomes of interest becomes large.
-We have proposed to take sub-samples formed by pairs (if needed triples, etc.) of the outcomes.
-Efficient combination rules were obtained and evaluated using simulations for linear and generalized linear models.
-The proposed methodology was also implemented in \textsf{R}.
-Simulation results shows up to 2500 times faster computations compared to existing procedures.
-
-\begin{itemize}
-\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}}, {\it Finite information limit variance-covariance structures: Is the entire dataset needed for analysis?}, In High Performance Computing \& Simulation (HPCS) 2016 International Conference (pp.
-736-742).
-IEEE.
-\end{itemize}
-We have introduced the concept of a finite information limit (FIL) estimator in this paper.
-This concerns the situation that the amount of information within each cluster is finite.
-Therefore, after some point adding new observations would not substantially improve efficiency.
-This means, in case of FIL estimators, one only needs to analyze a small fraction of the data to obtain results as efficient as analyzing the full dataset.
-This concept is introduced in this paper, and some procedures are proposed to detect this property as well as determining the size of this fraction.
-Compared to fitting the model to the full data, simulation results show about 1500 times faster computations using this idea.
-
-\begin{itemize}
-\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}}, {\it A simple, principled rule to determine the number of multiple imputations}, Submitted to The American Statistician.
-\end{itemize}
-We have discussed the need to take several sub-samples of the original data to deal with big data.
-In case that these sub-samples are taken with replacement, infinite number of sub-samples can be taken.
-Furthermore, for sub-sampling without replacement, it is very well possible that taking all possible sub-samples needs enormous amount of time.
-Thus, the need to use only a part of these sub-samples would emerge.
-This paper considers the problem of determining the number of sub-samples.
-Of course, as this problem is similar to the problem of determining the number of imputations when dealing with missing data using multiple imputations, the paper was written with that topic as a target.
-The idea is general and can be applied in many situations.
-It has been used in the FIL paper, and will be used in the IMO paper (see Section~\ref{sec_progress}).
-
-\begin{itemize}
-\item[--]{\textsf{Vahid Nassiri, Anik\'{o} Lovik, Geert Molenberghs, and Geert Verbeke}}, {\it On using multiple imputation for factor analysis of incomplete data}, Submitted to Behavior Research Methods.
-\end{itemize}
-When it comes to factor analysis, principal component analysis, or other methodologies dealing with  eigendecomposition of the covariance matrix, analyzing several datasets in place of the original one would cause some problems at the combination step (Step 3 in our proposed procedure).
-The reason lies in the fact that not all of the obtained factor loadings would follow the same order.
-In other words, there is no guarantee that the eigenvector corresponding to the largest eigenvalue of one of these sets of data is comparable with another.
-This could happen when missing data are multiply imputed, when sub-samples of clusters are taken to overcome computation issues with larger clusters, or when a multilevel factor analysis is of interest, but only at within cluster level.
-This paper considers this problem and proposed to first estimate the covariance matrix and then perform the factor analysis on the single estimated covariance matrix.
-The consequences of such an approach are discussed and some criteria are proposed to decide whether this is an appropriate solution or not.
-
-\section{Work in progress}
-\label{sec_progress}
-
-We now give an overview of work in progress with a provisional timeline, see Figure~\ref{fig_gant}.
-
-\begin{itemize}
-\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}}, {\it Iterative multiple outputation: a simple procedure to deal with large clustered data}, In progress.
-\end{itemize}
-Inspired by finite information limit estimators, we are trying to extend this idea to cases where this property does not hold, or the size of the one sub-sample needed to achieve it is large compared to available computation power.
-In this situation, we would suggest to take several smaller random sub-samples.
-That would involve appropriate rules to determine the number of sub-samples, as well as, combining the results after fitting the models to such sub-samples.
-Some simulations are already performed, also \textsf{R} functions are written to efficiently take the sub-samples and combine the results of analyses performed on them.
-Work is ongoing to apply this methodology on some of our motivating data.
-We expect to submit the paper before beginning of next academic year.
-
-\begin{itemize}
-\item[--]{\textsf{Vahid Nassiri, Anik\'{o} Lovik, Geert Molenberghs, and Geert Verbeke}}, {\it A simple multilevel factor analysis}, In progress.
-\end{itemize}
-The use of the sub-sampling from different clusters is not restricted to big data.
-This can also be useful in situations where the intra-cluster correlation is nuisance.
-Consider the divorce in Flanders data in Section~\ref{sec_data_sets}.
-There, the main interest, at validation stage, is to verify the BFI structure among the 44 items of the questionnaire removing the correlation imposed by the fact that some of the participants are coming from the same family.
-This paper considers this problem and try to propose a solutions based on random vertical splitting.
-The manuscript is almost finalized and we expect it to be submitted within this academic year.
-
-\begin{itemize}
-\item[--]{\textsf{Vahid Nassiri, Alvaro Florez Poveda, Luis Adrian Quintero Sarmiento, Wim Van der Elst, Geert Molenberghs, and Geert Verbeke}}, {\it Comparing different approaches for mixed-effects modelling of unbalanced clustered data}, In progress.
-\end{itemize}
-Although the methodologies we proposed and developed in this thesis are targeting big data, sometimes very small data and very large data suffer from the same issues.
-One common problem of large and small data is the convergence issues.
-In this article we try to study the performance of our proposed method to deal with small highly unbalanced data and compare it with other solutions developed in I-BioStat.
-
-\begin{itemize}
-\item[--]{\textsf{Vahid Nassiri, Geert Molenberghs, and Geert Verbeke}}, {\it Data splitting: an \textsf{R} package to analyze big clustered data.}, In progress.
-\end{itemize}
-As mentioned before, for many of the proposed and developed methods \textsf{R} functions have been prepared to make the implementation easier for the end-user.
-We are planning to make generic functions out of them and collect them in the form an \textsf{R} package which will be publicly available.
-This will need more time and effort, but we hope it can be accomplished by the end of the PhD.
-
-The last final step in the last year of this PhD research will be preparing the final manuscript of the PhD thesis.
-Figure~\ref{fig_gant} shows a Gantt chart of the fourth year PhD in order to complete the mentioned tasks.
-
-\begin{figure}
-\centering
-\begin{ganttchart}[vgrid={draw=none,draw=none},%
-             %today=15,%
-             %today offset=.5,%
-             %today label=Heute,%
-             %progress=today,%
-             x unit=0.7cm,
-             y unit title=1cm,
-             y unit chart=0.5cm,
-             bar incomplete/.append style={fill=red},%
-             progress label text=  {\quad\pgfmathprintnumber[precision=0,verbatim]{#1}\%},
-             milestone label font=\tiny,
-             group label font=\tiny,
-             title label font=\tiny,
-             bar label node/.style={text width=5cm,align=right,font=\scriptsize\RaggedLeft,anchor=east},
-             milestone label node/.style={text width=3cm,align=right,font=\scriptsize\RaggedLeft,anchor=east},
-             group label node/.style={text width=4cm,align=right,font=\scriptsize\RaggedLeft,anchor=east}
-             ]{1}{12}
-\gantttitlecalendar*[compress calendar,time slot format=isodate]{2017-05-1}{2018-4-30}{year, month} \\
-\gantttitlelist{1,...,12}{1}\\
-\ganttgroup{Total Duration}{1}{12} \\
-  %%%%%%%%%%%%%%%%%Phase-1
-\ganttgroup{Finishing research papers}{1}{6} \\
-\ganttbar{Paper 1}{1}{1} \\
-\ganttlinkedbar{Paper 2}{2}{3} \ganttnewline
-\ganttlinkedbar{Paper 3}{4}{6} \ganttnewline
-
-%%%%%%%%%%%%%%%%%Phase-2
-\ganttgroup{Preparing the \textsf{R} package}{4}{9} \\
-  %%%%%%%%%%%%%%%%%Phase-3
-\ganttgroup{Writing the thesis manuscript}{9}{12} \\
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\end{ganttchart}
-\label{fig_gant}
-\caption{Gantt chart of the provisional fourth year PhD plan}
-\end{figure}
-
-\section{Provisional structure of the thesis}
-\label{sec_structure}
-
-\begin{itemize}
-\item Chapter 1: Introduction and the state of art
-\item Chapter 2: Motivating datasets
-\item Chapter 3: Problem statement and preliminaries
-\item Chapter 4: Horizontal splitting
-\item Chapter 5: Vertical splitting
-\item Chapter 6: Related topics
-\item Chapter 7: Conclusions
-\end{itemize}
-
-\bibliography{bib_final_plan}{}
-\bibliographystyle{abbrvnat}
+For each of these splitting strategies appropriate and efficient combination rules are developed.
+We also have considered exceptional cases where the general mythologies would fail.
+Our findings are motivated and supported by various datasets.
+Furthermore, the methodologies we have developed are implemented in statistical computing language R.
+\vspace{1cm}
+\begin{center}
+{\large{Professional Background}}
+\end{center}
+\vspace{0.5cm}
+Vahid Nassiri  has obtained a Bachelor in Statistics (2002) from the Shahid Beheshti Univerity in Tehran, Iran, and a master in Mathematical Statistics (2008) from the Amirkabir University of Technology in Tehran, Iran.
+Afterwards, he has obtained a PhD in Applied Mathematics (2013) from Vrije Universiteit Brussel (VUB), Brussels, Belgium.
+His main research interests are in the field of Big Data and High-dimensional problems.
+While working on the sparse estimation of high-dimensional linear models in sense of least squares and quantile regression in his former PhD, his current work is concerned on the fast computation of models for correlated data with possible missing values.
+His research expertise is mainly concerned with analyzing incomplete big correlated data using mixed models, multiple imputation, penalized likelihood and related topics.
+He also has experiences in consultation involving projects from pharmacology, biology, medicine, engineering, psychology, sociology, linguistics, and market research.
+He is now working as a statistical consultant in Open Analytics NV, Antwerp, Belgium.
 

### 2019-11-13 summaryThesis.tex
PL is done + part of KM

@@
 In a nutshell, data splitting means to split the difficult (sometimes infeasible) job of modelling the big dataset by splitting it into smaller chunks of manageable sub-data, analyze each of them separately, and then combine the results of these analyses using appropriate combinations rules.
-This would break a difficult task into several, often so-called embarrassingly parallel by computer scientists, simpler tasks.
+This would break a difficult task into several, often so-called embarrassingly parallel, simpler tasks.
 The data splitting approach consists of 3 steps:
@@
 For each of these splitting strategies appropriate and efficient combination rules are developed.
-We also have considered exceptional cases where the general mythologies would fail.
+We also have considered exceptional cases where the general methodologies would fail.
 Our findings are motivated and supported by various datasets.
@@
 \begin{center}
-{\large{Professional Background}}
+{\large{\textbf{Professional Background}}}
 \end{center}

### 2019-11-16 appendixFile.tex
semi-final revised version, check for typos remained

\documentclass[11pt,a5paper,twoside]{book}
\usepackage[T1]{fontenc}
\usepackage[Glenn]{fncychap}
\usepackage{bbm}
\usepackage[font=small]{caption}
\usepackage{amsmath, amsthm, amscd, amsfonts}
\usepackage{mathtools}
\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
\usepackage{graphicx}
\usepackage{pstricks}
\usepackage{bm}
\usepackage{tabu}
\usepackage[authoryear]{natbib}
\usepackage{multirow}
\usepackage{tocbibind}
\usepackage{cleveref}
\usepackage{resizegather}
\usepackage{textcomp}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}{Lemma}[chapter]
\newtheorem{proposition}{Proposition}[chapter]
\newtheorem{corollary}{Corollary}[chapter]
\newtheorem{property}{Property}[chapter]
\newtheorem{definition}{Definition}[chapter]
\newtheorem{remark}{Remark}[chapter]
\usepackage{numprint}
\newcommand{\HRule}{\rule{\linewidth}{0.5mm}}
\usepackage[
  paperwidth=17cm,
  paperheight=24cm,
  % other options
]{geometry}
%\special{papersize=17cm,24cm}
%
%
%\textwidth=12.5cm
%\textheight=20cm
%\oddsidemargin=0.3cm
%\evensidemargin=-1cm
%\topmargin=-1.5cm

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%chapter on new page
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\clearemptydoublepage}%
        {\newpage{\pagestyle{empty}\cleardoublepage}}

\renewcommand*\listfigurename{List of figures}
\renewcommand*\listtablename{List of tables}

%\title{Data Splitting and Its Applications}
%\author{Vahid Nassiri}
%	options include 12pt or 11pt or 10pt
%	classes include article, report, book, letter, thesis
\usepackage{setspace}
\usepackage{numprint}
\usepackage{siunitx}
\sisetup{round-mode=places,       % enable rounding
         round-precision=4,       % amounts of digits for rounding
         table-format=-1.4}       % default numeric format
\newcommand\scalemath[2]{\scalebox{#1}{\mbox{\ensuremath{\displaystyle #2}}}}
\usepackage{url}
\usepackage{blindtext}
\usepackage{listings}
\usepackage{afterpage,lscape}
\usepackage{bbm}
\usepackage{amsmath, amsthm, amscd, amsfonts}
\usepackage{graphicx}
\usepackage{pstricks}
%\usepackage[natbib, maxcitenames=3, mincitenames=11, style=apa]{biblatex}
\usepackage{color}
\usepackage{multirow}
\usepackage{tocbibind}

\usepackage{textcomp}
\usepackage{lipsum}
\usepackage{fancyhdr}

\fancyfoot{}
\fancyhead[LE, RO]{\thepage}
\fancyhead[LO]{\leftmark}
\fancyhead[RE]{\rightmark}
%\usepackage{fancyhdr}
\pagestyle{fancy}
%\fancyhf{}
%\renewcommand{\headrulewidth}{0pt}
%\fancyhead[R]{\thepage}
%\fancyhead[RO,LE]{\thepage}
%\fancyhead[LO]{\leftmark}
%\fancyhead[RE]{\rightmark}
\usepackage{afterpage}

\newcommand\blankpage{%
    \null
    \thispagestyle{empty}%
    \addtocounter{page}{-1}%
    \newpage}
\usepackage{amssymb}
\usepackage{fixltx2e}
\usepackage{multirow}
\usepackage{verbatim}
\usepackage[normalem]{ulem}
\usepackage{amsmath}
\usepackage{soul}
\usepackage{color}
\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage{subfigure}% Support for small, `sub' figures and tables

\usepackage{kbordermatrix}% http://www.hss.caltech.edu/~kcb/TeX/kbordermatrix.sty
\usepackage{mathtools}
\usepackage{listings}
%\usepackage[a4paper, total={6.5in, 9in},top=37mm]{geometry}
%\usepackage{pgfgantt}
\usepackage{afterpage,lscape}
%\usepackage{bbm}
\usepackage{amsmath, amsthm, amscd, amsfonts}
\usepackage{natbib}
\setcitestyle{authoryear,open={(},close={)}}
%\usepackage{cleveref}
\usepackage{textcomp}
\usepackage{longtable}

\renewcommand{\baselinestretch}{1.4} 
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}

\usepackage{pdflscape} % provides the landscape environment
\usepackage{ragged2e} % provides \RaggedLeft

\usepackage{fancyhdr}

\newcommand{\bftheta}{\mbox{\boldmath $\theta$}}
\newcommand{\BY}{\mbox{\boldmath $Y$}}
\newcommand{\bn}{\mbox{\boldmath $n$}}
\newcommand{\bc}{\mbox{\boldmath $c$}}
  \newcommand{\BZ}{\mbox{\boldmath $Z$}}
 \newcommand{\J}{J\"{o}reskog}
\newcommand{\So}{S\"{o}rbom}
\newcommand{\bcx}{{\bf X}}
\newcommand{\bcy}{{\bf Y}}
\newcommand{\bcz}{{\bf Z}}
\newcommand{\bcu}{{\bf U}}
\newcommand{\bcv}{{\bf V}}
\newcommand{\bfv}{\mbox{\boldmath $v$}}
\newcommand{\bcw}{{\bf W}}
\newcommand{\bci}{{\bf I}}
\newcommand{\bch}{{\bf H}}
\newcommand{\bcb}{{\bf B}}
\newcommand{\bcr}{{\bf R}}
\newcommand{\bcm}{{\bf M}}
\newcommand{\bcf}{{\bf F}}
\newcommand{\bcg}{{\bf G}}
\newcommand{\bcs}{{\bf S}}
\renewcommand{\labelitemii}{$\triangleright$}
\newcommand{\tr}{\mbox{tr}}
\newcommand{\Zik}{\mbox{$Z_i^{(k)}$}}
\newcommand{\bca}{{\bf A}}
\newcommand{\bcd}{{\bf D}}
\newcommand{\bcc}{{\bf C}}
\newcommand{\bce}{{\bf E}}
\newcommand{\ba}{{\bf a}}
\newcommand{\bk}{\mbox{\boldmath{$k$}}}
\newcommand{\bb}{{\bf b}}
\newcommand{\bd}{{\bf d}}
\newcommand{\bx}{{\bf x}}
\newcommand{\by}{{\bf y}}
\newcommand{\bz}{{\bf z}}
\newcommand{\bu}{{\bf u}}
\newcommand{\bv}{{\bf v}}
\newcommand{\bh}{{\bf h}}
\newcommand{\bl}{{\bf l}}
\newcommand{\be}{{\bf e}}
\newcommand{\br}{{\bf r}}
\newcommand{\bw}{{\bf w}}
\newcommand{\bbi}{\bm{b_i}}
\newcommand{\Bb}{\mbox{\boldmath $b$}}
 \newcommand{\BW}{\mbox{\boldmath $W$}}
 \newcommand{\bfbeta}{\mbox{\boldmath $\beta$}}
\newcommand{\de}{\stackrel{D}{=}}
\newcommand{\bt}{\bigtriangleup}
 \newcommand{\BYi}{\mathbf{$\y_i$}}
\newcommand{\bfequiv}{\mbox{\boldmath $\equiv$}}
\newcommand{\bmu}{\mbox{\boldmath $\mu$}}
\newcommand{\bnu}{\mbox{\boldmath $\nu$}}
\newcommand{\bxi}{\mbox{\boldmath $\xi$}}
\newcommand{\btau}{\mbox{\boldmath $\tau$}}
\newcommand{\bgamma}{\mbox{\boldmath $\Gamma$}}
\newcommand{\bphi}{\mbox{\boldmath $\Phi$}}
\newcommand{\bfphi}{\mbox{\boldmath $\varphi$}}
\newcommand{\bfeta}{\mbox{\boldmath $\eta$}}
\newcommand{\bpi}{\mbox{\boldmath $\Pi$}}
\newcommand{\bequiv}{\mbox{\boldmath $\equiv$}}
\newcommand{\bvarepsilon}{\mbox{\boldmath $\varepsilon$}}
\newcommand{\btriangle}{\mbox{\boldmath $\triangle$}}
\newcommand{\bdelta}{\mbox{\boldmath $\Delta$}}
\newcommand{\beps}{\mbox{\boldmath $\epsilon$}}
\newcommand{\btheta}{\mbox{\boldmath $\theta$}}
\newcommand{\balpha}{\mbox{\boldmath $\alpha$}}
\newcommand{\bsphi}{\mbox{\boldmath $\varphi$}}
\newcommand{\bsig}{\mbox{\boldmath $\sigma$}}
\newcommand{\bfpsi}{\mbox{\boldmath $\psi$}}
\newcommand{\bfdelta}{\mbox{\boldmath $\delta$}}
\newcommand{\bsigma}{{\bf \Sigma}}
\newcommand{\bzero}{{\bf 0}}
\newcommand{\bpsi}{\mbox{\boldmath $\Psi$}}
\newcommand{\bep}{\mbox{\boldmath $\epsilon$}}
\newcommand{\bomega}{\mbox{\boldmath $\Omega$}}
\newcommand{\bfomega}{\mbox{\boldmath $\omega$}}
\newcommand{\blambda}{\mbox{\boldmath $\Lambda$}}
\newcommand{\bflambda}{\mbox{\boldmath $\lambda$}}
\newcommand{\bfsigma}{\mbox{\boldmath $\sigma$}}
\newcommand{\bfpi}{{\mbox{\boldmath $\pi$}}}
\newcommand{\bupsilon}{\mbox{\boldmath $\upsilon$}}
\newcommand{\obs}{{\rm obs}}
\newcommand{\mis}{{\rm mis}}
\newcommand{\Xik}{\mbox{$X_i^{(k)}$}}
\newcommand{\Aik}{\mbox{$A_i^{(k)}$}}
\newcommand{\Tik}{\mbox{$T_i^{(k)}$}}
\newcommand{\Qik}{\mbox{$Q_i^{(k)}$}}
 \newcommand{\bfalpha}{\mbox{\boldmath $\alpha$}}
 \newcommand{\bfgamma}{\mbox{\boldmath $\gamma$}}
 \newcommand{\bfSigma}{\mbox{\boldmath $\Sigma$}}
 \newcommand{\bfmu}{\mbox{\boldmath $\mu$}}
  \newcommand{\bfrho}{\mbox{\boldmath $\rho$}}

 \newcommand{\eenvec}{\mbox{\boldmath $1$}}

%%%% from rmarkdown
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{color}
\usepackage{fancyvrb}
\usepackage{cleveref}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
%\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%%
%\special{papersize=17cm,24cm}
%
%
%\textwidth=12.5cm
%\textheight=20cm
%\oddsidemargin=0.3cm
%\evensidemargin=-1cm
%\topmargin=-1.5cm
%END OWN DEFINITIONS

%\pagestyle{fancy}
%\fancyhf{}
%\rhead{{\textsf{Vahid Nassiri}}}
%\lhead{{\textsf{Data Splitting and its Applications}}}
%\rfoot{Page \thepage}

\renewcommand{\headrulewidth}{1pt}
\makeatletter
     \renewcommand*\l@figure{\@dottedtocline{1}{1em}{3.5em}}
          \renewcommand*\l@table{\@dottedtocline{1}{1em}{3.5em}}

\makeatother

\begin{document}

\backmatter

\appendix
\chapter{Appendices}
\setcounter{page}{273}

\setcounter{equation}{0}
\renewcommand{\thesection}{A.\arabic{section}}
\renewcommand{\theequation}{\thesection.\arabic{equation}}
\renewcommand\thefigure{\thesection.\arabic{figure}}    
\renewcommand\thetable{\thesection.\arabic{table}}

%% CS paper
In this chapter supplementary materials from different chapters are presented. Sections~\ref{appA}--\ref{bcim} will provide materials related to Section~\ref{sec_cs} (structured horizontal splitting for CS). The materials related to Section~\ref{sec_ar1} (structured horizontal splitting for AR1) will be given in Sections~\ref{ar1appendix}--\ref{suppdata}. The Fieller's method and delta method are explained in Section~\ref{app1}. The combination rules for random vertical splitting for a CS case will be derived in Section~\ref{app2}. In Section~\ref{app_mi}, combination rules of multiple imputation for $p$-values are given. And finally, the surrogate model we have used in thesis is presented in Section~\ref{sec_surrogate_model}.

\section[Incompleteness in the CS model]{Incompleteness in the compound-symmetry model} \label{appA}

{Based on (\ref{driestats})
%(\ref{driestats}) 
and the multiplicity of the cluster sizes, the sufficient statistics are:
\begin{eqnarray}
W_{1k}&=&\sum_{i=1}^{c_k}\sum_{j=1}^{n_k}Y_{ij}^{(k)},\label{keen}\\
W_2&=&\sum_{k=1}^L\sum_{i=1}^{c_k}\sum_{j=1}^{n_k}\left(Y_{ij}^{(k)}\right)^2,\label{ktwee}\\
W_{3k}&=&\sum_{i=1}^{c_k}\left(\sum_{j=1}^{n_k}Y_{ij}^{(k)}\right)^2,\label{kdrie}\\
W_{4k}&=&c_k.\label{kvier}
\end{eqnarray}
The conditional and marginal expectations of (\ref{keen})--(\ref{kvier}) are:
\[
\begin{aligned}
E(W_{1k}|c_k)&=&c_k n_k\mu,\\
E(W_{1k})&=&N\mu\pi_k n_k,\\
E(W_{2}|c_k)&=&\sum_{k=1}^Lc_k n_k(\sigma^2+d+\mu^2),\\
E(W_{2})&=&N(\sigma^2+d+\mu^2)\sum_{k=1}^L\pi_k n_k,\\
E(W_{3k}|c_k)&=&c_k\left\{
n_k(\sigma^2+d+\mu^2)+n_k(n_k-1)(d+\mu^2)\right\},\\
E(W_{3k})&=&N\pi_k n_k
\left\{
(\sigma^2+d+\mu^2)+(n_k-1)(d+\mu^2)
\right\}
,\\
E(W_{4k})&=&N\pi_k.
\end{aligned}
\]
Group all sufficient statistics in $\BW$ and define a function
\begin{equation}
\label{gcluster}
g(\bw)=\sum_{k=1}^L\lambda_k\frac{w_{1k}}{w_{4k}}.
\end{equation}
Then,
$$E\left\{g(\BW|\BW_4)\right\}=
\sum_{k=1}^L\lambda_k\frac{E(W_{1k}|W_{4k})}{W_{4k}}=
\mu\sum_{k=1}^L\lambda_k n_k,
$$
and hence
$$E\left\{g(\BW)\right\}=
\mu\sum_{k=1}^K\lambda_k n_k.$$
Thus, every solution $\bflambda \perp \bn$, where $\bn=(n_1,\dots,n_K)'$, provides a counterexample, establishing incompleteness.

Such a vector $\bflambda$ exists if and only if $K\ge 2$, for which it is assumed that at least two $c_k>0$ (i.e., at least two different cluster sizes occur).

\setcounter{equation}{0}
\section{Likelihood-based estimation of the CS model}
\label{likelihoodapp}

\subsection{Score functions}

The score function has components:
\begin{eqnarray}
\frac{\partial \ell}{\partial \mu_k} & =&
\frac{1}{\sigma_k^2+n_kd_k} \left( \sum_{i=1}^{c_k} \sum_{j=1}^{n_1} y_{ij}^{(k)}-c_k n_k \mu_k\right) ,
\label{jscore1} \\
\frac{\partial \ell}{\partial \sigma_k^2} & =&
 \frac{-c_kn_k}{2\sigma_k^2} \cdot \frac{\sigma_k^2+(n_k-1)d_k}{\sigma_k^2+n_k d_k} + \frac{c_kn_kS_k}{2\sigma_k^4}  \\
 &&-\frac{d_k(2\sigma_k^2+n_kd_k) c_kn_kT_k}{2\sigma_k^4(\sigma_k^2+n_kd_k)^2} ,
\label{jscore2}\\
\frac{\partial \ell}{\partial d_k} & = &
\frac{-c_kn_k}{2(\sigma_k^2+n_kd_k)}+\frac{c_kn_k T_k}{2(\sigma_k^2+n_kd_k)^2},
 \label{jscore3}
\end{eqnarray}
with
\begin{eqnarray}
\label{skfunc}\label{qkfunc}
S_k & =& \frac{1}{c_{kn_k}} Q_k= \frac{1}{c_kn_k}\sum_{i=1}^{c_k} \BZ_i^{(k)'}\BZ_i^{(k)}, \\
\label{tkfunc}\label{rkfunc}
T_k & =&\frac{1}{c_{kn_k}}R_k= \frac{1}{c_{kn_k}}\sum_{i=1}^{c_k} \BZ_i^{(k)'}\mathbf{J_{n_k}}\BZ_i^{(k)}.
\end{eqnarray}

\subsection{Lack of closed-form solution when $K\ge 2$}
\label{twodifferent}

The lack of a closed form when $K\ge 2$ is well known, but we highlight a few relevant features here. More detail is given in Supplementary Materials~\ref{appB}.
Function (4.4)
%(\ref{jloglik}) 
can be turned into the log-likelihood kernel for the conventional situation where there is a common mean parameter and common variance components across all cluster sizes, i.e., $\ell(\mu,\sigma^2,d)$. The score functions follow from summing the terms in (\ref{jscore1})--(\ref{jscore3}) across cluster sizes:
\begin{equation}
\label{jscorelik}
\begin{gathered}
\frac{\partial \ell}{\partial \mu}=
\sum_{k=1}^K\left.\frac{\partial \ell}{\partial \mu_k}\right|_{\mu_k=\mu},\\
\frac{\partial \ell}{\partial \sigma^2}=
\sum_{k=1}^K\left.\frac{\partial \ell}{\partial \sigma^2_k}\right|_{\sigma^2_k=\sigma^2},\\
\frac{\partial \ell}{\partial d}=
\sum_{k=1}^K\left.\frac{\partial \ell}{\partial d_k}\right|_{d_k=d}.
\end{gathered}
\end{equation}
Solving the score equation in (\ref{jscorelik}) for the mean, using that
$$\Sigma_{n_k}^{-1}=\frac{1}{\sigma^2}I_{n_k}-\frac{d}{\sigma^2(\sigma^2+n_kd)}J_{n_k},$$
leads to the identity:
\begin{equation}
\widehat{\mu}=\frac
{
\sum_{k=1}^K\frac{n_k c_k}{\sigma^2+n_k d} \overline{Y}^{(k)}
}
{
\sum_{k=1}^K\frac{n_k c_k}{\sigma^2+n_k d}
}
=
\frac
{
\sum_{k=1}^K\frac{n_k c_k}{\sigma^2+n_k d} \widehat{\mu_k}
}
{
\sum_{k=1}^K\frac{n_k c_k}{\sigma^2+n_k d}
},
\label{mulikapp}
\end{equation}
where $\widehat{\mu_k}$ as in (4.5).
%(\ref{muk}).
For the variance components, only implicit identities follow; they are functions of (\ref{skfunc})--(\ref{tkfunc}). These take the form of high-degree polynomials, for which no general explicit solution exists.  While (\ref{mulikapp}) is explicit, it is a weighted average of the cluster-size specific averages $\overline{Y}^{(k)}$, with weights depending on the variance components.  This, combined with the result for the variance components, implies that there is no explicit solution, unless the variance components are known or the cluster size is constant.

\setcounter{equation}{0}
\section{Full likelihood}
 \label{appB}

Referring to the conventional situations, i.e. $\ell(\mu,\sigma^2,d)$ in (4.2)
%(\ref{loglikcs}) 
and the score equation in (\ref{jscorelik}), also second derivatives can be calculated:
\begin{equation}
\frac{\partial^2\ell}{\partial \mu^2}  = \sum_{k=1}^K\frac{-c_kn_k}{\sigma^2+n_kd} \label{jhessian1}
\end{equation}
\begin{equation}
\frac{\partial^2\ell}{\partial \sigma^2 \partial \mu}  = \sum_{k=1}^K \frac{-1}{(\sigma^2+n_kd)^2} \left( \sum_{i=1}^{c_k} \sum_{j=1}^{n_k} y_{ij}^{(k)}-c_k n_k \mu_k\right) \label{jhessian2}
\end{equation}
\begin{equation}
\frac{\partial^2\ell}{\partial d \partial \mu_k}  = \sum_{k=1}^K\frac{-n_k}{(\sigma^2+n_kd)^2} \left( \sum_{i=1}^{c_k} \sum_{j=1}^{n_k} y_{ij}^{(k)}-c_k n_k \mu_k\right) \label{jhessian3}
\end{equation}
\begin{equation}
\frac{\partial^2\ell}{\partial \mu \partial \sigma^2} = \sum_{k=1}^K \left(\frac{-1}{\sigma^4} + \frac{d(2\sigma^2+n_kd)n_k}{\sigma^4(\sigma^2+n_kd)^2} \right) \sum_{i=1}^{c_k}\sum_{j=1}^{n_k}Z_{ij}^{(k)} \label{jhessian4} 
\end{equation}
\begin{multline}
\frac{\partial^2\ell}{(\partial \sigma^2)^2} = \sum_{k=1}^K\left( \frac{c_kn_k}{2\sigma^4} \cdot \frac{\sigma^2+(n_k-1)d}{\sigma^2+n_kd}- \frac{c_kn_k}{2\sigma^2} \cdot \frac{d}{(\sigma^2+n_kd)^2} - \frac{c_kn_k S_k}{\sigma^6} \right. \\
 \left.- d c_kn_kT_k \frac{\sigma^2(\sigma^2+n_kd)-(2\sigma^2+n_kd)^2}{\sigma^6(\sigma^2+n_kd)^3}\right) \label{jhessian5}
\end{multline}
\begin{multline}
\frac{\partial^2\ell}{\partial d \partial \sigma^2} = \\ \sum_{i=1}^K\left( \frac{c_kn_k}{2(\sigma^2+n_kd)}- \frac{c_kn_k T_k}{\sigma^4} \cdot \frac{(\sigma^2+n_kd)^2 - n_k d (2\sigma^2+n_kd)}{(\sigma^2+n_kd)^3} \right)\label{jhessian6}
\end{multline}
\begin{equation}
\frac{\partial^2\ell}{\partial \mu \partial d} =\sum_{k=1}^K \frac{-n_k}{(\sigma^2+n_kd)^2}\sum_{i=1}^{c_k} \sum_{j=1}^{n_k} Z_{ij}^{(k)} \label{jhessian7}
\end{equation}
\begin{equation}
\frac{\partial^2\ell}{\partial \sigma^2 \partial d} = \sum_{k=1}^K\left( \frac{c_kn_k}{2(\sigma^2+n_kd)^2}-\frac{c_kn_kT_k}{(\sigma^2+n_kd)^3} \right) \label{jhessian8}
\end{equation}
\begin{equation}
\frac{\partial^2\ell}{\partial d^2} = \sum_{k=1}^K \left(\frac{c_kn_k^2}{2(\sigma^2+n_kd)^2}-\frac{c_kn_k^2T_k}{(\sigma^2+n_kd)^3} \right) \label{jhessian9} 
\end{equation}
Should we use conditional likelihood, then the log-likelihood's kernel equals:
\begin{multline}
L  \propto \prod^k_{i=1} \frac{1}{(2\pi)^{n_1/2}|\Sigma_{n_1}|^{1/2}} \exp \left\{ -\frac{1}{2}(\by_{i1}-\bmu_{n_1})'\Sigma_{n_1}^{-1} (\by_{i1}-\bmu_{n_1}) \right\}  \\
\times \left[ \frac{\Phi (\alpha + \by'_{i1} \bfbeta)}{\Phi\left( \frac{\alpha + \bmu_{n_1}'\bfbeta}{\sqrt{1+\bfbeta'\Sigma_{n_1}\bfbeta/n_1}} \right)} \right] \\
\times \prod^N_{i=k+1} \frac{1}{(2\pi)^{n_2/2}|\Sigma_{n_2}|^{1/2}}\exp \left\{ -\frac{1}{2}(\by_i-\bmu_{n_2})'\Sigma_{n_2}^{-1} (\by_i-\bmu_{n_2}) \right\} \\
\qquad\times\left[ \frac{1-\Phi (\alpha + \by'_{i1} \bfbeta)}{1-\Phi\left( \frac{\alpha + \bmu_{n_1}'\bfbeta}{\sqrt{1+\bfbeta'\Sigma_{n_1}\bfbeta/n_1}} \right)} \right] \label{clik} 
\end{multline}
\begin{multline}
\ell  \propto - \frac{1}{2} \sum_{i=1}^k \left\{ \ln|\Sigma_{n_1}| + (\by_{i}-\bmu_{n_1})'\Sigma_{n_1}^{-1} (\by_{i1}-\bmu_{n_1}) \right\}  \\
 - k \ln \Phi(\tilde{\alpha} + \bmu_{n_1}'\tilde{\bfbeta})  \\
 -\frac{1}{2} \sum_{i=k+1}^N \left\{ \ln|\Sigma_{n_2}| + (\by_i-\bmu_{n_2})'\Sigma_{n_2}^{-1} (\by_i-\bmu_{n_2}) \right\}\\
 -(N-k) \ln \left\{1-\Phi(\tilde{\alpha} + \bmu_{n_1}' \tilde{\bfbeta})\right\},  \label{cloglik} 
\end{multline}
with $\tilde{\alpha}=\frac{\alpha}{\sqrt{1+\bfbeta'\Sigma_{n_1}\bfbeta/n_1}}$ and  $\tilde{\bfbeta}=\frac{\bfbeta}{\sqrt{1+\bfbeta'\Sigma_{n_1}\bfbeta/n_1}}$
The corresponding score equations are:
\begin{equation}
\begin{aligned}
\frac{\partial\ell}{\partial \mu} & =&\frac{1}{\sigma^2+n_1d} \left( \sum_{i=1}^k \sum_{j=1}^{n_1} \by_{ij}-k n_1 \mu\right)
 - k n_1 \mathbf{j_{n_1}'} \tilde{\bfbeta} \frac{\phi(\tilde{\alpha}+\bmu_{n_1}'\tilde{\bfbeta})}{\Phi(\tilde{\alpha}+\bmu_{n_1}'\tilde{\bfbeta})}  \\
&& + \frac{1}{\sigma^2+n_2d} \left( \sum_{i=k+1}^N \sum_{j=1}^{n_2} \by_{ij}-(N-k) n_2 \mu\right) \\
 &&- (N-k) n_2 \mathbf{j_{n_1}'} \tilde{\bfbeta} \frac{\phi(\tilde{\alpha}+\bmu_{n_1}'\tilde{\bfbeta})}{\Phi(\tilde{\alpha}+\bmu_{n_1}'\tilde{\bfbeta})} \label{cscore1} ,
\end{aligned}
\end{equation}
with $\frac{\partial\ell}{\partial \sigma^2}$ and $\frac{\partial\ell}{\partial d}$ identical to (\ref{jscore2}) and (\ref{jscore3}). The components of the Hessian are:
\begin{equation}
\begin{aligned}
\frac{\partial^2\ell}{\partial \mu^2} & = &\frac{-kn_1}{\sigma^2+n_1d}-\frac{(N-k)n_2}{\sigma^2 +n_2d}  \\
&& - kn_1 \mathbf{j_{n_1}'}\tilde{\bfbeta}\left[ \frac{-\Phi(\tilde{\alpha}+\bmu_{n_1}'\tilde{\bfbeta}) \cdot \phi(\tilde{\alpha}+\bmu_{n_1}'\tilde{\bfbeta})\cdot(\tilde{\alpha}+\bmu_{n_1}'\tilde{\bfbeta}) \cdot \mathbf{j_{n_1}'}\tilde{\bfbeta}}{\Phi^2(\tilde{\alpha}+\bmu_{n_1}'\tilde{\bfbeta})} \right.  \\
&& \left. - \frac{\phi^2(\tilde{\alpha}+\bmu_{n_1}'\tilde{\bfbeta}) \cdot \mathbf{j_{n_1}'}\tilde{\bfbeta}}{\Phi^2(\tilde{\alpha}+\bmu_{n_1}'\tilde{\bfbeta})}\right]  \\
&& +(N-k)n_2 \mathbf{j_{n_1}'}\tilde{\bfbeta}\\
&&\times \left[ \frac{-(1-\Phi(\tilde{\alpha}+\bmu_{n_1}'\tilde{\bfbeta})) \cdot \phi(\tilde{\alpha}+\bmu_{n_1}'\tilde{\bfbeta})\cdot(\tilde{\alpha}+\bmu_{n_1}'\tilde{\bfbeta}) \cdot \mathbf{j_{n_1}'}\tilde{\bfbeta}}{(1-\Phi(\tilde{\alpha}+\bmu_{n_1}'\tilde{\bfbeta}))^2} \right.  \\ 
&& + \left. \frac{\phi^2(\tilde{\alpha}+\bmu_{n_1}'\tilde{\bfbeta}) \cdot \mathbf{j_{n_1}'}\tilde{\bfbeta}}{(1-\Phi(\tilde{\alpha}+\bmu_{n_1}'\tilde{\bfbeta}))^2}\right]\label{chessian1}
\end{aligned}
\end{equation}
\begin{equation}
\begin{aligned}
\frac{\partial^2\ell}{\partial \sigma^2 \partial \mu} & = &\frac{-1}{(\sigma^2+n_1d)^2} \left( \sum_{i=1}^k \sum_{j=1}^{n_1} \by_{ij}-k n_1 \mu\right) \\
&& - \frac{1}{(\sigma^2+n_2d)^2} \left( \sum_{i=k+1}^N \sum_{j=1}^{n_2} \by_{ij}-(N-k) n_2 \mu\right) \label{chessian2} 
\end{aligned}
\end{equation}
\begin{equation}
\begin{aligned}
\frac{\partial^2\ell}{\partial d \partial \mu} & =& \frac{-n_1}{(\sigma^2+n_1d)^2} \left( \sum_{i=1}^k \sum_{j=1}^{n_1} \by_{ij}-k n_1 \mu\right) \\
& &- \frac{n_2}{(\sigma^2+n_2d)^2} \left( \sum_{i=k+1}^N \sum_{j=1}^{n_2} \by_{ij}-(N-k) n_2 \mu\right) \label{chessian3}
\end{aligned}
\end{equation}
with $\frac{\partial^2\ell}{\partial \mu \partial \sigma^2}$, $\frac{\partial^2\ell}{(\partial \sigma^2)^2}$, $\frac{\partial^2\ell}{\partial d \partial \sigma^2}$, $\frac{\partial^2\ell}{\partial \mu \partial d}$, $\frac{\partial^2\ell}{\partial \sigma^2 \partial d}$, and $\frac{\partial^2\ell}{\partial (d)^2}$ indentical to (\ref{jhessian4}),
(\ref{jhessian5}), (\ref{jhessian6}), (\ref{jhessian7}), (\ref{jhessian8}), and (\ref{jhessian9}).

\setcounter{equation}{0}
\section{Pseudo-likelihood for split samples}
\label{splitsample}

\subsection{General considerations}
Informally, a pseudo-likelihood function is one that replaces a computationally intractable or slow-to-maximize likelihood function, with another function that still produces a consistent and asymptotically normal estimator when maximised, i.e., by setting the first derivatives of the log pseudo-likelihood equal to zero and solving the resultant pseudo score equations. An early reference is \cite{arnold1991}, and details on various forms of pseudo-likelihood can be found in \cite[Ch.~9, 12, 21, 22, 24, and 25]{molenberghs2005}. In the current clustered setting, the likelihood contribution of a cluster is often replaced by a product of contributions for various sub-vectors. In some cases, such a sub-vector can be conditioned upon another sub-vector.

\cite{Verbeke2006} and \cite{Fetal06} used pseudo-likelihood to fit mixed models to high-dimensional multivariate longitudinal data. They supplemented the standard method with an additional device. They first replaced a set of $M$ longitudinal sequences by the $M(M-1)/2$ longitudinal pairs. This in itself is a standard application of pseudo-likelihood. They then assumed that each pair has its own parameter vector.
Symbolically, this can be written as:
\begin{equation}
p\ell({\bm{\theta}})  \equiv  p\ell(\mathbf{y}_{1i}, \mathbf{y}_{2i}, \ldots, \mathbf{y}_{Mi} |{\bm{\theta}})
 =  \sum_{r<s} \ell(\mathbf{y}_{ri}, \mathbf{y}_{si} | \bm{\theta}_{rs}), \label{mv23 eqn pseudo lik}
\end{equation}
where $\BY_{ri}$ is the $r$th sequence for subject~$i$. In (\ref{mv23 eqn pseudo lik}), $\bm{\theta}$ results from stacking all $M(M-1)/2$ pair-specific parameter vectors $\bm{\theta}_{rs}$. The actual parameter vector of interest is  $\bm{\theta}^*$, the set of non-redundant parameters is $\bm{\theta}$.

To obtain $\bm{\theta}^*$, \cite{Verbeke2006} take averages of all available estimates for that specific parameter, implying that $\mathbf{\widehat{\theta}^*} = A \mathbf{\widehat{\theta}}$ for an appropriate linear combination matrix $A$. Further, combining this step with general pseudo-likelihood inference, a sandwich estimator is used:
\begin{equation}
\sqrt{N}(\mathbf{\widehat{\theta}^*}- \mathbf{\theta^*} )   =  \sqrt{N}(A\widehat{\bm{\theta}}- A \\bm{\theta}) \stackrel{\mbox{\tiny approx.}}{\sim}  N(\mathbf{0}, A I_0^{-1}I_1 I_0^{-1} A'), \label{mv23 distr psister}
\end{equation}
where
\begin{equation}
I_0(\bftheta)=E\left[\frac{\partial^2p\ell(\bftheta)}{\partial\bftheta'\partial\bftheta}\right],\qquad
I_1(\bftheta)=E\left[
\left(\frac{\partial p\ell(\bftheta)}{\partial\bftheta}\right)'.
\frac{\partial p\ell(\bftheta)}{\partial\bftheta}
\right].
\end{equation}
\cite{Iddi2011} took a very similar route to partition a potentially large sample into sub-samples.

To fix ideas, consider log-likelihood (4.4).
%(\ref{jloglik}). 
When used as an instrument to estimate a single vector $(\mu,\sigma^2,d)$, this function can be viewed a pseudo-likelihood. This setting can be generalized by assuming that a dataset, consisting of repeated measures per subject, is divided into $K$ subgroups, each containing $c_k$ independent replicates. Consider the pseudo-likelihood:
\begin{equation}
\label{ourgenpl}
p\ell(\bftheta)=\sum_{k=1}^K\ell(\bftheta_k|\by_1^{(k)},\dots,\by_{c_k}^{(k)}).
\end{equation}
While the underlying principle is similar to (\ref{mv23 eqn pseudo lik}), it 
is not identical. The similarities are: (1) all $\bftheta_k$ are assumed to 
be different, allowing for separate, even parallel, estimation; (2) 
$\bftheta$ stacks all vectors $\bftheta_k$; (3) the parameter of 
interest $\bftheta^\ast$, is found from an appropriate combination 
of the $\bftheta_k$. Parallel estimation was also followed 
by \cite{scott2016bayes} and \cite{neiswanger2013asymptotically}.

\begin{sloppypar}
There are important differences, however. Here, and in the remainder 
of the article, we assume that
$\ell(\bftheta_k|\by_1^{(k)},\dots,\by_{c_k}^{(k)})$ is the likelihood 
that we would have, should group $k$ be the only one in the data. 
That is, the individual likelihood contributions are not altered, 
rather the data are partitioned. This is similar to the independent 
partitioning done by \cite{Iddi2011}. In line with their derivations, 
(\ref{mv23 distr psister}) can also be used here. Given that 
$\ell_k(\bftheta_k)$ is a genuine likelihood, its contributions 
to $I_0(\bftheta)$ and $I_1(\bftheta)$ are identical, up to the sign. 
As a result, $I_0(\bftheta)^{-1}I_1(\bftheta)I_0(\bftheta)^{-1}=-I_0(\bftheta)^{-1}$, 
a block-diagonal matrix with blocks of the form $I_0(\bftheta_k)$. We now turn to the split-sample case.
\end{sloppypar}

\subsection{Pseudo-likelihood for split sample }
\label{appC}

{\cite{Iddi2011}  chose
\begin{equation}
\label{amtrix}
A=\frac{1}{K}(I,\dots,I)
\end{equation}
to pass from $\bftheta$ to $\bftheta^\ast$. This is a sensible choice in the 
i.i.d.\ setting (e.g., when all clusters in a CS model have the same size) 
and with the same number of subjects per sub-sample. The estimator and 
precision estimator then become:
\begin{eqnarray}
\label{schatterindep}
\widehat{\bftheta}^*&=&\frac{1}{K}\sum_{k=1}^K\widehat{\bftheta}_k,\\
\label{asymptoticvarianceexpected}
\mbox{var}(\widehat{\bftheta}^*)&=&\frac{1}{K}H_{\widehat{\theta}}^{-1},
\end{eqnarray}
with $H_{\widehat{\theta}}^{-1}=-I_0(\bftheta_k)$. In this special case, 
the expected information matrices are identical. Alternatively, 
one can use the observed information matrices, and then use instead:
\begin{equation}
\label{asymptoticvarianceobserved}
\frac{1}{K^2}\sum_{k=1}^K{\cal H}_{\widehat{\theta}, k}^{-1}.
\end{equation}
where ${\cal H}_{\widehat{\theta}, k}$ is the observed 
information for sub-sample $k$.

In this particular case, pseudo-likelihood produces the same 
estimator as full likelihood. This stems from the fact that 
all subjects follow the same distribution, in contrast to, 
for example, the setting set out at the start of Section~4.
%\ref{model}. 
Should the subjects have identical distributions, but with $c_k$ variables, 
the above results can be modified accordingly. For example, 
(\ref{schatterindep}) would be replaced by
$$
\widehat{\bftheta}^*=\sum_{k=1}^K\frac{c_k}{N}\widehat{\bftheta}_k,
$$
with similar modification to precision estimation. In this case, 
full likelihood would be obtained.

In the next section, we will consider the more general case where 
different subjects may have a different distribution such as, 
for example, 
the CS case with a different number of measurements 
per cluster.}

\setcounter{equation}{0}
\section[Optimal scalar weights for CS]{Derivation of optimal scalar weights for compound-symmetry case}

To find the optimal scalar weight with minimum variance for $\mu$ we use the method of Lagrange with the constraint that the weights $a_k$ need to sum to 1:
\begin{equation}
Q=\sum_{k=1}^K a_k^2\frac{\sigma^2+n_k d}{c_kn_k}-\lambda\left(\sum_{k=1}^Ka_k-1\right).
\end{equation}
Solving the first partial derivative we become an expression for $a_k$ involving $\lambda$. Summing this one produces an expression for $\lambda$, leading to the complete formula for $a_k$. Precisely, the system of equations is:
\begin{eqnarray}
\frac{\partial Q}{\partial a_k}=2a_k\frac{\sigma^2+n_kd}{c_kn_k}-\lambda&=&0 ,\\
\frac{\partial Q}{\partial \lambda}= \sum_{k=1}^Ka_k-1&=&0,
\end{eqnarray}
or, alternatively:
\begin{eqnarray}
a_k&=&\frac{\lambda}{2}\frac{c_kn_k}{\sigma^2+n_k d},\\
\lambda&=& \left(\frac{1}{2} \sum_{k=1}^K \frac{c_kn_k}{\sigma^2+n_k d}\right)^{-1},
\end{eqnarray}
and hence
$$ a_k= \frac{\frac{c_kn_k}{\sigma^2+n_k d}}{\sum_{m=1}^K\frac{c_m n_m}{\sigma^2+n_m d}}
.$$
In the same manner, expressions for $b_k$ and $g_k$ can be found, as we will show next.
For $\sigma^2$:
$$
Q=2\sigma^4 \sum_{k=1}^K b_k^2\frac{1}{c_k(n_k-1)}-\lambda\left(\sum_{k=1}^Kb_k-1\right),
$$
producing the system:
\begin{eqnarray}
\frac{\partial Q}{\partial b_k}=\frac{4\sigma^4 b_k}{c_k(n_k-1)}-\lambda&=&0, \\
\frac{\partial Q}{\partial \lambda}= \sum_{k=1}^Kb_k-1&=&0,
\end{eqnarray}
which can be rewritten as:
\begin{eqnarray}
4\sigma^4 b_k&=&\lambda c_k (n_k-1),\\
\lambda&=& \frac{4\sigma^4}{\sum_{k=1}^Kc_k(n_k-1)},
\end{eqnarray}
and finally:
$$b_k= \frac{c_k(n_k-1)}{\sum_{m=1}^K c_m(n_m-1)}.
$$
For $d$, the objective function is:
$$
Q=\sum_{k=1}^K g_k^2 v_k-\lambda(\sum_{k=1}^Kg_k-1),
$$
with
$$v_k=\frac{2}{c_kn_k}\left( \frac{\sigma^4}{n_k-1}+2d\sigma^2+n_kd^2\right).$$
The system of equations now is:
\begin{eqnarray}
\frac{\partial Q}{\partial g_k}=2g_k v_k-\lambda&=&0, \\
\frac{\partial Q}{\partial \lambda}= \sum_{k=1}^K g_k-1&=&0,
\end{eqnarray}
leading to:
\begin{eqnarray}
g_k&=&\lambda \frac{1}{2 v_k},\\
\lambda&=& \frac{2}{\sum_{k=1}^K \frac{1}{v_k}},
\end{eqnarray}
giving the solution:
$$ g_k= \frac{\frac{c_kn_k}{\frac{\sigma^4}{n_k-1}+2d\sigma^2+n_kd^2}}{\sum_{m=1}^K\frac{c_m n_k}{n_m-1}+2d\sigma^2+n_md^2}.$$

\subsection{Cluster-by-cluster analysis}
\label{cbcaapp}

We study the case of the most extreme partitioning, i.e., where each of the clusters is analyzed separately. 
This can be relevant in cases with perhaps a limited number of very to extremely large clusters.  This 
means that $c_k\equiv 1$ throughout. Clearly, the $n_k$ will 
then no longer be unique. We will examine this case in detail, 
and contrast a first weighted estimator with an  
{\em ad hoc\/} one.

\subsubsection{The Weighted Estimator for the Cluster-by-cluster Case}

The estimator follows from setting $c_k\equiv 1$ and hence $K\equiv N$ 
throughout. For example, this special case can easily be considered 
for all expressions in Sections~6.1.1-6.1.3.
%\ref{owcs}--\ref{scalarcs}. 
Because $c_k$ 
enters the inverse of the variance-covariance matrix multiplicatively, as 
is seen from (4.8)-(4.9),
%(\ref{varexprmuk})--(\ref{varexprsigmakdk}), 
the optimal 
estimator that is obtained when each cluster is considered to be its own 
stratum, is identical to the one obtained when strata are defined in 
terms of all clusters of a given size. The same is true for the scalar 
weights.

It is insightful to consider in more detail the special case where 
further the cluster sizes are all identical to $n$. One then easily 
obtains:
\begin{eqnarray}
\label{muconstant}
\widehat{\mu}&=&\frac{1}{Nn}\sum_{i=1}^{N}\sum_{j=1}^{n}Y_{ij},\\[2mm]
\widehat{\sigma}^2&=&\frac{1}{Nn(n-1)}\left(
n\sum_{i=1}^{N}\BZ_{i}'\BZ_{i}-\sum_{i=1}^{N}\BZ_{i}'J_{n}\BZ_{i}
\right)  \\
&=&\frac{1}{Nn(n-1)}(nQ-R),\label{sigmaconstant}\\[2mm]
\widehat{d}&=&
\frac{1}{Nn(n-1)}\left(
\sum_{i=1}^{N}\BZ_{i}'J_{n}\BZ_i-\sum_{i=1}^{N}\BZ_{i}'\BZ_{i}
\right) \\
&=& \frac{1}{Nn(n-1)}(Q-R),\label{dconstant}
\end{eqnarray}
with obvious notation for $Q$ and $R$, inspired by
(\ref{qkfunc})--(\ref{rkfunc}). The corresponding variance-covariance 
elements, similar in spirit to (4.8)-(4.9),
%(\ref{varexprmuk})--(\ref{varexprsigmakdk}), 
are:
\begin{equation}
\label{varexprmucbyc}
\mbox{var}(\widehat{\mu})=\frac{\sigma^2+nd}{Nn},
\end{equation}
\begin{multline}
\label{varexprsigmadcbyc}
\mbox{var}\left(\begin{array}{c}\widehat{\sigma^2}\\ \widehat{d}\end{array}\right)=
\left(
\begin{array}{cc}
\frac{2\sigma^4}{N(n-1)}&-\frac{2\sigma^4}{Nn(n-1)}\\
-\frac{2\sigma^4}{Nn(n-1)}&
\frac{2}{Nn}\left[\frac{\sigma^4}{n-1}+2\sigma^2d+nd^2\right]
\end{array}
\right).
\end{multline}
This estimator coincides with the MLE, as is known from \cite{Iddi2011}.

\subsubsection{A Two-stage Estimator for Compound Symmetry}

In linear mixed models, there is a method of estimation, sometimes called 
the two-stage approach \citep{laird1982random,Molenberghs2000}, in which each cluster is 
analyzed separately to begin with, using linear regression, after 
which the cluster-specific parameters are summarized into fixed effects. 
Although the above cluster-by-cluster analysis is superficially similar to this,
it is not equivalent. In particular, there is no bias (as can be seen in 
the two stage method), and the maximum likelihood estimator is recovered.

This approach is most useful when cluster sizes are not constant, 
and in models that are more complex than compound symmetry. However, 
to gain some insight, we develop the details of the method for the CS model 
with constant cluster size.

For the mean, (\ref{muconstant}) is retained, as the average of the 
cluster-specific averages $\overline{Y}_i$. Further, define:
\begin{eqnarray}
\label{twostages1}
s^2&=&\frac{1}{Nn}\sum_{i=1}^N\sum_{j=1}^n(Y_{ij}-\overline{Y}_i)^2,\\
t^2&=&\frac{1}{N}\sum_{i1}^N(\overline{Y}_i-\widehat{\mu})^2.
\label{twostaget1}
\end{eqnarray}
Straightforward algebra shows:
\begin{eqnarray}
\label{exps1}
E(s^2)&=&\frac{n-1}{n}\sigma^2,\\
\label{expt1}
E(t^2)&=&\frac{N-1}{N}\left(d+\frac{1}{n}\sigma^2\right).
\end{eqnarray}
Should $n$ and $N$ approach infinity, then it follows that $s^2$ and $t^2$ 
are asymptotically unbiased estimators for $\sigma^2$ and $d$, 
respectively. However, this is not always reasonable. In applications 
such as the NTP data (Section~2),
%\ref{ntpintro}), 
it is fair to say that 
the cluster size has a biological upper limit. In other situations, 
however, such as meta-analyses, it is sensible to assume that both 
$n$ and $N$ approach infinity.

In the next section, we will study the consequences of removing the bias. 
For now, a small, obvious modification is:
\begin{eqnarray}
\label{twostages2}
s^2_\ast&=&\frac{n}{n-1}s^2,\\
t^2_\ast&=&\frac{N}{N-1}t^2.
\label{twostaget2}
\end{eqnarray}
Now, $s^2_\ast$ is unbiased, while $E(t^2_\ast)=d+\sigma^2/n$, 
the bias $\sigma^2/n$ can be made to disappear asymptotically 
provided it is sensible to let $n$ grow large.

It is of interest to consider the variance-covariance structure 
of the estimators $s^2$, $t^2$, $s^2_\ast$, and $t^2_\ast$, as 
well as to make relative efficiency considerations. This will 
be done next.

\subsubsection{Connections Between Estimators}

Comparing algebraic expressions (\ref{sigmaconstant})--(\ref{dconstant}) 
with (\ref{twostages1})--(\ref{twostaget1}), leads to the linear 
relationships:
\begin{eqnarray}
\label{rel1}
s^2&=&\frac{n-1}{n}\widehat{\sigma^2}+0\cdot\widehat{d},\\
t^2&=&\frac{N-1}{Nn}\widehat{\sigma^2}+\frac{N-1}{N}\widehat{d}.
\label{rel2}
\end{eqnarray}
Relationships (\ref{rel1})--(\ref{rel2}) can be combined with 
(\ref{varexprsigmadcbyc}) to produce:
\begin{equation}
\mbox{var}\left(\begin{array}{c}s^2\\ t^2\end{array}\right)=
\left(
\begin{array}{cc}
\frac{2(n-1)\sigma^4}{Nn^2}&0\\
0&
\frac{2(N-1)^2}{N^2n}\left[\frac{\sigma^4}{n}+2\sigma^2d+nd^2\right]
\end{array}
\right)
\end{equation}
and, similarly,
\begin{equation}
\mbox{var}\left(\begin{array}{c}s^2_\ast\\ t^2_\ast\end{array}\right)=
\left(
\begin{array}{cc}
\frac{2\sigma^4}{N(n-1)}&0\\
0&
\frac{2}{Nn}\left[\frac{\sigma^4}{n}+2\sigma^2d+nd^2\right]
\end{array}
\right).
\end{equation}
From its definition it follows that $s^2_\ast\equiv\widehat{\sigma^2}$. 
The same is not true for $t^2_\ast$. One reason to consider it 
nevertheless is its independence from $s^2_\ast$. Indeed, 
$(\widehat{\mu},s^2_\ast\equiv\widehat{\sigma^2}, t^2_\ast)'$ 
is an estimator with mutually independent components. While the 
same is true when $s^2$ and $t^2$ are used instead, the biases are 
larger.

For this case then, the choice between $\widehat{d}$ and $t^2_\ast$ 
is in terms of a trade-off between efficiency and independence.

To gauge the efficiency loss when using $t^2_\ast$, the mean 
squared error is:
$$MSE(t^2_\ast)=\frac{2}{Nn}\left(\frac{\sigma^4}{n}+2\sigma^2d+nd^2\right)+\frac{1}{n^2}\sigma^4,$$
and hence the relative MSE:
\begin{equation}
\label{releff}
RMSE(t^2;\widehat{d})=\frac
{
2\left(\frac{\sigma^4}{n}+2\sigma^2d+nd^2\right)+\frac{N}{n}\sigma^4
}
{
2\left(\frac{\sigma^4}{n-1}+2\sigma^2d+nd^2\right)
},
\end{equation}
which approaches infinity when $N$ does, while $n$ would 
remain constant. In other words, this estimator is inconsistent 
unless it is being applied in situations where $n$ can also be considered 
to be large.

There are three distinct situations. First, when $N/n=\lambda n+o(n)$, 
for some $\lambda$, i.e., when $N$ is of the order of $n^2$, then, 
based on (\ref{releff}), the ARE is $2(d^2+\lambda\sigma^4)/[2d^2]$. 
The magnitude of the efficiency loss depends on sizes of the parameters involved. 
Second, when ${\cal O}(N)<{\cal O}(n^2)$, the ARE equals~1. This includes 
the cases where $N$ is constant, $N=n^{1/2}$, $N=n$, and $N=n^{3/2}$, for 
example. A constant or slowly increasing $N$ is plausible in a meta-analytic 
context. Third,  if $N/n$ increases too quickly, i.e., 
${\cal O}(N/n)>{\cal O}(n)$, then the estimator $t^2_\ast$ is inconsistent. 
This is the case, in particular, for bounded $n$.

The estimators $s^2$ and $t^2$ can be combined 
linearly to produce unbiased estimators. In other words, 
based on (\ref{exps1})--(\ref{expt1}), the following 
corrections can be applied to (\ref{twostages1})--(\ref{twostaget1}):
\begin{eqnarray}
s^2_{\mbox{\scriptsize corr}}&=&\frac{n}{n-1}s^2,\\
t^2_{\mbox{\scriptsize corr}}&=&\frac{N}{N-1}t^2-\frac{N}{(n-1)(N-1)}s^2.
\end{eqnarray}
Interestingly, this requirement reproduces (\ref{rel1})--(\ref{rel2}): 
the requirement of an unbiased estimator reproduces $\widehat{\sigma^2}$ 
and $\widehat{d}$, presented in
(\ref{sigmaconstant})--(\ref{dconstant}) and hence also with their variance.

\setcounter{equation}{0}
\section{Details about the first simulation study}
\label{simapp}

The simulation study, summarized in Section~7,
%\ref{simulation}, 
is described in detail here.

\subsection{Simulation method}

The design of the simulation study is as follows.
\begin{itemize}
\item Each generated set of data consists of $c_k$ clusters of size $n_k$, for $k=1,\dots,K$. We choose $K=4$ throughout.
\item For a generated set of data, the splitting is done by placing all clusters of a given size in one sub-sample.
\item The CS model parameters are $\mu=0$, $d=1$, and $\sigma^2=2$.
\item After estimating the three model parameters within each sub-sample, they are combined using the following weighting methods: (a) equal, (b) proportional, where the weights are
$$w_k=\frac{c_k}{\sum_{\ell=1}^4 c_\ell},$$
and (c) size-proportional, where the weights for $\mu$ and $d$ are:
$$w_k\frac{c_k n_k}{\sum_{\ell=k}^4 c_\ell n_\ell},$$
while for $\sigma^2$ we take:
$$w_k=\frac{c_k (n_k-1)}{\sum_{\ell=1}^4 c_\ell (n_\ell-1)}.$$
\item Per setting, 100 replications are considered.
\end{itemize}
These settings are applied to various combinations of the $n_k$ and $c_k$, now described in turn.

\subsection[Setting 1]{Setting 1: equal $c_k\cdot n_k$, different $c_k$ and $n_k$.}

Consider  150 samples in each split, as follows: $(c_1,n_1)=(3,50)$, $(c_2,n_2)=(5,30)$, $(c_3,n_3)=(10,15)$, and $(c_4,n_4)=(15,10)$. The results are presented in Table~\ref{tab150}.  Graphical depictions can be found in Figures~\ref{CS150} and \ref{CS150COMB}. Figure~\ref{CS150} shows that there is a different amount of information in the various sub-samples. This is not a problem, rather a consequence of the way the splits are created and the different amounts of information carried in each. It reminds us that we need to be judicious how the information from the splits will be weighted. It is  not a surprise that equal weights are a poor choice. The other methods perform similarly, and all do very well. To varying degrees, the same will be seen in Settings~2 and~3.
\begin{table}[h]
\caption{\label{tab150}\label{tab150comb} \small \linespread{1.1} First simulation study. Setting 1. Average of split-specific and combined (weighted) parameters and their precision estimates.}

\vspace*{2mm}
\centering
\def\arraystretch{1}\begin{tabular}{rrrrrrr}
  \hline\hline
 & \multicolumn{1}{c}{$\mu$} & \multicolumn{1}{c}{\mbox{var}($\mu$)} & \multicolumn{1}{c}{$d$} & \multicolumn{1}{c}{\mbox{var}($d$)} & \multicolumn{1}{c}{$\sigma^2$} & \multicolumn{1}{c}{\mbox{var}($\sigma^2$)} \\
  \hline
split1 & -0.00396 & 0.05779 & 0.68143 & 0.41395 & 1.98676 & 0.00296 \\
  split2 & 0.05697 & 0.03071 & 0.80997 & 0.19712 & 1.98578 & 0.00304 \\
  split3 & -0.02111 & 0.01174 & 0.95161 & 0.07869 & 1.97690 & 0.00319 \\
  split4 & 0.01123 & 0.00626 & 0.98870 & 0.04677 & 1.98056 & 0.00347 \\
   \hline
Equal & 0.01078 & 0.03769 & 0.85793 & 0.09988 & 1.98250 & 0.01406 \\
  Prop & 0.00698 & 0.03230 & 0.92245 & 0.08568 & 1.98081 & 0.01907 \\
  Size prop & 0.01078 & 0.03769 & 0.85793 & 0.09988 & 1.98260 & 0.01405 \\
  Full & 0.00780 & 0.03513 & 0.98016 & 0.08614 & 1.98257 & 0.01392 \\
   \hline
\hline
\end{tabular}
\end{table}

\begin{figure}[!t]
\centering
\includegraphics[width=\textwidth]{aCS150.eps}
\caption{\small \linespread{1.1} First simulation study. Setting 1. Split-specific results.} \label{CS150}
\end{figure}
\begin{figure}[!t]
\centering
\includegraphics[width=\textwidth]{aCS150COMB.eps}
\caption{\small \linespread{1.1} First simulation study. Setting 1. Combining the results from the four splits, using equal, proportional, and size proportional weights. This is compared with full maximum likelihood.} \label{CS150COMB}
\end{figure}

\subsection[Setting 2]{Setting 2: different $c_k\cdot n_k$, equal $c_k$, different $n_k$}

To see the effect of split size, the following choices are made:
$(c_1,n_1)=(4,25)$,  $(c_2,n_2)=(4,50)$,  $(c_3,n_3)=(4,125)$, and $(c_4,n_4)=(4,250)$.
As a consequence, the size of the splits will be $100$, $200$, $500$, and $1000$, respectively. Table~\ref{tab_diff1} summarized the results, with graphical displays presented in Figures~\ref{CS_diff1} and \ref{CS_diff_comb}.
\begin{table}[t]
\centering
\caption{\small \linespread{1.1} First simulation study. Setting 2. Average of split-specific and combined (weighted) parameters and their precision estimates.}
\label{tab_diff1}

\vspace*{2mm}

\def\arraystretch{1} \begin{tabular}{rrrrrrr}
\hline  \hline
  & \multicolumn{1}{c}{$\mu$} & \multicolumn{1}{c}{\mbox{var}($\mu$)} & \multicolumn{1}{c}{$d$} & \multicolumn{1}{c}{\mbox{var}($d$)} & \multicolumn{1}{c}{$\sigma^2$} & \multicolumn{1}{c}{\mbox{var}($\sigma^2$)} \\
\hline
split1 & -0.02515 & 0.05227 & 0.83440 & 0.52423 & 2.00347 & 0.00730 \\
  split2 & 0.01287 & 0.05157 & 0.86891 & 0.57904 & 1.97285 & 0.00160 \\
  split3 & 0.06812 & 0.03586 & 0.74147 & 0.23681 & 2.00165 & 0.00026 \\
  split4 & -0.03676 & 0.02979 & 0.68241 & 0.14117 & 1.99216 & 0.00006 \\
  \hline
 Equal & 0.00477 & 0.05111 & 0.78180 & 0.14770 & 1.99253 & 0.00935 \\
  Prop & 0.00477 & 0.05111 & 0.78180 & 0.14770 & 1.99253 & 0.00935 \\
  Size prop & -0.00147 & 0.07139 & 0.72798 & 0.16585 & 1.99328 & 0.00447 \\
  Full & 0.00530 & 0.06339 & 0.89599 & 0.14604 & 1.99333 & 0.00446 \\
 \hline
\hline
\end{tabular}
\end{table}

\begin{figure}[!t]
\centering
\includegraphics[width=\textwidth]{aCS_diff1.eps}
\caption{\small \linespread{1.1} First simulation study. Setting 2. Split-specific results.} \label{CS_diff1}
\end{figure}
\begin{figure}[!t]
\centering
\includegraphics[width=\textwidth]{aCS_diff_comb.eps}
\caption{\small \linespread{1.1} First simulation study. Setting 2. Combining the results from the four splits, using equal, proportional, and size proportional weights. This is compared with full maximum likelihood.} \label{CS_diff_comb}
\end{figure}

\subsection[Setting 3]{Setting 3: different $c_k\cdot n_k$, different $c_k$, equal $n_k$}
We now choose: $(c_1,n_1)=(10,20)$,  $(c_2,n_2)=(20,20)$, $(c_3,n_3)=(50,20)$, and
$(c_4,n_4)=(100,20)$. Table~\ref{tab_diff2} summarizes the results. Graphs can be found in Figures~\ref{CS_diff2} and \ref{CS_diff_comb2}.
\begin{table}[h]
\centering
\caption{\small \linespread{1.1} First simulation study. Setting 3. Average of split-specific and combined (weighted) parameters and their precision estimates.}
\label{tab_diff2}

\vspace*{2mm}

\def\arraystretch{1} \begin{tabular}{rrrrrrr}
  \hline\hline
  & \multicolumn{1}{c}{$\mu$} & \multicolumn{1}{c}{\mbox{var}($\mu$)} & \multicolumn{1}{c}{$d$} & \multicolumn{1}{c}{\mbox{var}($d$)} & \multicolumn{1}{c}{$\sigma^2$} & \multicolumn{1}{c}{\mbox{var}($\sigma^2$)} \\
  \hline
split1 & 0.00343 & 0.00900 & 0.84739 & 0.05169 & 2.02445 & 0.00190 \\
  split2 & 0.02553 & 0.00304 & 1.00224 & 0.01754 & 2.01962 & 0.00047 \\
  split3 & -0.00010 & 0.00045 & 0.95794 & 0.00212 & 1.99765 & 0.00007 \\
  split4 & 0.01151 & 0.00012 & 1.01226 & 0.00064 & 1.98944 & 0.00002 \\
   \hline
 \hline
Equal & 0.01009 & 0.01139 & 0.95496 & 0.02694 & 2.00779 & 0.00486 \\
  Prop & 0.00939 & 0.00604 & 0.98690 & 0.01369 & 1.99702 & 0.00234 \\
  Size prop & 0.00939 & 0.00604 & 0.98690 & 0.01369 & 1.99702 & 0.00234 \\
  Full & 0.00939 & 0.00614 & 1.00487 & 0.01372 & 1.99702 & 0.00233 \\
 \hline\hline
\end{tabular}
\end{table}

\begin{figure}[!t]
\centering
\includegraphics[width=\textwidth]{aCS_diff2.eps}
\caption{\small \linespread{1.1} First simulation study. Setting 3. Split-specific results. \label{CS_diff2}}
\end{figure}
\begin{figure}[!t]
\centering
\includegraphics[width=\textwidth]{aCS_diff_comb2.eps}
\caption{\small \linespread{1.1} First simulation study. Setting 3. Combining the results from the four splits, using equal, proportional, and size proportional weights. This is compared with full maximum likelihood.} \label{CS_diff_comb2}
\end{figure}

\subsection{Optimal, approximate optimal, and iterated optimal weights}

Optimal weights were discussed in Section~6.1.1.
%\ref{owcs}. 
When we plug the MLE's into the optimal weights, the result of using these weights is the MLE's itself. Of course, this is a circular reasoning, which is why one needs to resort to, for example, the approximate or iterated optimal weights derived in Section~6.1.2.
%\ref{iaow}. 
For both of these, using Settings~1--3, we conducted simulations. They are reported in Figures~\ref{Setting1_opt}--\ref{Setting3_opt}.
\begin{figure}[!t]
\centering
\includegraphics[width=\textwidth]{aSetting1_opt.eps}
\caption{\small \linespread{1.1} First simulation study. Setting 1. (Size) proportional, approximate, and iterated optimal weights, as well as full maximum likelihood.} \label{Setting1_opt}
\end{figure}
\begin{figure}[!t]
\centering
\includegraphics[width=\textwidth]{aSetting2_opt.eps}
\caption{\small \linespread{1.1} First simulation study. Setting 2. (Size) proportional, approximate, and iterated optimal weights, as well as full maximum likelihood.} \label{Setting2_opt}
\end{figure}
\begin{figure}[!t]
\centering
\includegraphics[width=\textwidth]{aSetting3_opt.eps}
\caption{\small \linespread{1.1} First simulation study. Setting 3. (Size) proportional, approximate, and iterated optimal weights, as well as full maximum likelihood.} \label{Setting3_opt}
\end{figure}

It is noteworthy that the behavior of the iterated optimal weights depends on $c_k$ and $n_k$. First, they often but not always converge in a single iteration; the maximum number of iterations observed in our simulations being~6. Second, the iterated optimal weights  converge to size optimal weights for $\sigma^2$ and to proportional weights for $d$.

Taken together, it follows that both approximately optimal and iterated optimal weights provide excellent results. The specific attraction of the approximate optimal weights is that they obviate the need for iteration, which is a factor of stability and speed.

\setcounter{equation}{0}
\section{Details about the second simulation study}
\label{simapptwee}

The aim of this study is to compare the proposed method to two alternatives: 
\begin{enumerate}
\item full maximum likelihood;
\item the proposed  sample-splitting method, allowing for closed forms;
\item using multiple imputation (MI) first, to render the clusters of equal sizes, and then apply closed-form solutions to the augmented balanced data, together with the combination rules.
\end{enumerate}

\subsection{Simulation plan}

In order to study the effect of cluster sizes ($n_k$) and number of clusters of each size ($c_k$), 5 different configurations are considered:
\begin{description}
\item[Config.~1.] $c_k=(15,25,30,20,10)$, $n_k=(8,5,3,9,15)$;
\item[Config.~2.] $c_k=(150,250,300,200,100)$, $n_k=(8,5,3,9,15)$;
\item[Config.~3.] $c_k=(1500,2500,3000,2000,1000)$, $n_k=(8,5,3,9,15)$;
\item[Config.~4.] $c_k=(15,25,30,20,10)$, $n_k=(80,50,30,90,150)$;
\item[Config.~5.] $c_k=(15,25,30,20,10)$, $n_k=(800,500,300,900,1500)$.
\end{description}
 Each configuration is repeated $100$ times.

Each cluster is generated from a CS model with $\mu_0=0$, $d_0=1$, and $\sigma^2_0=4$.

For estimating the parameters using the full unbalanced data, PROC MIXED in SAS (Version 9.4) is used with the covariance structure in the REPEATED statement set to {\tt{type=cs}}.

The closed form solutions and their variances are implemented in {\tt R} in three different ways. First, the formulas  are implemented directly using `for' loops. Following the ideas in \cite{sikorska2013_2}, it might be faster to replace `for' loops with vectorized computation. For $\widehat{\mu}_k$ it is straightforward, since one just needs to compute an arithmetic average. If $Z$ is a $n_k$ times $c_k$ matrix with its $i$th column defined as $Z_i^{(k)}=\left(Y_i^{(k)} - \mu_k \mathbf{1}_{n_k} \right)$, then computing $\sum_{i=1}^ {c_k}  Z_i^{(k)'} Z_i^{(k)}$  is equivalent to replacing each element in matrix $Z$ by its square, and then sum over the sum of its columns. Furthermore, $J_{n_k} Z_i^{(k)}$ would simply compute the sum of columns in matrix $Z$. Therefore, $\sum_{i=1}^{c_k} Z_i^{(k)'} J_{n_k} Z_I^{(k)}$ is equivalent to post-multiplying $Z$ by the sum of its columns and then sum over this vector. In this way, within each split, the parameters can be estimated avoiding `for' loops.

Second, it is also possible to find the estimates for all of the splits at once instead of computing them separately.

A third way consists of calculating all the estimates together and not split by split in a `for' loop. This approach is possible via imposing balance through adding missing values in the matrix but, when multiplying and summing, ignoring the missing values. This is very easy in {\tt R}.

We will compare computation time between these three approaches, for the five configurations.

Additionally, to combine the results from sample splitting, the same
weights as used in the case study are considered here as well: equal, proportional, approximate scalar, scalar, and approximate optimal. In the case of the approximate optimal weights both simple and proper variances are calculated.

For the multiple imputation based approach, $M=20$ imputations are considered and the conventional combination rules applied.

 Note that the MI approach cannot be used with configurations 1, 4, and 5, because the number of available subjects in the observed dataset is less than the number of repeated measurements, leading to a singular covariance matrix. From the remaining configurations 2 and 3, we have chosen \#2, which implies smaller numbers and hence is more challenging.

For each configuration we report three results: the estimated parameters, their standard errors, and the mean square error (MSE). Furthermore, we report computation time.

\subsection{Simulation results}

\begin{table}[ht]
\centering
\caption{\small \linespread{1.1} Second simulation study. Mean, standard deviation (S.D.) and MSE for $\mu$ among 100 replications for each configuration using different combination weights comparing with full sample MLE.}
\label{tab_mu_est}

\vspace*{2mm}

\resizebox{\textwidth}{!}{%
\def\arraystretch{1} \begin{tabular}{cccccccc}
  \hline\hline
Config.& & Equal & Prop & Approx.~sc. & Scalar & Approx.~opt. & ML \\ 
  \hline
 & Mean & -1.72277E-02 & -1.51605E-02 & -1.51605E-02 & -1.21296E-02 & -1.21296E-02 & -1.56028E-02 \\ 
1 & S.D. & (1.32751E-01) & (1.28989E-01) & (1.28989E-01) & (1.32435E-01) & (1.32435E-01) & (1.29150E-01) \\ 
 & MSE & 1.77434E-02 & 1.67015E-02 & 1.67015E-02 & 1.75108E-02 & 1.75108E-02 & 1.67564E-02 \\ 
 & Mean & -9.39926E-04 & 6.93419E-04 & 6.93419E-04 & 1.27613E-03 & 1.27613E-03 & 7.59533E-04 \\ 
2 & S.D. & (3.93526E-02) & (3.95534E-02) & (3.95534E-02) & (3.84321E-02) & (3.84321E-02) & (3.83996E-02) \\ 
 & MSE & 1.53403E-03 & 1.54930E-03 & 1.54930E-03 & 1.46389E-03 & 1.46389E-03 & 1.46036E-03 \\ 
 & Mean & -8.30934E-04 & -1.25609E-03 & -1.25609E-03 & -1.26810E-03 & -1.26810E-03 & -1.31356E-03 \\ 
3 & S.D.& (1.44839E-02) & (1.47545E-02) & (1.47545E-02) & (1.41310E-02) & (1.41310E-02) & (1.41704E-02) \\ 
 & MSE & 2.08376E-04 & 2.17095E-04 & 2.17095E-04 & 1.99298E-04 & 1.99298E-04 & 2.00517E-04 \\ 
 & Mean & 9.30928E-03 & 2.53713E-03 & 2.53713E-03 & 8.29367E-03 & 8.29367E-03 & 2.82086E-03 \\ 
4 & S.D. & (9.26881E-02) & (8.32009E-02) & (8.32009E-02) & (9.76672E-02) & (9.76672E-02) & (8.28999E-02) \\ 
 & MSE & 8.59183E-03 & 6.85960E-03 & 6.85960E-03 & 9.51227E-03 & 9.51227E-03 & 6.81163E-03 \\ 
 & Mean & 9.77532E-03 & 1.02173E-02 & 1.02173E-02 & 8.72381E-03 & 8.72381E-03 & 1.02422E-02 \\ 
5 & S.D. & (1.09769E-01) & (1.04876E-01) & (1.04876E-01) & (1.04982E-01) & (1.04982E-01) & (1.04847E-01) \\ 
 & MSE & 1.20243E-02 & 1.09934E-02 & 1.09934E-02 & 1.09872E-02 & 1.09872E-02 & 1.09880E-02 \\ 
   \hline\hline
\end{tabular}}
\end{table}

\begin{table}[ht]
\centering
\caption{\small \linespread{1.1} Second simulation study. Mean and standard deviation (S.D.) for standard errors of $\mu$ estimates in 100 replications for each configuration using different combination weights comparing with full sample MLE.}
\label{tab_mu_std}

\vspace*{2mm}

\resizebox{\textwidth}{!}{%
\def\arraystretch{1} \begin{tabular}{ccccccccc}
  \hline\hline
Config.& & Equal & Prop & Approx.~sc. & Scalar & Simple opt. & Proper opt. & ML \\ 
  \hline
1  &Mean & 1.29844E-01 & 1.29733E-01 & 1.29733E-01 & 1.20593E-01 & 1.20593E-01 & 1.41085E-01 & 1.29648E-01 \\ 
  &S.D.& (1.18175E-02) & (1.07496E-02) & (1.07496E-02) & (1.09122E-02) & (1.09122E-02) & (2.25155E-02) & (9.95614E-03) \\ 
2  &Mean & 4.23655E-02 & 4.23056E-02 & 4.23056E-02 & 4.11657E-02 & 4.11657E-02 & 4.12635E-02 & 4.14298E-02 \\ 
  &S.D.& (1.08386E-03) & (8.88747E-04) & (8.88747E-04) & (8.71465E-04) & (8.71465E-04) & (8.90259E-04) & (8.66432E-04) \\ 
3  &Mean & 1.34008E-02 & 1.33822E-02 & 1.33822E-02 & 1.30725E-02 & 1.30725E-02 & 1.30728E-02 & 1.30799E-02 \\ 
  &S.D.& (1.21880E-04) & (9.92324E-05) & (9.92324E-05) & (9.68871E-05) & (9.68871E-05) & (9.68974E-05) & (9.62652E-05) \\ 
4  &Mean & 1.06373E-01 & 1.01232E-01 & 1.01232E-01 & 9.60807E-02 & 9.60807E-02 & 3.30358E-01 & 1.03382E-01 \\ 
  &S.D.& (9.80278E-03) & (7.26031E-03) & (7.26031E-03) & (8.36242E-03) & (8.36242E-03) & (1.39042E-01) & (7.17708E-03) \\ 
4  &Mean & 1.05176E-01 & 9.84615E-02 & 9.84615E-02 & 9.45066E-02 & 9.45066E-02 & 2.81427E+00 & 1.00533E-01 \\ 
  &S.D.& (1.10711E-02) & (8.18259E-03) & (8.18259E-03) & (8.14229E-03) & (8.14229E-03) & (1.41211E+00) & (8.28537E-03) \\ 
\hline\hline  
\end{tabular}}
\end{table}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{fig_mu.eps}
\caption{\small \linespread{1.1}  Second simulation study. Estimates for $\mu$ (first row) and its standard error (second row).} \label{fig_mu}
\end{figure}

\begin{table}[ht]
\centering
\caption{\small \linespread{1.1} Second simulation study. Mean, standard deviation (S.D.) and MSE for $d$ estimates in 100 replications for each configuration using different combination weights comparing with full sample MLE.}
\label{tab_d_est}

\vspace*{2mm}

\resizebox{\textwidth}{!}{%
\def\arraystretch{1} \begin{tabular}{cccccccc}
  \hline\hline
Config.& & Equal & Prop & Approx.~sc. & Scalar & Approx.~opt & ML \\ 
  \hline
 & Mean & 9.09580E-01 & 9.00788E-01 & 9.00788E-01 & 6.63293E-01 & 6.71111E-01 & 9.86549E-01 \\ 
1 & S.D. & (2.66885E-01) & (2.99736E-01) & (2.99736E-01) & (2.70556E-01) & (2.79060E-01) & (2.55874E-01) \\ 
 & MSE & 7.86910E-02 & 9.87862E-02 & 9.87862E-02 & 1.85840E-01 & 1.85264E-01 & 6.49975E-02 \\ 
 & Mean & 9.98984E-01 & 9.97534E-01 & 9.97534E-01 & 9.72269E-01 & 9.73771E-01 & 1.00712E+00 \\ 
2 & S.D. & (7.44958E-02) & (8.28143E-02) & (8.28143E-02) & (7.28186E-02) & (7.22762E-02) & (7.08463E-02) \\ 
 & MSE & 5.49515E-03 & 6.79570E-03 & 6.79570E-03 & 6.01854E-03 & 5.85958E-03 & 5.01969E-03 \\ 
 & Mean & 1.00004E+00 & 9.99697E-01 & 9.99697E-01 & 9.97218E-01 & 9.97153E-01 & 1.00019E+00 \\ 
3 & S.D. & (2.61046E-02) & (2.82480E-02) & (2.82480E-02) & (2.48989E-02) & (2.48068E-02) & (2.44715E-02) \\ 
 & MSE & 6.74637E-04 & 7.90063E-04 & 7.90063E-04 & 6.21496E-04 & 6.17332E-04 & 5.92900E-04 \\ 
 & Mean & 9.40257E-01 & 9.50756E-01 & 9.50756E-01 & 7.65219E-01 & 7.65362E-01 & 9.96005E-01 \\ 
4 & S.D. & (1.53414E-01) & (1.48216E-01) & (1.48216E-01) & (1.91865E-01) & (1.91614E-01) & (1.49451E-01) \\ 
 & MSE & 2.68697E-02 & 2.41734E-02 & 2.41734E-02 & 9.15661E-02 & 9.14038E-02 & 2.21282E-02 \\ 
 & Mean & 9.63459E-01 & 9.68182E-01 & 9.68182E-01 & 8.21266E-01 & 8.21270E-01 & 1.00958E+00 \\ 
5 & S.D. & (1.73767E-01) & (1.63471E-01) & (1.63471E-01) & (1.72162E-01) & (1.72163E-01) & (1.69235E-01) \\ 
 & MSE & 3.12283E-02 & 2.74677E-02 & 2.74677E-02 & 6.12894E-02 & 6.12881E-02 & 2.84459E-02 \\ 
\hline
   \hline
\end{tabular}}
\end{table}

\begin{table}[ht]
\centering
\caption{\small \linespread{1.1}Second simulation study. Mean and standard deviation (S.D.) for standard errors of $d$ estimates in 100 replications for each configuration using different combination weights comparing with full sample MLE.}
\label{tab_d_std}

\vspace*{2mm}

\resizebox{\textwidth}{!}{%
\def\arraystretch{1} \begin{tabular}{ccccccccc}
  \hline\hline
Config.& & Equal & Prop. & Approx.~sc. & Scalar & Simple opt. & Proper opt. & ML \\ 
  \hline
1 & Mean & 5.30888E-01 & 6.76159E-01 & 6.76159E-01 & 4.83519E-01 & 1.95328E-01 & 4.66561E+00 & 2.36408E-01 \\ 
 & S.D. & (1.00495E-01) & (1.66365E-01) & (1.66365E-01) & (1.10589E-01) & (4.01066E-02) & (1.40173E+00) & (3.91752E-02) \\ 
2 & Mean & 1.60365E-01 & 2.03126E-01 & 2.03126E-01 & 1.45284E-01 & 7.48056E-02 & 1.47955E+00 & 7.60798E-02 \\ 
 & S.D. & (8.29444E-03) & (1.42616E-02) & (1.42616E-02) & (8.70034E-03) & (3.35411E-03) & (1.31587E-01) & (3.26483E-03) \\ 
3 & Mean & 5.02596E-02 & 6.34861E-02 & 6.34861E-02 & 4.56061E-02 & 2.39463E-02 & 4.67989E-01 & 2.39748E-02 \\ 
 & S.D. & (8.43976E-04) & (1.44201E-03) & (1.44201E-03) & (8.34510E-04) & (3.68251E-04) & (1.24414E-02) & (3.66865E-04) \\ 
4 & Mean & 1.62410E-01 & 1.59656E-01 & 1.59656E-01 & 1.34256E-01 & 1.24552E-01 & 2.35351E-01 & 1.51740E-01 \\ 
 & S.D. & (2.83009E-02) & (2.05482E-02) & (2.05482E-02) & (2.31559E-02) &(2.51091E-02) & (3.00109E-02) & (2.11541E-02) \\ 
5 & Mean & 1.54122E-01 & 1.43313E-01 & 1.43313E-01 & 1.22117E-01 & 1.22024E-01 & 1.70868E-01 & 1.43764E-01 \\ 
 & S.D. & (3.48725E-02) & (2.50371E-02) & (2.50371E-02) & (2.30972E-02)& (2.31064E-02) & (3.70951E-02) & (2.38992E-02) \\ 
   \hline\hline
\end{tabular}}
\end{table}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{fig_d.eps}
\caption{\small \linespread{1.1} Second simulation study. Estimates for $d$ (first row) and standard errors ( second row).} \label{fig_d}
\end{figure}

\begin{table}[ht]
\centering
\caption{\small \linespread{1.1}Second simulation study. Mean, standard deviation (S.D.) and MSE for $\sigma^2$ estimates in 100 replications for each configuration using different combination weights comparing with full sample MLE.}
\label{tab_sigma2_est}

\vspace*{2mm}

\resizebox{\textwidth}{!}{%
\def\arraystretch{1} \begin{tabular}{cccccccc}
  \hline\hline
Config.& & Equal & Prop. & Approx.~sc. & Scalar & Approx.~opt. & ML \\ 
  \hline
 & Mean & 3.98608E+00 & 3.99739E+00 & 3.98571E+00 & 3.98364E+00 & 3.87487E+00 & 3.98075E+00 \\ 
1 & S.D. & (2.50882E-01) & (2.95821E-01) & (2.35138E-01) & (2.33650E-01) & (2.38167E-01) & (2.30477E-01) \\ 
 & MSE & 6.25062E-02 & 8.66420E-02 & 5.49414E-02 & 5.43140E-02 & 7.18141E-02 & 5.29588E-02 \\ 
 & Mean & 4.00184E+00 & 4.00681E+00 & 4.00177E+00 & 4.00087E+00 & 3.99027E+00 & 4.00064E+00 \\ 
2 & S.D. & (7.85190E-02) & (9.05849E-02) & (7.57443E-02) & (7.56486E-02) & (7.50477E-02) & (7.51739E-02) \\ 
 & MSE & 6.10698E-03 & 8.16998E-03 & 5.68296E-03 & 5.66624E-03 & 5.67055E-03 & 5.59502E-03 \\ 
 & Mean & 4.00509E+00 & 4.00491E+00 & 4.00575E+00 & 4.00590E+00 & 4.00472E+00 & 4.00587E+00 \\ 
3 & S.D. & (2.45760E-02) & (2.82395E-02) & (2.36027E-02) & (2.36983E-02) & (2.37694E-02) & (2.36697E-02) \\ 
 & MSE & 6.23828E-04 & 8.13646E-04 & 5.84605E-04 & 5.90806E-04 & 5.81652E-04 & 5.89081E-04 \\ 
 & Mean & 4.01402E+00 & 4.01190E+00 & 4.01302E+00 & 4.01304E+00 & 4.00363E+00 & 4.01306E+00 \\ 
4 & S.D. & (7.69655E-02) & (8.78962E-02) & (7.32088E-02) & (7.31202E-02) & (7.20589E-02) & (7.31207E-02) \\ 
 & MSE & 6.06101E-03 & 7.79021E-03 & 5.47558E-03 & 5.46319E-03 & 5.15372E-03 & 5.46370E-03 \\ 
 & Mean & 4.00346E+00 & 4.00338E+00 & 4.00292E+00 & 4.00292E+00 & 4.00192E+00 & 4.00292E+00 \\ 
5 & S.D. & (2.56561E-02) & (2.79741E-02) & (2.46670E-02) & (2.46664E-02) & (2.46901E-02) & (2.46669E-02) \\ 
 & MSE & 6.63599E-04 & 7.86128E-04 & 6.10884E-04 & 6.10854E-04 & 6.07187E-04 & 6.10909E-04 \\ 
   \hline\hline
\end{tabular}}
\end{table}

\begin{table}[ht]
\centering
\caption{\small \linespread{1.1} Second simulation study. Mean and standard deviation (S.D.) for standard errors of $\sigma^2$ estimates in 100 replications for each configuration using different combination weights comparing with full sample MLE.}
\label{tab_sigma2_std}

\vspace*{2mm}

\resizebox{\textwidth}{!}{%
\def\arraystretch{1} \begin{tabular}{ccccccccc}
  \hline\hline
Config.& & Equal & Prop. & Approx.~sc. & Scalar & Simple opt. & Proper opt. & ML \\ 
  \hline
1  &Mean & 2.53758E-01 & 2.95524E-01 & 2.40482E-01 & 2.38699E-01 & 2.30855E-01 & 2.88730E+01 & 2.35804E-01 \\ 
  &S.D. & (1.98892E-02) & (3.33124E-02) & (1.45355E-02) & (1.39849E-02) & (1.38813E-02) & (6.93951E+00) & (1.34905E-02) \\ 
2  &Mean  & 7.98210E-02 & 9.26570E-02 & 7.57927E-02 & 7.53242E-02 & 7.48395E-02 & 8.82494E+00 & 7.49872E-02 \\ 
  &S.D. & (1.84469E-03) & (3.08841E-03) & (1.45443E-03) & (1.42354E-03) & (1.40575E-03) & (6.92140E-01) & (1.39989E-03) \\ 
3  &Mean  & 2.52202E-02 & 2.92034E-02 & 2.39687E-02 & 2.38353E-02 & 2.37400E-02 & 2.78559E+00 & 2.37444E-02 \\ 
  &S.D. & (1.85701E-04) & (3.12856E-04) & (1.42587E-04) & (1.40888E-04) & (1.40071E-04) & (6.70930E-02) & (1.39784E-04) \\ 
4  &Mean  & 7.22378E-02 & 8.07793E-02 & 7.01684E-02 & 7.01663E-02 & 6.99970E-02 & 7.23107E+00 & 7.01238E-02 \\ 
  &S.D. & (1.54384E-03) & (2.35043E-03) & (1.28712E-03) & (1.28385E-03) & (1.26263E-03) & (5.28710E-01) & (1.27765E-03) \\ 
5  &Mean  & 2.25782E-02 & 2.52054E-02 & 2.19702E-02 & 2.19702E-02 & 2.19647E-02 & 2.23651E+00 & 2.19689E-02 \\ 
  &S.D. & (1.55999E-04) & (2.29649E-04) & (1.35364E-04) & (1.35358E-04) & (1.35462E-04) & (5.50362E-02) & (1.35379E-04) \\ 
   \hline\hline
\end{tabular}}
\end{table}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{fig_sigma2.eps}
\caption{\small \linespread{1.1}  Second simulation study. Estimates for $\sigma^2$ (first row) and standard errors (second row).} \label{fig_sigma2}
\end{figure}

\begin{table}[ht]
\centering
\caption{\small \linespread{1.1} Second simulation study. Computation time (in seconds) using closed-form solutions with different implementation forms, compared to PROC MIXED.}

\vspace*{2mm}
\resizebox{\textwidth}{!}{%
\def\arraystretch{1} 
\begin{tabular}{cccccc}
  \hline\hline
&&\multicolumn{2}{c}{Split by split} & & PROC \\ 
\cline{3-4}
Config.& & without loops &  using loops & Together & MIXED \\ 
  \hline
1& Mean & \textbf{0.00520} & 0.00980 & 0.00640 & 0.34155 \\ 
 & S.D. & (0.00882) & (0.00943) & (0.00823) & (0.17711) \\ 
2  &Mean & \textbf{0.02150} & 0.07340 & 0.05660 & 0.35575 \\ 
  &S.D. & (0.01617) & (0.02006) & (0.09662) & (0.03073) \\ 
3  &Mean & \textbf{0.17980} & 0.73480 & 0.43400 & 0.81783 \\ 
  &S.D. & (0.02292) & (0.05835) & (0.11543) & (0.02582) \\ 
4  &Mean & \textbf{0.00220} & 0.00610 & 0.00360 & 2.58808 \\ 
  &S.D. & (0.00579) & (0.01497) & (0.00689) & (0.40720) \\ 
5  &Mean & 0.04030 & 0.27490 &\textbf{ 0.02130} & 629.58333 \\ 
  &S.D. & (0.01521) & (0.01941) & (0.00872) & (116.09435) \\ 
\hline\hline
\end{tabular}}
\end{table}

\begin{table}[ht]
\centering
\caption{\small \linespread{1.1} Second simulation study. Mean, standard deviation (S.D.) and MSE for CS parameter estimates in 100 replications for configuration 2 using different combination weights comparing with full sample MLE and MI-MLE.}
\label{tab_mi_est}

\vspace*{2mm}

\resizebox{\textwidth}{!}{%
\def\arraystretch{1} \begin{tabular}{rcccccccc} 
  \hline\hline
& & Equal & Prop. & Approx.~sc.\ & Scalar & Approx.~opt. & MI & ML \\ 
  \hline
&Mean & -5.09643E-03 & -2.62061E-03 & -2.62061E-03 & -2.88933E-03 & -2.88933E-03 & -8.04195E-03 & -3.28224E-03 \\ 
$\mu$&  S.D. & (4.85424E-02) & (4.91772E-02) & (4.91772E-02) & (4.68032E-02) & (4.68032E-02) & (6.00662E-02) & (4.71723E-02) \\ 
&  MSE & 2.35877E-03 & 2.40108E-03 & 2.40108E-03 & 2.17698E-03 & 2.17698E-03 & 3.63654E-03 & 2.21375E-03 \\ 
&  Mean & 9.98392E-01 & 9.95216E-01 & 9.95216E-01 & 9.70589E-01 & 9.71627E-01 & 3.51123E-02 & 9.92960E-01 \\ 
$d$&   S.D. & (7.33193E-02 & (7.46946E-02) & (7.46946E-02) & (7.59516E-02) & (7.53362E-02) & (1.83983E-02) & (7.50610E-02) \\ 
&   MSE & 5.32455E-03 & 5.54638E-03 & 5.54638E-03 & 6.57598E-03 & 6.42380E-03 & 9.31343E-01 & 5.62737E-03 \\ 
& Mean & 4.00782E+00 & 4.00791E+00 & 4.00581E+00 & 4.00544E+00 & 3.99400E+00 & 5.32627E+00 & 4.00347E+00 \\ 
$\sigma^2$&   S.D. & (7.66500E-02) & (8.98881E-02) & (7.19708E-02) & (7.13441E-02) & (7.19080E-02) & (1.55770E-01) & (7.56968E-02) \\ 
&  MSE & 5.87763E-03 & 8.06169E-03 & 5.16181E-03 & 5.06871E-03 & 5.15505E-03 & 1.78301E+00 & 5.68472E-03 \\ 
   \hline\hline
\end{tabular}}
\end{table}

\begin{table}[ht]
\caption{\small \linespread{1.1} Second simulation study. Mean and standard deviation (S.D.) for the standard error of CS parameter estimates in 100 replications for configuration 2 using different combination weights comparing with full sample MLE and MI-MLE.}
\label{tab_mi_std}

\vspace*{2mm}

\centering
\resizebox{\textwidth}{!}{%
\def\arraystretch{1} \begin{tabular}{rccccccccc}
  \hline\hline
 & & Equal & Prop. & Approx.~sc. & Scalar & Simple opt. & Proper opt. & MI &ML \\   
  \hline
$\mu$ &Mean & 4.24162E-02 & 4.22766E-02 & 4.22766E-02 & 4.11356E-02 & 4.11356E-02 & 4.12439E-02 & 4.95431E-02 & 4.14004E-02 \\ 
 & S.D. & (1.21548E-03) & (8.59430E-04) & (8.59430E-04) & (9.31222E-04) & (9.31222E-04) & (9.42598E-04) & (7.65032E-03) & (9.11785E-04) \\ 
$d$ & Mean & 1.60545E-01 & 2.02922E-01 & 2.02922E-01 & 1.45623E-01 & 7.47531E-02 & 1.48865E+00 & 9.10211E-02 & 7.54391E-02 \\ 
 &  S.D. & (8.88524E-03) & (1.47416E-02) & (1.47416E-02) & (9.84170E-03) & (3.76197E-03) & (1.29914E-01) & (5.18621E-03) & (3.38044E-03) \\ 
$\sigma^2$ & Mean & 7.99442E-02 & 9.26315E-02 & 7.58626E-02 & 7.54139E-02 & 7.49171E-02 & 8.86450E+00 & 1.64310E-01 & 7.50377E-02 \\ 
 &  S.D. & (1.87358E-03) & (3.14842E-03) & (1.39531E-03) & (1.34244E-03) & (1.33807E-03) & (6.56714E-01) & (2.27104E-02) & (1.42688E-03) \\ 
   \hline\hline
\end{tabular}}
\end{table}

\begin{figure}[t]
\centering
\includegraphics[width=0.9\textwidth]{fig_mi.eps}
\caption{\small \linespread{1.1}  Second simulation study. Estimated CS parameters (first row) and their standard error (second row) using sample splitting, MI-MLE, and MLE.} \label{fig_mi}
\end{figure}

Based on the simulation results, it appears that  using equal weights is not recommended, while using proportional weights produces results comparable with ML. Of course, in case of $\sigma^2$ the approximate scalar weights  work better comparing with ML. An interesting outcome of the simulation is that by keeping the number of clusters of different sizes constant, but allowing the cluster sizes to increase, improves estimation of  $\sigma^2$, while increasing the number of clusters and keeping their sizes constant improves the estimation of  $d$. This is not surprising, because $d$ is the between-cluster variability, which is easier to estimate from a larger number of clusters. This should be seen against the background of relatively small differences anyway.

The results based on MI are not comparable with sample-splitting results. In particular, the variance component $d$ is underestimated using MI, while $\sigma^2$ is overestimated.  The larger standard errors in this case suggest that the sample-splitting methods use information more efficiently.

Comparing computation times, the closed-form approaches are the clear winners, further enhanced by smaller standard errors.  Furthermore, it follows that computing the estimates in a semi-parallel fashion, thus avoiding `for' loops within the splits but using then  between splits, is most efficient. Unless the $n_k$'s are very large, computing all  estimates at once is the most efficient. Of course, if the estimates for different splits can be done in parallel (without `for' loops), this is more efficient than estimating them all at once.

%% Ar1
\setcounter{equation}{0}
\section{The balanced conditionally independent model\label{bcim}}

In this case, one imposes the following structure on (\ref{lmmhier}):
\begin{itemize}
\item \Xik\ can be rewritten in terms of a first matrix that imposes structure between clusters (e.g., treatment effect), termed \Aik, and a second one that imposes structure within clusters (e.g., time evolution), $\Tik'=(\Zik',\Qik')'$.
\item The matrices \Aik, \Zik, and \Qik\ are constant among all clusters of size~$n_k$.
\item The matrix $\Sigma_i^{(k)}=\sigma^2I_{n_k}$.
\end{itemize}
This is the general, balanced growth-curve model as studied by \cite{lange1989effect} and \cite{verbeke2007effect}.
\label{setting1est}
Building on their development, we will now derive sufficient statistics and associated maximum likelihood estimators for the parameters in this model. This can be expressed
$$\BY=A(\bfbeta_1,\bfbeta_2)\left(
\begin{array}{c}
Z\\
Q
\end{array}
\right)
 +
BZ
+\beps.
$$
Here, $\BY$ is an $N\times n$ matrix stacking the outcomes of all clusters of size $c$, $A$, $Z$, and $Q$ group the designs mentioned in Section~\ref{defsetting1}, the vectors $\bfbeta_1$ and $\bfbeta_2$ contain the fixed effects, $B$ contains $N$ rows of length $q$, representing the $q$-dimensional random-effects vector, and $\beps$ shares its dimensions with $\BY$.

Now, define $K$ the projection matrix such that $K'K=I_{r-q}$, for an appropriate dimension $r$, and $ZK=0$. Then, set $P=QK$ and consider the projection model:
$$\BY_1\equiv YK=A\bfbeta_2P+\beps K.$$
The variance of a cluster is $\sigma^2I_{r-q}$. Next, define $H$ such that $H'H=I_q$ and $QH=0$. A second projection model emerges:
$$\BY_2\equiv YH=A\bfbeta_1+B+\beps H.$$
The variance of a cluster is $\sigma^2I_q+H'DH$, with $D$ the variance-covariance matrix of the vector of random effects. Importantly, projections $\BY_1\perp \BY_2$.

Conventional algebra leads from these to the following set of sufficient statistics:
\begin{eqnarray}
\label{suffstat1}
T_1&=&(A'A)^{-1}A'\BY_1P'(PP')^{-1},\\
T_2&=&\tr\left\{
\BY_1'\left[I-A(A'A)^{-1}A'\right]\BY_1
\right\},
\\
T_3&=&(A'A)^{-1}A'\BY_2,\\
T_4&=&\BY_2'\left[I-A(A'A)^{-1}A'\right]\BY_2.
\label{suffstat4}
\end{eqnarray}
Sufficient statistics (\ref{suffstat1})--(\ref{suffstat4}) lead to the maximum likelihood estimators:
\begin{eqnarray}
\label{est1}
\widehat{\bfbeta}_1&=&T_1,\\
\widehat{\bfbeta}_2&=&T_3,\\
\widehat{\sigma}^2&=&\frac{1}{N(n-q)}T_2,\\
\widehat{D}&=&\frac{1}{N}T_4-\widehat{\sigma}^2I_q.
\label{est4}
\end{eqnarray}
Note that the estimators for the fixed effects do not involve the variance components.

\setcounter{equation}{0}
\section{Algebraic derivations in the AR(1) case\label{ar1appendix}}

Here, we present more detailed derivations of the key algebraic expressions presented in Sections~\ref{secmodel} and \ref{secestimators}.

\subsection{Some useful expressions}\label{suppexp}

Consider,
\begin{equation}
\label{C}
C =
 \begin{pmatrix}
  1 & \rho & \rho^2 & \rho^3&\ldots \\
   & 1 & \rho & \rho^2 &\ldots \\
   &   & \ddots & \ddots &\ddots  \\
   & & & 1 & \rho\\
   &  & & & 1
 \end{pmatrix},
\end{equation}
then,
\begin{equation}
\label{suppSigma}
\Sigma= \sigma^2 C.
\end{equation}
It can be shown that:
\begin{equation}
\label{det_C}
\mathrm{det}(C)=\left(1-\rho^2 \right)^{n-1}
\end{equation}
The inverse of $C$ can be calculated as follows:
\begin{equation}
\label{inv_C}
C^{-1} =\frac{1}{1-\rho^2}
 \begin{pmatrix}
  1 & -\rho & \ldots & 0 \\
  -\rho & 1+\rho^2 & \ddots & \ldots \\
   &   \ddots & \ddots &\ddots  \\
   &  & 1+\rho^2& -\rho\\
   &   &-\rho& 1
 \end{pmatrix},
\end{equation}
as one may see $C^{-1}$ is a symmetric-tridiagonal matrix with constant diagonal except for the outer entries, and constant first off-diagonal.

Consider:
\begin{equation}
\label{G}
C^{-1}=\frac{1}{1-\rho^2} G.
\end{equation}
Then, by taking the derivative with respect to $\rho$:
\begin{equation}
\label{derivative_c_inv1}
\frac{\partial C^{-1}}{\partial \rho} = \frac{2\rho}{(1-\rho^2)^2}G + \frac{1}{1-\rho^2} H,
\end{equation}
where, $H=\frac{\partial G}{\partial \rho}$ and has the form:
\begin{equation}
\label{H}
H =
 \begin{pmatrix}
  0 & -1 & \ldots & 0 \\
  -1 & 2\rho & \ddots & \ldots \\
  \ddots &   \ddots & \ddots &\ddots  \\
   &\ddots  & 2\rho& -1\\
   0&   &-1& 0
 \end{pmatrix}.
\end{equation}
Also, considering the fact $CC^{-1}=I$, one can derive:
\begin{equation}
\label{derivative_c_inv2}
\frac{\partial C^{-1}}{\partial \rho} = -C^{-1} \frac{\partial C}{\partial \rho} C^{-1}.
\end{equation}

\subsection{The likelihood estimators in a given cluster}

The likelihood function for an $n_k$-dimensional multivariate normal sample of size $c_k$ has the following form:
\begin{equation}
\label{likelihood_general}
L=\prod_{i=1}^{C_k} \frac{1}{|\Sigma|^{1/2} (2\pi)^{n_k/2}}\exp\left\{ -\frac{1}{2} (y_i-\mu_i )' \Sigma^{-1} (y_i-\mu_i) \right\}.
\end{equation}
Therefore, the non-constant terms of the log-likelihood are as follows:
\begin{equation}
\label{log_general}
\ell \propto -\frac{C_k}{2}\ln |\Sigma|. \frac{1}{2}\sum_{i=1}^{C_k} (y_i-\mu_i)' \Sigma^{-1} (y_i-\mu_i),
\end{equation}
which considering (\ref{det_C}) for AR(1):
\begin{equation}
\label{det_simga}
|\Sigma|=\left(\sigma^2 \right)^{n_k} \left(1-\rho^2 \right)^{n_k-1}.
\end{equation}
As a general case, if we consider the mean as linear model with the form $\mu_i=X_i\beta$, one can derive:
\begin{equation}
\label{betahat}
\frac{\partial \ell}{\partial \mu_i}=\Sigma^{-1} \sum_{i=1}^{C_k} (y_i-\mu_i) =0 \Rightarrow \widehat{\beta}=(X'X)^{-1} X'y.
\end{equation}
Now expanding the log-likelihood for $\sigma^2$ and $\rho$, we have:
\begin{equation}
\label{log_likelihood_variance_comp}
\ell \propto -\frac{C_k}{2}n_k \ln\sigma^2 - \frac{C_k}{2} (n_k-1) \ln (1-\rho^2) -\frac{1}{2}\sum_{i=1}^{C_k} (y_i-\mu_i)' \Sigma^{-1} (y_i-\mu_i).
\end{equation}
Considering $\Sigma=\sigma^2 C$ and (\ref{inv_C}), the derivative with respect to $\sigma^2$ is as follows:
\begin{equation}
\label{derivative_sigma_2}
\frac{\partial \ell}{\partial \sigma^2} = -\frac{C_kn_k}{2} \frac{1}{\sigma^2} +\frac{1}{2}\frac{1}{(\sigma^2)^2}\sum_{i=1}^{C_k} (y_i-\mu_i)' C^{-1} (y_i-\mu_i).
\end{equation}
Solving
$\frac{\partial \ell}{\partial \sigma^2} =0$ gives:
\begin{equation}
\label{sigma_2_hat1}
\widehat{\sigma}^2=\frac{1}{C_kn_k}\sum_{i=1}^{C_k} (y_i-\mu_i)' C^{-1} (y_i-\mu_i).
\end{equation}
One may notice that $C^{-1}$ contains the parameter $\rho$.

Taking the derivative of (\ref{log_likelihood_variance_comp}) with respect to $\rho$ gives:
\begin{equation}
\label{derivative_rho}
\frac{\partial \ell}{\partial \rho} =\frac{C_k (n_k-1)}{2}\frac{2\rho}{1-\rho^2} -\frac{1}{\sigma^2}\sum_{i=1}^{C_k} (y_i-\mu_i)' \frac{\partial C^{-1}}{\partial \rho} (y_i-\mu_i).
\end{equation}
Setting $\frac{\partial \ell}{\partial \rho} =0$ gives:
\begin{equation}
\label{rho_hat1}
\widehat{\sigma}^2 \frac{2\widehat{\rho}}{1-\widehat{\rho}^2} = \frac{1}{C_k (n_k-1)} \sum_{i=1}^{C_k} (y_i-\mu_i)' \frac{\partial C^{-1}}{\partial \rho}  (y_i-\mu_i).
\end{equation}
Solving (\ref{sigma_2_hat1}) and (\ref{rho_hat1}) gives $\widehat{\sigma}^2$ and $\widehat{\rho}$. For any ($n_k\times n_k$) matrix $Q$,  $\sum_i (y_i- \mu_i)' Q (y_i-\mu_i)$ 
equals $\mathrm{tr}\left\{ S Q \right\}$,
where $\mathrm{tr}$ denotes the trace of a matrix, and $S=\sum_i (y_i-\mu_i)(y_i-m_i)'$. 
Hence, from (\ref{sigma_2_hat1}), (\ref{rho_hat1}), (\ref{G}), (\ref{derivative_c_inv1}), and (\ref{derivative_c_inv2}), one can write:
\begin{equation}
\label{two_equations}
\begin{cases}
(1-\widehat{\rho}^2) \widehat{\sigma}^2 = \frac{1}{C_k n_k} \mathrm{tr}\{ SG\},\\
(1-\widehat{\rho}^2)\widehat{\sigma}^2 = \frac{1}{C_k n_k}\mathrm{tr}\{SG\} + \frac{1-\widehat{\rho}^2}{2\widehat{\rho}}\frac{1}{C_k(n_k-1)} \frac{1}{C_k(n_k-1} \mathrm{tr}\{SH\}.
\end{cases}
\end{equation}
Set $g=\mathrm{tr}\{SG\}$ and $h=\mathrm{tr}\{SH\}$, it follows that
\begin{equation}
\label{finding_rho}
\frac{g}{n_k} + \frac{1-\widehat{\rho}^2}{2\widehat{\rho}}h =0.
\end{equation}
Given that both $g$ and $h$ are functions of $\rho$ only, $\rho$ can be estimated using ($\ref{finding_rho}$). Given$\rho$, one can use one of equations in (\ref{two_equations}) to estimate $\sigma^2$.

Let us consider some special cases. For $n_k=2$:
\begin{equation*}
\label{n_2_1}
G=
 \begin{pmatrix}
  1 & -\rho\\
 - \rho & 1
 \end{pmatrix},\qquad H=
  \begin{pmatrix}
  0 & -1\\
  -1 & 0
 \end{pmatrix}.
\end{equation*}
Therefore, $g$ and $h$ can be computed as:
\begin{equation*}
g=\mathrm{tr}\left[ \begin{pmatrix}
S_{11} & S_{12} \\
S_{21} & S_{22}
\end{pmatrix}\begin{pmatrix}
1 & -\rho \\
-\rho & 1
\end{pmatrix} \right ] = S_{11} - 2\rho S_{12} + S_{22}.
\end{equation*}
\begin{equation*}
h=\mathrm{tr}\left[ \begin{pmatrix}
S_{11} & S_{12} \\
S_{21} & S_{22}
\end{pmatrix}\begin{pmatrix}
0 & -1 \\
-1 & 0
\end{pmatrix} \right ] = -2S_{12}.
\end{equation*}
Now using (\ref{finding_rho}):
$$\widehat{\rho}(S_{11}- 2\widehat{\rho}S_{12}+S{22}+(1-\widehat{\rho}^2)(-2S_{12}),$$
which gives:
\begin{equation}
\label{n2_hat_rho}
\widehat{\rho}=\frac{2S_{12}}{S_{11}+S_{22}}.
\end{equation}
Then, using first equation in (\ref{two_equations}):
$$(1-\widehat{\rho}^2)\widehat{\sigma}^2 = \frac{1}{2C_k}(S_{11}-2\widehat{\rho}S_{12} + S_{22}),$$
which gives:
\begin{equation}
\label{n2_hat_sigma2}
\widehat{\sigma}^2=\frac{S_{11}+S{22}}{2C_k}.
\end{equation}
For $n_k=3$:
\[
\begin{aligned}
g&=&\mathrm{tr}\left[ \begin{pmatrix}
S_{11} & S_{12} & S_{13} \\
S_{21} & S_{22} & S_{23}\\
S_{31} & S_{32} & S_{33}
\end{pmatrix}\begin{pmatrix}
1 & -\rho & 0\\
-\rho & 1+\rho^2 -\rho\\
0 & -\rho & 1
\end{pmatrix} \right ] \\ &=& S_{11}+S_{22}+S_{33} -2\rho (S_{12} +S_{23})  + \rho^2 S_{22},\\
h&=&\mathrm{tr}\left[ \begin{pmatrix}
S_{11} & S_{12} & S_{13} \\
S_{21} & S_{22} & S_{23}\\
S_{31} & S_{32} & S_{33}
\end{pmatrix}\begin{pmatrix}
0 & -1 &0 \\
-1 & 2\rho & -1\\
0 & -1 & 0
\end{pmatrix} \right ]\\ &=& -2 (S_{12} +S_{23}) + 2 \rho S_{22}.
\end{aligned}
\]
Let,
\begin{equation*}
\begin{cases}
S=S_{11}+S{22}+S{33}\\
R=S_{12}+S_{23}
\end{cases} \Rightarrow \begin{cases}
g=S+\rho^2 S_{22} - 2\rho R\\
h=-2R + 2\rho S_{22}
\end{cases}
\end{equation*}
Using (\ref{finding_rho}):
\begin{equation}
\label{n3}
2S_{22}\rho^3 - R\rho^2 - (S+3 S_{22})\rho + 3R =0.
\end{equation}
Considering the results for $n_k=2$ and $n_k=3$, one can calculate (\ref{finding_rho}) for the general case $n_k=n$ as follows.
\begin{equation}
\label{nk_n_rho}
(n-1) \tilde{S} \rho^3 - (n-2) R \rho^2 - (n\tilde{S}+S)\rho + nR=0
\end{equation}
with:
\begin{equation*}
\begin{cases}
S=S_{11}+\ldots+S_{nn},\\
\tilde{S}=S_{22}+\ldots+S_{n-1,n-1},\\
R=S_{12}+S_{23}+\ldots+S_{n-1,n}.
\end{cases}
\end{equation*}
Then using (\ref{two_equations}):
\begin{equation}
\label{nk_n_sigma2}
\widehat{\sigma}^2=\frac{1}{C_n}\frac{1}{(1-\widehat{\rho}^2)} (S+\widehat{\rho}^2 \tilde{S} - 2\widehat{\rho} R).
\end{equation}
For $n_k>2$,  (\ref{nk_n_rho}) is a third-degree polynomial. One can show that  this equation has only one root in $[-1,1]$.

\begin{proof}
Consider:
\begin{equation*}
\begin{aligned}
 f(\rho) &= (n-1)\tilde{S}\rho^3 - (n-2) R \rho^2 - (n\tilde{S}+S)\rho + nR\\
 f'(\rho) &= 3(n-1)\tilde{S}\rho^2 - 2(n-2) R\rho - (n\tilde{S}+S)\\
  f''(\rho) &= 6(n-1)\tilde{S}\rho - 2(n-2) R
\end{aligned}
\end{equation*}
The discriminant of $f'(\rho)$ is as follows:
\begin{equation*}
\Delta_{f'(\rho)}=(n-2)^2 R^2 + 3 (n\tilde{S}+S)(n-1) \tilde{S}\geq 0.
\end{equation*}
Therefore $f'(\rho)$ has no root and hence $f(\rho)$ is monotone. One may see $f'(0)\leq 0$, therefore, $f(\rho)$ is a monotonically decreasing function (I). One can show $f(1)\leq 0$ and $f(-1)\geq 0$ (II). Considering (I) and (II) together, one may conclude $f(\rho)$ must necessarily cross the horizontal line only once between $[-1,1]$.
\end{proof}
This shows the unique $\widehat{\rho}$ can be easily estimated solving (\ref{nk_n_rho})  using Cardano's formula \citep{Cardano}.

\subsection{Hessians, covariance matrices, and optimal weights}

Given the MLEs for the AR(1) covariance structure,  the Hessians and covariance matrices of the MLEs can be derived. Following the general results obtained about optimal weights, they can be used to compute the exact optimal weights in the case of the AR(1) structure. As mean and variance parameters are orthogonal in the normal case, we can consider the second derivative for fixed effects and variance components separately.

\subsubsection{Second derivative with respect to fixed effects}
As $$\frac{\partial \ell}{\partial \beta}=\sum_{i=1}^{C_k} X'_i\ \Sigma^{-1} (y_i-\mu_i),$$
we have:
\begin{equation*}
\begin{aligned}
\mathrm{E}\left[\frac{\partial \ell}{\partial \beta} \left( \frac{\partial \ell}{\partial \beta}\right)' \right] & = \sum_{i=1}^{C_k} X'_i \Sigma^{-1} \mathrm{E}(y_i-\mu_i)(y_i-\mu_i)' \Sigma^{-1}X_i \\
&= \sum_{i=1}^{C_k} X'_i \Sigma^{-1} X_i.
\end{aligned}
\end{equation*}
For the special case of just an intercept $X_i=\mathbf{1}$:
\begin{multline}
\label{var_mu}
\mathrm{E}\left[\frac{\partial \ell}{\partial \beta} \left( \frac{\partial \ell}{\partial \beta}\right)' \right]= \sum_{i=1}^{C_k} \mathbf{1}' \Sigma^{-1} \mathbf{1}= \\ \frac{C_k}{\sigma^2 (1-\rho^2)}\left[(n_k-2)\rho^2 - 2(n_k-1)\rho +n_k \right].
\end{multline}
Therefore, the variance for $\widehat{\mu}$ can be computed as inverse of (\ref{var_mu}).
\subsubsection{Second derivative with respect to variance components}
To calculate the derivatives with respect to variance components rather than
$\frac{\partial C^{-1}}{\partial \rho}$, we need $K=\frac{\partial C^{-1}}{\partial \rho^2}$. Using these derivatives:
\begin{equation}
\label{useful_deriv}
\begin{cases}
\frac{\partial}{\partial \rho} 2\left(\frac{\rho}{1-\rho^2} \right)= 2\frac{1+\rho^2}{(1-\rho^2)^2},\\
\frac{\partial}{\partial \rho} \frac{\rho}{(1-\rho^2)^2}=\frac{1+3\rho^2}{(1-\rho^2)^3},\\
\frac{\partial}{\partial \rho} \frac{1+\rho^2}{(1-\rho^2)^2} = \frac{2\rho(3+\rho^2)}{(1-\rho^2)^3}.
\end{cases}
\end{equation}
it follows that
\begin{multline}
\label{K}
\frac{\partial C^{-1}}{\partial \rho^2} = K =  \frac{1}{(1-\rho^2)^3} \\
\begin{pmatrix}
2(1+3\rho^2) & -2\rho(3+\rho^2) &   & 0\\
-2\rho(3+\rho^2) & 4(1+3\rho^2) & \ddots &  \ddots\\
 & \ddots & \ddots  & \ddots\\
 0&  &  -2\rho(3+\rho^2) & 2(1+3\rho^2).
\end{pmatrix}
\end{multline}
The second-derivatives are:
\begin{equation}
\label{second_deriv_variance_comp}
\begin{cases}
\frac{\partial^2 \ell}{\partial (\sigma^2)^2} = \frac{C_kn_k}{2}\frac{1}{(\sigma^2)^2}- \frac{1}{(\sigma^2)^3}\sum_{i=1}^{C_k} (y_i-\mu_i)' C^{-1} (y_i-\mu_i),\\
\frac{\partial^2 \ell}{\partial \rho^2} = \frac{C_k(n_k-1)(1+\rho^2)}{(1-\rho^2)^2}-\frac{1}{2\sigma^2}\sum_{i=1}^{C_k} (y_i-\mu_i)' K (y_i-\mu_i),\\
\frac{\partial^2 \ell}{\partial \rho \partial \sigma^2} =\frac{1}{2(\sigma^2)^2}\sum_{i=1}^{C_k} (y_i-\mu_i)'\frac{\partial C^{-1}}{\partial \rho} (y_i-\mu_i).
\end{cases}
\end{equation}
To construct the expected Hessian and covariance matrix, one needs to find the expectations of the expressions in (\ref{second_deriv_variance_comp}).
\begin{equation}
\label{expectation_second_deriv_seigma2}
\mathrm{E}\left(\frac{\partial^2 \ell}{\partial (\sigma^2)^2} \right)= -\frac{C_kn_k}{2}\frac{1}{(\sigma^2)^2}.
\end{equation}
This follows from the fact that: $$\mathrm{E}\left(\sum_{i=1}^ {C_k}  (y_i-\mu_i)' C^{-1} (y_i-\mu_i) \right)=C_k \mathrm{tr}\left\{\mathrm{E}\left[(y_i-\mu_i)'(y_i-\mu_i) \right ] C^{-1}\right\},$$ and $\mathrm{E}\left[(y_i-\mu_i)(y_i-\mu_i)'\right]=\sigma^2 C$.

For the second derivative with respect to $\rho$:
\begin{equation}
\label{expectation_second_deriv_rho1}
\mathrm{E}\left[\frac{\partial^2\ell}{\partial \rho^2} \right]=\frac{C_k(n_k-1)(1+\rho^2)}{(1-\rho^2)^2}-\frac{C_k}{2}\mathrm{tr}\{KS\}.
\end{equation}
Likewise:
\begin{equation}
\label{expectation_deriv_rho_sigma2_1}
\mathrm{E}\left[\frac{\partial^2 \ell}{\partial \rho \partial \sigma^2} \right]=\frac{C_k}{2\sigma^2}\mathrm{tr}\left\{C \frac{\partial C^{-1}}{\partial \rho} \right\}.
\end{equation}
Substituting for $\mathrm{tr}\{KS\}$ and $\mathrm{tr}\left\{C \frac{\partial C^{-1}}{\partial \rho} \right\}$ we get:
\begin{equation}
\label{expectation_deriv_rho_sigma2}
\mathrm{E}\left[\frac{\partial^2 \ell}{\partial \rho \partial \sigma^2} \right]=\frac{C_k(n_k-1)}{\sigma^2}\frac{\rho}{1-\rho^2}.
\end{equation}
\begin{equation}
\label{expectation_second_deriv_rho}
\mathrm{E}\left[\frac{\partial^2\ell}{\partial \rho^2} \right]=-C_k(n_k-1)\frac{1+\rho^2}{(1-\rho^2)^2}.
\end{equation}
Using (\ref{expectation_second_deriv_seigma2}), (\ref{expectation_deriv_rho_sigma2}), and (\ref{expectation_second_deriv_rho}) one obtains the $2\times 2$ Hessian matrix as follows:
\begin{equation}
\label{Hessian}
H=-C_k
\begin{pmatrix}
\frac{n_k}{2(\sigma^2)^2} & -\frac{n_k-1}{\sigma^2} \frac{\rho}{1-\rho^2}\\
-\frac{n_k-1}{\sigma^2} \frac{\rho}{1-\rho^2} & (n_k-1)\frac{1+\rho^2}{(1-\rho^2)^2}
\end{pmatrix}.
\end{equation}
The determinant of the Hessian in (\ref{Hessian}) is as follows:
\begin{equation}
\label{det_hessian}
\mathrm{det}(H)=\frac{C_k^2(n_k-1)(n_k-(n_k-2)\rho^2)}{2(\sigma^2)^2 (1-\rho^2)^2}.
\end{equation}
So,
\begin{multline}
\label{Covariance_biased}
-H^{-1}=\\ \frac{1}{C_k(n_k-(n_k-2)\rho^2)} \begin{pmatrix}
2(\sigma^2)^2 (1+\rho^2) & 2\rho \sigma^2 (1-\rho^2)\\
2\rho \sigma^2 (1-\rho^2) & \frac{n_k}{n_k-1} (1-\rho^2)^2
\end{pmatrix}.
\end{multline}
The Hessian for the unbiased estimator differs slightly from its MLE counterpart:
\begin{eqnarray}
\label{tilde_H}
\tilde{H}&=& -C_k \begin{pmatrix}
\frac{n_k-1}{2(\sigma^2)^2} & -\frac{n_k-1}{\sigma^2} \frac{\rho}{1-\rho^2}\\
-\frac{n_k-1}{\sigma^2} \frac{\rho}{1-\rho^2} & (n_k-1)\frac{1+\rho^2}{(1-\rho^2)^2}
\end{pmatrix},\\
\label{det_tilde_H}
\mathrm{det}(\tilde{H})&=&\frac{C_k^2 (n_k-1)^2}{2(\sigma^2)^2 (1-\rho^2)}.
\end{eqnarray}
Therefore,
\begin{multline}
\label{covariance_unbiased}
-\tilde{H}^{-1}=\\ \frac{1}{C_k(n_k-(n_k-2)\rho^2)} \begin{pmatrix}
2(\sigma^2)^2 (1+\rho^2) & 2\rho \sigma^2 (1-\rho^2)\\
2\rho \sigma^2 (1-\rho^2) &  (1-\rho^2)^2
\end{pmatrix}.
\end{multline}
Having the covariance matrix, one may easily find the optimal weights using
\begin{equation}
\label{optimal_weights}
W_{opt.}=\frac{V^{-1}_k}{\sum_{i=1}^{K} V^{-1}_i}
\end{equation}
The variance of an estimator obtained using the optimal weights in (\ref{optimal_weights}) can be calculated as $\left(\sum_{i=1}^{K} V^{-1}_i\right)^{-1}$.

\subsection{Proof of Proposition~\ref{prop1}\label{suppproof}}

\newcommand{\mua}{\widetilde{\mu}_\alpha}

{\bfseries Proof.} Consider an estimator of the form:
\begin{equation}
\widetilde{\mu}_\alpha=\frac{1}{c}\sum_{i=1}^c\sum_{j=1}^n\alpha_jY_{ij},
\end{equation}
for a vector of weights $\bfalpha=(\alpha_1,\dots,\alpha_n)'$. Because the clusters are i.i.d.\ it is evident that the components of $\bfalpha$ do not depend on the cluster index~$i$. Clearly, the requirement that $E(\mua)=\mu$ implies the condition
\begin{equation}
\label{alphasum}
\sum_{j=1}^n\alpha_j=1.
\end{equation}
 An expression of the variance of $\mua$ combined with this requirement produces the objective function:
\begin{equation}
\label{objfunc1}
Q=\sigma^2\left(
\sum_{j=1}^n\alpha_j^2+2\sum_{j<k}\alpha_j\alpha_k\rho^{|j-k|}
\right)
-\lambda\left(
\sum_{j=1}^n\alpha_j-1
\right),
\end{equation}
with $\lambda$ a Lagrange multiplier. Taking the derivative of (\ref{objfunc1}) w.r.t.\ $\bfalpha$ leads to, after rearrangement:
$$\bfalpha=\frac{\lambda}{2\sigma^2}C^{-1}\eenvec.$$
Given that we have an explicit form for $C^{-1}$, it follows that
\begin{equation}
\label{alphatemp}
\bfalpha=\frac{\lambda}{2\sigma^2(1+\rho)}\bfrho^{(1)},
\end{equation}
with $\bfrho^{(1)}=(1,1-\rho,\dots,1-\rho,1)'$.
Combining (\ref{alphatemp}) with constraint (\ref{alphasum}) leads to
$\lambda=2\sigma^2(1+\rho)/[2+(n-2)(1-\rho)]$, hence
$$\alpha=
\frac{1}{[2+(n-2)(1-\rho)]}\bfrho^{(1)},
$$
establishing the MLE. This completes the proof.

\subsection{Optimal weights in case of a general mean structure $X_i^{(k)}\bfbeta$}\label{suppgenmean}

Cluster size specific expressions are:
\begin{equation}
\widehat{\bfbeta_k}= \left( \sum_{i=1}^{c_k} X_{i}^{(k)'}\Sigma_k^{-1}X_{i}^{(k)}\right)^{-1} \left( \sum_{i=1}^{c_k} X_{i}^{(k)'}\Sigma_k^{-1}Y_{i}^{(k)}\right) \label{betagenmean}
\end{equation}
and
\begin{equation}
\mbox{var}(\widehat{\bfbeta_k})=V_k=\left( \sum_{i=1}^{c_k} X_{i}^{(k)'}\Sigma_k^{-1}X_{i}^{(k)}\right)^{-1}.
\end{equation}
The combination ruleis 
\begin{equation}
\tilde{\bfbeta_k}=\sum_{i=1}^K A_k \widehat{\bfbeta_k},
\end{equation}
with
\begin{equation}
V_k^{-1}=\frac{1}{\sigma^2}\sum_{i=1}^{c_k} X_{i}^{(k)'}C_k^{-1}X_{i}^{(k)}
\end{equation}
and $C_k$ is as described in Supplementary Materials~\ref{suppexp}.

The first factor in (\ref{betagenmean}) can be split into three parts:
$$
\begin{aligned}
(1-\rho^2) X_{i}^{(k)'}C_k^{-1}X_{i}^{(k)} &= X_{i}^{(k)'} (1+\rho^2)I_k X_{i}^{(k)} \label{betaffft}\\
&-\rho^2 X_{i}^{(k)'} \begin{pmatrix}
1 & 0 & \hdots & \hdots & 0 \\
0 & 0 & & & \vdots \\
\vdots & & \ddots & & \\
\vdots & & & 0 & 0 \\
0 & \hdots & \hdots & 0 & 1\\
\end{pmatrix} X_{i}^{(k)} \label{betaffst} \\
&-\rho X_{i}^{(k)'} \begin{pmatrix}
0 & 1 & 0 & \hdots & 0 \\
1 & 0 & \ddots & & \vdots \\
0 & \ddots & \ddots & \ddots & 0 \\
\vdots & & \ddots & 0 & 1 \\
0 & \hdots & 0 & 1 & 0 \\
\end{pmatrix} X_{i}^{(k)}.\label{betafftt}
\end{aligned}
$$

(\ref{betaffft}) simplifies to $(1+\rho^2)X_{i}^{(k)'}X_{i}^{(k)}$, while (\ref{betaffst}) equals
\begin{equation}
\scalemath{0.8}{
\rho^2 \begin{pmatrix}
x_{ki11}^2+x_{kin_k1}^2 & x_{ki11}\cdot x_{ki22}+x_{kin_k1}\cdot x_{kin_k2} & \hdots & x_{ki11}\cdot x_{ki1p} + x_{kin_k1}\cdot x_{kin_kp} \\
 & x_{ki12}^2 + x_{kin_k2}^2 & & \\
 & & \ddots & \\
  & & & x_{ki1p}^2 + x_{kin_kp}^2.
\end{pmatrix}.
}
\end{equation}
Defining $\boldsymbol{X}_{i1}^{(k)}=(x_{ki11} \ \hdots \ x_{ki1p})^t$ and $\boldsymbol{X}_{in_k}^{(k)}=(x_{kin_k1} \ \hdots \ x_{kin_kp})^t$, (\ref{betaffst}) equals
\begin{equation}
\rho^2[\boldsymbol{X}_{i1}^{(k)} \ 0 \hdots 0 \ \boldsymbol{X}_{in_k}^{(k)}] X_i^{(k)}.
\end{equation}
For the third term, define $\boldsymbol{X}_{ij}^{(k)-}=(x_{ki2j} \ \hdots \ x_{kin_kj})^t$ and $\boldsymbol{X}_{ij}^{(k)+}=(x_{ki1j} \ \hdots \ x_{kin_{k-1}j})^t$. As a consequence, (\ref{betafftt}) will equal
\begin{equation}
-\rho [X_i^{(k)-'} \cdot X_i^{(k)+}+X_i^{(k)+'} \cdot X_i^{(k)-}].
\end{equation}
In summary:
\begin{equation}
\begin{aligned}
(1-\rho^2) X_{i}^{(k)'}C_k^{-1}X_{i}^{(k)} &= (1+\rho^2)X_{i}^{(k)'}X_{i}^{(k)} \\
&-\rho^2[\boldsymbol{X}_{i1}^{(k)} \ 0 \hdots 0 \ \boldsymbol{X}_{in_k}^{(k)}] X_i^{(k)} \\
&-\rho [X_i^{(k)-'} \cdot X_i^{(k)+}+X_i^{(k)+'} \cdot X_i^{(k)-}] \\
&\overset{\text{notation}}= F_{1k}. \label{betagenmeanff}
\end{aligned}
\end{equation}
The second factor in (\ref{betagenmean}), using the same notations for $Y_i^{(k)}$ as described above, can be rewritten as:
\begin{equation}
\begin{aligned}
(1-\rho^2) X_{i}^{(k)'}C_k^{-1}Y_{i}^{(k)} =& (1+\rho^2)X_{i}^{(k)'}Y_{i}^{(k)} \\
&- \rho^2 \begin{pmatrix}
x_{ki11}\cdot y_{ki1} + x_{kin_k1}\cdot y_{kin_k}  \\
x_{ki12}\cdot y_{ki1} + x_{kin_k2}\cdot y_{kin_k}  \\
\vdots \\
x_{ki1p}\cdot y_{ki1} + x_{kin_kp}\cdot y_{kin_k} \\
\end{pmatrix}  \\
&-\rho [X_i^{(k)-'} \cdot Y_i^{(k)+}+X_i^{(k)+'} \cdot Y_i^{(k)-}]  \\
&\overset{\text{not.}}= F_{2k}.\label{betagenmeansf}
\end{aligned}
\end{equation}
Combining (\ref{betagenmeanff}) and (\ref{betagenmeansf}) the overall estimate equals:
\begin{equation}
\begin{aligned}
\tilde{\bfbeta_k}&=\sum_{i=1}^K A_k \widehat{\bfbeta_k} \\
&=\sum_{i=1}^K \left( \sum_{m=1}^K F_{1m} \right)^{-1} F_{2k}
\end{aligned}
\end{equation}
\subsection{Delta method for the mean estimator}\label{suppdelta}

By plugging in $\rho_k$ and defining $a_k'=c_k[n_k-(n_k-2)\rho_k]$, equation~(\ref{meanweight}) simplifies to
\begin{equation}
\label{meanweight}
a_k=
\frac
{
a_k'
}
{
\sum_{m=1}^K a_m'
}
,
\end{equation}
and (\ref{varmuk})
becomes
\begin{equation}
\mbox{var}(\widehat{\mu}_k)=v_k=\frac{\sigma_k^2(1+\rho_k)}{a_k'}.
\end{equation}
The first derivatives equal
\begin{equation}
\begin{aligned}
\frac{\partial \tilde{\mu}}{\partial\mu_k} &= a_k = \frac{a_k'}{\sum_{m=1}^K a_m'},  \\
\frac{\partial\tilde{\mu}}{\partial\sigma_k^2}&=0,  \\
\frac{\partial\tilde{\mu}}{\partial\rho_k}&=\frac{-c_k(n_k-2)\sum_{m=1}^Ka_m'(\mu_k-\mu_m)}{\left( \sum_{m=1}^K a_m' \right)^2},
\end{aligned}
\end{equation}
and these can be combined using the delta method, resulting in (\ref{deltavareen}):
\[
\begin{aligned}
\mbox{var}(\tilde{\mu})=& \sum_{i=1}^K \frac{a_k^{'2}}{\left( \sum_{k=1}^K a_k' \right)^2} \cdot \frac{\sigma_k^2(1-\rho_k^2)}{a_k'}\\
& + \frac{\sum_{k=1}^K\left[ c_k(n_k-2)\sum_{m=1}^Ka_m'(\mu_k-\mu_m)\right]^2}{\left( \sum_{k=1}^K a_k' \right)^4}\cdot \frac{1-\rho_k^2}{c_k(n_k-1)}.
\end{aligned}
\]

\subsection{Calculating $\widehat{\rho}$ and $\widehat{\sigma}^2$ in R}

In this section, we consider the implementation of the calculations for the variance components via the R packahe {\tt{fastAR1}}. This can be done with a few simple lines of code. For fixed $C_k=C$ and $n_k=n$, and given the data $y$, the function {\tt{est.ar1}} estimates the variance components and provides a plot for the third-degree polynomial in (\ref{nk_n_rho}). This visually underscores that there is only one root in $[-1,1]$. Figure \ref{3rd_poly} shows (\ref{nk_n_rho}) for 10 simulated data sets;  clearly, there is a single root only in $[-1,1]$. For convenience, the R code can be obtained via R package {\tt{fastAR1}}, see Section~\ref{fastar1}. Other functions to find variances and iterated optimal weights are also available.
\begin{figure}[t]
\centering
\includegraphics[width=\textwidth]{3rd_poly.eps}
\caption{The third degree polynomial in (\ref{nk_n_rho}) for $10$ different generated data. The red veertical line shows $\widehat{\rho}$.} \label{3rd_poly}
\end{figure}

\setcounter{equation}{0}
\section{Details on additional simulations\label{suppsimul}}

\subsection{Simulations with proportional and size-proportional weights}

Here we consider $C_1=500,  C_2=250 ,C_3=250, C_4= 500$, and $n_1=5,n_2= 10,n_3= 10,n_4=  5$. Parameters are set to $\mu=0$, $\sigma=2$ and $\rho=(0.1,0.5,0.8)$. The data are generated 100 times and the model is fitted using PROC MIXED in SAS (for a single overall intercept). For combining the results from different sub-samples we have used proportional weights and size-proportional weights:
\begin{equation}
\label{weights}
\begin{cases}
\mbox{Prop}=\frac{C_k}{\sum_l C_l}, \\
\mbox{Size.Prop}=\frac{C_kn_k}{\sum_l C_l n_l}.
\end{cases}.
\end{equation}
The results are compared with full likelihood (Table~\ref{table_comp1}). In contrast to the compound-symmetry case, the size-proportional weights show much better results than the proportional weights. Furthermore, the size-proportional weights in the current simulation are identical with the equal weights. The $n_k$'s have a much larger influence in the AR(1) case compared to CS. 
 Figures~\ref{fig_rho1}, \ref{fig_rho2}, and \ref{fig_rho3} make the comparisons easier.
\begin{table}[t]
\centering
\caption{Comparing proportional, size-proportional and iterated optimal weights with full likelihood for AR(1) covariance structure.}
\label{table_comp1}
 \resizebox{\textwidth}{!}{%
\begin{tabular}{clcccccc}
  \hline\hline
 &&$\widehat{\mu}$ & Sd & $\widehat{\rho}$ & Sd & $\widehat{\sigma}^2$ & Sd \\
 \hline
& Prop. & -0.00190 & 0.01615 & 0.10027 & 0.01158 & 1.99642 & 0.03002 \\
  $\rho=0.1$ & Size.Prop.& -0.00207 & 0.01538 & 0.10024 & 0.01080 & 1.99793 & 0.02853 \\
&It.Opt.  & -0.00206 & 0.01538 & 0.10024 & 0.00993 & 1.99792 & 0.02850\\
&ML   & -0.00207 & 0.01538 & 0.10032 & 0.01078 & 1.99793 & 0.02850 \\
  \hline
& Prop.    & -0.00212 & 0.02221 & 0.49966 & 0.00954 & 2.00349 & 0.03652 \\
  $\rho=0.5$ & Size.Prop& -0.00155 & 0.02156 & 0.49955 & 0.00898 & 2.00257 & 0.03494 \\
  &It.Opt.&  -0.00168 & 0.02149  &0.49956 & 0.00826 & 2.00265 & 0.03486\\
 &ML  & -0.00170 & 0.02150 & 0.49986 & 0.00896 & 2.00259 & 0.03488 \\
    \hline
  & Prop. & 0.00195 & 0.02890 & 0.79923 & 0.00549 & 1.99529 & 0.04989 \\
 $\rho=0.8$ &Size.Prop& 0.00234 & 0.02904 & 0.79911 & 0.00530 & 1.99542 & 0.04907 \\
  &It.Opt.& 0.00213 &0.02855& 0.79915& 0.00486&1.99538& 0.04859\\
&ML   & 0.00212 & 0.02855 & 0.79937 & 0.00527 & 1.99519 & 0.04861 \\
   \hline\hline
\end{tabular}}
\end{table}

\begin{figure}[ht!]
\centering
\includegraphics[width=\textwidth]{rho1.eps}
\caption{Comparing proportional and size-proportional weights with full likelihood for $100$ replications with $\mu=0$, $\sigma^2=2$ and $\rho=0.1$.} \label{fig_rho1}
\end{figure}

\begin{figure}[ht!]
\centering
\includegraphics[width=\textwidth]{rho2.eps}
\caption{Comparing proportional and size-proportional weights with full likelihood for $100$ replications with $\mu=0$, $\sigma^2=2$ and $\rho=0.5$.} \label{fig_rho2}
\end{figure}

\begin{figure}[ht!]
\centering
\includegraphics[width=\textwidth]{rho3.eps}
\caption{Comparing proportional and size-proportional weights with full likelihood for $100$ replications with $\mu=0$, $\sigma^2=2$ and $\rho=0.8$.} \label{fig_rho3}
\end{figure}

\subsection{Simulations with proportional and size-proportional weights: $\rho$ near 0/1}

We now present a comparison between proportional and size-proportional weights. We see that,  for $\rho$'s near $1$ (i.e., near CS), size-proportional weights are worse than proportional weights.

We consider $c_1=500$,  $c_2=250$, $c_3=250$, $c_4= 500$, and $n_1=5$, $n_2= 10$, $n_3= 10$, $n_4=  5$. Parameters are set as $\mu=0$, $\sigma=2$ and $\rho\in\{0.01,0.2,0.5,0.8,0.9,0.95,0.99\}$. The data are generated 100 times and the model is fitted using PROC MIXED in SAS (for a single overall intercept). For combining results from different sub-samples we have used proportional weights and size-proportional weights as in (\ref{weights}). The results are compared with full likelihood results.

In Figure~\ref{fig_rho1b}, for $\rho=0.99$ and $0.95$, the size-proportional weights perform  worse than the proportional weights. This is expected, because in this case AR(1) approaches CS. This result is clearer in the left panel of Figure~\ref{fig_rho1b}, where the standard deviations are shown. For $\rho$'s near $1$, the proportional weights are as efficient as  full likelihood, while as $\rho$ moves further from $1$ this would happen for size-proportional weights.

Figure~\ref{fig_rho2b} shows this phenomenon more clearly, as for some selected $\rho$'s ($0.01,0.5,0.95$) the density plot for all $100$ simulated datasets is plotted rather than a  boxplot. The size-proportional weights are better than proportional weights if $\rho$ is not very close to $1$. As soon as $\rho$ becomes $0.95$ or $0.99$, the size-proportional weights become worse.

\begin{figure}[ht!]
\centering
\includegraphics[angle=270]{comp_rho.eps}
\caption[Simulation study. Boxplots comparing proportional and size-proportional weights with full likelihood for $100$ replications with $\mu=0$, $\sigma^2=2$ and $\rho=0.99,0.95,0.9,0.8,0.5,0.2,0.01$]{Simulation study. Boxplots comparing proportional and size-proportional weights with full likelihood for $100$ replications with $\mu=0$, $\sigma^2=2$ and $\rho=0.99,0.95,0.9,0.8,0.5,0.2,0.01$. In every section of the boxplots (which are separated by dashed lines) the first out of three represents the proportional weights, the middle of is size-propotional weights and the one on the right shows the results for the full likelihood. The first row presents the estimates while the second row shows the standard deviations of these estimates.} \label{fig_rho1b}
\end{figure}
\begin{figure}[ht!]
\centering
\includegraphics[angle=270]{comp_rho2.eps}
\caption[Simulation study. Comparing proportional, size-proportional and full likelihood results via their empirical density for the 100 replications]{Simulation study. Comparing proportional, size-proportional and full likelihood results via their empirical density for the 100 replications. In all of the figures $\mu=0$ and $\sigma^2=2$. The first row is for $\rho=0.01$, the middle one is for $\rho=0.5$ and last one corresponds to $\rho=0.99$. In each figure the ticker dotted line corresponds to full likelihood, the dashed line is for size-proportional weights and the solid line is for proportional weights.} \label{fig_rho2b}
\end{figure}

\subsection{Simulations with optimal weights}

Given the  covariance matrix of the parameter estimators, finding the optimal weights is straightforward, but in practice the unknown parameters therein need to be estimated.  Here we compare the iterative weights with size-proportional weights and ML. See Figures \ref{fig_rho1_opt}, \ref{fig_rho2_opt}, \ref{fig_rho3_opt}, and Table \ref{table_comp1}.
As expected, the optimal weights lead to estimates very close to the MLE; the difference between them being numerical.

Size-proportional weights are used as the initial weights to begin the iterative procedure. One interesting outcome of this simulation is that the iterative procedure always converged after just one iteration. This means, iterated optimal weights are just like the approximated optimal weights, but there, instead of using $\widehat{\theta}_k$ from each sub-sample, one may use $\tilde{\theta}$ obtained from all sub-samples using a non-optimal but good weight.

\begin{figure}[ht!]
\centering
\includegraphics[width=\textwidth]{rho1_opt.eps}
\caption{Comparing iterated optimal and size-proportioanl weights with full likelihood for $100$ replications with $\mu=0$, $\sigma^2=2$ and $\rho=0.1$.} \label{fig_rho1_opt}
\end{figure}

\begin{figure}[ht!]
\centering
\includegraphics[width=\textwidth]{rho2_opt.eps}
\caption{Comparing iterated optimal and size-proportioanl weights with full likelihood for $100$ replications with $\mu=0$, $\sigma^2=2$ and $\rho=0.5$.} \label{fig_rho2_opt}
\end{figure}

\begin{figure}[ht!]
\centering
\includegraphics[width=\textwidth]{rho3_opt.eps}
\caption{Comparing iterated optimal and size-proportioanl weights with full likelihood for $100$ replications with $\mu=0$, $\sigma^2=2$ and $\rho=0.8$.} \label{fig_rho3_opt}
\end{figure}

\subsection{Simulations on computation  time}
Here, some summary tables are presented to summarize the results which are already presented via figures earlier. Furthermore, a table and a figure are added to compare computation time for closed form solutions to numerical ones.

In each table the mean of the estimated parameter and its standard deviation using the $100$ replications are given, together with the standard deviation of those 100 numbers (in parentheses). If $\theta$ is the parameter of interest, $\widehat{\theta}$ is its estimate and $\theta_0$ is its real value, then the MSE is computed as follows:
\begin{equation}
\label{MSE}
\mathrm{MSE}(\widehat{\theta})=\frac{1}{100}\sum_{i=1}^{100} (\widehat{\theta}-\theta_0)^2.
\end{equation}
Table~\ref{sim_mu} summarizes the results for $\mu$. The sample splitting estimates are computed using proportional and size-proportional (identical to equal weights in this case) weights. The results using the full sample are also given. The third column in Table~\ref{sim_mu} presents the averaged (over 100 replications) estimated $\mu$ and its standard deviation. The fourth column presents the averaged estimated standard deviation for $\widehat{\mu}$ (over 100 replications) and its standard deviation. The last column shows the MSE computed using (\ref{MSE}) for $\mu_0=1$. Tables~\ref{sim_rho} and \ref{sim_sigma2} shows the same results for $\rho$ and $\sigma^2$ ($\sigma^2_0=2$), respectively.

Table~\ref{comp_time} compares the computation time 
between closed-form and interative methods.  The closed-form solutions are implemented in R and for the numerical methods the MIXED procedure in SAS is used, with error covariance structure set to AR(1).

The data are generated using $n=10$ for all clusters, with $c$ is varying from 100 to 1000000. Therefore, the design is balanced and the point of this comparison is to see how the computation time is reduced in each split.

As one may see in Table~\ref{comp_time} and Figure~\ref{fig_comp_time}, using closed form solutions  significantly reduces the computation time. This means, as well as the computation time reduction due to splitting the data, using closed form solutions within each split the computation time reduction is also huge: for example, for a million clusters, the reduction is from  almost one hour to less than 5 seconds. Figure~\ref{fig_comp_time} shows that computation time using closed form solution changes linearly with the number of clusters, while this will be exponential using an iterative solution.

To assess the effect of the overall size of the dataset, the model is fitted to two concatenated copies of the same set. Computation time results are presented in Table~\ref{comp_time} and Figure~\ref{fig_comp_time}.  The data are generated with $\mu=0$, $\sigma^2=2$, and $\rho=0.25$.
\begin{table}[ht]
\centering
\caption[Simulation study. Estimating $\mu$ and its standard deviation]{Simulation study. Estimating $\mu$ and its standard deviation. The mean (standard deviation) of the $100$ replications are given together with mean squared errors for $\rho=0.01,0.2,0.5,0.8,0.9,0.99$ using proportional and size-proportional weights comparing with the full likelihood results.}
\label{sim_mu}

\vspace*{2mm}
\resizebox{\textwidth}{!} 
{
\begin{tabular}{cllcccc}
\hline\hline
$\rho_0$&method & mean($\widehat{\mu}$) (s.d.)  & mean(s.e.($\widehat{\mu}$)) (s.d.)  & MSE$\times 10^4$ \\
  \hline
  &Prop. & -0.00271 (0.01462) & 0.01503 (0.00020) & 2.19000 \\
0.01  &Size Prop. & -0.00277 (0.01320) & 0.01429 (0.00018) & 1.80172 \\
  &Full & -0.00275 (0.01319) & 0.01428 (0.00018) & 1.79828 \\ \hline
  &Prop. & 0.00158 (0.01677) & 0.01752 (0.00025) & 2.81056 \\
0.2   &Size Prop. & 0.00085 (0.01616) & 0.01673 (0.00021) & 2.59147 \\
  &Full & 0.00090 (0.01615) & 0.01672 (0.00021) & 2.58880 \\ \hline
  &Prop. & 0.00391 (0.02244) & 0.02217 (0.00037) & 5.13770 \\
0.5   &Size Prop. & 0.00397 (0.02191) & 0.02155 (0.00035) & 4.91201 \\
  &Full & 0.00396 (0.02182) & 0.02148 (0.00034) & 4.87038 \\ \hline
  &Prop. & 0.00130 (0.02790) & 0.02894 (0.00050) & 7.72450 \\
0.8   &Size Prop. & -0.00049 (0.02710) & 0.02912 (0.00045) & 7.27464 \\
  &Full & 0.00053 (0.02713) & 0.02862 (0.00045) & 7.29130 \\ \hline
  &Prop. & -0.00828 (0.03006) & 0.03221 (0.00056)&  9.63393 \\
0.9   &Size Prop. & -0.00727 (0.03145) & 0.03306 (0.00070) & 10.3224 \\
  &Full & -0.00803 (0.02998) & 0.03207 (0.00057) & 9.54258 \\ \hline
  &Prop. & 0.00162 (0.03663) & 0.03597 (0.00064) & 13.3123e \\
0.99   &Size Prop. & -0.00007 (0.03930) & 0.03797 (0.00088) & 15.2876 \\
  &Full & 0.00156 (0.03666) & 0.03597 (0.00064) &  13.3305 \\
   \hline
\hline
\end{tabular}}
\end{table}

\begin{table}[ht]
\centering
\caption[Simulation study. Estimating $\rho$ and its standard deviation]{Simulation study. Estimating $\rho$ and its standard deviation. The mean (standard deviation) of the $100$ replications are given together with mean squared errors for $\rho=0.01,0.2,0.5,0.8,0.9,0.99$ using proportional and size-proportional weights comparing with the full likelihood results.}
\label{sim_rho}

\vspace*{2mm}
 \resizebox{\textwidth}{!}{%
\begin{tabular}{lllcccc}
\hline\hline
$\rho_0$&method & mean($\widehat{\rho}$) (s.e.)  & mean(s.e.($\widehat{\rho}$)) (s.e.)  & MSE \\
  \hline
  &Prop. & 0.01077 (0.01237) & 0.01165 (0.00006) &  1.52178e-04 \\
  0.01&Size Prop. & 0.01115 (0.01200) & 0.01087 (0.00004) & 1.43974e-04 \\
  &Full & 0.01123 (0.01203) & 0.01084 (0.00004) & 1.44675e-04 \\  \hline
  &Prop. & 0.19960 (0.01213) & 0.01133 (0.00007) & 1.45806e-04  \\
 0.2 &Size Prop. & 0.19973 (0.01145) & 0.01058 (0.00005) &  1.29974e-04 \\
  &Full& 0.19986 (0.01142) & 0.01056 (0.00005) & 1.29174e-04 \\  \hline
  &Prop. & 0.49956 (0.00963) & 0.00954 (0.00011) & 9.19119e-05 \\
0.5  &Size Prop. & 0.49965 (0.00904) & 0.00898 (0.00008) & 8.11057e-05 \\
  &Full & 0.49990 (0.00904) & 0.00896 (0.00008) & 8.08877e-05 \\  \hline
  &Prop. & 0.79973 (0.00541) & 0.00548 (0.00012) & 2.90660e-05 \\
0.8  &Size Prop & 0.79990 (0.00483) & 0.00529 (0.00009) & 2.31132e-05 \\
  &Full & 0.80018 (0.00489) & 0.00525 (0.00009) &  2.36726e-05 \\  \hline
  &Prop. & 0.90017 (0.00286) & 0.00321 (0.00008) & 8.14862e-06 \\
0.9  &Size Prop. & 0.90013 (0.00297) & 0.00318 (0.00008) & 8.76257e-06 \\
  &Full & 0.90040 (0.00292) & 0.00312 (0.00008) & 8.60114e-06 \\  \hline
  &Prop. & 0.98994 (0.00038) & 0.00039 (0.00001) & 1.45292e-07 \\
0.99  &Size Prop. & 0.98992 (0.00042) & 0.00041 (0.00002) & 1.77848e-07 \\
  &Full & 0.98997 (0.00037) & 0.00039 (0.00001) & 1.37289e-07  \\
   \hline\hline
\end{tabular}}
\end{table}

\begin{table}[ht]
\centering
\caption[Simulation study. Estimating $\sigma^2$ and its standard deviation]{Simulation study. Estimating $\sigma^2$ and its standard deviation. The mean (standard deviation) of the $100$ replications are given together with mean squared errors for $\rho=0.01,0.2,0.5,0.8,0.9,0.99$ using proportional and size-proportional weights comparing with the full likelihood results.}
\label{sim_sigma2}

\vspace*{2mm}
 \resizebox{\textwidth}{!}{%
\begin{tabular}{lllcccc}
\hline\hline
$\rho_0$&method& mean($\widehat{\sigma}^2$) (s.e.)  & mean(s.e.($\widehat{\sigma}^2$)) (s.e.)  & MSE \\
  \hline
  &Prop. & 1.99964 (0.02960) & 0.02981 (0.00049) & 8.67280e-04 \\
0.01  &Size Prop. & 2.00165 (0.02836) & 0.02834 (0.00040) & 7.98842e-04 \\
  &Full & 2.00167 (0.02832) & 0.02832 (0.00040) & 7.97002e-04 \\   \hline
  &Prop. & 2.00581 (0.02907) & 0.03093 (0.00055) & 8.70077e-04  \\
0.2  &Size Prop. & 2.00484 (0.02778) & 0.02936 (0.00044) & 7.87298e-04 \\
  &Full & 2.00473 (0.02772) & 0.02933 (0.00044) & 7.83248e-04 \\   \hline
  &Prop. & 1.99783 (0.03860) & 0.03638 (0.00097) & 1.47960e-03 \\
0.5  &Size Prop. & 1.99890 (0.03747) & 0.03488 (0.00085) & 1.39116e-03 \\
  &Full & 1.99897 (0.03748) & 0.03481 (0.00085) & 1.39152e-03 \\   \hline
  &Prop. & 2.00013 (0.04900) & 0.05002 (0.00166) & 2.37744e-03 \\
0.8  &Size Prop. & 2.00136 (0.04423) & 0.04930 (0.00137) & 1.93881e-03 \\
  &Full & 2.00101 (0.04569) & 0.04880 (0.00142) & 2.06754e-03 \\   \hline
  &Prop. & 2.00122 (0.05767) & 0.05872 (0.00196) & 3.29407e-03 \\
0.9  &Size Prop. & 2.00089 (0.06037) & 0.05915 (0.00230) & 3.60829e-03 \\
  &Full & 2.00115 (0.05876) & 0.05793 (0.00198) &  3.41986e-03 \\   \hline
  &Prop. & 1.99683 (0.06941) & 0.07117 (0.00254) & 4.77911e-03\\
0.99  &Size Prop. & 1.99527 (0.07598) & 0.07484 (0.00344) & 5.73813e-03 \\
  &Full & 1.99641 (0.06940) & 0.07093 (0.00253) & 4.78099e-03 \\
   \hline\hline
\end{tabular}}
\end{table}

\addtolength{\tabcolsep}{-.9mm}

\begin{table}[ht]
\centering
\caption[Simulation study. The computation time]{Simulation study. The computation time for a sample with $n=10$ and $c=$1e+02, 1e+03, 1e+04, 5e+04, 1e+05, 3e+05, 5e+05, 7e+05, 9e+05, 1e+06. The closed form solution is obtained by implementing the results of this paper in R, and the numerical solution is obtained using PROC MIXED in SAS to estimate a repeated measurement model with AR(1) covariance structure.}
\label{comp_time}

\vspace*{2mm}
 \resizebox{\textwidth}{!}{%
\begin{tabular}{lcccccccccc}
  \hline\hline
 time (s) & 1e+02 &1e+03&1e+04& 5e+04& 1e+05& 3e+05 &5e+05 &7e+05& 9e+05 &1e+06\\
   \hline
  Closed form & 0.00 & 0.00 & 0.03 & 0.23 & 0.34 & 1.45 & 2.07 & 3.37 & 4.40 & 4.89 \\
  Numerical & 0.08 & 0.13 & 1.04 & 10.45 & 34.74 & 268.96 & 770.74 & 1611.43 & 2724.31 & 3399.47 \\
   \hline
\hline
\end{tabular}}
\end{table}

\addtolength{\tabcolsep}{.9mm}

\begin{figure}[t]
\centering
\includegraphics[width=\textwidth, trim= 0cm 4cm 0cm 4cm]{comp_time.eps}
\caption[Simulation study. Comparing computation time using closed form and numerical solutions]{Simulation study. Comparing computation time using closed form (left) and numerical (right) solutions. The horizontal axis shows number of clusters ($c$) and the vertical axis shows the computation time in seconds.} \label{fig_comp_time}
\end{figure}

\setcounter{equation}{0}
\section{Details on PANSS data analysis}\label{suppdata}
\vspace{-3.5mm}
As one may see from Table~\ref{PANSSTab_freq}, by far the majority of the study subjects have complete data and hence belong to the first pattern.

 Figure~\ref{fig_all_first} presents boxplots for the entire set of data, for the subjects from the first pattern only, and for various split samples.

\begin{table}[ht!]
\centering
\caption{PANSS data. Number of clusters in each trial for each cluster pattern.}
\label{PANSSTab_freq}\vspace*{2mm}
 \resizebox{\textwidth}{!}{%
\begin{tabular}{rcccccccccccccccc}
\hline\hline
 & &&&&&&&&&&  \multicolumn{5}{c}{Trial} &  \\
  \cline{12-16}
$n$ & \multicolumn{10}{c}{Pattern}  & FIN-1 & FRA-3 & INT-2 & INT-3 & INT-7 &Total  \\
  \hline
 &  $\ast$& $\ast$& $\cdot$& $\cdot$& $\cdot$& $\cdot$& $\cdot$& $\cdot$&  $\cdot$&  $\cdot$&  17 & 8 & 71 & 43 & 3 & 142 \\
  2 &  $\ast$& $\cdot$& $\ast$& $\cdot$& $\cdot$& $\cdot$& $\cdot$& $\cdot$&  $\cdot$&  $\cdot$ & 0 & 0 & 2 & 0 & 1 & 3 \\
   &  $\ast$& $\cdot$& $\cdot$& $\cdot$& $\ast$& $\cdot$& $\cdot$& $\cdot$&  $\cdot$&  $\cdot$ & 0 & 0 & 1 & 0 & 0 & 1 \\ \hline
   &  $\ast$& $\ast$& $\ast$& $\cdot$& $\cdot$& $\cdot$& $\cdot$& $\cdot$&  $\cdot$&  $\cdot$ & 8 & 4 & 83 & 41 & 7 & 143 \\
  3 &  $\ast$& $\cdot$& $\ast$& $\cdot$& $\ast$& $\cdot$& $\cdot$& $\cdot$&  $\cdot$&  $\cdot$ & 0 & 0 & 2 & 0 & 0 & 2 \\
   &  $\ast$& $\ast$& $\cdot$& $\cdot$& $\ast$& $\cdot$& $\cdot$& $\cdot$&  $\cdot$&  $\cdot$ & 1 & 0 & 3 & 1 & 0 & 5 \\ \hline
   &  $\ast$& $\ast$& $\ast$& $\cdot$& $\ast$& $\cdot$& $\cdot$& $\cdot$&  $\cdot$&  $\cdot$ & 11 & 0 & 85 & 66 & 5 & 167 \\
   &  $\ast$& $\cdot$& $\ast$& $\cdot$& $\ast$& $\cdot$& $\ast$& $\cdot$&  $\cdot$&  $\cdot$ & 0 & 0 & 1 & 0 & 1 & 2 \\
  4 &  $\ast$& $\cdot$& $\ast$& $\cdot$& $\ast$& $\cdot$& $\cdot$& $\ast$&  $\cdot$&  $\cdot$ & 0 & 0 & 1 & 0 & 0 & 1 \\
   &  $\ast$& $\ast$& $\ast$& $\cdot$& $\cdot$& $\cdot$& $\ast$& $\cdot$&  $\cdot$&  $\cdot$ & 0 & 0 & 3 & 0 & 0 & 3 \\
   &  $\ast$& $\ast$& $\ast$& $\ast$& $\cdot$& $\cdot$& $\cdot$& $\cdot$&  $\cdot$&  $\cdot$ & 0 & 4 & 1 & 0 & 0 & 5 \\
   &  $\ast$& $\ast$& $\cdot$& $\ast$& $\cdot$& $\cdot$& $\ast$& $\cdot$&  $\cdot$&  $\cdot$ & 0 & 1 & 0 & 0 & 0 & 1 \\
  &  $\ast$& $\cdot$& $\ast$& $\cdot$& $\cdot$& $\cdot$& $\ast$& $\ast$&  $\cdot$&  $\cdot$ & 0 & 0 & 0 & 0 & 1 & 1 \\ \hline
   &  $\ast$& $\ast$& $\ast$& $\cdot$& $\ast$& $\cdot$& $\ast$& $\cdot$&  $\cdot$&  $\cdot$ & 58 & 0 & 85 & 35 & 6 & 184 \\
   &  $\ast$& $\ast$& $\ast$& $\cdot$& $\ast$& $\cdot$& $\cdot$& $\ast$&  $\cdot$&  $\cdot$ & 0 & 0 & 8 & 0 & 1 & 9 \\
   &  $\ast$& $\ast$& $\cdot$& $\cdot$& $\ast$& $\cdot$& $\ast$& $\ast$&  $\cdot$&  $\cdot$ & 0 & 0 & 6 & 0 & 0 & 6 \\
  5 &  $\ast$& $\ast$& $\ast$& $\cdot$& $\cdot$& $\cdot$& $\ast$& $\ast$&  $\cdot$&  $\cdot$ & 0 & 0 & 8 & 0 & 0 & 8 \\
   & $\ast$& $\cdot$& $\ast$& $\cdot$& $\ast$& $\cdot$& $\ast$& $\ast$&  $\cdot$&  $\cdot$ & 0 & 0 & 3 & 0 & 2 & 5 \\
   & $\ast$& $\cdot$& $\ast$& $\cdot$& $\ast$& $\cdot$& $\cdot$& $\ast$&  $\cdot$&  $\ast$ & 0 & 0 & 1 & 0 & 0 & 1 \\
   & $\ast$& $\ast$& $\ast$& $\ast$& $\ast$& $\cdot$& $\cdot$& $\cdot$&  $\cdot$&  $\cdot$ & 0 & 44 & 0 & 0 & 0 & 44 \\
   & $\ast$& $\ast$& $\cdot$& $\ast$& $\ast$& $\ast$& $\cdot$& $\cdot$&  $\cdot$&  $\cdot$ & 0 & 1 & 0 & 0 & 0 & 1 \\ \hline
   & $\ast$& $\ast$& $\ast$& $\cdot$& $\ast$& $\cdot$& $\ast$& $\ast$&  $\cdot$&  $\cdot$ & 0 & 0 & 986 & 240 & 74 & 1300 \\
   & $\ast$& $\ast$& $\cdot$& $\cdot$& $\ast$& $\cdot$& $\ast$& $\ast$&  $\ast$&  $\cdot$ & 0 & 0 & 1 & 0 & 0 & 1 \\
  6 & $\ast$& $\ast$& $\ast$& $\cdot$& $\cdot$& $\cdot$& $\ast$& $\ast$&  $\cdot$&  $\ast$ & 0 & 0 & 1 & 0 & 0 & 1 \\
   & $\ast$& $\ast$& $\ast$& $\cdot$& $\ast$& $\cdot$& $\ast$& $\cdot$&  $\ast$&  $\cdot$ & 0 & 0 & 1 & 0 & 0 & 1 \\
  & $\ast$& $\ast$& $\ast$& $\cdot$& $\ast$& $\ast$& $\ast$& $\cdot$&  $\cdot$&  $\cdot$ & 0 & 0 & 2 & 0 & 0 & 2 \\
   \hline\hline
\end{tabular}}
\end{table}

\begin{figure}[ht]
\centering
\includegraphics[width=0.8\textwidth, trim=2cm 5cm 2cm 5cm]{all_first.eps}
\caption{PANSS data. Boxplots for the entire set of data, for the subject from the first pattern only, and for various split samples. } \label{fig_all_first}
\end{figure}

To examine the choice of an AR(1) covariance structure, Table~\ref{fig_model_comp} shows three model selection criteria  to compare different error covariance structures. Changing from independence structure ($R=\sigma^2 I$) to compound-symmetry ($R=\sigma^2 I$) the criteria decrease with a large amount, and the same when changing to AR(1). The step to an unstructured covariance does not make a big difference (considering that the unstructured covariance would has $21$ parameters to estimate compared to $2$ parameters in the AR(1) model). Therefore, AR(1) seems to be a good choice.

\begin{table}[t]
\centering
\caption[PANSS data. Comparing different error covariance structures using three model comparison criteria for without trial model]{PANSS data. Comparing different error covariance structures using three model comparison criteria for model (\ref{model_no_trial}) (residual log-likelihood value; AIC; BIC). Three $R$ structures: Ind. : independence structure ($R=\sigma^2 I$), CS: compound-symmetry structure ($R= \sigma^2 I +dJ $), AR(1): AR(1) structure ($R_{ij}=\sigma^2 \rho^{|i-j|}$), UN: unstructured ($R_{ij}=\sigma^2_{ij}$).} \label{fig_model_comp}

\begin{tabular}{lccc}
\hline
\hline
Model&-2 Res.log.lik.&AIC&BIC\\
\hline
Unstructured  &80005.1&80047.1&80164.1\\
AR(1)         &80522.6&80526.6&80537.8\\
Compound symm.&82683.1&82687.1&82698.3\\
Independence  &89546.1&89548.1&89553.7\\
\hline\hline
\end{tabular}
\end{table}

The $95\%$ confidence intervals, accompanying (\ref{model_no_trial}), are presented in Figure~\ref{fig_without_trial}. In order to give more insight in these results, Figure~\ref{fig_without_trial_split} shows the $95\%$ confidence interval in each split, comparing with the full sample splits (the horizontal dashed line in the figure).

The $95\%$ confidence intervals, accompanying (\ref{model_trial}), are presented in Figure~\ref{fig_with_trial}. Figure~\ref{fig_with_trial_split} shows the $95\%$ confidence intervals for the parameter estimates in each split comparing with the full sample estimate (the horizontal dashed-like in the figure.)

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{without_trial.eps}
\caption[PANSS data. $95\%$ confidence intervals for fixed effects and variance components estimates and the standard deviations of these estimates using sample splitting, combined with proportional and size-proportional weights, and full likelihood (without trial model)]{PANSS data. $95\%$ confidence intervals for fixed effects and variance components estimates and the standard deviations of these estimates using sample splitting, combined with proportional (Pr - first) and size-proportional (S-Pr - second) weights, and full likelihood (Fl - third). The dashed horizontal line shows the full likelihood estimate. The model used in here is without trial effect (\ref{model_no_trial}).} \label{fig_without_trial}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{without_trial_split_by_split.eps}
\caption[PANSS data. $95\%$ confidence intervals for fixed effects and variance components estimates and the standard deviations of these estimates within each split (without trial model)]{PANSS data. $95\%$ confidence intervals for fixed effects and variance components estimates and the standard deviations of these estimates within each split. The dashed horizontal line shows the full likelihood estimate. The model used in here is without trial effect (\ref{model_trial}).} \label{fig_without_trial_split}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{with_trial.eps}
\caption[PANSS data. $95\%$ confidence intervals for fixed effects and variance components estimates and the standard deviations of these estimates using sample splitting, combined with proportional and size-proportional weights, and full likelihood (with trial model)]{PANSS data. $95\%$ confidence intervals for fixed effects and variance components estimates and the standard deviations of these estimates using sample splitting, combined with proportional (Pr - first) and size-proportional (S-Pr - second) weights, and full likelihood (Fl - third). The dashed horizontal line shows the full likelihood estimate. The model used in here is with trial effect (\ref{model_trial}).} \label{fig_with_trial}
\end{figure}

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{with_trial_split_by_split.eps}
\caption[PANSS data. $95\%$ confidence intervals for fixed effects and variance components estimates and the standard deviations of these estimates within each split (with trial model)]{PANSS data. $95\%$ confidence intervals for fixed effects and variance components estimates and the standard deviations of these estimates within each split. The dashed horizontal line shows the full likelihood estimate. The model used in here is with trial effect (\ref{model_trial}).} \label{fig_with_trial_split}
\end{figure}

\setcounter{equation}{0}
\section{Fieller's method and delta method}
\label{app1}

Here, the Fieller's method to construct a confidence interval for (\ref{nf_var_min_cs}) as well as the Delta method  are discussed. The ratio which we require a confidence interval for is:
$$n_f=\frac{\hat{\sigma^2}}{\widehat{\tau} \varepsilon}.$$
Suppose $n_1$ is a small-feasible sub-sampling size chosen to estimate the parameters in the model (e.g., $n_1=5$). From \cite{Iddi2011}, one finds:

\begin{multline}
%\label{cov_sigma2_tau}
\mathrm{Var}  \left(
\begin{array}{c}
\widehat{\sigma^2}\\
\widehat{\tau\varepsilon}
\end{array} \right) =  \\ \frac{2 \sigma^4}{Nn_1(n_1-1)}  \left(
\begin{array}{cc}
n_1 & -\varepsilon \\
-\varepsilon & \varepsilon^2\frac{\sigma^4 + 2(n_1-1) \tau \sigma^2 + n_1 (n_1-1) \tau^2}{\sigma^4}
\end{array} \right) \\ = \left(
\begin{array}{cc}
s_{11} &  s_{12} \\ s_{12} & s_{22}
\end{array} \right).
\end{multline}

The Fieller's confidence interval for (\ref{nf_var_min_cs}) can be calculated following these steps.
\begin{equation}
\begin{aligned}
C_1^2 =&\frac{s_{11}}{\widehat{\sigma^2}^2}, \\
C_2^2=& \frac{s_{22}}{\left(\varepsilon\widehat{\tau}\right)^2},\\
r= & \frac{s_{12}}{\sqrt{s_{11} s_{22}}},\\
A= &C_1^2 + C_2^2 - 2r C_1C_2,\\
B=& t^2 C_1^2 C_2^2 (1-r^2),\\
L=& \frac{\widehat{\sigma^2}}{\varepsilon\widehat{\tau}} \frac{1-z_{\alpha/2}^2 r C_1C_2 - z_{\alpha/2}\sqrt{A-B}}{1-z_{\alpha/2}^2C_2^2},\\
U=& \frac{\widehat{\sigma^2}}{\varepsilon\widehat{\tau}} \frac{1-z_{\alpha/2}^2 r C_1C_2 + z_{\alpha/2}\sqrt{A-B}}{1-z_{\alpha/2}^2C_2^2}.
\end{aligned}
\end{equation}
One may take $n_f=\max\{L,U\}$. Of course, in case were $z_{\alpha/2}^2 C_2^2 <1$, there would be no finite interval. As it was mentioned, one may also use the Delta method as follows:
\begin{equation}
\label{var_delta_method}
\mathrm{Var}(\frac{\widehat{\sigma^2}}{\varepsilon \widehat{\tau}}) \approx \frac{1}{\varepsilon^2 \widehat{\tau}^2} \left( s_{11} - 2 \frac{\sigma^2}{\tau} s_{12} + \frac{\sigma^4}{\tau^2} s_{22} \right).
\end{equation}
\setcounter{equation}{0}
\section[Random vertical data splitting for CS]{Random vertical data splitting for compound-symmetry structure}
\label{app2}
Assume a single cluster of size $N$ with multivariate normal distribution and compound-symmetry structure for its covariance matrix, see Section~\ref{sec_CS_effect}. Suppose we take $M_0$ sub-samples of size $n_0$ from this cluster. Following \cite{hoffman2001}, the parameter $\mu$ should be estimated within each sub-sample, $\widehat{\mu}^{(m)}$ ($m=1,\ldots,M_0$) and then the overall estimate can be obtained by averaging these estimates. Furthermore, one needs the within and between sub-samples variabilities to compute the overall variability, see (\ref{mo_cov}).

Within each sub-sample the variance of $\widehat{\mu}^{(m)}$ can be computed using the variance formula for a mean estimator in CS-multivariate normal \cite{Iddi2011} :
\begin{equation}
\label{within}
S_W= \frac{\sigma^2}{n_0} \left[ 1+ (n_0-1) \rho \right ].
\end{equation}
Estimating the between variability requires computing $\mathrm{E}\left\{ (\widehat{\mu}^{(m)} - \bar{\mu})^2 \right\}$, where $\bar{\mu}$ is the average of $\widehat{\mu}^{(m)}$'s ($m=1,\ldots,M_0$). Set $S_m$ as the index of members of sub-sample $m$ and $S_{m'}$ the rest of the sample, then,
\begin{equation*}
\widehat{\mu}^{(m)} - \bar{\mu} = \frac{1}{n_0}\left\{(1-\frac{1}{M_0})\sum_{i\in S_m} y_i - \frac{1}{M_0} \sum_{m=1}^ {M_0} \sum_{i\in S_{m'}} y_i \right\}.
\end{equation*}
Therefore,
\begin{equation}
\label{T1_T4}
\mathrm{E}\left\{(\widehat{\mu}^{(m)} - \bar{\mu})^2 \right\} = \frac{1}{n_0} \left( T_1 + T_2 + T_3+ T_4\right),
\end{equation}
where,
\begin{equation}
\label{eq_T1_T4}
\begin{cases}
T_1= \left( 1- \frac{1}{M_0} \right) \mathrm{E} \left [ \sum_{i\in S_m} y_i \right]^2\\
T_2= \frac{1}{M_0^2} (M_0-1) \mathrm{E} \left[ \sum_{i \in S_m} y_i \right]^2\\
T_3=  -2 \left(1-\frac{1}{M_0} \right) \frac{1}{M_0} (M_0-1) \mathrm{E} \left\{ \left(\sum_{i\in S_{m'}} y_i \right) \left( \sum_{i'\in S_{m'}} y_{i'} \right) \right\} \\
T_4= 2 \frac{1}{M_0} \frac{(M_0-1) (M_0-2)}{2} \mathrm{E} \left\{ \left(\sum_{i \in S_m} y_i \right) \left(\sum_{i'\in S_{m'}} y_{i'} \right) \right\}.
\end{cases}
\end{equation}
From (\ref{eq_T1_T4}) one may find,
\begin{equation}
\label{sum_T}
\begin{cases}
T_1+ T_2 = \frac{M_0-1}{M_0} n_0 \sigma^2 \left [1+ (n_0-1) \rho \right]\\
T_3 + T_4= - \frac{M_0-1}{M_0} \frac{\sigma^2 n_0^2}{N} \left[ 1+ (N_1) \rho \right].
\end{cases}
\end{equation}
Therefore, the between variability can be computed as follows,
\begin{multline}
\label{between}
S_B=\mathrm{E} \left\{ (\widehat{\mu}^{(m)} - \bar{\mu} )^2 \right\} = \\ \frac{M_0-1}{M_0} \left[\frac{1+(n_0-1)\rho}{n_0} - \frac{1+(N_1)\rho}{N} \right].
\end{multline}
Using (\ref{within}) and (\ref{between}), the total variability can be computed as follows,
\begin{equation}
\label{total_variability}
S_T= \frac{1}{M_0} \sigma^2 \frac{1+ (n_0-1) \rho }{n_0} + \frac{M_0-1}{M_0} \sigma^2 \frac{1+ (N-1) \rho}{N}
\end{equation}
Comparing the variance in (\ref{total_variability}) with the variance when estimating $\mu$ using the full sample would give,
\begin{equation}
\label{ARE}
\mathrm{ARE} = \frac{M_0-1}{M_0} + \frac{1}{M_0} \frac{N}{n_0} \frac{1+(n_0-1) \rho}{1+ (N-1) \rho}
\end{equation}
For example, if sub-sampling is done only once ($M_0=1$), then,
\begin{equation}
\label{ARE_M1}
\mathrm{ARE} = \frac{N}{n_0} \frac{1+(n_0-1) \rho}{1+(N-1) \rho},
\end{equation}
The same calculation is possible for $M_0=2$,
\begin{equation}
\label{are_M02}
\mathrm{ARE}= \frac{1}{2} + \frac{1}{2} \frac{N}{n_0} \frac{1+(n_0-1) \rho}{1+ (N-1) \rho}.
\end{equation}
By considering the desired ARE as $1+\epsilon$, one may find $n_0$.

\setcounter{equation}{0}
\section{Combination rule for $p$-values for MI}
\label{app_mi}

The Kruskal-Wallis test statistic asymptotically follows a $\chi^2$ distribution. Therefore, in order to find the combined $p$-value, we may follow the procedure proposed by \cite{li1991}. More details can be found in \cite{rubin2004} and \cite{enders2010}. The combination is done using {\tt{micombine.chisquare}} in package {\tt{miceadds}} in \textsf{R}, \cite{miceadds}.

\begin{itemize}
	\item \textbf{Averaging.} Compute the test statistic for each imputed data and take their average: $\overline{\chi^2} = \frac{1}{M} \sum_{m=1}^M \chi^2_m$.
	\item \textbf{Relative variance increase.} $r=\left(1+\frac{1}{M} \right) \frac{\sum_{m=1}^M (\sqrt{\chi^2_m} - \sqrt{\overline{\chi^2}})^2}{M-1}$
	\item \textbf{Test statistic.} $D=\frac{\frac{\overline{\chi^2}}{\kappa} - \frac{M+1}{M-1}r}{1+r}$, where $\kappa$ is the degrees-of-freedom of the $\chi^2_m$. For a $\chi^2$-test of independence, it is $(k_1-1)(k_2-1)$ where $k_1$ and $k_2$ are the number of levels of two categorical variables. Take five diagnosis groups and variable \textsf{Eye} with two levels, then $\kappa=(5-1)(2-1)=4$. For the Kruskal-Wallis test, $\kappa$ is the number of groups minus 1. So, when comparing values of a continuous variable among three types of diagnosis, $\kappa=2$. When this comparison is done for all five types, $\kappa=4$. 
	\item \textbf{Computing $p$-value.} $p$-value$=P(F_{\kappa,\nu} > D)$, where $F$ is the $F$-distribution and $\nu=\kappa^{-3/M} (M-1) \left(1+\frac{1}{r}\right)^2$
\end{itemize}
\newpage~
\setcounter{equation}{0}
\setcounter{page}{355}
\section{The surrogate model}
\label{sec_surrogate_model}
The model takes the form:
\begin{equation}
\label{surrogatemodel}
\left\{
\begin{array}{l}
S_{ij}=\mu_S + m_{S_i} + (\alpha + a_i) Z_{ij} + \varepsilon_{S_{ij}},\\
T_{ij}= \mu_T + m_{T_i} + (\beta + b_i) Z_{ij} + \varepsilon_{T_{ij}},
\end{array} \right.
\end{equation}
where $S_{ij}$ and $T_{ij}$ are the surrogate and true endpoints, and $Z_{ij}$ is the treatment indicator (1 for active; 0 for placebo), respectively. Index $i$ refers to the trial and $j$ to the subject. Further, $\mu_{S}$ and $\mu_{T}$ are fixed intercepts and $m_{S_i}$ and $m_{T_i}$ are  random intercepts, for surrogate and true endpoint, respectively. The fixed treatment effects are $\alpha$ (for $S$) and $\beta$ (for $T$), and the centre-specific treatment effects are $a_i$ (for $S$) and $b_i$ (for $T$). 
Random intercepts and slopes are assumed to follow
$(m_{S_i}, m_{T_i}, a_i, b_i)'\sim N(0,D)$,
where 
\begin{equation}
\label{D}
D = \left(
\begin{array}{cccc}
d_{SS} & & &\\
d_{ST} & d_{TT} & &\\
d_{Sa} & d_{Ta} & d_{aa} &\\
d_{Sb} & d_{Tb} & d_{ab} & d_{bb}
\end{array} \right).
\end{equation}
Furthermore, for the error terms,
\begin{equation}
\label{surrogateSigma}
\left(\begin{array}{c}
\varepsilon_{S_{ij}}\\
\varepsilon_{T_{ij}}
\end{array} \right)
\sim N(0,\Sigma),\; \Sigma = \left(
\begin{array}{cc}
\sigma_{SS} & \\
\sigma_{ST} & \sigma_{TT}
\end{array} \right).
\end{equation}
The parameters of interest for the surrogate evaluation are trial- and individual-level coefficients of determination, $R^2_{\mbox{\tiny trial}}$ and $R^2_{\mbox{\tiny indiv}}$, respectively:
\begin{equation}
\label{R2}
\begin{gathered}
R^2_{\mbox{\tiny trial}} = \frac{1}{d_{ab}}
\left(
\begin{array}{c}
d_{Sb}  \\
d_{ab}
\end{array} \right)^{\prime} 
\left(
\begin{array}{cc}
d_{SS} & d_{Sa}\\
d_{Sa} & d_{aa}
\end{array} \right)^{-1}
\left(
\begin{array}{c}
d_{Sb}  \\
d_{ab}
\end{array} \right),\\ R^2_{\mbox{\tiny indiv}}=\frac{\sigma_{ST}^2}{\sigma_{SS} \sigma_{TT}}.
\end{gathered}
\end{equation}

\bibliography{thesis_leuven_bib}{}
\bibliographystyle{abbrvnat}

\end{document}

### 2019-11-16 summaryThesis02.tex
semi-final revised version, check for typos remained

--- 2019-11-13 summaryThesis.tex
@@
 
-In this thesis, we propose methodologies to deal with such problems using the so-called data splitting.
-In a nutshell, data splitting means to split the difficult (sometimes infeasible) job of modelling the big dataset by splitting it into smaller chunks of manageable sub-data, analyze each of them separately, and then combine the results of these analyses using appropriate combinations rules.
+In this thesis, we propose methodologies to deal with such problems using  so-called data splitting.
+In a nutshell, data splitting refers to splitting the difficult (sometimes infeasible) job of modelling a big dataset by splitting it into smaller chunks of manageable sub-data, analyzing each of them separately, and then combining the results of these analyses using appropriate combination rules.
 This would break a difficult task into several, often so-called embarrassingly parallel, simpler tasks.
@@
 \begin{enumerate}
-\item Splitting the data.
-\item Performing the analysis on each sub-dataset.
+\item Splitting the data;
+\item Performing the analysis on each sub-dataset;
 \item Efficiently combine the results from each sub-dataset to form an overall estimate with its precision.
@@
 This would lead to balanced sub-samples.
-We have shown the existence of closed-form solutions, leading to faster convergence of model fitting procedures.
-\item \textbf{Random vertical data splitting.} This approach is useful when we are dealing with data with large number of measurements for each cluster per outcome.
-In this situation one can take sub-samples of each cluster with a manageable size.
+We have established the existence of closed-form solutions, leading to faster convergence of model fitting procedures.
+\item \textbf{Random vertical data splitting.} This approach is useful when we are dealing with data with large numbers of measurements for each cluster per outcome.
+In this situation one can take sub-samples of each cluster of manageable size.
 \item \textbf{Structured vertical data splitting.} This approach is useful when the number of outcomes of interest is large.
@@
 Our findings are motivated and supported by various datasets.
-Furthermore, the methodologies we have developed are implemented in statistical computing language R.
+Furthermore, the methodologies we have developed are implemented in statistical computing language~R.
 \vspace{1cm}
+
+\begin{center}
+{\Huge{Samenvatting}}
+\end{center}
+\vspace{1cm}
+\begin{sloppypar}
+Steekproefgrootte is altijd een belangrijk thema geweest in de statistiek.
+Terwijl de problemen zich lange tijd situeerden ter hoogte van kleine steekproeven, is het nu zo dat ook (zeer) grote steekproeven voor problemen zorgen bij statistische analyse.
+Wanneer zeer grote steekproeven gebruikt worden samen met heel complexe modellen, dan kunnen er zich een reeks problemen voordoen, zoals bijvoorbeeld onrealistisch lange berekeningstijden, convergentieproblemen, tot en met de onmogelijkheid om de voorkeurstechniek (bijv.\ meest aannemelijke schatters) te gebruiken.
+\end{sloppypar}
+
+\begin{sloppypar}
+Verder, terwijl in de eenvoudigste studies alle metingen onafhankelijk zijn, zien we vandaag heel veel zogenaamd hi\"erarchische gegevens (ook gekend onder de termen: clusters, gecorreleerde gegevens, afhankelijke gegevens).
+Met de term `cluster' verwijzen we naar metingen herhaald genomen aan eenzelfde eenheid (bijv.\ proefpersoon, huishouden, studie, enz.).
+Tegen die achtergrond verwijzen we met het aantal herhalingen naar het aantal metingen per cluster.
+Enkele voorbeelden: pati\"enten die herhaald doorheen de tijd worden gemeten; verscheidene pati\"enten van dezelfde huisarts of dezelfe kliniek, verscheidene web-links aangeklikt door dezelfde gebruiker, allerlei genetische informatie beschikbaar voor dezelfde persoon, enz.
+Belangrijke deelgebieden van cluster gegevens zijn herhaalde metingen, longitudinale gegevens, multi-centrische studies en meta-analyse.
+Ook vandaag is het in rekening brengen van correlatie binnen clusters een uitdagend probleem, vooral wanneer het totaal aantal clusters zeer groot is en/of wanneer een groot aantal metingen per cluster beschikbaar is.
+We worden dan veelal geconfronteerd met zo goed als onoverkomelijke problemen van berekenbaarheid.
+\end{sloppypar}
+
+\begin{sloppypar}
+In deze thesis stellen we methdologie voor om de bovenstaande problemen aan te pakken op basis van zogenaamde data-splitsing.
+In een notendop verwijst data-splitsing naar het opdelen van een moeilijke (soms zelfs onmogelijke) taak zoals het modelleren van een zeer grote dataset, in kleinere deeltaken die elk wel overzichtelijk zijn.
+Een voorbeeld is het analyseren van delen van de dataset, \'e\'en voor \'e\'en, waarna de resultaten worden gecombineerd tot \'e\'en enkel resultaat.
+Als het aantal deeltaken zeer groot wordt spreken we in het Engels van {\em embarrassingly parallel tasks.} Data-splitsing bestaat uit drie deeltaken:
+\begin{enumerate}
+\item Splitsen van de data;
+\item Elk van de delen analyseren;
+\item Het effici\"ent combineren van de resultaten tot \'e\'en enkel eindresultaat. 
+\end{enumerate}
+\end{sloppypar}
+
+\begin{sloppypar}
+Er zijn verscheidene splitsings-strategie\"en. We kunnen ze in vier grote klasen opdelen:
+\begin{itemize}
+\item \textbf{Random horizontale splitsing.} 
+Deze methode is nutting wanneer het aantal clusters groot is; idealiter zijn de clusters van ongeveer gelijke grootte, d.w.z.\ een (bijna) gebalanceerd steekproefopzet.
+Deze methode leidt tot een aantal sub-datasets van kleinere omvang, die elk makkelijker te manipuleren zijn.
+\item \textbf{Gestructureerde horizontale data splitsing.} 
+Deze methode is nuttig wanneer sommige clusters makkelijker individueel te hanteren zijn dan in samenhang.
+Daarom, eerder dan willekeurige deel-steekproeven te nemen, maakt men hier gebruik van de vooraf aanwezige structuur.
+Een belangrijk voorbeeld, beschouwd in dit werk, is splitsing op basis van cluster-grootte.
+We vormen dus deel-datasets waarbinnen alle clusters dezelfde grootte hebben.
+Elk van deze deelsteekproeven is dan gebalanceerd.
+Voor dergelijke gebalanceerde deelsteekproeven hebben we het bestaan van gesloten schattingsvormen aangetoond.
+Dit leidt daarom tot bijzonder effici\"ente procedures.
+\item \textbf{Random verticale data splitsing.} Deze methode is nuttig wanneer we een zeer groot aantal herhalingen hebben binnen een cluster.
+We splitsen clusters dus in deelclusters van een hanteerbare grootte.
+\item \textbf{Gestructureerde verticale data splitsing.} Deze aanpak is nuttig wanneer het aantal te bestuderen responsveranderlijken groot is.
+Het gemeenschappelijk analyseren van alle responsveranderlijken samen is vaak ondoenbaar, maar dat geldt niet voor alle mogelijke paren die we vormen uit de responsveranderlijken (of tripels, quadrupels, enz.).
+De deelsteekproeven bestaan dan uit alle gegevens beschikbaar voor het paar van veranderlijken dat wordt beschouwd.
+Deelsteekproeven respecteren dus deze specifieke structuur.
+\end{itemize}
+\end{sloppypar}
+
+\begin{sloppypar}
+Voor elk van de splitsings-strategie\"en werden gepaste en effici\"ente combinatieregels opgesteld.
+Daarbij werden ook uitzonderlijke gevallen waar de algemene methodologie niet werkt in ogenschouw genomen.
+Onze resultaten werden gemotiveerd en ge\"{\i}llustreerd aan de hand van een reeks datasets.
+Tenslotte hebben we de voorgesltede methodologie ge\"{\i}mplementeerd in de statistische softwaretaal~R.
+\end{sloppypar}
+\vspace{1cm}
+
 \begin{center}
@@
 \vspace{0.5cm}
-Vahid Nassiri  has obtained a Bachelor in Statistics (2002) from the Shahid Beheshti Univerity in Tehran, Iran, and a master in Mathematical Statistics (2008) from the Amirkabir University of Technology in Tehran, Iran.
+Vahid Nassiri  has obtained a Bachelor in Statistics (2002) from the Shahid Beheshti Univerity in Tehran, Iran, and a Master in Mathematical Statistics (2008) from the Amirkabir University of Technology in Tehran, Iran.
 Afterwards, he has obtained a PhD in Applied Mathematics (2013) from Vrije Universiteit Brussel (VUB), Brussels, Belgium.
 His main research interests are in the field of Big Data and High-dimensional problems.
-While working on the sparse estimation of high-dimensional linear models in sense of least squares and quantile regression in his former PhD, his current work is concerned on the fast computation of models for correlated data with possible missing values.
-His research expertise is mainly concerned with analyzing incomplete big correlated data using mixed models, multiple imputation, penalized likelihood and related topics.
-He also has experiences in consultation involving projects from pharmacology, biology, medicine, engineering, psychology, sociology, linguistics, and market research.
-He is now working as a statistical consultant in Open Analytics NV, Antwerp, Belgium.
+While working on the sparse estimation of high-dimensional linear models in the sense of least squares and quantile regression in his former PhD, his current work is concerned on the fast computation of models for correlated data with possible missing values.
+His research expertise is mainly concerned with analyzing incomplete big correlated data using mixed models, multiple imputation, penalized likelihood, and related topics.
+He also has experience in consulting involving projects from pharmacology, biology, medicine, engineering, psychology, sociology, linguistics, and market research.
+He is now working as a statistical consultant at Open Analytics NV, Antwerp, Belgium.
 

### 2019-11-16 rebuttal.tex
PL is done + part of KM
reply to referees updated (All included)

@@
 \usepackage{fullpage}
-
+\newcommand\Chapter[2]{
+  \chapter[#1: {\itshape#2}]{#1\\[2ex]\Large\itshape#2}
+}
 % import Eq and Section references from the main manuscript where needed
@@
 \begin{document}
+\begin{center}
+{\Large{{PhD Thesis}}}\\
+\vspace{2mm}
+{\LARGE{\textbf{Data Splitting and Its Applications}}}\\
+\vspace{5mm}
+{\Large{\textbf{Vahid Nassiri}}}
+\end{center}
+\vspace{1cm}
+\section*{Response to the reviewers}
 
-\section*{Response to the reviewers}
 % General intro text goes here
@@
 % Let's start point-by-point with Reviewer 1
-\reviewersection
+\reviewersection{\textbf{\large{Prof. Helena Geys}}}
 
@@
 
-\reviewersection
+\reviewersection{\textbf{\large{Prof. Katrijn Van Deun}}}
 
@@
 \begin{itemize}
-\item P3, first line: add ‘that’ -> when we realize THAT increasing….
+\item P3, first line: add ‘that’ $\rightarrow$ when we realize THAT increasing….
  \item Top p4: replace “are” with ‘can be done’ in “and analyzing these different sub-samples are independent of each other”
@@
  \item P18 replace ‘that’ by ‘where’ in ‘An example in our motivation datasets WHERE random horizontal splitting can be beneficial is the Divorce in Flanders’
- \item P18, insert ‘s’ in ‘deign’ -> design
+ \item P18, insert ‘s’ in ‘deign’ $\rightarrow$ design
  \item P20, top: insert ‘is’ in ‘it IS worth …’
@@
 
-\begin{response}
+\begin{reply}
 We totally agree with the reviewer. The revised manuscript includes necessary references now:
@@
 
-\reviewersection
+\end{reply}
+
+\reviewersection{\textbf{\large{Prof. Karl Meerbergen}}}
+
+\begin{point}
+Table 2.1 contains too few nonzero numbers to draw conclusions about the complexity of the faster methods. It may be better to use more digits behind the comma.
+\end{point}
+
+\begin{reply}
+We thank the reviewer for point out to this issue.
+In the revised manuscript, bench-marking is done again and now the computation times are displayed in microseconds, so the comparison becomes possible also for faster methods.
+\end{reply}
+
+\begin{point}
+The font in Figure 3.1 is too small for my eyes.   
+\end{point}
+
+\begin{reply}
+The font size is made larger in the revised manuscript.
+\end{reply}
 
@@
 
-\begin{response}
+\begin{reply}
 $J$ indicates a matrix of ones. The revised manuscript now includes this information: `where $I$ indicates an identity matrix and $J$ indicates a matrix of ones.'
-\end{response}
+\end{reply}
+
+\reviewersection{\textbf{\large{Prof. Philippe Lemey}}}
+
+\begin{point}
+The PhD candidate may consider including a list of abbreviations, but this is not absolutely necessary, so I leave this to his discretion.
+I will send the PhD candidate an annotated pdf version of thesis that marks a number of typos that can be corrected.
+\end{point}
+
+\begin{reply}
+The list of abbreviations are added to the revised manuscript. We also thank the reviewer for all his comments in the annotated PDF file. We have made sure all of the corrections are applied.
+\end{reply}
+
+\reviewersection{\textbf{\large{Prof. Peter Hoet}}}
+
+\begin{point}
+The manuscript form is not (clearly) in line with the instructions provided by the doctoral school.
+The manuscript is much longer than the suggested 150 pages, do not want to be too strict on this but 402 is more than two time exceeding.
+The candidate can consider e.g.
+whether chapter 7 is strictly needed (?) and/or to place the appendix after the bibliography (in fact the appendix is not a part of the manuscript).
+\end{point}
+
+\begin{reply}
+Chapter 7 is actually based on two of our publication, do despite its title, it is indeed necessary. But we totally agree with the reviewer and move the appendix after the bibliography.
+\end{reply}
+
+\begin{point}
+The final manuscript should contain a scientific abstract in Dutch and English and a short description of the professional background of the PhD researcher.
+\end{point}
+
+\begin{reply}
+The revised manuscript now contains a scientific abstract both in Dutch and English as well as a short description of professional background.
+\end{reply}
 

### 2019-11-17 reply2reviewersVahidNassiriManuscript.tex
final version

--- 2019-11-16 rebuttal.tex
@@
 In fact, $i$ goes from 1 to 15, as we have 5 personality factors, each for mother, father, and child.
-Then $j$ runs for various items in the 44-item BFI questionnaire correspond to each of these aspects.
-The number of items correspond to each personality aspect are not equal.
-Now, instead of $m_j$ we show these by $m_i$ as they corresponds to factor-role $i$.
+Then, $j$ runs for various items in the 44-item BFI questionnaire correspond to each of these aspects.
+The numbers of items corresponding	to each personality aspect are not equal.
+Now, instead of $m_j$ we show these by $m_i$ as they correspond to factor-role $i$.
 This is changed in the revised version as follows:
@@
 \begin{reply}
-The matrix matrix on page 138 show the index corresponds to each element in the sorted matrix on page 137.
+The matrix matrix on page 138 shows the index corresponds to each element in the sorted matrix on page 137.
 We agree with the reviewer that currently this connection is not sufficiently clear, so we have changed it in the revised manuscript as follows:
@@
 \begin{quote}
-Order the columns of the matrix in the previous step. For example, the following matrix shows a permutation which rearranges columns of the matrix in the previous step in an ascending order:
+Order the columns of the matrix in the previous step. For example, the following matrix shows a permutation which rearranges columns of the matrix in the previous step in an ascending order.
 \end{quote}
@@
 We thank the reviewer to bring this up, indeed, the current sentence is vague.
-We meant the generated sample in the fist example as the predictor, but now it is changed as follows in the revised manuscript:
+We meant the generated sample in the first example is used as the predictor, but now it is changed as follows in the revised manuscript:
 
@@
 \begin{point}
-P4: please check the sentence containing “but for others, it would even speeds down with a sub-linear rate” -> the ‘s’ in speedS is grammatically not correct and you probably mean ‘slow down’
+P4: please check the sentence containing “but for others, it would even speeds down with a sub-linear rate” $\rightarrow$ the ‘s’ in speedS is grammatically not correct and you probably mean ‘slow down’
 \end{point}
@@
 \begin{reply}
-Chapter 7 is actually based on two of our publication, do despite its title, it is indeed necessary. But we totally agree with the reviewer and move the appendix after the bibliography.
+Chapter 7 is actually based on two of our publications, do despite its title, it is indeed necessary. But we totally agree with the reviewer and move the appendix after the bibliography.
 \end{reply}
